### 2023-05-11 noise_covid_v2.Rmd
add Rmd

---
title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
author: "Rachael Aber"
header-includes: \usepackage{setspace}\doublespacing
output:
  html_document:
    df_print: paged
  pdf_document: default
geometry: margin=2.54cm
fontsize: 12pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F, 
                      fig.align = "center", fig.width = 5, fig.height = 5)
set.seed(12345)
devtools::load_all()
source("R/simulation.R")
```

# Introduction

Core challenges in epidemiology include estimating the efficacy of control measures, and understanding the impact of events of interest on incidence case counts. Allocating healthcare resources toward individuals with higher-than-average transmission risk can disproportionately reduce the population-level effective reproduction number (i.e., the average number of secondary infections caused by a randomly selected individual; Althouse et al. 2020). In other words, leveraging transmission variability among individuals to target interventions can be a highly effective strategy for reducing population spread. However, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable. Conversely, evaluating control strategies using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. The impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.

Metrics of population-level *variability* may be overlooked and useful ways to understand epidemic dynamics and control. By population-level variability, I mean - loosely - dispersion of data around the time-averaged trajectory. The ability to estimate changing rates of *dispersion* in epidemic time series would be a step towards a more complete view and predictive understanding of NPIs and gatherings at the individual as well as the population level. This is because individual-level heterogeneity in transmission scales up to affect population-level dynamics (Lloyd-Smith 2005), so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.

From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects has been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003). Similarly, inference based on variability in epidemic time series' are emerging: for instance, Graham et al. (2019) use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen. Sun et al. (2021) found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability. Analyzing variability in epidemic dynamics in terms of bursts of incidence is also important for planning surge capacity (Wallinga 2018).

Since evaluating control strategies using only course-scale changes in incidence over time may fail to account for parts of the context of transmission and control, we should aim to learn about individual-level heterogeneity as a means to understand control strategies more comprehensively. Similarly, we should aim to learn about individual-level heterogenity following congregation events, as estimates of individual-level variability are used to identify so-called "superspreading" dynamics: high levels of individual-level variability indicate that some individuals cause more onward transmission (via secondary infections) than others, and this in turn scales up to influence resulting incidence time series following these events.

## Causes of overdispersed disease incidence

*Overdispersed disease incidence* is suggestive of underlying biological processes of superspreading (overdispersed *individual reproductive number*), demographic/environmental stochasticity affecting cases, or changes in propagation of the pathogen at the population level (i.e., population effective reproduction number changing in time). Note that some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992).

Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992). However, issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992).

Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.

Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions. Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al. 2002). The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.

## Specific aim

**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint. Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **.

It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a reduction in dispersion of cases, potentially modulated by changes in $k_t$. Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.

# Data

For this project, we use county-level COVID-19 incidence data. Source: [USAFacts](https://usafacts.org/). The data is weekly case counts by US county.

Approximately two months of data on either side of November 24th, 2020 was used in the analysis. This step was done to include a set number of "waves" of incidence in the model fit to each side of the hypothesized change. However, we acknowledge that there is a trade-off between including few waves and increasing the power of the proposed test by including more time series data.

```{r, echo=FALSE, fig.cap="COVID-19 county-level incidence in New York surrounding Thanksgiving 2020."}
# USA case/population size data to work with

data <- read.csv("data/confirmed_and_death_1209.csv")
data$DateTime <- as.Date(data$DateTime)

# NY dataframe for Thanksgiving analysis
data <- data[data$DateTime >= "2020-10-26" & data$DateTime <= "2020-12-24", ]
inds <- seq(as.Date("2020-10-26"), as.Date("2020-12-24"), by = "day")
days <- weekdays(inds)
inc <- aggregate(DailyNewConfirmedCases ~ countyFIPS, data, FUN = list)
inc$populations <- tapply(X = data$population, INDEX = data$countyFIPS, FUN = mean)
inc$stateFIPS <- as.numeric(substr(inc$countyFIPS, 1, 2))

# all_zero <- c()
# for (i in 1:dim(inc)[1]){
#   if (all(unlist(inc$DailyNewConfirmedCases[i]) == 0)){
#     print(i)
#     all_zero <- c(all_zero, i)
#   }
# }
# inc <- inc[-all_zero,] 
ny_df_t <- inc[inc$stateFIPS == 36,]
fl_df_t <- inc[inc$stateFIPS == 12,]
ny_fl_df_t <- rbind(ny_df_t, fl_df_t)
```

# Methods

## Detecting variability changes in count data: the problem of population size

One might expect that in segments where the mean parameter is approximately the same, inference based on the Poisson distribution might be effective in detecting overdispersed incidence. There are several approaches to testing the Poisson assumption for count data, including testing variance-to-mean relationship, testing the probability-generating function (PGF) against an alternative, and standard goodness-of-fit tests (Karlis and Xekalaki 2000). However, an important consideration should be addressed for modeling time series of counts: the Poisson model has a spread of values that is comparatively more narrow for a process with larger mean (the mean/standard deviation ratio of a Poisson process is larger for larger total counts). For example, consider the mean to standard deviation ratios of the Poisson processes tabulated below: $\mu = \sigma^2 = 5$ and another with $\mu = \sigma^2 = 10$. For the process with a smaller mean, the mean to standard deviation ratio is smaller, whereas for the process with a larger mean, the mean to standard deviation ratio is larger. This presents an issue for using tests of Poisson variance to detect overdispersion or changes in dispersion in count time series, such as (importantly for epidemiology) incidence time series from populations of different sizes.

```{r}
rat <- data.frame(Mean = c(5, 10), Ratio = c(2.236068, 3.162278))
kable(rat, align = "cc", caption = "Poisson mean to standard deviation ratio", digits=2)
```

To illustrate this issue, I identified segments with approximately steady mean by mean changepoint detection (performed using an information criterion, MBIC). The cpt.mean function in the changepoint package (Killick and Eckley 2014) assumes that the variance of the process is equal to one, so each New York county time series was scaled beforehand. Once these segments were identified, we used a goodness-of-fit test assuming that the Poisson mean parameter of each segment, $\lambda_i$ is unknown. This is called the variance test, and it has an optimal power property against the negative binomial alternative. The rejection region is demarcated by large values of the test statistic (Potthoff and Whittinghill 1966), and values of the test statistic increase with population size.

```{r, fig.cap="Average Pothoff-Whittinghill (PW) statistic for each NY county plotted against county population size."}

scaled_cases <- map(ny_df_t$DailyNewConfirmedCases, ~ as.vector(scale(.)))

ny_df_t$ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE)))

which(sapply(ny_df_t$ch_pts, length) == 0)

segments <- map2(.x = ny_df_t$DailyNewConfirmedCases[-c(2, 8, 21, 49, 54, 62)], .y = ny_df_t$ch_pts[-c(2, 8, 21, 49, 54, 62)], ~ get_segments(series = .x, changepoints = .y))

mean_pw <- map(segments, ~ get_mean_pw(.))

# Plot the mean PW statistic v. population size

plot(log(ny_df_t$populations[-c(2, 8, 21, 49, 54, 62)]), log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2)
```

Methods designed specifically for count data show a similar dependence on population size. The cpt.meanvar function (changepoint package) is designed to detect changes in mean and variance in count data. The number of changepoints identified for a time series increases with its population size.

```{r, fig.cap="Log of the number of changepoints identified by cpt.meanvar in each NY county plotted against log of county population size."}

ny_df_t$ch_pts2 <- map(ny_df_t$DailyNewConfirmedCases, ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2))) 
ny_df_t$n_ch_pts2 <- map(ny_df_t$ch_pts2, ~length(unlist(.)))

plot(log(ny_df_t$populations), log(unlist(ny_df_t$n_ch_pts2)), xlab="Log Population Size", ylab=" Log Number of Changepoints", col=2)
```

## Modeling proccess mean

We propose a generalized linear model (GLM) approach to quantify incidence dispersion in segments of a time series, using a single parameter, $\theta$, while separately accounting for population size in the model. Another approach is to fit one model to the entire time series accounting for autocorrelation using a quasi-likelihood approach using a specified matrix V (Barron 1992). We account for unobserved heterogeneity, contagion, and time dependence in rate through specifying a changing process mean (to account for autocorrelation) and specifying a overdispersed conditional distribution (to account for overdispersion).

A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a generalized additive model (GAM) with one explanatory variable, time. In the GAM framework, the function for a predictor can be represented using spline functions:

 $$g(E[Y_i]) = f(t_i) = \sum_m^M\beta_m h_m(t_i)$$
 $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$

Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). The degrees of freedom were chosen due to **

$$g(E[Y_i]) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
$$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
```{r, fig.cap="County case count time series for each New York county with the hypothesized Thanksgiving changepoint represented by the black vertical line (L), and total New York COVID-19 case count time series (R)."}

par(mfrow=c(1,2))
# By visual inspection 
plot(ny_df_t$DailyNewConfirmedCases[[1]] ~ inds, type="l", ylim=c(0, 5000), xaxt="n", ylab="Daily Incidence", xlab="Date")
for (i in 2:dim(ny_df_t)[1]){
  lines(ny_df_t$DailyNewConfirmedCases[[i]] ~ inds, col=sample(rainbow(10)))
}
abline(v=inds[30], lwd=2)
axis(1, inds, inds, cex.axis = .7)

ny_col_sum <- matrix(unlist(ny_df_t$DailyNewConfirmedCases), nrow=62,
                     ncol=length(ny_df_t$DailyNewConfirmedCases[[1]]))
ny_cases <- colSums(ny_col_sum)
plot(ny_cases ~ inds, type="l", col=6, xlab="Date", ylab="Daily Incidence")
abline(v=inds[30], lwd=2)
```

```{r, fig.cap="County case count time series for each Florida county with the hypothesized Thanksgiving changepoint represented by the black vertical line (L), and total Florida COVID-19 case count time series (R)."}

par(mfrow=c(1,2))
# By visual inspection 
plot(fl_df_t$DailyNewConfirmedCases[[1]] ~ inds, type="l", ylim=c(0, 5000), xaxt="n", ylab="Daily Incidence", xlab="Date")
for (i in 2:dim(fl_df_t)[1]){
  lines(fl_df_t$DailyNewConfirmedCases[[i]] ~ inds, col=sample(rainbow(10)))
}
abline(v=inds[30], lwd=2)
axis(1, inds, inds, cex.axis = .7)

fl_col_sum <- matrix(unlist(fl_df_t$DailyNewConfirmedCases), nrow=62,
                     ncol=length(fl_df_t$DailyNewConfirmedCases[[1]]))
fl_cases <- colSums(fl_col_sum)
plot(fl_cases ~ inds, type="l", col=6, xlab="Date", ylab="Daily Incidence")
abline(v=inds[30], lwd=2)
```
## IRLS

Briefly, what iteratively reweighted least squares (IRLS) does is compute an adjusted dependent variable, compute a weight matrix, and compute (using weighted least squares) coefficients and therefore the linear predictor as well.

In irls.nb.1 in the NBPSeq R package (*cite):

1. Initial mu (either supplied or computed) and eta = log(mu/s).
2. z = eta - offset + (y - mu)/mu
3. w = drop(\mu/sqrt(varmu))
4. call Fortran code to perform weighted LS. Weighted least squares produces: $\hat{\beta} = (X^TWX)^{-1}X^TWY$, where weight is based on the variance of the observation. 
5. Then new values of mu and eta from new coefficients.

Within diy_spl_fit, fit.nb.regression is run and the standard error of param theta is computed by inverting info matrix, take [1,1] entry and take square root. Within fit.nb.regression, kappa is optimized, then irls.nb.1 (the IRLS portion) is called, then additional output - like the information matrix - is added.

## Inclusion of an offset term in the model

Regardless of whether the data is modeled as negative binomial or Poisson around the process mean (and regardless of whether quasi-likelihood methods are used), it's crucial to incorporate an offset term - sometimes called an exposure variable - in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual). Essentially, instead of focusing on the infection rate, our model allows the focus to be on infection rate per person. In other words, an offset term was added to the model to account for case counts resulting from different population sizes - an offset for population size means that incidence dispersion properties may be assessed *while accounting for a population effect*.

Note that if the offset model were perfect, the estimated coefficient on log population would be exactly one:
$$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
, where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.

In this case, a log link function is used:

  $$log(E[Y_i])-log(n_i) = \sum_m^M\beta_m h_m(t_i)$$
  $$log(E[Y_i]) =  \sum_m^M\beta_m h_m(t_i) + log(n_i)$$

## Identifying an appropriate model of the conditional distribution

```{r}
disp_stats <- c()
for (i in 1:dim(ny_fl_df_t)[1]){
  Y <- ny_fl_df_t$DailyNewConfirmedCases[[i]]
  spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
  disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
  #print(summary(spline_mod.1.nb)$coefficients[,4])
}
```

```{r, fig.cap="Average squared residual per bin v. fitted bin mean for Poisson model (L) and negative binomial model (R) with number of bins = 25 (observations per bin = 11). Visualizations displayed for a randomly chosen counties"}
par(mfrow=c(1,2))
i <- 2
Y <- ny_fl_df_t$DailyNewConfirmedCases[[i]]
spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
spline_mod.1.nb <- diy_spl_fit(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2), df = 3)
bin_it(fits=spline_mod.1.pois$fitted.values, y=Y[1:(length(Y)/2)], nbins=25)
abline(0,1)
abline(0,2, col=3)
bin_it(fits=spline_mod.1.nb$mu, y=Y[1:(length(Y)/2)], nbins=25)
abline(0,1)
abline(0, 2, col=3)

```

To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only `r sum(disp_stats <=1)` dispersion statistics were less than or equal to one. Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.

Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical.

```{r, fig.cap="Poisson and negative binomial fits to incidence data prior to Thanksgiving 2020 in a single county time series."}

time <- 1:length(spline_mod.1.nb$mu)
plot(time, Y[1:(length(Y)/2)], xlab="Day Index (Segment 1)", ylab="Incidence") 
lines(time, spline_mod.1.pois$fitted.values, col="red")
lines(time, spline_mod.1.nb$mu, col="blue")
legend(10, 10, legend=c("Poisson", "NB"),
       col=c("red", "blue"), lty=1)
```

Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), I inspected the mean-variance relationships using a diagnostic plot as in Ver Hoef and Boveng (2007).

In the quasi-Poisson model (called the linear negative binomial specification by Barron (1992)), the variance is a linear function of the mean, with a slope of one representing the Poisson special case. By contrast, the negative binomial model may more appropriately represent the data when the variance has approximately a quadratic relationship with the mean. Since the true mean and variance of the data set is unknown, we approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively. These quantities are estimated using both the Poisson and the negative binomial model. Variance can be estimated by averaging squared residuals in each fitted mean category. The number of fitted mean categories (bins) was chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present. Visualization of individual squared residuals is also possible, but these may not be reliable estimates of the variance.

## Proposed hypothesis test

```{r, fig.cap="Example of empirical increase in dispersion at the hypothesized Thanksgiving 2020 changepoint."}
plot(ny_df_t$DailyNewConfirmedCases[[14]] ~ inds, xaxt="n", ylab="Daily Incidence", xlab="Date")
abline(v=inds[30], col=2)
axis(1, inds, inds, cex.axis = .7)
```

We tested the hypothesis that the incidence dispersion parameter decreases at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$. Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020. The Wald test is used, where the estimates of $\theta$ in different segments are considered independent. This approach is possible because we use maximum likelihood estimates of $\theta$ (found by alternating between fitting a GLM for a given theta and then (for fixed means) estimating $\theta$). The standard error of $\theta$ is found utilizing the observed information (Venables and Ripley 2002).

$$Cov(\hat{\theta_1},\hat{\theta_2}) = 0 $$
The form of the Wald test statistic is:

$$\frac{\hat{\theta_1} -\hat{\theta_2}}{SE(\hat{\theta_1} -\hat{\theta_2})}$$
$$ = \frac{\hat{\theta_1} -\hat{\theta_2}}{\sqrt{\widehat{Var(\hat{\theta_1})} + \widehat{Var(\hat{\theta_2})}}}$$
# Results

## Thanksgiving 2020: New York and Florida

First, for every county, the model was fit separately to either side of November 25th, 2020 - where the 25th was included in the second segment -  in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$. A test of the one-sided upper alternative that the difference between before and after is positive (lower theta after Thanksgiving) was performed.

```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York and Florida."}

# Wald test for NY Thanksgiving in 2020 - doesn't converge for 13th and 50th NY counties
diff_thetas <- c()
t_ps <- c()
for (i in 1:dim(ny_df_t)[1]){ 
  if (i == 13 | i ==50){
    p = NA
    theta = NA
  }
  else{
    p <- W(y=ny_df_t$DailyNewConfirmedCases[[i]], pop=ny_df_t$populations[i], split=30)
    theta1 <- diy_spl_fit(Y=ny_df_t$DailyNewConfirmedCases[[i]], population =
                            ny_df_t$populations[i],
                          inds=1:(length(ny_df_t$DailyNewConfirmedCases[[i]])/2), df = 3)$theta
    theta2 <- diy_spl_fit(Y=ny_df_t$DailyNewConfirmedCases[[i]], population =
                            ny_df_t$populations[i],
                          inds=(length(ny_df_t$DailyNewConfirmedCases[[i]])/2
                                +1):length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$theta
  }
  t_ps <- c(t_ps, p)
  diff_thetas <- c(diff_thetas, abs(theta1 - theta2))
}
ny_df_t$p <- t_ps
ny_df_t$diff_thetas <- diff_thetas

# Wald test for FL Thanksgiving in 2020
diff_thetas <- c()
t_ps <- c()
for (i in 1:dim(fl_df_t)[1]){ 
  p <- W(y=fl_df_t$DailyNewConfirmedCases[[i]], pop=fl_df_t$populations[i], split=30)
  theta1 <- diy_spl_fit(Y=fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i],
                       inds=1:(length(fl_df_t$DailyNewConfirmedCases[[i]])/2), df = 3)$theta
  theta2 <- diy_spl_fit(Y=fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i],
                       inds=(length(fl_df_t$DailyNewConfirmedCases[[i]])/2 +
                               1):length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$theta
  t_ps <- c(t_ps, p)
  diff_thetas <- c(diff_thetas, abs(theta1-theta2))
}
fl_df_t$p <- t_ps
fl_df_t$diff_thetas <- diff_thetas

#p_adj_t <- p.adjust(t_ps, method = "BH")
par(mfrow=c(2,2))
# Histogram of p-values NY
hist(ny_df_t$p, col=2, breaks=20, xlab="p-values", main=NULL)
# Power lower based on population? NY
plot(ny_df_t$p ~ ny_df_t$populations)
# Histogram of p-values FL
hist(fl_df_t$p, col=2, breaks=20, xlab="p-values", main=NULL)
# Power lower based on population? FL
plot(fl_df_t$p ~ fl_df_t$populations)

```

```{r, fig.width=7, fig.height=7}
# Plot fitted curves for the counties with significant p-values

ny_indeces <- which(ny_df_t$p < 0.05)

par(mar=c(1,1,1,1))
par(mfrow=c(6,2))

for (i in ny_indeces){

mu <- diy_spl_fit(Y = ny_df_t$DailyNewConfirmedCases[[i]], population = ny_df_t$populations[i], inds=1:length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu

plot(1:length(ny_df_t$DailyNewConfirmedCases[[i]]), ny_df_t$DailyNewConfirmedCases[[i]], ylim=c(0,500))
lines(mu, col=2)
}

fl_indeces <- which(fl_df_t$p < 0.05)

par(mar=c(1,1,1,1))
par(mfrow=c(6,2))

for (i in fl_indeces){

mu <- diy_spl_fit(Y = fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i], inds=1:length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu

plot(1:length(fl_df_t$DailyNewConfirmedCases[[i]]), fl_df_t$DailyNewConfirmedCases[[i]])
lines(mu, col=2)
}

```
https://cran.r-project.org/web/packages/tigris/readme/README.html

```{r, echo=F, fig.cap="P-values for each New York county resulting from the test of the one-sided upper alternative that time series dispersion is greater after Thanksgiving 2020"}

ny <- counties("New York")

p <- round(ny_df_t$p, digits=1)
p[is.na(p)] <- "na"
ny$p <- p

ggplot(ny) +
  geom_sf(aes(fill = p))

fl <- counties("Florida")

p <- round(fl_df_t$p, digits=1)
p[is.na(p)] <- "na"
fl$p <- p

ggplot(fl) +
  geom_sf(aes(fill=p))

```
# Simulation

## Validity of the hypothesis test

To ensure that the hypothesis testing framework is valid, a simulation with known (equal) values of theta on either side of Thanksgiving 2020 was run in order to ensure that the distribution of resulting p-values from the test of the upper alternative is approximately uniform on the interval from zero to one. In both the validity and power simulations, representative epidemic curves are used. The area under these curves (final outbreak size) is set to be the population size divided by ten.

To implement these constraints on the integral:
we use a normalizing constant of (n/10) * sqrt(2*pi* $\sigma^2$)

$$ \int_0^\infty {(n/10)/\sqrt{2\pi(\sigma^2))} e^{-(t-61)^2/(2\sigma^2)}}dt = n/10$$
So, we have the desired Gaussian function:
$$ \mu_t = (n/10)/\sqrt{2\pi\sigma^2} e^{-(t-61)^2/(2\sigma^2)} $$
So that integration will result in a final outbreak size of n/10.

```{r, fig.cap = "Representative epidemic curve to inform the validity simulation and the power simulation."}

curve_parms <- data.frame(n = rep(mean(ny_df_t$populations), times = 300),  b = rep(30, times = 300), c = rep(200, times = 300), theta1 = c(seq(4, 10, length=100), seq(4, 10, length=100), seq(9, 3, length=100)), theta2 = c(seq(4, 10, length=100), seq(12, 6, length=100), seq(2, 8, length=100)))

plot(1:60, u_t(n = curve_parms$n[1], t = 1:60, b = curve_parms$b[1], c=curve_parms$c[1]), ylim = c(0, 1000))
```

```{r, fig.cap="Validity simulation p-values."}
# Wald test for simulated data - same process as was done to real time series from NY counties

splits <- c(20, 30, 40)

p_vals <- c()
diff_vals <- c()

for (k in splits){
  for (i in 1:100){

  sides <- rerun(sim_sides(parms=curve_parms[i,], time=1:60), .n=10)
  ps <- map(sides, ~ W(y=., pop=curve_parms$n[i], split=k, verbose=FALSE))

  p_vals <- c(p_vals, ps)
  diff_vals <-c(diff_vals, rep(abs(curve_parms[i,4]-curve_parms[i,5]), times=10))

  }
}

hist(unlist(p_vals), col=2, breaks=20, xlab="p-values", main=NULL)

```

## Power of the hypothesis test

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.

```{r, fig.cap="Power simulation p-values."}
p_pows <- c()
diff_pows <- c()

for (k in splits){
  for (i in 100:300){ # change to 100:300 for two-sided hyp test

    sides <- rerun(sim_sides(parms = curve_parms[i,], time=1:60), .n=10)
    ps <- map(sides, ~ W(y=., pop=curve_parms$n[i], split=k, verbose=FALSE))

    p_pows <- c(p_pows, ps)
    diff_pows <- c(diff_pows, rep((abs(curve_parms[i,4]-curve_parms[i,5])), times=10))

  }
}

# Histogram of p-values
hist(unlist(p_pows), col=3, breaks=20, xlab="p-values", main=NULL)
```

```{r}
plot(c(diff_vals, diff_pows), c(p_vals, p_pows))
```

The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean. As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.

## References

Adam, D., Gostic, K., Tsang, T., Wu, P., Lim, W. W., Yeung, A., Wong, J., Lau, E., Du, Z., Chen, D., Ho, L.-M., Martín-Sánchez, M., Cauchemez, S., Cobey, S., Leung, G., & Cowling, B. (2022). Time-varying transmission heterogeneity of SARS and COVID-19 in Hong Kong. https://doi.org/10.21203/rs.3.rs-1407962/v1

Cook, J. D. (n.d.). Notes on the Negative Binomial Distribution. 5.

Davis, R. A., & Wu, R. (2009). A negative binomial model for time series of counts. Biometrika, 96(3), 735–749.

Graham, M., Winter, A. K., Ferrari, M., Grenfell, B., Moss, W. J., Azman, A. S., Metcalf, C. J. E., & Lessler, J. (2019). Measles and the canonical path to elimination. Science, 364(6440), 584–587. https://doi.org/10.1126/science.aau6299

Karlis, D., & Xekalaki, E. (2000). A Simulation Comparison of Several Procedures for Testing the Poisson Assumption. Journal of the Royal Statistical Society: Series D (The Statistician), 49(3), 355–382. https://doi.org/10.1111/1467-9884.00240

Killick, R., & Eckley, I. A. (2014). changepoint: An R Package for Changepoint Analysis. Journal of Statistical Software, 58, 1–19. https://doi.org/10.18637/jss.v058.i03

Lloyd-Smith, J. O., Schreiber, S. J., Kopp, P. E., & Getz, W. M. (2005). Superspreading and the effect of individual variation on disease emergence. Nature, 438(7066), Article 7066. https://doi.org/10.1038/nature04153

Perperoglou, A., Sauerbrei, W., Abrahamowicz, M., & Schmid, M. (2019). A review of spline function procedures in R. BMC Medical Research Methodology, 19(1), 46. https://doi.org/10.1186/s12874-019-0666-3

Potthoff, R. F., & Whittinghill, M. (1966). Testing for homogeneity. II. The Poisson distribution. Biometrika, 53(1), 183–190.
Transmission heterogeneities, kinetics, and controllability of SARS-CoV-2. (n.d.). https://doi.org/10.1126/science.abe2424

Transmission heterogeneities, kinetics, and controllability of SARS-CoV-2. (n.d.). https://doi.org/10.1126/science.abe2424

Venables W.N., Ripley B.D. (2002). Modern Applied Statistics with S, Fourth edition. Springer, New York. ISBN 0-387-95457-0, https://www.stats.ox.ac.uk/pub/MASS4/.

Ver Hoef, J. M., & Boveng, P. L. (2007). QUASI-POISSON VS. NEGATIVE BINOMIAL REGRESSION: HOW SHOULD WE MODEL OVERDISPERSED COUNT DATA? Ecology, 88(11), 2766–2772. https://doi.org/10.1890/07-0043.1

Wallinga, J. (2018). Metropolitan versus small-town influenza. Science. https://doi.org/10.1126/science.aav1003

### 2023-05-11 noise_covid_v2.Rmd
remove excess code

@@
 ---
+
 ```{r setup, include=FALSE}
@@
                       fig.align = "center", fig.width = 5, fig.height = 5)
-set.seed(12345)
-devtools::load_all()
-source("R/simulation.R")
 ```
@@
 
-Approximately two months of data on either side of November 24th, 2020 was used in the analysis.
+Approximately one month of data on either side of November 24th, 2020 was used in the analysis.
 This step was done to include a set number of "waves" of incidence in the model fit to each side of the hypothesized change.
 However, we acknowledge that there is a trade-off between including few waves and increasing the power of the proposed test by including more time series data.
-
-```{r, echo=FALSE, fig.cap="COVID-19 county-level incidence in New York surrounding Thanksgiving 2020."}
-# USA case/population size data to work with
-
-data <- read.csv("data/confirmed_and_death_1209.csv")
-data$DateTime <- as.Date(data$DateTime)
-
-# NY dataframe for Thanksgiving analysis
-data <- data[data$DateTime >= "2020-10-26" & data$DateTime <= "2020-12-24", ]
-inds <- seq(as.Date("2020-10-26"), as.Date("2020-12-24"), by = "day")
-days <- weekdays(inds)
-inc <- aggregate(DailyNewConfirmedCases ~ countyFIPS, data, FUN = list)
-inc$populations <- tapply(X = data$population, INDEX = data$countyFIPS, FUN = mean)
-inc$stateFIPS <- as.numeric(substr(inc$countyFIPS, 1, 2))
-
-# all_zero <- c()
-# for (i in 1:dim(inc)[1]){
-#   if (all(unlist(inc$DailyNewConfirmedCases[i]) == 0)){
-#     print(i)
-#     all_zero <- c(all_zero, i)
-#   }
-# }
-# inc <- inc[-all_zero,] 
-ny_df_t <- inc[inc$stateFIPS == 36,]
-fl_df_t <- inc[inc$stateFIPS == 12,]
-ny_fl_df_t <- rbind(ny_df_t, fl_df_t)
-```
 
@@
 
-scaled_cases <- map(ny_df_t$DailyNewConfirmedCases, ~ as.vector(scale(.)))
+library(tidyverse)
+library(changepoint)
 
-ny_df_t$ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE)))
+filename <- "data/processed_dat.Rdata"
+load(filename)
+rm(filename)
 
-which(sapply(ny_df_t$ch_pts, length) == 0)
+scaled_cases <- map(as.data.frame(t(new_cases_subset)), ~ as.vector(scale(.)))
+
+ch_pts <- map(scaled_cases, ~ cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE))
 
@@
 
-ny_df_t$ch_pts2 <- map(ny_df_t$DailyNewConfirmedCases, ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))
+ny_df_t$ch_pts2 <- map(as.data.frame(new_cases_subset), ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))
 ny_df_t$n_ch_pts2 <- map(ny_df_t$ch_pts2, ~length(unlist(.)))
@@
 
-par(mfrow=c(1,2))
-# By visual inspection 
-plot(ny_df_t$DailyNewConfirmedCases[[1]] ~ inds, type="l", ylim=c(0, 5000), xaxt="n", ylab="Daily Incidence", xlab="Date")
-for (i in 2:dim(ny_df_t)[1]){
-  lines(ny_df_t$DailyNewConfirmedCases[[i]] ~ inds, col=sample(rainbow(10)))
+inds <- seq(start_date, end_date, by = "day")
+
+plot(new_cases_subset[state_index[1],], type = "l", xaxt = "n", ylab = "Daily Incidence", xlab = "Date")
+for (i in 2:length(state_index)){
+  lines(new_cases_subset[state_index[i],], col=sample(rainbow(10)))
 }
@@
 
+```{r}
+source("code/squared.res_v_squared.fitted.R")
+```
+
 ```{r, fig.cap="Average squared residual per bin v.
@@
 
-time <- 1:length(spline_mod.1.nb$mu)
-plot(time, Y[1:(length(Y)/2)], xlab="Day Index (Segment 1)", ylab="Incidence") 
-lines(time, spline_mod.1.pois$fitted.values, col="red")
-lines(time, spline_mod.1.nb$mu, col="blue")
-legend(10, 10, legend=c("Poisson", "NB"),
-       col=c("red", "blue"), lty=1)
+source("code/check_nb_estimation.R")
 ```

### 2023-05-12 noise_covid_v2.Rmd
prepare draft
source scripts within Rmd file to produce first draft

@@
 output:
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
 geometry: margin=2.54cm
@@
 
-For this project, we use county-level COVID-19 incidence data. Source: [USAFacts](https://usafacts.org/). The data is weekly case counts by US county.
+For this project, we use county-level COVID-19 incidence data.
+Source: [USAFacts](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/).
+The data is weekly case counts for each US county.
 
-Approximately one month of data on either side of November 24th, 2020 was used in the analysis.
+Approximately one month of data on either side of November 26th, 2020 was used in the analysis.
 This step was done to include a set number of "waves" of incidence in the model fit to each side of the hypothesized change.
@@
 ```{r}
+library(knitr)
 rat <- data.frame(Mean = c(5, 10), Ratio = c(2.236068, 3.162278))
@@
 ```{r, fig.cap="Average Pothoff-Whittinghill (PW) statistic for each NY county plotted against county population size."}
-
 library(tidyverse)
 library(changepoint)
+source("code/functions_intro.R")
 
@@
 
-ch_pts <- map(scaled_cases, ~ cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE))
+ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE)))
 
-segments <- map2(.x = ny_df_t$DailyNewConfirmedCases[-c(2, 8, 21, 49, 54, 62)], .y = ny_df_t$ch_pts[-c(2, 8, 21, 49, 54, 62)], ~ get_segments(series = .x, changepoints = .y))
+segments <- map2(.x = as.data.frame(t(new_cases_subset[which(sapply(ch_pts, length) != 0),])), .y = ch_pts[which(sapply(ch_pts, length) != 0)], ~ get_segments(series = .x, changepoints = .y))
 
@@
 
-plot(log(ny_df_t$populations[-c(2, 8, 21, 49, 54, 62)]), log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2)
+plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0,1000000))
 ```
-
 Methods designed specifically for count data show a similar dependence on population size.
@@
 
-ny_df_t$ch_pts2 <- map(as.data.frame(new_cases_subset), ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))
-ny_df_t$n_ch_pts2 <- map(ny_df_t$ch_pts2, ~length(unlist(.)))
+ch_pts2 <- map(as.data.frame(t(new_cases_subset)), ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))
 
-plot(log(ny_df_t$populations), log(unlist(ny_df_t$n_ch_pts2)), xlab="Log Population Size", ylab=" Log Number of Changepoints", col=2)
+n_ch_pts2 <- map(ch_pts2, ~length(unlist(.)))
+
+plot(populations$population, log(unlist(n_ch_pts2)), xlab="Log Population Size", ylab=" Log Number of Changepoints", col=2)
 ```
-
 ## Modeling proccess mean
@@
 We propose a generalized linear model (GLM) approach to quantify incidence dispersion in segments of a time series, using a single parameter, $\theta$, while separately accounting for population size in the model.
-Another approach is to fit one model to the entire time series accounting for autocorrelation using a quasi-likelihood approach using a specified matrix V (Barron 1992).
-We account for unobserved heterogeneity, contagion, and time dependence in rate through specifying a changing process mean (to account for autocorrelation) and specifying a overdispersed conditional distribution (to account for overdispersion).
+Here we account for unobserved heterogeneity, and time dependence in rate - potentially due to contagion - (Barron 1992) by specifying an overdispersed conditional distribution.
 
 A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
-The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a generalized additive model (GAM) with one explanatory variable, time.
-In the GAM framework, the function for a predictor can be represented using spline functions:
+The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time.
+Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al.
+2019).
 
- $$g(E[Y_i]) = f(t_i) = \sum_m^M\beta_m h_m(t_i)$$
- $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
+  $$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$ $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
+  ,where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
 
-Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). The degrees of freedom were chosen due to **
-
-$$g(E[Y_i]) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+In this case, we use a log link function:
 $$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-```{r, fig.cap="County case count time series for each New York county with the hypothesized Thanksgiving changepoint represented by the black vertical line (L), and total New York COVID-19 case count time series (R)."}
-
-inds <- seq(start_date, end_date, by = "day")
-
-plot(new_cases_subset[state_index[1],], type = "l", xaxt = "n", ylab = "Daily Incidence", xlab = "Date")
-for (i in 2:length(state_index)){
-  lines(new_cases_subset[state_index[i],], col=sample(rainbow(10)))
-}
-abline(v=inds[30], lwd=2)
-axis(1, inds, inds, cex.axis = .7)
-
-ny_col_sum <- matrix(unlist(ny_df_t$DailyNewConfirmedCases), nrow=62,
-                     ncol=length(ny_df_t$DailyNewConfirmedCases[[1]]))
-ny_cases <- colSums(ny_col_sum)
-plot(ny_cases ~ inds, type="l", col=6, xlab="Date", ylab="Daily Incidence")
-abline(v=inds[30], lwd=2)
-```
-
-```{r, fig.cap="County case count time series for each Florida county with the hypothesized Thanksgiving changepoint represented by the black vertical line (L), and total Florida COVID-19 case count time series (R)."}
-
-par(mfrow=c(1,2))
-# By visual inspection 
-plot(fl_df_t$DailyNewConfirmedCases[[1]] ~ inds, type="l", ylim=c(0, 5000), xaxt="n", ylab="Daily Incidence", xlab="Date")
-for (i in 2:dim(fl_df_t)[1]){
-  lines(fl_df_t$DailyNewConfirmedCases[[i]] ~ inds, col=sample(rainbow(10)))
-}
-abline(v=inds[30], lwd=2)
-axis(1, inds, inds, cex.axis = .7)
-
-fl_col_sum <- matrix(unlist(fl_df_t$DailyNewConfirmedCases), nrow=62,
-                     ncol=length(fl_df_t$DailyNewConfirmedCases[[1]]))
-fl_cases <- colSums(fl_col_sum)
-plot(fl_cases ~ inds, type="l", col=6, xlab="Date", ylab="Daily Incidence")
-abline(v=inds[30], lwd=2)
-```
+  $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+  $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
 ## IRLS
 
-Briefly, what iteratively reweighted least squares (IRLS) does is compute an adjusted dependent variable, compute a weight matrix, and compute (using weighted least squares) coefficients and therefore the linear predictor as well.
-
-In irls.nb.1 in the NBPSeq R package (*cite):
-
-1. Initial mu (either supplied or computed) and eta = log(mu/s).
-2. z = eta - offset + (y - mu)/mu
-3. w = drop(\mu/sqrt(varmu))
-4. call Fortran code to perform weighted LS. Weighted least squares produces: $\hat{\beta} = (X^TWX)^{-1}X^TWY$, where weight is based on the variance of the observation. 
-5. Then new values of mu and eta from new coefficients.
-
-Within diy_spl_fit, fit.nb.regression is run and the standard error of param theta is computed by inverting info matrix, take [1,1] entry and take square root.
-Within fit.nb.regression, kappa is optimized, then irls.nb.1 (the IRLS portion) is called, then additional output - like the information matrix - is added.
+Iteratively reweighted least squares (IRLS) is used to obtain parameter estimates of the model.
+Briefly, IRLS works by computing an adjusted dependent variable, computing a weight matrix, and then calculating (using weighted least squares) coefficients estimates of the model used to calculate the linear predictor as well.
+This procedure is implemented via the irls.nb.1 and functions from the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al.
+(2011).
 
@@
 
-Note that if the offset model were perfect, the estimated coefficient on log population would be exactly one:
-$$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
-, where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
-
-In this case, a log link function is used:
-
-  $$log(E[Y_i])-log(n_i) = \sum_m^M\beta_m h_m(t_i)$$
-  $$log(E[Y_i]) =  \sum_m^M\beta_m h_m(t_i) + log(n_i)$$
+Note that if the offset model were perfect, the estimated coefficient on log population would be exactly one.
 
@@
 
-```{r}
-disp_stats <- c()
-for (i in 1:dim(ny_fl_df_t)[1]){
-  Y <- ny_fl_df_t$DailyNewConfirmedCases[[i]]
-  spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
-  disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
-  #print(summary(spline_mod.1.nb)$coefficients[,4])
-}
-```
+# ```{r}
+# disp_stats <- c()
+# for (i in 1:dim(new_cases_subset[state_index])[1]){
+#   Y <- new_cases_subset[i,]
+#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
+#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
+# }
+# ```
+
+To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only a few dispersion statistics were less than or equal to one.
+Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
 
@@
 
-```{r, fig.cap="Average squared residual per bin v.
-fitted bin mean for Poisson model (L) and negative binomial model (R) with number of bins = 25 (observations per bin = 11).
-Visualizations displayed for a randomly chosen counties"}
-par(mfrow=c(1,2))
-i <- 2
-Y <- ny_fl_df_t$DailyNewConfirmedCases[[i]]
-spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
-spline_mod.1.nb <- diy_spl_fit(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2), df = 3)
-bin_it(fits=spline_mod.1.pois$fitted.values, y=Y[1:(length(Y)/2)], nbins=25)
-abline(0,1)
-abline(0,2, col=3)
-bin_it(fits=spline_mod.1.nb$mu, y=Y[1:(length(Y)/2)], nbins=25)
-abline(0,1)
-abline(0, 2, col=3)
-
-```
-
-To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only `r sum(disp_stats <=1)` dispersion statistics were less than or equal to one.
-Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
-
 Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical.
-
-```{r, fig.cap="Poisson and negative binomial fits to incidence data prior to Thanksgiving 2020 in a single county time series."}
-
-source("code/check_nb_estimation.R")
-```
 
@@
 ```{r, fig.cap="Example of empirical increase in dispersion at the hypothesized Thanksgiving 2020 changepoint."}
-plot(ny_df_t$DailyNewConfirmedCases[[14]] ~ inds, xaxt="n", ylab="Daily Incidence", xlab="Date")
-abline(v=inds[30], col=2)
-axis(1, inds, inds, cex.axis = .7)
+plot(new_cases_subset[1842,], xaxt="n", ylab="Daily Incidence", xlab="Date") #~ inds
+abline(v=30, col=2)
+#axis(1, inds, inds, cex.axis = .7)
 ```
@@
 The Wald test is used, where the estimates of $\theta$ in different segments are considered independent.
-This approach is possible because we use maximum likelihood estimates of $\theta$ (found by alternating between fitting a GLM for a given theta and then (for fixed means) estimating $\theta$).
-The standard error of $\theta$ is found utilizing the observed information (Venables and Ripley 2002).
+This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information.
 
@@
 
-## Thanksgiving 2020: New York and Florida
+## Thanksgiving 2020: Wald test
 
-First, for every county, the model was fit separately to either side of November 25th, 2020 - where the 25th was included in the second segment -  in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
+First, for every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
 A test of the one-sided upper alternative that the difference between before and after is positive (lower theta after Thanksgiving) was performed.
 
-```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York and Florida."}
-
-# Wald test for NY Thanksgiving in 2020 - doesn't converge for 13th and 50th NY counties
-diff_thetas <- c()
-t_ps <- c()
-for (i in 1:dim(ny_df_t)[1]){ 
-  if (i == 13 | i ==50){
-    p = NA
-    theta = NA
-  }
-  else{
-    p <- W(y=ny_df_t$DailyNewConfirmedCases[[i]], pop=ny_df_t$populations[i], split=30)
-    theta1 <- diy_spl_fit(Y=ny_df_t$DailyNewConfirmedCases[[i]], population =
-                            ny_df_t$populations[i],
-                          inds=1:(length(ny_df_t$DailyNewConfirmedCases[[i]])/2), df = 3)$theta
-    theta2 <- diy_spl_fit(Y=ny_df_t$DailyNewConfirmedCases[[i]], population =
-                            ny_df_t$populations[i],
-                          inds=(length(ny_df_t$DailyNewConfirmedCases[[i]])/2
-                                +1):length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$theta
-  }
-  t_ps <- c(t_ps, p)
-  diff_thetas <- c(diff_thetas, abs(theta1 - theta2))
-}
-ny_df_t$p <- t_ps
-ny_df_t$diff_thetas <- diff_thetas
-
-# Wald test for FL Thanksgiving in 2020
-diff_thetas <- c()
-t_ps <- c()
-for (i in 1:dim(fl_df_t)[1]){ 
-  p <- W(y=fl_df_t$DailyNewConfirmedCases[[i]], pop=fl_df_t$populations[i], split=30)
-  theta1 <- diy_spl_fit(Y=fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i],
-                       inds=1:(length(fl_df_t$DailyNewConfirmedCases[[i]])/2), df = 3)$theta
-  theta2 <- diy_spl_fit(Y=fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i],
-                       inds=(length(fl_df_t$DailyNewConfirmedCases[[i]])/2 +
-                               1):length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$theta
-  t_ps <- c(t_ps, p)
-  diff_thetas <- c(diff_thetas, abs(theta1-theta2))
-}
-fl_df_t$p <- t_ps
-fl_df_t$diff_thetas <- diff_thetas
-
-#p_adj_t <- p.adjust(t_ps, method = "BH")
-par(mfrow=c(2,2))
-# Histogram of p-values NY
-hist(ny_df_t$p, col=2, breaks=20, xlab="p-values", main=NULL)
-# Power lower based on population? NY
-plot(ny_df_t$p ~ ny_df_t$populations)
-# Histogram of p-values FL
-hist(fl_df_t$p, col=2, breaks=20, xlab="p-values", main=NULL)
-# Power lower based on population? FL
-plot(fl_df_t$p ~ fl_df_t$populations)
-
+```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
+source("code/wald_counties.R")
+```
+```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
+source("code/LRT_counties.R")
 ```
 
-```{r, fig.width=7, fig.height=7}
-# Plot fitted curves for the counties with significant p-values
-
-ny_indeces <- which(ny_df_t$p < 0.05)
-
-par(mar=c(1,1,1,1))
-par(mfrow=c(6,2))
-
-for (i in ny_indeces){
-
-mu <- diy_spl_fit(Y = ny_df_t$DailyNewConfirmedCases[[i]], population = ny_df_t$populations[i], inds=1:length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
-
-plot(1:length(ny_df_t$DailyNewConfirmedCases[[i]]), ny_df_t$DailyNewConfirmedCases[[i]], ylim=c(0,500))
-lines(mu, col=2)
-}
-
-fl_indeces <- which(fl_df_t$p < 0.05)
-
-par(mar=c(1,1,1,1))
-par(mfrow=c(6,2))
-
-for (i in fl_indeces){
-
-mu <- diy_spl_fit(Y = fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i], inds=1:length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
-
-plot(1:length(fl_df_t$DailyNewConfirmedCases[[i]]), fl_df_t$DailyNewConfirmedCases[[i]])
-lines(mu, col=2)
-}
-
-```
-https://cran.r-project.org/web/packages/tigris/readme/README.html
-
-```{r, echo=F, fig.cap="P-values for each New York county resulting from the test of the one-sided upper alternative that time series dispersion is greater after Thanksgiving 2020"}
-
-ny <- counties("New York")
-
-p <- round(ny_df_t$p, digits=1)
-p[is.na(p)] <- "na"
-ny$p <- p
-
-ggplot(ny) +
-  geom_sf(aes(fill = p))
-
-fl <- counties("Florida")
-
-p <- round(fl_df_t$p, digits=1)
-p[is.na(p)] <- "na"
-fl$p <- p
-
-ggplot(fl) +
-  geom_sf(aes(fill=p))
-
-```
 # Simulation
 
-## Validity of the hypothesis test
+## Validity and power of the Wald test
 
@@
 
-```{r, fig.cap = "Representative epidemic curve to inform the validity simulation and the power simulation."}
-
-curve_parms <- data.frame(n = rep(mean(ny_df_t$populations), times = 300),  b = rep(30, times = 300), c = rep(200, times = 300), theta1 = c(seq(4, 10, length=100), seq(4, 10, length=100), seq(9, 3, length=100)), theta2 = c(seq(4, 10, length=100), seq(12, 6, length=100), seq(2, 8, length=100)))
-
-plot(1:60, u_t(n = curve_parms$n[1], t = 1:60, b = curve_parms$b[1], c=curve_parms$c[1]), ylim = c(0, 1000))
+```{r, fig.cap="Wald simulation p-values."}
+source("code/validity_and_power_Wald.R")
 ```
 
-```{r, fig.cap="Validity simulation p-values."}
-# Wald test for simulated data - same process as was done to real time series from NY counties
-
-splits <- c(20, 30, 40)
-
-p_vals <- c()
-diff_vals <- c()
-
-for (k in splits){
-  for (i in 1:100){
-
-  sides <- rerun(sim_sides(parms=curve_parms[i,], time=1:60), .n=10)
-  ps <- map(sides, ~ W(y=., pop=curve_parms$n[i], split=k, verbose=FALSE))
-
-  p_vals <- c(p_vals, ps)
-  diff_vals <-c(diff_vals, rep(abs(curve_parms[i,4]-curve_parms[i,5]), times=10))
-
-  }
-}
-
-hist(unlist(p_vals), col=2, breaks=20, xlab="p-values", main=NULL)
-
-```
-
-## Power of the hypothesis test
-
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
-
-```{r, fig.cap="Power simulation p-values."}
-p_pows <- c()
-diff_pows <- c()
-
-for (k in splits){
-  for (i in 100:300){ # change to 100:300 for two-sided hyp test
-
-    sides <- rerun(sim_sides(parms = curve_parms[i,], time=1:60), .n=10)
-    ps <- map(sides, ~ W(y=., pop=curve_parms$n[i], split=k, verbose=FALSE))
-
-    p_pows <- c(p_pows, ps)
-    diff_pows <- c(diff_pows, rep((abs(curve_parms[i,4]-curve_parms[i,5])), times=10))
-
-  }
-}
-
-# Histogram of p-values
-hist(unlist(p_pows), col=3, breaks=20, xlab="p-values", main=NULL)
-```
-
-```{r}
-plot(c(diff_vals, diff_pows), c(p_vals, p_pows))
-```
 
@@
 Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.
+
+## Validity and power of the LRT
+
+```{r, fig.cap="Wald simulation p-values."}
+source("code/validity_and_power_LRT.R")
+```
 

### 2023-05-12 code/functions_intro.R

@@ -1,0 +1,29 @@
+# Divide a time series given a set of changepoints
+
+get_segments <- function(series, changepoints){
+  pts <- c(1,changepoints)
+  segments <- rep(NA, times = length(pts)-1)
+  for (i in 1:length(segments)){
+    segments[i] <- list(series[pts[i]:(pts[i+1]-1)])
+  }
+  return(segments)
+}
+
+# Get (mean) pw stat. (could pass a known set of means) given list of vectors
+get_mean_pw <- function(segments){  
+  pws = rep(NA, length(segments))
+  for (j in 1:length(segments)){
+    segment = segments[[j]]
+    n = length(segment)
+    pws[j] = sum(segment^2)/mean(segment) - sum(segment)
+  } 
+  return(mean(pws, na.rm = TRUE))  
+}
+
+# Fit Poisson model to time series segment
+spl_fit_pois <- function(Y, population, inds){ 
+  y = Y[inds]
+  pop_vec = rep(population, times=length(y))
+  fit = glm(y ~ gam(Y[inds] ~ s(inds))$fitted.values + offset(log(pop_vec)), family=poisson(link=log))
+  return(fit)
+}

### 2023-05-12 code/lrt_counties.R

@@ -20,3 +20,6 @@ for (i in state_index){
 # Save NY p-values
 filename <- "data/LRT_pvals_NY.Rdata"
 save(pvals, file = filename)
+
+# Visualize p-values
+hist(pvals, col=2)

### 2023-05-12 code/nb_estimation.R

@@ -1,6 +1,6 @@
 library(NBPSeq)
 
-# TODO: citation or link
+# Di et al. (2011)
 fit.nb.regression = function(y, s, x, beta0=rep(NA, dim(x)[2]), kappa=NA) {
   
   ## Find preliminary estiamtes of mu assuming phi=1. Will serve as

### 2023-05-12 code/process_COVID_data.R

@@ -49,7 +49,9 @@ end_date <- american_thanksgiving + 30
 dates_subset <- dates[dates >= start_date & dates <= end_date]
 new_cases_subset <- new_cases[, dates >= start_date & dates <= end_date]
 new_cases_subset[which(new_cases_subset < 0)] <- 0
+
 new_cases_subset <- new_cases_subset[which(rowSums(new_cases_subset) != 0),]
+populations <- populations[which(rowSums(new_cases_subset) != 0),]
 
 # Save curves
 filename <- "data/processed_dat.Rdata"

### 2023-05-12 code/squared.res_v_squared.fitted.R

@@ -21,7 +21,8 @@ colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 
 # Plot squared residuals against squared fitted values
-plot(state_fits$mu[[1]]^2, (unlist(state_fits[1,8:67]) - state_fits$mu[[1]])^2)
+plot(state_fits$mu[[1]]^2, (unlist(state_fits[1,8:67]) - state_fits$mu[[1]])^2, xlab = 
+       "Squared Fitted Values", ylab = "Squared Residuals")
 for (i in 2:length(state_index)){
   points(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
          col = sample(rainbow(i)))

### 2023-05-12 code/statemap_pvals.R

@@ -1,0 +1,25 @@
+# Map a state's p-values
+
+https://cran.r-project.org/web/packages/tigris/readme/README.html
+
+```{r, echo=F, fig.cap="P-values for each New York county resulting from the test of the one-sided upper alternative that time series dispersion is greater after Thanksgiving 2020"}
+
+ny <- counties("New York")
+
+p <- round(ny_df_t$p, digits=1)
+p[is.na(p)] <- "na"
+ny$p <- p
+
+ggplot(ny) +
+  geom_sf(aes(fill = p)) 
+
+fl <- counties("Florida")
+
+p <- round(fl_df_t$p, digits=1)
+p[is.na(p)] <- "na"
+fl$p <- p 
+
+ggplot(fl) +
+  geom_sf(aes(fill=p))
+
+```

### 2023-05-12 code/validity_and_power_LRT.R

@@ -19,3 +19,8 @@ for (i in 1:ncurve) {
 # Save p-values
 filename <- "data/pvals_sim_LRT.Rdata"
 save(pvals, file = filename)
+
+# Visualize p-values
+par(mfrow = c(1,2))
+hist(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
+hist(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)

### 2023-05-12 code/validity_and_power_Wald.R

@@ -19,3 +19,8 @@ for (i in 1:ncurve) {
 # Save p-values
 filename <- "data/pvals_sim_Wald.Rdata"
 save(pvals, file = filename)
+
+# Visualize p-values
+par(mfrow = c(1,2))
+hist(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
+hist(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)

### 2023-05-12 code/vis_signific_state.R

@@ -1,0 +1,27 @@
+# Visualize fitted curves in a state's counties where significant Wald result observed
+
+ny_indeces <- which(ny_df_t$p < 0.05)
+
+par(mar=c(1,1,1,1))
+par(mfrow=c(6,2))
+
+for (i in ny_indeces){
+  
+  mu <- diy_spl_fit(Y = ny_df_t$DailyNewConfirmedCases[[i]], population = ny_df_t$populations[i], inds=1:length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
+  
+  plot(1:length(ny_df_t$DailyNewConfirmedCases[[i]]), ny_df_t$DailyNewConfirmedCases[[i]], ylim=c(0,500))
+  lines(mu, col=2)
+}
+
+fl_indeces <- which(fl_df_t$p < 0.05)
+
+par(mar=c(1,1,1,1))
+par(mfrow=c(6,2))
+
+for (i in fl_indeces){
+  
+  mu <- diy_spl_fit(Y = fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i], inds=1:length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
+  
+  plot(1:length(fl_df_t$DailyNewConfirmedCases[[i]]), fl_df_t$DailyNewConfirmedCases[[i]])
+  lines(mu, col=2)
+}

### 2023-05-12 code/wald_counties.R

@@ -19,3 +19,6 @@ for (i in state_index){
 # Save NY p-values
 filename <- "data/W_pvals_NY.Rdata"
 save(pvals, file = filename)
+
+# Visualize p-values
+hist(pvals, col=2)

### 2023-05-12 noise_covid_v2.Rmd
scripts for visualization

@@
 
-# ```{r}
-# disp_stats <- c()
-# for (i in 1:dim(new_cases_subset[state_index])[1]){
-#   Y <- new_cases_subset[i,]
-#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
-#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
-# }
-# ```
-
 To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only a few dispersion statistics were less than or equal to one.
@@
 ```{r}
+par(mfrow = c(1,2))
+state <- "NY"
+source("code/squared.res_v_squared.fitted.R")
+
+state <- "FL"
 source("code/squared.res_v_squared.fitted.R")
@@
 The Wald test is used, where the estimates of $\theta$ in different segments are considered independent.
-This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information.
+This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information matrix.
 
@@
 ```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
+par(mfrow = c(1,2))
+state <- "NY"
+source("code/wald_counties.R")
+
+state <- "FL"
 source("code/wald_counties.R")
@@
 ```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
-source("code/LRT_counties.R")
+par(mfrow = c(1,2))
+state <- "NY"
+source("code/lrt_counties.R")
+NY_sig <- (which(pvals < 0.05))
+
+state <- "FL"
+source("code/lrt_counties.R")
+FL_sig <- print(which(pvals < 0.05))
+
 ```
@@
 ```
-
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
-
 The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean.
@@
 ```
+There appears to be slight Type I error inflation in both tests.
 

### 2023-05-12 code/disp_stats.R

@@ -1,0 +1,8 @@
+# ```{r}
+# disp_stats <- c()
+# for (i in 1:dim(new_cases_subset[state_index])[1]){
+#   Y <- new_cases_subset[i,]
+#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
+#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
+# }
+# `

### 2023-05-12 code/lrt_counties.R

@@ -5,8 +5,6 @@ filename <- "data/processed_dat.RData"
 load(filename)
 rm(filename)
 
-state <- "NY"
-
 # Loop through counties and calculate p values
 state_index <- which(populations$State == state)
 pvals <- rep(NA, times = length(state_index))

### 2023-05-12 code/squared.res_v_squared.fitted.R

@@ -1,32 +1,40 @@
-# Check whether dispersion well represented by negative binomial model 
+# Check whether dispersion well represented by negative binomial model - just using the counties 
+# with significant lrt result
 
 source("code/W.R")
 filename <- "data/processed_dat.RData"
 load(filename)
 rm(filename)
 
-state <- "NY"
-state_index <- which(populations$State == state)
+signif_index <- c(1842, 1848, 1853, 1862, 1874, 1881, 1883, 1885, 323, 324, 329, 330, 364, 375)
 state_fits <- c()
 
-for (i in state_index){
+for (i in signif_index){
   state_fits <- c(state_fits, my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
                               inds = 1:60, df = 3))
 }
 
-state_fits <- matrix(state_fits, nrow = length(state_index), ncol=7, byrow = TRUE)
-state_fits <- cbind(state_fits, new_cases_subset[state_index,])
+state_fits <- matrix(state_fits, nrow = length(signif_index), ncol=7, byrow = TRUE)
+state_fits <- cbind(state_fits, new_cases_subset[signif_index,])
 state_fits <- as.data.frame(state_fits)
 colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 
 # Plot squared residuals against squared fitted values
-plot(state_fits$mu[[1]]^2, (unlist(state_fits[1,8:67]) - state_fits$mu[[1]])^2, xlab = 
-       "Squared Fitted Values", ylab = "Squared Residuals")
-for (i in 2:length(state_index)){
-  points(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
-         col = sample(rainbow(i)))
+par(mfrow = c(2,4))
+for (i in 1:8){   #NY
+  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
+       col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
+  
 }
+
+par(mfrow = c(2,3))
+for (i in 9:14){   #FL
+  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
+       col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
+  
+}
+
[…]

### 2023-05-12 code/wald_counties.R

@@ -5,10 +5,14 @@ filename <- "data/processed_dat.RData"
 load(filename)
 rm(filename)
 
-state <- "NY"
-
 # Loop through counties and calculate p values
 state_index <- which(populations$State == state)
+
+# Estimation issue for state_index = 385 (FL)
+if (state == "FL"){
+  state_index <- state_index[-c(which(state_index == 385))]
+}
+
 pvals <- rep(NA, times = length(state_index))
 
 for (i in state_index){

### 2023-05-18 noise_covid_v2.Rmd
Edit scripts sourced in Rmd file

@@
   $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
+
 ## IRLS
@@
 Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
-
-```{r}
-par(mfrow = c(1,2))
-state <- "NY"
-source("code/squared.res_v_squared.fitted.R")
-
-state <- "FL"
-source("code/squared.res_v_squared.fitted.R")
-```
 
@@
 source("code/lrt_counties.R")
-FL_sig <- print(which(pvals < 0.05))
+FL_sig <- (which(pvals < 0.05))
+```
 
+```{r}
+signif_index <- c(NY_sig, FL_sig)
+
+state <- "NY"
+source("code/squared.res_v_squared.fitted.R")
+
+state <- "FL"
+source("code/squared.res_v_squared.fitted.R")
+
+source("code/vis_signific_state.R")
+```
+```{r}
+# disp_stats <- c()
+# for (i in 1:dim(new_cases_subset[state_index])[1]){
+#   Y <- new_cases_subset[i,]
+#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
+#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
+# }
+```
+
+```{r}
+# library(tigris)
+# p <- round(fl_df_t$p, digits=1)
+# p[is.na(p)] <- "na"
+# fl$p <- p 
+# 
+# ggplot(fl) +
+#   geom_sf(aes(fill=p))
 ```

### 2023-05-18 code/disp_stats.R

@@ -1,8 +1,0 @@
-# ```{r}
-# disp_stats <- c()
-# for (i in 1:dim(new_cases_subset[state_index])[1]){
-#   Y <- new_cases_subset[i,]
-#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
-#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
-# }
-# `

### 2023-05-18 code/lrt_counties.R

@@ -15,8 +15,8 @@ for (i in state_index){
                   i2 = 31:60, df1 = 3, df2 = 3)$p
 }
 
-# Save NY p-values
-filename <- "data/LRT_pvals_NY.Rdata"
+# Save p-values
+filename <- paste("data/LRT_pvals", state, ".Rdata", sep = "")
 save(pvals, file = filename)
 
 # Visualize p-values

### 2023-05-18 code/squared.res_v_squared.fitted.R

@@ -6,7 +6,6 @@ filename <- "data/processed_dat.RData"
 load(filename)
 rm(filename)
 
-signif_index <- c(1842, 1848, 1853, 1862, 1874, 1881, 1883, 1885, 323, 324, 329, 330, 364, 375)
 state_fits <- c()
 
 for (i in signif_index){

### 2023-05-18 code/statemap_pvals.R

@@ -1,25 +1,0 @@
-# Map a state's p-values
-
-https://cran.r-project.org/web/packages/tigris/readme/README.html
-
-```{r, echo=F, fig.cap="P-values for each New York county resulting from the test of the one-sided upper alternative that time series dispersion is greater after Thanksgiving 2020"}
-
-ny <- counties("New York")
-
-p <- round(ny_df_t$p, digits=1)
-p[is.na(p)] <- "na"
-ny$p <- p
-
-ggplot(ny) +
-  geom_sf(aes(fill = p)) 
-
-fl <- counties("Florida")
-
-p <- round(fl_df_t$p, digits=1)
-p[is.na(p)] <- "na"
-fl$p <- p 
-
-ggplot(fl) +
-  geom_sf(aes(fill=p))
-
-```

### 2023-05-18 code/vis_signific_state.R

@@ -1,27 +1,10 @@
 # Visualize fitted curves in a state's counties where significant Wald result observed
 
-ny_indeces <- which(ny_df_t$p < 0.05)
-
-par(mar=c(1,1,1,1))
-par(mfrow=c(6,2))
-
-for (i in ny_indeces){
+for (i in signif_index){
   
-  mu <- diy_spl_fit(Y = ny_df_t$DailyNewConfirmedCases[[i]], population = ny_df_t$populations[i], inds=1:length(ny_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
+  mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
+                   inds=1:length(new_cases_subset[i,]), df = 3)$mu
   
-  plot(1:length(ny_df_t$DailyNewConfirmedCases[[i]]), ny_df_t$DailyNewConfirmedCases[[i]], ylim=c(0,500))
-  lines(mu, col=2)
-}
-
-fl_indeces <- which(fl_df_t$p < 0.05)
-
-par(mar=c(1,1,1,1))
-par(mfrow=c(6,2))
-
-for (i in fl_indeces){
-  
-  mu <- diy_spl_fit(Y = fl_df_t$DailyNewConfirmedCases[[i]], population = fl_df_t$populations[i], inds=1:length(fl_df_t$DailyNewConfirmedCases[[i]]), df = 3)$mu
-  
-  plot(1:length(fl_df_t$DailyNewConfirmedCases[[i]]), fl_df_t$DailyNewConfirmedCases[[i]])
+  plot(1:length(new_cases_subset[i,]), new_cases_subset[i,])
   lines(mu, col=2)
 }

### 2023-05-18 code/wald_counties.R

@@ -20,8 +20,8 @@ for (i in state_index){
                 deg_free = 3, fn = my_spl_fit, verbose = FALSE)
 }
 
-# Save NY p-values
-filename <- "data/W_pvals_NY.Rdata"
+# Save p-values
+filename <- paste("data/W_pvals", state, ".Rdata", sep ="")
 save(pvals, file = filename)
 
 # Visualize p-values

### 2023-05-20 noise_covid_v2.Rmd
debug all

@@
 
-plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0,1000000))
+plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0, 1000000))
 ```
@@
 ```
+```{r}
+library(tigris)
+fl <- counties("Florida")
+load("data/W_pvalsFL.Rdata")
+p <- round(pvals, digits=1)
+p[is.na(p)] <- "na"
+fl$p <- p
+
+ggplot(fl) +
+  geom_sf(aes(fill = p)) 
+```
+
 ```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
@@
 ```
-
+Note that in these visualizations, an NB2 fit would be:
+A Poisson fit would be:
 ```{r}
@@
 source("code/vis_signific_state.R")
-```
-```{r}
-# disp_stats <- c()
-# for (i in 1:dim(new_cases_subset[state_index])[1]){
-#   Y <- new_cases_subset[i,]
-#   spline_mod.1.pois <- spl_fit_pois(Y = Y, population=ny_fl_df_t$populations[i], inds=1:(length(Y)/2))
-#   disp_stats <- c(disp_stats, spline_mod.1.pois$deviance/spline_mod.1.pois$df.residual)
-# }
-```
-
-```{r}
-# library(tigris)
-# p <- round(fl_df_t$p, digits=1)
-# p[is.na(p)] <- "na"
-# fl$p <- p 
-# 
-# ggplot(fl) +
-#   geom_sf(aes(fill=p))
 ```
@@
 There appears to be slight Type I error inflation in both tests.
+
+```{r}
+load("data/pvals_sim_Wald.Rdata")
+mean(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
+mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)
+
+load("data/pvals_sim_LRT.Rdata")
+mean(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
+mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)
+
+```
 

### 2023-05-20 code/lrt_counties.R

@@ -7,11 +7,12 @@ rm(filename)
 
 # Loop through counties and calculate p values
 state_index <- which(populations$State == state)
+
 pvals <- rep(NA, times = length(state_index))
 
-for (i in state_index){
-  pvals[i] <- lrt(y1 = new_cases_subset[i,][1:30], y2 = new_cases_subset[i,][31:60], 
-                  s1 = populations$population[i], s2 = populations$population[i], i1 = 1:30, 
+for (i in 1:length(state_index)){
+  pvals[i] <- lrt(y1 = new_cases_subset[state_index[i],][1:30], y2 = new_cases_subset[state_index[i],][31:60], 
+                  s1 = populations$population[state_index[i]], s2 = populations$population[state_index[i]], i1 = 1:30, 
                   i2 = 31:60, df1 = 3, df2 = 3)$p
 }
 

### 2023-05-20 code/squared.res_v_squared.fitted.R

@@ -8,13 +8,13 @@ rm(filename)
 
 state_fits <- c()
 
-for (i in signif_index){
+for (i in W_signif_index){
   state_fits <- c(state_fits, my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
                               inds = 1:60, df = 3))
 }
 
-state_fits <- matrix(state_fits, nrow = length(signif_index), ncol=7, byrow = TRUE)
-state_fits <- cbind(state_fits, new_cases_subset[signif_index,])
+state_fits <- matrix(state_fits, nrow = length(W_signif_index), ncol=7, byrow = TRUE)
+state_fits <- cbind(state_fits, new_cases_subset[W_signif_index,])
 state_fits <- as.data.frame(state_fits)
 colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
@@ -22,14 +22,14 @@ colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 # Plot squared residuals against squared fitted values
 par(mfrow = c(2,4))
 for (i in 1:8){   #NY
-  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
+  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
        col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
   
 }
 
 par(mfrow = c(2,3))
 for (i in 9:14){   #FL
-  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i,8:67]) - state_fits$mu[[i]])^2, 
+  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
        col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
   
 }

### 2023-05-20 code/validity_and_power_LRT.R

@@ -21,6 +21,6 @@ filename <- "data/pvals_sim_LRT.Rdata"
 save(pvals, file = filename)
 
 # Visualize p-values
-par(mfrow = c(1,2))
+par(mfrow = c(2, 1))
 hist(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
 hist(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)

### 2023-05-20 code/validity_and_power_Wald.R

@@ -12,8 +12,9 @@ ncurve <- nrow(curves)
 pvals <- rep(NA, ncurve)
 
 for (i in 1:ncurve) {
-  pvals[i] <- W(curves[i, ], population_size = curve_parms$population[i], breakpoint = curve_parms$breakpoint[i], deg_free = 3, 
-           fn = my_spl_fit, verbose = FALSE)
+  pvals[i] <- W(curves[i, ], population_size = curve_parms$population[i], 
+                breakpoint = curve_parms$breakpoint[i], deg_free = 3, 
+                fn = my_spl_fit, verbose = FALSE)
 }
 
 # Save p-values
@@ -21,6 +22,6 @@ filename <- "data/pvals_sim_Wald.Rdata"
 save(pvals, file = filename)
 
 # Visualize p-values
-par(mfrow = c(1,2))
+par(mfrow = c(2, 1))
 hist(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
 hist(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)

### 2023-05-20 code/vis_signific_state.R

@@ -1,6 +1,6 @@
 # Visualize fitted curves in a state's counties where significant Wald result observed
 
-for (i in signif_index){
+for (i in W_signif_index){
   
   mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
                    inds=1:length(new_cases_subset[i,]), df = 3)$mu

### 2023-05-20 code/wald_counties.R

@@ -8,16 +8,14 @@ rm(filename)
 # Loop through counties and calculate p values
 state_index <- which(populations$State == state)
 
-# Estimation issue for state_index = 385 (FL)
-if (state == "FL"){
-  state_index <- state_index[-c(which(state_index == 385))]
-}
-
 pvals <- rep(NA, times = length(state_index))
 
-for (i in state_index){
-  pvals[i] <- W(new_cases_subset[i,], population_size = populations$population[i], breakpoint = 30, 
-                deg_free = 3, fn = my_spl_fit, verbose = FALSE)
+for (i in 1:length(state_index)){
+  pvals[i] <- tryCatch(W(new_cases_subset[state_index[i],], 
+                         population_size = populations$population[state_index[i]],
+                         breakpoint = 30, deg_free = 3, 
+                         fn = my_spl_fit, verbose = FALSE), 
+                       error = function(e) return(NA))
 }
 
 # Save p-values

### 2023-05-20 noise_covid_v2.Rmd
edit visualization for significant counties

@@
 knitr::opts_chunk$set(echo = F, warning = F, message = F, 
-                      fig.align = "center", fig.width = 5, fig.height = 5)
+                      fig.align = "center", fig.width = 5,
+                      fig.height = 5)
 ```
@@
 source("code/wald_counties.R")
+signif_index <- state_index[(which(pvals < 0.05))]
+source("code/squared.res_v_squared.fitted.R")
+source("code/vis_signific_state.R")
 
@@
 ```
+
+```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
+par(mfrow = c(1,2))
+state <- "NY"
+source("code/lrt_counties.R")
+
+state <- "FL"
+source("code/lrt_counties.R")
+```
+
 ```{r}
@@
 
-```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
-par(mfrow = c(1,2))
-state <- "NY"
-source("code/lrt_counties.R")
-NY_sig <- (which(pvals < 0.05))
-
-state <- "FL"
-source("code/lrt_counties.R")
-FL_sig <- (which(pvals < 0.05))
-```
-Note that in these visualizations, an NB2 fit would be:
+Note that in these visualizations to check the relationship between mean and dispersion, an NB2 fit would be:
 A Poisson fit would be:
-```{r}
-signif_index <- c(NY_sig, FL_sig)
-
-state <- "NY"
-source("code/squared.res_v_squared.fitted.R")
-
-state <- "FL"
-source("code/squared.res_v_squared.fitted.R")
-
-source("code/vis_signific_state.R")
-```
 
@@
 load("data/pvals_sim_Wald.Rdata")
-mean(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
-mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)
+mean(pvals[which(curve_parms$theta1 == curve_parms$theta2)] < 0.05)
+mean(pvals[which(curve_parms$theta1 != curve_parms$theta2)] < 0.05)
 
 load("data/pvals_sim_LRT.Rdata")
-mean(pvals[which(curve_parms$theta1==curve_parms$theta2)], col=1)
-mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)], col=2)
-
+mean(pvals[which(curve_parms$theta1==curve_parms$theta2)] < 0.05)
+mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)] < 0.05)
 ```

### 2023-05-20 code/squared.res_v_squared.fitted.R

@@ -8,33 +8,22 @@ rm(filename)
 
 state_fits <- c()
 
-for (i in W_signif_index){
+for (i in signif_index){
   state_fits <- c(state_fits, my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
                               inds = 1:60, df = 3))
 }
 
-state_fits <- matrix(state_fits, nrow = length(W_signif_index), ncol=7, byrow = TRUE)
-state_fits <- cbind(state_fits, new_cases_subset[W_signif_index,])
+state_fits <- matrix(state_fits, nrow = length(signif_index), ncol=7, byrow = TRUE)
+state_fits <- cbind(state_fits, new_cases_subset[signif_index,])
 state_fits <- as.data.frame(state_fits)
 colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 
 # Plot squared residuals against squared fitted values
-par(mfrow = c(2,4))
-for (i in 1:8){   #NY
+for (i in 1:length(signif_index)){   
   plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
        col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
-  
-}
-
-par(mfrow = c(2,3))
-for (i in 9:14){   #FL
-  plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
-       col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
-  
 }
 
 # If the constant dispersion model (NB2) model is OK, then we should see a linear trend. 
 # If that checks out, then we don’t have to consider more complicated models.
-
-

### 2023-05-20 code/vis_signific_state.R

@@ -1,6 +1,6 @@
-# Visualize fitted curves in a state's counties where significant Wald result observed
+# Visualize fitted curves in a state's counties where significant result observed
 
-for (i in W_signif_index){
+for (i in signif_index){
   
   mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
                    inds=1:length(new_cases_subset[i,]), df = 3)$mu

### 2023-05-30 noise_covid_v2.Rmd
add figures as PDF's and incorporate

@@
 
-## Specific aim
-
 **Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint.
 Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **.
-
 It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al.
@@
 
-```{r}
-library(knitr)
-rat <- data.frame(Mean = c(5, 10), Ratio = c(2.236068, 3.162278))
-kable(rat, align = "cc", caption = "Poisson mean to standard deviation ratio", digits=2)
-```
-
 To illustrate this issue, I identified segments with approximately steady mean by mean changepoint detection (performed using an information criterion, MBIC).
@@
 
-```{r, fig.cap="Average Pothoff-Whittinghill (PW) statistic for each NY county plotted against county population size."}
-library(tidyverse)
-library(changepoint)
-source("code/functions_intro.R")
-
-filename <- "data/processed_dat.Rdata"
-load(filename)
-rm(filename)
-
-scaled_cases <- map(as.data.frame(t(new_cases_subset)), ~ as.vector(scale(.)))
-
-ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE)))
-
-segments <- map2(.x = as.data.frame(t(new_cases_subset[which(sapply(ch_pts, length) != 0),])), .y = ch_pts[which(sapply(ch_pts, length) != 0)], ~ get_segments(series = .x, changepoints = .y))
-
-mean_pw <- map(segments, ~ get_mean_pw(.))
-
-# Plot the mean PW statistic v. population size
-
-plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0, 1000000))
-```
 Methods designed specifically for count data show a similar dependence on population size.
@@
 
-```{r, fig.cap="Log of the number of changepoints identified by cpt.meanvar in each NY county plotted against log of county population size."}
-
-ch_pts2 <- map(as.data.frame(t(new_cases_subset)), ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))
-
-n_ch_pts2 <- map(ch_pts2, ~length(unlist(.)))
-
-plot(populations$population, log(unlist(n_ch_pts2)), xlab="Log Population Size", ylab=" Log Number of Changepoints", col=2)
-```
 ## Modeling proccess mean
@@
 2019).
+The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 60 days around Thanksgiving.
+In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.
 
@@
 
-```{r, fig.cap="Example of empirical increase in dispersion at the hypothesized Thanksgiving 2020 changepoint."}
-plot(new_cases_subset[1842,], xaxt="n", ylab="Daily Incidence", xlab="Date") #~ inds
-abline(v=30, col=2)
-#axis(1, inds, inds, cex.axis = .7)
-```
+![Figure 1. Empirical increase in dispersion](./figures/empirical_increase.pdf)
 
@@
 A test of the one-sided upper alternative that the difference between before and after is positive (lower theta after Thanksgiving) was performed.
-
-```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
-par(mfrow = c(1,2))
-state <- "NY"
-source("code/wald_counties.R")
-signif_index <- state_index[(which(pvals < 0.05))]
-source("code/squared.res_v_squared.fitted.R")
-source("code/vis_signific_state.R")
-
-state <- "FL"
-source("code/wald_counties.R")
-```
-
-```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
-par(mfrow = c(1,2))
-state <- "NY"
-source("code/lrt_counties.R")
-
-state <- "FL"
-source("code/lrt_counties.R")
-```
-
-```{r}
-library(tigris)
-fl <- counties("Florida")
-load("data/W_pvalsFL.Rdata")
-p <- round(pvals, digits=1)
-p[is.na(p)] <- "na"
-fl$p <- p
-
-ggplot(fl) +
-  geom_sf(aes(fill = p)) 
-```
 
@@
 
-```{r, fig.cap="Wald simulation p-values."}
-source("code/validity_and_power_Wald.R")
-```
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
@@
 
-```{r, fig.cap="Wald simulation p-values."}
-source("code/validity_and_power_LRT.R")
-```
+{![Figure 2A. (1) Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf){width=33%}![(2) Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf){width=33%}!
+}
+
 There appears to be slight Type I error inflation in both tests.
 
-```{r}
-load("data/pvals_sim_Wald.Rdata")
-mean(pvals[which(curve_parms$theta1 == curve_parms$theta2)] < 0.05)
-mean(pvals[which(curve_parms$theta1 != curve_parms$theta2)] < 0.05)
+![Figure 2B. Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)
 
-load("data/pvals_sim_LRT.Rdata")
-mean(pvals[which(curve_parms$theta1==curve_parms$theta2)] < 0.05)
-mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)] < 0.05)
-```
+![Figure 3. Scanning for breakpoints using the LRT statistic](./figures/LRT_scan_figure.pdf)
 

### 2023-05-30 code/LRT_scan_figure.R

@@ -1,0 +1,29 @@
+# LRT scan figure
+source("code/lrt.R")
+load("./data/processed_long_dat.Rdata")
+load("./data/processed_dat.Rdata")
+
+# Use first full time series
+test <- new_cases[1,]
+
+lrt_stat <- c()
+
+for (i in 30:(length(test) - 30 + 1)){
+  lrt_stat <- c(lrt_stat, tryCatch(lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
+                           s1 = populations$population[1], 
+                           s2 = populations$population[1],
+                           i1 = (i - 29):i,
+                           i2 = (i + 1):(i + 30), 
+                           df1 = 3, 
+                           df2 = 3)$lambda, error = function(e) return(NA)))
+}
+
+dates <- dates[30:(length(test) - 30 + 1)]
+
+filename <- "figures/LRT_scan_figure.pdf"
+pdf(filename, width = 6, height = 6)
+
+plot(lrt_stat ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
+abline(h = qchisq(0.95, df = 1), col = 2, lwd = 2)
+
+dev.off()

### 2023-05-30 code/changepoint_figures.R

@@ -1,0 +1,35 @@
+library(tidyverse)
+library(changepoint)
+source("code/functions_intro.R")
+
+filename <- "data/processed_dat.Rdata"
+load(filename)
+rm(filename)
+
+# Changepoint detection using cpt.mean
+scaled_cases <- map(as.data.frame(t(new_cases_subset)), ~ as.vector(scale(.)))
+ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", 
+                                            pen.value = 0, method = "PELT", param.estimates = TRUE)))
+segments <- map2(.x = as.data.frame(t(new_cases_subset[which(sapply(ch_pts, length) != 0),])), 
+                 .y = ch_pts[which(sapply(ch_pts, length) != 0)],
+                 ~ get_segments(series = .x, changepoints = .y))
+mean_pw <- map(segments, ~ get_mean_pw(.))
+
+# Changepoint detection using cpt.meanvar 
+ch_pts2 <- map(as.data.frame(t(new_cases_subset)), 
+               ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, 
+                                  method = "PELT", test.stat = "Poisson", 
+                                  param.estimates = TRUE, minseglen = 2))) 
+n_ch_pts2 <- map(ch_pts2, ~length(unlist(.)))
+
+# Plots
+filename <- "figures/changepoint_figures.pdf"
+pdf(filename, width = 6, height = 6)
+
+plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), 
+     xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0, 1000000))
+
+plot(populations$population, log(unlist(n_ch_pts2)), xlab="Log Population Size", 
+     ylab=" Log Number of Changepoints", col=2)
+
+dev.off()

### 2023-05-30 code/fl_map_Wald.R

@@ -1,0 +1,14 @@
+# Create a map of p-values resulting from the Wald test on the FL data
+
+library(tigris)
+all_counties <- counties(state = NULL)
+
+load("./data/W_pvals_and_thetas_allcounties.Rdata")
+p_values <- data.frame(cbind(pvals, fips))
+
+p <- round(pvals, digits=1)
+p[is.na(p)] <- "na"
+fl$p <- p
+
+ggplot(fl) +
+  geom_sf(aes(fill = p)) 

### 2023-05-30 code/map_Wald_US.R

@@ -1,0 +1,27 @@
+# Create a map of p-values resulting from the Wald test on the FL data
+
+library(tigris)
+library(tidyverse)
+
+all_counties <- counties(state = NULL)
+
+# Remove leading zeroes from GEOID column
+all_counties$GEOID <- as.character(as.numeric(all_counties$GEOID)) 
+
+load("./data/W_pvals_and_thetas_allcounties.Rdata")
+pvals <- round(pvals, digits = 1)
+pvals[is.na(pvals)] <- "na"
+
+p_values <- data.frame(cbind(pvals, fips))
+p_values$fips <- as.character(p_values$fips)
+
+map_df <- merge(all_counties, p_values, by.x = "GEOID", by.y = "fips",
+                all.x = TRUE)
+
+filename <- "figures/map_Wald_US.pdf"
+pdf(filename, width = 6, height = 6)
+
+ggplot(map_df) +
+  geom_sf(aes(fill = pvals)) 
+
+dev.off()

### 2023-05-30 code/process_COVID_data.R

@@ -39,6 +39,10 @@ dates <- names(dat)[5:ncol(dat)]
 dates <- gsub("X", "", dates)
 dates <- as.Date(dates, format = "%Y.%m.%d")
 
+# Save curves
+filename <- "data/processed_long_dat.Rdata"
+save(new_cases, dates, file = filename)
+
 # Subset to 60 days centered on Thanksgiving
 canadian_thanksgiving <- as.Date("2020-10-12")
 american_thanksgiving <- as.Date("2020-11-26")

### 2023-05-30 code/squared.res_v_squared.fitted.R

@@ -1,11 +1,19 @@
-# Check whether dispersion well represented by negative binomial model - just using the counties 
-# with significant lrt result
+# Check whether dispersion well represented by negative binomial model - just using the NY and FL counties 
+# with significant LRT result
 
 source("code/W.R")
-filename <- "data/processed_dat.RData"
-load(filename)
-rm(filename)
 
+load("./data/processed_dat.Rdata")
+
+load("./data/LRT_pvalsFL.Rdata")
+FL_index <- which(populations$State == "FL")
+FL_index <- FL_index[(which(pvals < 0.05))]
+
+load("./data/LRT_pvalsNY.Rdata")
+NY_index <- which(populations$State == "NY")
+NY_index <- NY_index[(which(pvals < 0.05))]
+
+signif_index <- c(NY_index, FL_index)
 state_fits <- c()
 
 for (i in signif_index){
@@ -13,13 +21,14 @@ for (i in signif_index){
                               inds = 1:60, df = 3))
 }
 
-state_fits <- matrix(state_fits, nrow = length(signif_index), ncol=7, byrow = TRUE)
+state_fits <- matrix(state_fits, nrow = length(signif_index), ncol = 7, byrow = TRUE)
 state_fits <- cbind(state_fits, new_cases_subset[signif_index,])
 state_fits <- as.data.frame(state_fits)
 colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 
 # Plot squared residuals against squared fitted values
+par(mfrow = c(2, 7))
 for (i in 1:length(signif_index)){   
   plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
        col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")

### 2023-05-30 code/vis_empirical_increase.R

@@ -1,0 +1,19 @@
+load("./data/processed_dat.Rdata")
+load("./data/simulated_curves.Rdata")
+
+filename <- "figures/empirical_increase.pdf"
+
+pdf(filename, width = 6, height = 6)
+
+american_thanksgiving <- as.Date("2020-11-26")
+start_date <- american_thanksgiving - 29
+end_date <- american_thanksgiving + 30
+days <- seq(start_date, end_date, by = "day")
+
+par(mfrow = c(1,2))
+plot(new_cases_subset[1842,] ~ days,  ylab="Daily Incidence", xlab="Date") 
+abline(v = american_thanksgiving, col=2)
+plot(curves[10000,] ~ days, ylab="Daily Incidence", xlab="Date") # theta_1 = theta_2 = 12
+abline(v = american_thanksgiving, col=2)
+
+dev.off()

### 2023-05-30 code/vis_signific_state.R

@@ -1,5 +1,19 @@
-# Visualize fitted curves in a state's counties where significant result observed
+# Visualize fitted curves in NY and FL counties where significant LRT result observed
 
+source("./code/W.R")
+load("./data/processed_dat.Rdata")
+
+load("./data/LRT_pvalsFL.Rdata")
+FL_index <- which(populations$State == "FL")
+FL_index <- FL_index[(which(pvals < 0.05))]
+
+load("./data/LRT_pvalsNY.Rdata")
+NY_index <- which(populations$State == "NY")
+NY_index <- NY_index[(which(pvals < 0.05))]
+
+signif_index <- c(NY_index, FL_index)
+
+par(mfrow = c(2, 7))
 for (i in signif_index){
   
   mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 

### 2023-05-30 code/wald_counties.R

@@ -9,6 +9,7 @@ rm(filename)
 pvals <- rep(NA, nrow(populations))
 theta1 <- rep(NA, nrow(populations))
 theta2 <- rep(NA, nrow(populations))
+fips <- rep(NA, nrow(populations))
 
 for (i in seq_len(nrow(populations))) {
   out <- tryCatch(
@@ -29,6 +30,7 @@ for (i in seq_len(nrow(populations))) {
     pvals[i] <- out$p
     theta1[i] <- out$theta1
     theta2[i] <- out$theta2
+    fips[i] <- populations$countyFIPS[i]
   }
 
   cat("County", i, "of", nrow(populations), "p-value:", pvals[i], "\n")
@@ -36,9 +38,10 @@ for (i in seq_len(nrow(populations))) {
 
 # Save p-values
 filename <- "data/W_pvals_and_thetas_allcounties.Rdata"
-save(pvals, theta1, theta2, file = filename)
+save(pvals, fips, theta1, theta2, file = filename)
 
 # Visualize thetas and p-values
+
 theta_max <- 100
 p_max <- 0.05
 

### 2023-06-03 noise_covid_v2.Rmd
incorporate info about LRT scan below figure

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = F, warning = F, message = F, 
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                       fig.align = "center", fig.width = 5,
@@
 
-{![Figure 2A. (1) Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf){width=33%}![(2) Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf){width=33%}!
-}
+![Figure 2A. (1) Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf)
+![(2) Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf)
 
@@
 ![Figure 3. Scanning for breakpoints using the LRT statistic](./figures/LRT_scan_figure.pdf)
+
+In some time windows, the Likelihood-Ratio statistic can't be computed because (TODO**: the weight matrix in the IRLS can't be calculated?).
+
+\newpage
 

### 2023-06-03 code/LRT_scan_figure.R

@@ -9,15 +9,14 @@ test <- new_cases[1,]
 lrt_stat <- c()
 
 for (i in 30:(length(test) - 30 + 1)){
-  lrt_stat <- c(lrt_stat, tryCatch(lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
+  lrt_stat <- c(lrt_stat, lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
                            s1 = populations$population[1], 
                            s2 = populations$population[1],
                            i1 = (i - 29):i,
                            i2 = (i + 1):(i + 30), 
                            df1 = 3, 
-                           df2 = 3)$lambda, error = function(e) return(NA)))
+                           df2 = 3)$lambda)
 }
-
 dates <- dates[30:(length(test) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"

### 2023-06-03 code/vis_empirical_increase.R

@@ -10,10 +10,14 @@ start_date <- american_thanksgiving - 29
 end_date <- american_thanksgiving + 30
 days <- seq(start_date, end_date, by = "day")
 
-par(mfrow = c(1,2))
-plot(new_cases_subset[1842,] ~ days,  ylab="Daily Incidence", xlab="Date") 
-abline(v = american_thanksgiving, col=2)
-plot(curves[10000,] ~ days, ylab="Daily Incidence", xlab="Date") # theta_1 = theta_2 = 12
-abline(v = american_thanksgiving, col=2)
+par(mfrow = c(2,2))
+plot(new_cases_subset[1862, ] ~ days,  ylab = "Daily Incidence in FIPS 36067 (Wald test p-value = 0.017)", xlab = "Date") 
+abline(v = american_thanksgiving, col = 2)
+plot(new_cases_subset[1863, ] ~ days, ylab = "Daily Incidence in FIPS 36069 (Wald test p_value = 0.099)", xlab ="Date") 
+abline(v = american_thanksgiving, col = 2)
+plot(curves[220, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.004)", xlab = "Date")
+abline(v = american_thanksgiving, col = 2)
+plot(curves[212, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.260)", xlab = "Date")
+abline(v = american_thanksgiving, col = 2)
 
 dev.off()

### 2023-06-03 noise_covid_v2.Rmd
updates for map

@@
 
-![Figure 1. Empirical increase in dispersion](./figures/empirical_increase.pdf)
+![Empirical increase in dispersion](./figures/empirical_increase.pdf)
 
@@
 
-![Figure 2A. (1) Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf)
-![(2) Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf)
+![Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf)
+![Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf)
 
@@
 
-![Figure 2B. Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)
+![Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)
 
-![Figure 3. Scanning for breakpoints using the LRT statistic](./figures/LRT_scan_figure.pdf)
+![Scanning for breakpoints using the LRT statistic](./figures/map_diff_theta_US.pdf)
 

### 2023-06-03 code/fl_map_Wald.R

@@ -1,14 +1,0 @@
-# Create a map of p-values resulting from the Wald test on the FL data
-
-library(tigris)
-all_counties <- counties(state = NULL)
-
-load("./data/W_pvals_and_thetas_allcounties.Rdata")
-p_values <- data.frame(cbind(pvals, fips))
-
-p <- round(pvals, digits=1)
-p[is.na(p)] <- "na"
-fl$p <- p
-
-ggplot(fl) +
-  geom_sf(aes(fill = p)) 

### 2023-06-03 code/map_Wald_US.R

@@ -1,27 +1,0 @@
-# Create a map of p-values resulting from the Wald test on the FL data
-
-library(tigris)
-library(tidyverse)
-
-all_counties <- counties(state = NULL)
-
-# Remove leading zeroes from GEOID column
-all_counties$GEOID <- as.character(as.numeric(all_counties$GEOID)) 
-
-load("./data/W_pvals_and_thetas_allcounties.Rdata")
-pvals <- round(pvals, digits = 1)
-pvals[is.na(pvals)] <- "na"
-
-p_values <- data.frame(cbind(pvals, fips))
-p_values$fips <- as.character(p_values$fips)
-
-map_df <- merge(all_counties, p_values, by.x = "GEOID", by.y = "fips",
-                all.x = TRUE)
-
-filename <- "figures/map_Wald_US.pdf"
-pdf(filename, width = 6, height = 6)
-
-ggplot(map_df) +
-  geom_sf(aes(fill = pvals)) 
-
-dev.off()

### 2023-06-03 code/map_thanksgiving.R

@@ -7,12 +7,7 @@ library(maps)
 library(mapproj)
 library(ggplot2)
 library(viridis)
-library(magrittr) #I will pipe from outside the damn tidyverse thank you
-
-
-
-
-
+library(magrittr) 
 
 # This from
 # https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
@@ -41,9 +36,7 @@ p1 <- ggplot(choropleth, aes(long, lat, group = group)) +
     scale_fill_viridis(option = "magma")
 
 
-
-
-# Adapt the above to plot thanksgiving results
+# Adapt the above to plot Thanksgiving results
 load("data/W_pvals_and_thetas_allcounties.Rdata")
 load("data/processed_dat.RData")
 
@@ -74,7 +67,7 @@ p2 <- ggplot(choropleth2, aes(long, lat, group = group)) +
     geom_polygon(data = state_df, colour = "white", fill = NA) +
     coord_fixed() +
     theme_minimal() +
-    ggtitle("theta2-theta1 (clamped)") +
+    ggtitle("Difference in k over Thanksgiving 2020") +
     theme(
         axis.line = element_blank(), axis.text = element_blank(),
         axis.ticks = element_blank(), axis.title = element_blank()
@@ -82,7 +75,13 @@ p2 <- ggplot(choropleth2, aes(long, lat, group = group)) +
     scale_fill_viridis(option = "magma")
 
 
-
-# Plot the 2 maps together
+# Save map 
 library(gridExtra)
-grid.arrange(p1, p2, nrow = 2)
+
+filename <- "./figures/map_difftheta_US.pdf"
+
+pdf(filename, width = 6, height = 6)
+
+p2
+
+dev.off()

### 2023-06-03 code/validity_and_power_LRT.R

@@ -35,11 +35,19 @@ for (i in 1:ncurve) {
 filename <- "data/pvals_sim_LRT.Rdata"
 save(pvals, file = filename)
 
+# Tabulate p-values
+X <- data.frame("Type I" =  mean(pvals[which(curve_parms$theta2 == curve_parms$theta1)]),
+                "Power at 3" = mean(pvals[which(abs(curve_parms$theta2 - curve_parms$theta1) == 3)]),
+                "Power at 9" = mean(pvals[which(abs(curve_parms$theta2 - curve_parms$theta1) == 9)]))
 
+filename <- "figures/pvals_sim_LRT_table.pdf"
+
+pdf(filename, width = 6, height = 6)
+
+dev.off()
 
 # Visualize p-values
 dtheta <- curve_parms$theta2 - curve_parms$theta1
-
 
 filename <- "figures/pvals_sim_LRT.pdf"
 

### 2023-06-03 code/validity_and_power_Wald.R

@@ -28,7 +28,16 @@ for (i in 1:ncurve) {
 filename <- "data/pvals_sim_Wald.Rdata"
 save(pvals, file = filename)
 
+# Tabulate p-values
+X <- data.frame("Type I" =  mean(pvals[which(curve_parms$theta2 == curve_parms$theta1)]),
+                "Power at 3" = mean(pvals[which(abs(curve_parms$theta2 - curve_parms$theta1) == 3)]),
+                "Power at 9" = mean(pvals[which(abs(curve_parms$theta2 - curve_parms$theta1) == 9)]))
 
+filename <- "figures/pvals_sim_Wald_table.pdf"
+
+pdf(filename, width = 6, height = 6)
+
+dev.off()
 
 # Visualize p-values
 dtheta <- curve_parms$theta2 - curve_parms$theta1

### 2023-06-09 noise_covid_v2.Rmd
update figures

@@
 
-Core challenges in epidemiology include estimating the efficacy of control measures, and understanding the impact of events of interest on incidence case counts.
-Allocating healthcare resources toward individuals with higher-than-average transmission risk can disproportionately reduce the population-level effective reproduction number (i.e., the average number of secondary infections caused by a randomly selected individual; Althouse et al.
-2020).
-In other words, leveraging transmission variability among individuals to target interventions can be a highly effective strategy for reducing population spread.
-However, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
-Conversely, evaluating control strategies using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control.
-The impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy.
+Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control.
+For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy.
 Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.
+
+Conversely, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
+We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.
 
@@
 Analyzing variability in epidemic dynamics in terms of bursts of incidence is also important for planning surge capacity (Wallinga 2018).
-
-Since evaluating control strategies using only course-scale changes in incidence over time may fail to account for parts of the context of transmission and control, we should aim to learn about individual-level heterogeneity as a means to understand control strategies more comprehensively.
-Similarly, we should aim to learn about individual-level heterogenity following congregation events, as estimates of individual-level variability are used to identify so-called "superspreading" dynamics: high levels of individual-level variability indicate that some individuals cause more onward transmission (via secondary infections) than others, and this in turn scales up to influence resulting incidence time series following these events.
 

### 2023-06-09 code/LRT_scan_figure.R

@@ -9,20 +9,21 @@ test <- new_cases[1,]
 lrt_stat <- c()
 
 for (i in 30:(length(test) - 30 + 1)){
-  lrt_stat <- c(lrt_stat, lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
+  lrt_stat <- c(lrt_stat, tryCatch(lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
                            s1 = populations$population[1], 
                            s2 = populations$population[1],
                            i1 = (i - 29):i,
                            i2 = (i + 1):(i + 30), 
                            df1 = 3, 
-                           df2 = 3)$lambda)
+                           df2 = 3)$lambda, error = function(e) return(NA)))
 }
 dates <- dates[30:(length(test) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"
 pdf(filename, width = 6, height = 6)
 
+par(mfrow = c(2, 1))
 plot(lrt_stat ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
-abline(h = qchisq(0.95, df = 1), col = 2, lwd = 2)
+plot(test[30:(length(test) - 30 + 1)] ~ dates, col = "red", type = "l")
 
 dev.off()

### 2023-06-09 code/vis_empirical_increase.R

@@ -1,4 +1,5 @@
 load("./data/processed_dat.Rdata")
+
 load("./data/simulated_curves.Rdata")
 
 filename <- "figures/empirical_increase.pdf"
@@ -6,18 +7,29 @@ filename <- "figures/empirical_increase.pdf"
 pdf(filename, width = 6, height = 6)
 
 american_thanksgiving <- as.Date("2020-11-26")
+
 start_date <- american_thanksgiving - 29
+
 end_date <- american_thanksgiving + 30
+
 days <- seq(start_date, end_date, by = "day")
 
 par(mfrow = c(2,2))
-plot(new_cases_subset[1862, ] ~ days,  ylab = "Daily Incidence in FIPS 36067 (Wald test p-value = 0.017)", xlab = "Date") 
+
+plot(new_cases_subset[1862, ] ~ days,  ylab = "Daily Incidence in FIPS 36067 (Wald test p-value = 0.017)", 
+     xlab = "Date", ylim = c(0, 1400)) 
 abline(v = american_thanksgiving, col = 2)
-plot(new_cases_subset[1863, ] ~ days, ylab = "Daily Incidence in FIPS 36069 (Wald test p_value = 0.099)", xlab ="Date") 
+
+plot(new_cases_subset[1863, ] ~ days, ylab = "Daily Incidence in FIPS 36069 (Wald test p_value = 0.099)", 
+     xlab ="Date", ylim = c(0, 1400)) 
 abline(v = american_thanksgiving, col = 2)
-plot(curves[220, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.004)", xlab = "Date")
+
+plot(curves[220, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.004)", 
+     xlab = "Date", ylim = c(0, 1400))
 abline(v = american_thanksgiving, col = 2)
-plot(curves[212, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.260)", xlab = "Date")
+
+plot(curves[212, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.260)", 
+     xlab = "Date", ylim = c(0, 1400))
 abline(v = american_thanksgiving, col = 2)
 
 dev.off()

### 2023-06-09 noise_covid_v2.Rmd
add text about V(mu) = 0

@@
 
-In some time windows, the Likelihood-Ratio statistic can't be computed because (TODO**: the weight matrix in the IRLS can't be calculated?).
+In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
 

### 2023-06-27 noise_covid_v2.Rmd
add Earn ref
edit text

@@
 
-From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects has been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003).
+From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003).
 Similarly, inference based on variability in epidemic time series' are emerging: for instance, Graham et al.
@@
 Note that some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992).
+Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - in other words, the effects of age structure can be captured by a model that includes an infection rate that varies over time (Earn et al.
+1998).
 
@@
 **Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint.
-Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **.
+Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly changes **.
 It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al.
 2022).
-Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a reduction in dispersion of cases, potentially modulated by changes in $k_t$.
+Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a change in dispersion of cases, potentially modulated by changes in $k_t$.
 Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.
@@
 Approximately one month of data on either side of November 26th, 2020 was used in the analysis.
-This step was done to include a set number of "waves" of incidence in the model fit to each side of the hypothesized change.
-However, we acknowledge that there is a trade-off between including few waves and increasing the power of the proposed test by including more time series data.
+This step was done to include a set "wave" of incidence in the model fit to each side of the hypothesized change.
+However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data.
 
@@
 Briefly, IRLS works by computing an adjusted dependent variable, computing a weight matrix, and then calculating (using weighted least squares) coefficients estimates of the model used to calculate the linear predictor as well.
-This procedure is implemented via the irls.nb.1 and functions from the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al.
+This procedure is implemented via the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al.
 (2011).
@@
 
-We tested the hypothesis that the incidence dispersion parameter decreases at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$.
+We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$.
 Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020.
@@
 First, for every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the one-sided upper alternative that the difference between before and after is positive (lower theta after Thanksgiving) was performed.
-
-Note that in these visualizations to check the relationship between mean and dispersion, an NB2 fit would be:
-A Poisson fit would be:
+A test of the two-sided alternative that the theta before is different from the theta afte (different theta after Thanksgiving) was performed.
 
@@
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
-The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean.
-As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05).
+The resulting probabilities of rejection are visualized below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05).
 Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.
+We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (see Supplement).
 

### 2023-06-27 code/lrt.R

@@ -29,10 +29,10 @@ lrt = function(y1, y2, s1, s2, i1, i2, df1, df2){
   l0 =  function(lkappa) {
     kappa = exp(lkappa);
     phi = 1/kappa;
-    res1 = irls.nb.1(y1, s1, x1, phi, mustart=mustart1);
-    res2 = irls.nb.1(y2, s2, x2, phi, mustart=mustart2);
-    sum(dnbinom(y1, size=kappa, mu=res1$mu, log=TRUE)) + 
-      sum(dnbinom(y2, size=kappa, mu=res2$mu, log=TRUE));
+    res1 = irls.nb.1(y1, s1, x1, phi, mustart = mustart1);
+    res2 = irls.nb.1(y2, s2, x2, phi, mustart = mustart2);
+    sum(dnbinom(y1, size = kappa, mu = res1$mu, log = TRUE)) + 
+      sum(dnbinom(y2, size = kappa, mu = res2$mu, log = TRUE));
   }
   
   obj0 = optimize(l0, c(log(1e-20), log(1e20)), maximum=TRUE);
@@ -67,10 +67,10 @@ lrt = function(y1, y2, s1, s2, i1, i2, df1, df2){
   kappa2 = exp(obj2$maximum);
   l12 = obj2$objective;
   phi12 = 1/kappa2;
-  res12 = irls.nb.1(y2, s2, x2, phi12, mustart=mustart2);
+  res12 = irls.nb.1(y2, s2, x2, phi12, mustart = mustart2);
   
   lambda = (l11 + l12 - l0) * 2;
-  p = pchisq(lambda, df=1, lower.tail=FALSE);
+  p = pchisq(lambda, df = 1, lower.tail = FALSE);
   
   list(p = p, lambda = lambda, l0 = l0, l11 = l11, l12= l12, 
        phi0 = phi0, phi11 = phi11, phi12 = phi12, 

### 2023-07-06 code/LRT_scan_figure.R
changes

@@ -1,9 +1,32 @@
-# LRT scan figure
+# Compute LRT scan statistics for all counties
+
 source("code/lrt.R")
 load("./data/processed_long_dat.Rdata")
 load("./data/processed_dat.Rdata")
 
-# Use first full time series
+lrt_stats <- matrix(NA, nrow = nrow(new_cases), ncol = 1124)
+
+for (j in 1:nrow(new_cases)){
+  
+  test = new_cases[j,]
+  
+  lrt_stat <- c()
+  
+  for (i in 30:(length(test) - 30 + 1)){
+    lrt_stat = c(lrt_stat, tryCatch(lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
+                                        s1 = populations$population[1], 
+                                        s2 = populations$population[1],
+                                        i1 = (i - 29):i,
+                                        i2 = (i + 1):(i + 30), 
+                                        df1 = 3, 
+                                        df2 = 3)$lambda, error = function(e) return(NA)))
+  }
+  
+  lrt_stats[j,] <- lrt_stat
+}
+
+# Produce figure for one county - use first full time series
+
 test <- new_cases[1,]
 
 lrt_stat <- c()
@@ -20,10 +43,15 @@ for (i in 30:(length(test) - 30 + 1)){
 dates <- dates[30:(length(test) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"
+
 pdf(filename, width = 6, height = 6)
 
-par(mfrow = c(2, 1))
 plot(lrt_stat ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
+
 plot(test[30:(length(test) - 30 + 1)] ~ dates, col = "red", type = "l")
+
+abline(h = quantile(lrt_stat, probs = 0.95, na.rm =TRUE), col = "black")
+
+abline(h = qchisq(0.95, 1), col = "blue")
 
 dev.off()

### 2023-07-06 code/process_COVID_data.R

@@ -39,7 +39,12 @@ dates <- names(dat)[5:ncol(dat)]
 dates <- gsub("X", "", dates)
 dates <- as.Date(dates, format = "%Y.%m.%d")
 
-# Save curves
+# Process full data
+new_cases[which(new_cases < 0)] <- 0
+new_cases <- new_cases[which(rowSums(new_cases) != 0),]
+populations <- populations[which(rowSums(new_cases) != 0),]
+
+# Save full data 
 filename <- "data/processed_long_dat.Rdata"
 save(new_cases, dates, file = filename)
 
@@ -52,11 +57,12 @@ end_date <- american_thanksgiving + 30
 
 dates_subset <- dates[dates >= start_date & dates <= end_date]
 new_cases_subset <- new_cases[, dates >= start_date & dates <= end_date]
-new_cases_subset[which(new_cases_subset < 0)] <- 0
 
+# Process subsetted data
+new_cases_subset[which(new_cases_subset < 0)] <- 0
 new_cases_subset <- new_cases_subset[which(rowSums(new_cases_subset) != 0),]
 populations <- populations[which(rowSums(new_cases_subset) != 0),]
 
-# Save curves
+# Save subsetted data
 filename <- "data/processed_dat.Rdata"
 save(populations, new_cases_subset, file = filename)

### 2023-07-07 noise_covid_v2.Rmd
changes

@@
 
-Conversely, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
+Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable <-(exclude this sentence?**).
 We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.
@@
 Note that some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992).
-Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - in other words, the effects of age structure can be captured by a model that includes an infection rate that varies over time (Earn et al.
+Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time (Earn et al.
 1998).
@@
 
-Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.
-
-Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
+Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
 Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al.
@@
 **Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint.
-Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly changes **.
+Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion changes significantly**.
 It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al.
@@
 However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data.
+In both the full data and the subset of data, rows that were all zeroes were omitted.
 
@@
 A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
-The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time.
+The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time (say something about how this addresses autocorrelation**).
 Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al.
@@
 Regardless of whether the data is modeled as negative binomial or Poisson around the process mean (and regardless of whether quasi-likelihood methods are used), it's crucial to incorporate an offset term - sometimes called an exposure variable - in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual).
-Essentially, instead of focusing on the infection rate, our model allows the focus to be on infection rate per person.
+Essentially, instead of focusing on the infection rate, our model allows the focus to be on per person infection rate.
 In other words, an offset term was added to the model to account for case counts resulting from different population sizes - an offset for population size means that incidence dispersion properties may be assessed *while accounting for a population effect*.
-
 Note that if the offset model were perfect, the estimated coefficient on log population would be exactly one.
@@
 Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
+(add**)
 
@@
 
-![Empirical increase in dispersion](./figures/empirical_increase.pdf)
+![Empirical change in dispersion](./figures/empirical_increase.pdf)
 
-We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$.
-Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020.
-The Wald test is used, where the estimates of $\theta$ in different segments are considered independent.
-This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information matrix.
+We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 \ne \theta_2$.
+Hypothesis tests were conducted to assess whether $\theta$ changes after Thanksgiving 2020.
+The Wald test was used, where the estimates of $\theta$ in different segments are considered independent.
+This approach is possible because we used IRLS estimates of $\theta$ - the standard error of $\theta$ was found utilizing the observed information matrix.
 
@@
 First, for every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the two-sided alternative that the theta before is different from the theta afte (different theta after Thanksgiving) was performed.
+A test of the two-sided alternative that theta before is different from theta after (different theta after Thanksgiving) was performed.
 
@@
 
-To ensure that the hypothesis testing framework is valid, a simulation with known (equal) values of theta on either side of Thanksgiving 2020 was run in order to ensure that the distribution of resulting p-values from the test of the upper alternative is approximately uniform on the interval from zero to one.
+To ensure that the hypothesis testing framework is valid, a simulation with known (equal) values of theta on either side of Thanksgiving 2020 was run in order to ensure that the distribution of resulting p-values from the test of the two-sided alternative is approximately uniform on the interval from zero to one.
 In both the validity and power simulations, representative epidemic curves are used.

### 2023-07-07 code/LRT_scan_figure.R

@@ -4,9 +4,13 @@ source("code/lrt.R")
 load("./data/processed_long_dat.Rdata")
 load("./data/processed_dat.Rdata")
 
-lrt_stats <- matrix(NA, nrow = nrow(new_cases), ncol = 1124)
+lg_pop <- which(populations$population > quantile(populations$population, 0.90))
 
-for (j in 1:nrow(new_cases)){
+lrt_stats <- matrix(NA, nrow = length(lg_pop), ncol = 1124)
+
+for (j in lg_pop){
+  
+  print(j)
   
   test = new_cases[j,]
   
@@ -25,33 +29,28 @@ for (j in 1:nrow(new_cases)){
   lrt_stats[j,] <- lrt_stat
 }
 
-# Produce figure for one county - use first full time series
+# Produce figure for one county - use first (of large populations) full time series 
 
-test <- new_cases[1,]
+test <- new_cases[lg_pop[1],]
 
-lrt_stat <- c()
-
-for (i in 30:(length(test) - 30 + 1)){
-  lrt_stat <- c(lrt_stat, tryCatch(lrt(y1 = test[(i - 29):i], y2 = test[(i + 1):(i + 30)], 
-                           s1 = populations$population[1], 
-                           s2 = populations$population[1],
-                           i1 = (i - 29):i,
-                           i2 = (i + 1):(i + 30), 
-                           df1 = 3, 
-                           df2 = 3)$lambda, error = function(e) return(NA)))
-}
 dates <- dates[30:(length(test) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"
 
 pdf(filename, width = 6, height = 6)
 
-plot(lrt_stat ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
+plot(lrt_stats[1, ] ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
 
 plot(test[30:(length(test) - 30 + 1)] ~ dates, col = "red", type = "l")
 
-abline(h = quantile(lrt_stat, probs = 0.95, na.rm =TRUE), col = "black")
+abline(h = quantile(lrt_stats[lg_pop[1], ], probs = 0.95, na.rm =TRUE), col = "black")
 
 abline(h = qchisq(0.95, 1), col = "blue")
 
 dev.off()
+
+# Find the approx date of the max lrt stat for each county 
+# Histogram
+# Top k lrt stats as points in time and space ? The issue is that might not be comparable
[…]

### 2023-07-07 code/diagnostic_plot.R

@@ -1,4 +1,5 @@
 source("code/W.R")
+load("data/processed_dat.Rdata")
 
 diagnostic <- function(y = new_cases_subset[i, ], 
                        population = populations$population[i]){
@@ -38,8 +39,14 @@ diagnostic <- function(y = new_cases_subset[i, ],
   
 }
 
+filename <- "figures/diagnostic_plot.pdf"
+
+pdf(filename, width = 6, height = 6)
+
 par(mfrow = c(2,2))
 
 for (i in c(2, 40, 300, 777)){
   diagnostic()
 }
+
+dev.off()

### 2023-07-07 code/disp_stats.R

@@ -1,0 +1,18 @@
+# Compute dispersion stat for each county
+
+disp_stats <- rep(NA, times = nrow(new_cases_subset))
+
+for (i in 1:nrow(new_cases_subset)){
+  
+  y = new_cases_subset[i,]
+  
+  inds = 1:length(y)
+  
+  pop = rep(populations$population[i], times = length(y))
+  
+  pois.fit <- glm(y ~ ns(inds, df = 3, intercept = TRUE) + offset(log(pop)), 
+                  family = poisson)
+  print(pois.fit)
+  #disp_stats[i] = pois.fit$deviance/(length(y) - 3)
+  
+}

### 2023-07-26 code/dem_disp_association.R
add figure, Waller approach

@@ -34,5 +34,5 @@ dat <- data.frame(fips = fips, dtheta_clamped = dtheta_clamped)
 
 reg_df <- merge(unemp, dat, by = "fips")
 
-# Regression
+# Spatial regression
 

### 2023-07-26 code/lrt_surface_figure.R

@@ -17,7 +17,9 @@ z <- t(lrt_stats)
 x <- seq_len(nrow(z))
 y <- seq_len(ncol(z))
 
+filename <- "figures/lrt_surface_figure.pdf"
 
+pdf(filename, width = 8, height = 8)
 
 # Plot
 par(pin = c(7, 5))
@@ -30,3 +32,5 @@ image(x, y, z,
 
 par(las = 2)
 axis(1, x, dates, tick = FALSE)
+
+dev.off()

### 2023-07-26 code/map_inprogress.R

@@ -1,0 +1,16 @@
+library(here)
+library(sf)
+library(maps)
+
+load("data/processed_dat.Rdata")
+
+county.map = map("county", region = "florida", fill = TRUE, plot = FALSE)
+county.map = sf::st_as_sf(county.map) # Convert from map object to sf
+
+plot (st_geometry(county.map)) # Only plot the polygons
+
+data(county.fips)
+county.map$FIPS = county.fips$fips[match(county.map$ID, county.fips$polyname)]
+dat.clor = merge(county.map, populations, by.x = "FIPS", by.y = "countyFIPS", all.x = TRUE)
+
+plot(dat.clor$population, main = "Pop")

### 2023-08-10 chpt/changepoint_approach.Rmd
clean text

--- 2023-06-09 noise_covid_v2.Rmd
@@
 ---
-title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
-author: "Rachael Aber"
-header-includes: \usepackage{setspace}\doublespacing
-output:
-  pdf_document: default
-  html_document:
-    df_print: paged
-geometry: margin=2.54cm
-fontsize: 12pt
+title: "Changepoint Approach"
+output: pdf_document
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
-                      fig.align = "center", fig.width = 5,
-                      fig.height = 5)
-```
-
-# Introduction
-
-Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control.
-For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy.
-Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.
-
-Conversely, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
-We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.
-
-Metrics of population-level *variability* may be overlooked and useful ways to understand epidemic dynamics and control.
-By population-level variability, I mean - loosely - dispersion of data around the time-averaged trajectory.
-The ability to estimate changing rates of *dispersion* in epidemic time series would be a step towards a more complete view and predictive understanding of NPIs and gatherings at the individual as well as the population level.
-This is because individual-level heterogeneity in transmission scales up to affect population-level dynamics (Lloyd-Smith 2005), so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
-
-From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects has been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003).
-Similarly, inference based on variability in epidemic time series' are emerging: for instance, Graham et al.
-(2019) use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
-Sun et al.
-(2021) found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-Analyzing variability in epidemic dynamics in terms of bursts of incidence is also important for planning surge capacity (Wallinga 2018).
-
-## Causes of overdispersed disease incidence
-
-*Overdispersed disease incidence* is suggestive of underlying biological processes of superspreading (overdispersed *individual reproductive number*), demographic/environmental stochasticity affecting cases, or changes in propagation of the pathogen at the population level (i.e., population effective reproduction number changing in time).
-Note that some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992).
-
-Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009).
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
-However, issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992).
-
-Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.
-
-Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
-Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al.
-2002).
-The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.
-
-**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint.
-Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **.
-It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al.
-2022).
-Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a reduction in dispersion of cases, potentially modulated by changes in $k_t$.
-Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.
-
-# Data
-
-For this project, we use county-level COVID-19 incidence data.
-Source: [USAFacts](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/).
-The data is weekly case counts for each US county.
-
-Approximately one month of data on either side of November 26th, 2020 was used in the analysis.
-This step was done to include a set number of "waves" of incidence in the model fit to each side of the hypothesized change.
-However, we acknowledge that there is a trade-off between including few waves and increasing the power of the proposed test by including more time series data.
-
-# Methods
 
@@
 
-To illustrate this issue, I identified segments with approximately steady mean by mean changepoint detection (performed using an information criterion, MBIC).
+To illustrate this issue, we identified segments with approximately steady mean by mean changepoint detection (performed using an information criterion, MBIC).
 The cpt.mean function in the changepoint package (Killick and Eckley 2014) assumes that the variance of the process is equal to one, so each New York county time series was scaled beforehand.
@@
 
-## Modeling proccess mean
-
-We propose a generalized linear model (GLM) approach to quantify incidence dispersion in segments of a time series, using a single parameter, $\theta$, while separately accounting for population size in the model.
-Here we account for unobserved heterogeneity, and time dependence in rate - potentially due to contagion - (Barron 1992) by specifying an overdispersed conditional distribution.
-
-A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
-The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time.
-Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al.
-2019).
-The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 60 days around Thanksgiving.
-In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.
-
-  $$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$ $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
-  ,where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
-
-In this case, we use a log link function:
-$$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-  $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-  $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
-
-## IRLS
-
-Iteratively reweighted least squares (IRLS) is used to obtain parameter estimates of the model.
-Briefly, IRLS works by computing an adjusted dependent variable, computing a weight matrix, and then calculating (using weighted least squares) coefficients estimates of the model used to calculate the linear predictor as well.
-This procedure is implemented via the irls.nb.1 and functions from the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al.
-(2011).
-
-## Inclusion of an offset term in the model
-
-Regardless of whether the data is modeled as negative binomial or Poisson around the process mean (and regardless of whether quasi-likelihood methods are used), it's crucial to incorporate an offset term - sometimes called an exposure variable - in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual).
-Essentially, instead of focusing on the infection rate, our model allows the focus to be on infection rate per person.
-In other words, an offset term was added to the model to account for case counts resulting from different population sizes - an offset for population size means that incidence dispersion properties may be assessed *while accounting for a population effect*.
-
-Note that if the offset model were perfect, the estimated coefficient on log population would be exactly one.
-
-## Identifying an appropriate model of the conditional distribution
-
-To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only a few dispersion statistics were less than or equal to one.
-Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
-
-Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical.
-
-Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), I inspected the mean-variance relationships using a diagnostic plot as in Ver Hoef and Boveng (2007).
-
-In the quasi-Poisson model (called the linear negative binomial specification by Barron (1992)), the variance is a linear function of the mean, with a slope of one representing the Poisson special case.
-By contrast, the negative binomial model may more appropriately represent the data when the variance has approximately a quadratic relationship with the mean.
-Since the true mean and variance of the data set is unknown, we approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively.
-These quantities are estimated using both the Poisson and the negative binomial model.
-Variance can be estimated by averaging squared residuals in each fitted mean category.
-The number of fitted mean categories (bins) was chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present.
-Visualization of individual squared residuals is also possible, but these may not be reliable estimates of the variance.
-
-## Proposed hypothesis test
-
-![Empirical increase in dispersion](./figures/empirical_increase.pdf)
-
-We tested the hypothesis that the incidence dispersion parameter decreases at the Thanksgiving changepoint of interest: $\theta_1 > \theta_2$.
-Hypothesis tests were conducted to assess whether $\theta$ is smaller (more variable incidence) after Thanksgiving 2020.
-The Wald test is used, where the estimates of $\theta$ in different segments are considered independent.
-This approach is possible because we use IRLS estimates of $\theta$ - the standard error of $\theta$ is found utilizing the observed information matrix.
-
-$$Cov(\hat{\theta_1},\hat{\theta_2}) = 0 $$
-The form of the Wald test statistic is:
-
-$$\frac{\hat{\theta_1} -\hat{\theta_2}}{SE(\hat{\theta_1} -\hat{\theta_2})}$$
-$$ = \frac{\hat{\theta_1} -\hat{\theta_2}}{\sqrt{\widehat{Var(\hat{\theta_1})} + \widehat{Var(\hat{\theta_2})}}}$$
-# Results
-
-## Thanksgiving 2020: Wald test
-
-First, for every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the one-sided upper alternative that the difference between before and after is positive (lower theta after Thanksgiving) was performed.
-
-Note that in these visualizations to check the relationship between mean and dispersion, an NB2 fit would be:
-A Poisson fit would be:
-
-# Simulation
-
-## Validity and power of the Wald test
-
-To ensure that the hypothesis testing framework is valid, a simulation with known (equal) values of theta on either side of Thanksgiving 2020 was run in order to ensure that the distribution of resulting p-values from the test of the upper alternative is approximately uniform on the interval from zero to one.
-In both the validity and power simulations, representative epidemic curves are used.
-The area under these curves (final outbreak size) is set to be the population size divided by ten.
-
-To implement these constraints on the integral:
-we use a normalizing constant of (n/10) * sqrt(2*pi* $\sigma^2$)
-
-$$ \int_0^\infty {(n/10)/\sqrt{2\pi(\sigma^2))} e^{-(t-61)^2/(2\sigma^2)}}dt = n/10$$
-So, we have the desired Gaussian function:
-$$ \mu_t = (n/10)/\sqrt{2\pi\sigma^2} e^{-(t-61)^2/(2\sigma^2)} $$
-So that integration will result in a final outbreak size of n/10.
-
-To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
-The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean.
-As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05).
-Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.
-
-## Validity and power of the LRT
-
-![Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf)
-![Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf)
-
-There appears to be slight Type I error inflation in both tests.
-
-![Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)
-
-![Scanning for breakpoints using the LRT statistic](./figures/map_diff_theta_US.pdf)
-
-In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
-
-\newpage
-
-## References
-
-Adam, D., Gostic, K., Tsang, T., Wu, P., Lim, W.
-W., Yeung, A., Wong, J., Lau, E., Du, Z., Chen, D., Ho, L.-M., Martín-Sánchez, M., Cauchemez, S., Cobey, S., Leung, G., & Cowling, B.
-(2022).
-Time-varying transmission heterogeneity of SARS and COVID-19 in Hong Kong.
-https://doi.org/10.21203/rs.3.rs-1407962/v1
-
-Cook, J. D. (n.d.). Notes on the Negative Binomial Distribution. 5.
-
-Davis, R. A., & Wu, R. (2009). A negative binomial model for time series of counts. Biometrika, 96(3), 735–749.
-
-Graham, M., Winter, A.
-K., Ferrari, M., Grenfell, B., Moss, W.
-J., Azman, A.
-S., Metcalf, C.
-J.
-E., & Lessler, J.
-(2019).
-Measles and the canonical path to elimination.
-Science, 364(6440), 584–587.
-https://doi.org/10.1126/science.aau6299
-
-Karlis, D., & Xekalaki, E.
-(2000).
-A Simulation Comparison of Several Procedures for Testing the Poisson Assumption.
-Journal of the Royal Statistical Society: Series D (The Statistician), 49(3), 355–382.
-https://doi.org/10.1111/1467-9884.00240
-
-Killick, R., & Eckley, I. A. (2014). changepoint: An R Package for Changepoint Analysis. Journal of Statistical Software, 58, 1–19. https://doi.org/10.18637/jss.v058.i03
-
-Lloyd-Smith, J.
-O., Schreiber, S.
-J., Kopp, P.
-E., & Getz, W.
-M.
-(2005).
-Superspreading and the effect of individual variation on disease emergence.
-Nature, 438(7066), Article 7066.
-https://doi.org/10.1038/nature04153
-
-Perperoglou, A., Sauerbrei, W., Abrahamowicz, M., & Schmid, M.
-(2019).
-A review of spline function procedures in R.
-BMC Medical Research Methodology, 19(1), 46.
-https://doi.org/10.1186/s12874-019-0666-3
-
-Potthoff, R. F., & Whittinghill, M. (1966). Testing for homogeneity. II. The Poisson distribution. Biometrika, 53(1), 183–190.
-Transmission heterogeneities, kinetics, and controllability of SARS-CoV-2. (n.d.). https://doi.org/10.1126/science.abe2424
-
-Transmission heterogeneities, kinetics, and controllability of SARS-CoV-2. (n.d.). https://doi.org/10.1126/science.abe2424
-
-Venables W.N., Ripley B.D. (2002). Modern Applied Statistics with S, Fourth edition. Springer, New York. ISBN 0-387-95457-0, https://www.stats.ox.ac.uk/pub/MASS4/.
-
-Ver Hoef, J. M., & Boveng, P. L. (2007). QUASI-POISSON VS. NEGATIVE BINOMIAL REGRESSION: HOW SHOULD WE MODEL OVERDISPERSED COUNT DATA? Ecology, 88(11), 2766–2772. https://doi.org/10.1890/07-0043.1
-
-Wallinga, J. (2018). Metropolitan versus small-town influenza. Science. https://doi.org/10.1126/science.aav1003
+![Changepoint figures](./figures/changepoint_figures.pdf)

### 2023-08-10 noise_covid_v2.Rmd
add figure, Waller approach
clean text

@@
 Metrics of population-level *variability* may be overlooked and useful ways to understand epidemic dynamics and control.
-By population-level variability, I mean - loosely - dispersion of data around the time-averaged trajectory.
+Population-level variability can be thought of as dispersion of data around the time-averaged trajectory.
 The ability to estimate changing rates of *dispersion* in epidemic time series would be a step towards a more complete view and predictive understanding of NPIs and gatherings at the individual as well as the population level.
@@
 
-*Overdispersed disease incidence* is suggestive of underlying biological processes of superspreading (overdispersed *individual reproductive number*), demographic/environmental stochasticity affecting cases, or changes in propagation of the pathogen at the population level (i.e., population effective reproduction number changing in time).
+*Overdispersed disease incidence* is suggestive of the underlying biological processes of superspreading (overdispersed *individual reproductive number*), demographic/environmental stochasticity affecting cases, or changes in propagation of the pathogen at the population level (i.e., population effective reproduction number changing in time).
 Note that some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992).
@@
 
-## Detecting variability changes in count data: the problem of population size
-
-One might expect that in segments where the mean parameter is approximately the same, inference based on the Poisson distribution might be effective in detecting overdispersed incidence.
-There are several approaches to testing the Poisson assumption for count data, including testing variance-to-mean relationship, testing the probability-generating function (PGF) against an alternative, and standard goodness-of-fit tests (Karlis and Xekalaki 2000).
-However, an important consideration should be addressed for modeling time series of counts: the Poisson model has a spread of values that is comparatively more narrow for a process with larger mean (the mean/standard deviation ratio of a Poisson process is larger for larger total counts).
-For example, consider the mean to standard deviation ratios of the Poisson processes tabulated below: $\mu = \sigma^2 = 5$ and another with $\mu = \sigma^2 = 10$.
-For the process with a smaller mean, the mean to standard deviation ratio is smaller, whereas for the process with a larger mean, the mean to standard deviation ratio is larger.
-This presents an issue for using tests of Poisson variance to detect overdispersion or changes in dispersion in count time series, such as (importantly for epidemiology) incidence time series from populations of different sizes.
-
-To illustrate this issue, I identified segments with approximately steady mean by mean changepoint detection (performed using an information criterion, MBIC).
-The cpt.mean function in the changepoint package (Killick and Eckley 2014) assumes that the variance of the process is equal to one, so each New York county time series was scaled beforehand.
-Once these segments were identified, we used a goodness-of-fit test assuming that the Poisson mean parameter of each segment, $\lambda_i$ is unknown.
-This is called the variance test, and it has an optimal power property against the negative binomial alternative.
-The rejection region is demarcated by large values of the test statistic (Potthoff and Whittinghill 1966), and values of the test statistic increase with population size.
-
-Methods designed specifically for count data show a similar dependence on population size.
-The cpt.meanvar function (changepoint package) is designed to detect changes in mean and variance in count data.
-The number of changepoints identified for a time series increases with its population size.
-
-## Modeling proccess mean
+## Modeling the proccess mean
 
@@
 A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
-The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time (say something about how this addresses autocorrelation**).
+The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time to account for autocorrelation in case counts.
 Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al.
 2019).
-The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 60 days around Thanksgiving.
+The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 30 days.
 In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.
@@
 
-Iteratively reweighted least squares (IRLS) is used to obtain parameter estimates of the model.
+Iteratively reweighted least squares (IRLS) is used to obtain parameter estimates for the model.
 Briefly, IRLS works by computing an adjusted dependent variable, computing a weight matrix, and then calculating (using weighted least squares) coefficients estimates of the model used to calculate the linear predictor as well.
@@
 
-To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion at the time point level is present under this model and found that only a few dispersion statistics were less than or equal to one.
+To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion is present under this model and found that only a few dispersion statistics were less than or equal to one.
 Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
-(add**)
+
+```{r}
+load("data/dispersion_stats.Rdata")
+length(which(disp_stats > 1))/length(disp_stats)
+```
 
 Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical.
-
-Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), I inspected the mean-variance relationships using a diagnostic plot as in Ver Hoef and Boveng (2007).
-
+Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007).
 In the quasi-Poisson model (called the linear negative binomial specification by Barron (1992)), the variance is a linear function of the mean, with a slope of one representing the Poisson special case.
 By contrast, the negative binomial model may more appropriately represent the data when the variance has approximately a quadratic relationship with the mean.
-Since the true mean and variance of the data set is unknown, we approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively.
-These quantities are estimated using both the Poisson and the negative binomial model.
+Since the true mean and variance of the data set is unknown, one could approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively.
+These quantities should be estimated using both the Poisson and the negative binomial model.
 Variance can be estimated by averaging squared residuals in each fitted mean category.
-The number of fitted mean categories (bins) was chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present.
+The number of fitted mean categories (bins) should be chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present.
 Visualization of individual squared residuals is also possible, but these may not be reliable estimates of the variance.
 
+We use a diagnostic plot approach in which we rely on the expected value of the residuals to infer "p" using the slope of a regression of log(Y) on log(X), where X is r^2 - Y.
+The expectation of X is phi * mu^p, so
+ log(E(X)) ~ log(phi) + p log(E(Y))
+
+![Diagnostic Plot](figures/diagnostic_plot.pdf)
+
 ## Proposed hypothesis test
-
-![Empirical change in dispersion](./figures/empirical_increase.pdf)
 
@@
 This approach is possible because we used IRLS estimates of $\theta$ - the standard error of $\theta$ was found utilizing the observed information matrix.
+
+![Empirical change in dispersion](./figures/empirical_increase.pdf)
 
@@
 In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
+
+![The LRT surface](./figures/map_diff_theta_US.pdf)
 

### 2023-08-10 code/disp_stats.R

@@ -1,5 +1,9 @@
 # Compute dispersion stat (deviance divided by degrees of freedom) for each county
 
+library(splines)
+
+load("data/processed_dat.Rdata")
+
 disp_stats <- rep(NA, times = nrow(new_cases_subset))
 
 for (i in 1:nrow(new_cases_subset)){
@@ -8,10 +12,15 @@ for (i in 1:nrow(new_cases_subset)){
   
   inds = 1:length(y)
   
-  pop = rep(populations$population[i], times = length(y))
+  pop = rep(populations_subset$population[i], times = length(y))
   
   pois.fit <- glm(y ~ ns(inds, df = 3, intercept = TRUE) + offset(log(pop)), 
                   family = poisson)
+  
   disp_stats[i] = pois.fit$deviance/(length(y) - 3)
   
 }
+
+filename <- "data/dispersion_stats.Rdata"
+
+save(disp_stats, file = filename)

### 2023-08-11 noise_covid_v2.Rmd
more changes for text

@@
 
-$$Cov(\hat{\theta_1},\hat{\theta_2}) = 0 $$
+$$Cov(\hat{\theta_1}, \hat{\theta_2}) = 0 $$
 The form of the Wald test statistic is:
@@
 
-First, for every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the two-sided alternative that theta before is different from theta after (different theta after Thanksgiving) was performed.
+For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
+A test of the two-sided alternative that $\theta$ before is different from theta after (different theta after Thanksgiving) was performed.
 
@@
 To ensure that the hypothesis testing framework is valid, a simulation with known (equal) values of theta on either side of Thanksgiving 2020 was run in order to ensure that the distribution of resulting p-values from the test of the two-sided alternative is approximately uniform on the interval from zero to one.
-In both the validity and power simulations, representative epidemic curves are used.
+In both the validity and power simulations, Gaussian and uniform epidemic curves (see figure above which demonstrates correct behavior of the test in either case) are used.
 The area under these curves (final outbreak size) is set to be the population size divided by ten.
 
-To implement these constraints on the integral:
+To implement these constraints on the integral in the Gaussian case:
+
 we use a normalizing constant of (n/10) * sqrt(2*pi* $\sigma^2$)
@@
 
+![Type I Error Rate and Power - Wald Test](figures/pvals_sim_Wald_table.pdf)
+
+![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf)
+
 ## Validity and power of the LRT
 
-![Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf)
-![Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf)
+![Type I Error Rate and Power - LRT Test](figures/pvals_sim_LRT_table.pdf)
 
-There appears to be slight Type I error inflation in both tests.
+![LRT  p-values v. Theta Difference](figures/pvals_sim_LRT.pdf)
 
-![Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)
+## Biogeography of Superspreading Dynamics
 
-![Scanning for breakpoints using the LRT statistic](./figures/map_diff_theta_US.pdf)
+![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf)
+
+![LRT Scan Figure](./figures/LRT_scan_figure.pdf)
+
+![The LRT surface](./figures/lrt_surface_figure.pdf)
+
+![The Theta Difference Surface](./figures/thetadiff_surface.pdf)
 
 In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
-
-![The LRT surface](./figures/map_diff_theta_US.pdf)
 

### 2023-08-11 code/LRT_scan_figure.R

@@ -76,94 +76,94 @@ filename <- "data/thetadiff_lg_pops.Rdata"
 
 save(thetadiffs, file = filename)
 
-# # Produce figure for one county - use first (of large populations) full time series 
-# 
-# this_series <- new_cases_lg[1, ]
-# 
-# dates <- dates[30:(length(this_series) - 30 + 1)]
-# 
-# filename <- "figures/LRT_scan_figure.pdf"
-# 
-# pdf(filename, width = 6, height = 6)
-# 
-# plot(lrt_stats[1, ] ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
-# 
-# plot(this_series[30:(length(this_series) - 30 + 1)] ~ dates, col = "red", type = "l")
-# 
-# abline(h = quantile(lrt_stats[1, ], probs = 0.95, na.rm =TRUE), col = "black")
-# 
-# abline(h = qchisq(0.95, 1), col = "blue")
-# 
-# dev.off()
-# 
-# # Create surfaces 
-# 
-# # Plot likelihood ratios as a surface over counties and time
-# 
-# rm(list = ls())
-# 
-# graphics.off()
-# 
-# library(viridis)
-# 
-# 
-# # Load data
-# 
-# load("data/lrt_lg_pops.Rdata")
-# 
-# load("data/processed_long_dat.Rdata")   # for dates
-# 
-# # Data setup
-# dates <- dates[30:(length(dates) - 30 + 1)]
-# z <- t(lrt_stats)
-# x <- seq_len(nrow(z))
-# y <- seq_len(ncol(z))
-# 
-# filename <- "figures/lrt_surface_figure.pdf"
-# 
-# pdf(filename, width = 8, height = 8)
-# 
-# # Plot
-# par(pin = c(7, 5))
-# image(x, y, z,
-#       xaxt = "n",
-#       yaxt = "n",
-#       xlab = "",
-#       ylab = "Locations",
-#       col = rev(mako(32)))
-# 
-# par(las = 2)
-# axis(1, x, dates, tick = FALSE)
-# 
-# dev.off()
-# 
-# # Dates of bands: 
-# # Just after 2021-03-01; just after 2021-05-15; 2021-04-05
-# # 2021-01-10 seen across many of the counties
-# 
-# # Color by direction of theta change by making same matrix w/
-# # theta change
-# 
-# load("data/thetadiff_lg_pops.Rdata")
-# 
[…]

### 2023-08-11 code/check_nb_estimation.R

@@ -1,13 +1,0 @@
-load("data/simulated_curves.Rdata")
-source("code/W.R")
-
-i <- 5000
-Y <- curves[i, ]
-population <- curve_parms$population[i]
-
-out <- my_spl_fit(Y = Y, population = population, inds = 1:30, df = 3)
-
-# Check results visually
-ylim <- range(curves)
-plot(Y, ylim = ylim)
-lines(out$mu, col = "red", lwd = 2)

### 2023-08-11 code/dem_disp_association.R

@@ -1,38 +1,0 @@
-# Assess associates between dtheta_clamped and demographic variables
-library(stringi)
-
-
-# Format FIPS in unemp data
-
-data(unemp, package = "viridis")
-
-statefips <-stri_pad_left(unemp$state_fips, 2, 0)
-
-countyfips <- sprintf("%03d", unemp$county_fips)
-
-unemp$fips <- paste(statefips, countyfips, sep = "")
-
-# Format dtheta data
-
-load("data/W_pvals_and_thetas_allcounties.Rdata")
-
-fips <- stri_pad_left(fips, 5, 0)
-
-dtheta <- theta1 - theta2
-
-clamp <- function(x, c) {
-  x[x > c] <- c
-  x[x < -c] <- -c
-  return(x)
-}
-
-dtheta_clamped <- clamp(dtheta, 20)
-
-dat <- data.frame(fips = fips, dtheta_clamped = dtheta_clamped)
-
-# Merge datasets
-
-reg_df <- merge(unemp, dat, by = "fips")
-
-# Spatial regression
-

### 2023-08-11 code/generate_time_series_b.R

@@ -43,7 +43,7 @@ rm(theta1, theta2, breakpoint, population, curve_type)
 
 # Generate epidemic curves
 ncurve <- nrow(curve_parms)
-curves <- matrix(NA, nrow = ncurve, ncol = nstep)
+curves_b <- matrix(NA, nrow = ncurve, ncol = nstep)
 for (i in 1:ncurve) {
 
     breakpoint <- curve_parms$breakpoint[i]
@@ -64,10 +64,10 @@ for (i in 1:ncurve) {
     c1 <- rnbinom(n = length(means1), mu = means1, size = k1 * means1)
     c2 <- rnbinom(n = length(means2), mu = means2, size = k2 * means2)
 
-    curves[i, ] <- c(c1, c2)
+    curves_b[i, ] <- c(c1, c2)
 
 }
 
 # Save curves
-filename <- "data/simulated_curves.Rdata"
-save(curves, curve_parms, file = filename)
+filename <- "data/simulated_curves_b.Rdata"
+save(curves_b, curve_parms, file = filename)

### 2023-08-11 code/histogram_LRT_time.R

@@ -1,14 +1,0 @@
-# Find the approx date of the max lrt stat for each large county 
-
-load("data/lrt_lg_pops.Rdata")
-
-max_inds <- rep(NA, times = nrow(lrt_stats))
-
-for (j in 1:nrow(lrt_stats)){
-  max_inds[j] = which.max(lrt_stats[j, ])
-} 
-
-# Histogram
-
-hist(max_inds) 
-# seems like many max stats are around index 300 (index 281 is Thanksgiving 2020)

### 2023-08-11 code/map_thanksgiving_RA.R

@@ -5,6 +5,7 @@ library(spdep)
 library(spatialreg)
 
 load("data/W_pvals_and_thetas_allcounties.Rdata")
+
 load("data/processed_dat.Rdata")
 
 county.map = map("county", fill = TRUE, plot = FALSE)
@@ -12,7 +13,7 @@ county.map = sf::st_as_sf(county.map) # Convert from map object to sf
 plot(st_geometry(county.map)) # Only plot the polygons
 
 # Prepare the dataframe to be merged w/polygon data
-dat <- cbind(populations, theta1, theta2, pvals)
+dat <- cbind(populations_subset, theta1, theta2, pvals)
 dat$county <- gsub(" County", "", dat$County.Name)
 dat$county <- gsub(" Parish", "", dat$county)
 dat$county <- tolower(dat$county)
@@ -32,30 +33,35 @@ county.map$FIPS = county.fips$fips[match(county.map$ID, county.fips$polyname)]
 dat.clor = merge(county.map, dat, by.x = "FIPS", by.y = "countyFIPS", all.x = TRUE)
 
 # Save map 
+
 filename <- "./figures/map_difftheta_US.pdf"
+
 pdf(filename, width = 6, height = 6)
+
 plot(dat.clor["dtheta_clamped"], main = "dtheta_clamped")
+
 dev.off()
 
-# Spatial (Poisson) regression for dtheta_clamped
-dat_clor = st_transform(dat.clor, crs = "ESRI:102004")
-nb = poly2nb(dat_clor)
-nb = nblag(nb, 2) #Create 2nd order and this makes nb into a list
-
-#2nd-order has about twice as many links
-nb[[1]]
-nb[[2]] 
-
-#Create adjacency matrix: 
-#Style B = binary for edges
-W = nb2listw(nb[[1]], style="B", zero.policy = TRUE)
-W2 = nb2listw(nb[[2]], style="B", zero.policy = TRUE) 
-
-## Standard linear regression
-Y = dat_clor$dtheta_clamped
-X = (dat_clor$population - mean(dat_clor$population, na.rm = TRUE)) #center (scale?**)
-fit0 = lm(Y ~ X)
-
-## Fit CAR model
-fit1.car = spautolm(Y ~ X, family = "CAR", listw = W)
-fit2.car = spautolm (Y ~ X, family = "CAR", listw = W2)
[…]

### 2023-08-11 code/validity_and_power_LRT.R

@@ -48,15 +48,19 @@ filename <- "data/pvals_sim_LRT.Rdata"
 save(pvals, file = filename)
 
 # Tabulate Type I error rate and Power
+
+p_vals.0 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 0)]
+pow.0 <- mean(p_vals.0 < 0.05)
+
 p_vals.3 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 3)]
 pow.3 <- mean(p_vals.3 < 0.05)
 
 p_vals.9 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 9)]
 pow.9 <- mean(p_vals.9 < 0.05)
 
-X <- data.frame("Type I Error Rate" =  round(mean(pvals[which(curve_parms$theta2 == curve_parms$theta1)]), 2),
-                "Power at 3" = pow.3,
-                "Power at 9" = pow.9)
+X <- data.frame("Type I Error Rate" =  round(pow.0, 2),
+                "Power at 3" = round(pow.3, 2),
+                "Power at 9" = round(pow.9, 2))
 
 filename <- "figures/pvals_sim_LRT_table.pdf"
 

### 2023-08-11 code/validity_and_power_Wald.R

@@ -31,15 +31,20 @@ filename <- "data/pvals_sim_Wald.Rdata"
 save(pvals, file = filename)
 
 # Tabulate Type I error rate and Power
+
+p_vals.0 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 0)]
+pow.0 <- mean(p_vals.0 < 0.05)
+
 p_vals.3 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 3)]
 pow.3 <- mean(p_vals.3 < 0.05)
 
 p_vals.9 <- pvals[which(curve_parms$theta2 - curve_parms$theta1 == 9)]
 pow.9 <- mean(p_vals.9 < 0.05)
 
-X <- data.frame("Type I Error Rate" =  round(mean(pvals[which(curve_parms$theta2 == curve_parms$theta1)]), 2),
-                "Power at 3" = pow.3,
-                "Power at 9" = pow.9)
+
+X <- data.frame("Type I Error Rate" =  round(pow.0, 2),
+                "Power at 3" = round(pow.3, 2),
+                "Power at 9" = round(pow.9, 2))
 
 filename <- "figures/pvals_sim_Wald_table.pdf"
 

### 2023-08-15 noise_covid_v2.Rmd
changes

@@
 
+![Fitted Curves With Three Spline Degrees of Freedom](figures/FL_signif_fittedcurves.pdf)
+
   $$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$ $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
@@
 
+We use a second diagnostic plot, squared residuals versus squared fitted values to look for a linear trend in the Florida counties with a significant LRT test result for the Thanksgiving hypothesis.
+If the constant dispersion model (NB2) model is ok, then we should see a linear trend.
+If that checks out, then we don’t have to consider more complicated models.
+
 ![Diagnostic Plot](figures/diagnostic_plot.pdf)
+
+![Diagnostic Plot 2](figures/diagnostic_plot.2.pdf)
 

### 2023-08-15 code/diagnostic_plot.2.R

@@ -1,38 +1,40 @@
-# Check whether dispersion well represented by negative binomial model - just using the NY and FL counties 
-# with significant LRT result
+# Check whether dispersion is well represented by NB model 
+# using FL counties with significant LRT result
 
 source("code/W.R")
 
 load("./data/processed_dat.Rdata")
 
 load("./data/LRT_pvalsFL.Rdata")
-FL_index <- which(populations$State == "FL")
+FL_index <- which(populations_subset$State == "FL")
 FL_index <- FL_index[(which(pvals < 0.05))]
 
-load("./data/LRT_pvalsNY.Rdata")
-NY_index <- which(populations$State == "NY")
-NY_index <- NY_index[(which(pvals < 0.05))]
-
-signif_index <- c(NY_index, FL_index)
 state_fits <- c()
 
-for (i in signif_index){
-  state_fits <- c(state_fits, my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
+for (i in FL_index){
+  state_fits <- c(state_fits, my_spl_fit(Y = new_cases_subset[i,], population = populations_subset$population[i], 
                               inds = 1:60, df = 3))
 }
 
-state_fits <- matrix(state_fits, nrow = length(signif_index), ncol = 7, byrow = TRUE)
-state_fits <- cbind(state_fits, new_cases_subset[signif_index,])
+state_fits <- matrix(state_fits, nrow = length(FL_index), ncol = 7, byrow = TRUE)
+state_fits <- cbind(state_fits, new_cases_subset[FL_index,])
 state_fits <- as.data.frame(state_fits)
 colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 
 # Plot squared residuals against squared fitted values
-par(mfrow = c(2, 7))
-for (i in 1:length(signif_index)){   
+
+filename <- "figures/diagnostic_plot.2.pdf"
+
+pdf(file = filename, width = 6, height = 6)
+
+par(mfrow = c(2, 3))
+
+for (i in 1:length(FL_index)){   
   plot(state_fits$mu[[i]]^2, (unlist(state_fits[i, 8:67]) - state_fits$mu[[i]])^2, 
        col = 2, xlab = "Squared Fitted Values", ylab = "Squared Residuals")
 }
 
+dev.off()
 # If the constant dispersion model (NB2) model is OK, then we should see a linear trend. 
 # If that checks out, then we don’t have to consider more complicated models.

### 2023-08-15 code/vis_signific_state.R

@@ -1,24 +1,28 @@
-# Visualize fitted curves in NY and FL counties where significant LRT result observed
+# Visualize fitted curves in FL counties where significant LRT result observed
 
 source("./code/W.R")
+
 load("./data/processed_dat.Rdata")
 
 load("./data/LRT_pvalsFL.Rdata")
-FL_index <- which(populations$State == "FL")
+
+FL_index <- which(populations_subset$State == "FL")
+
 FL_index <- FL_index[(which(pvals < 0.05))]
 
-load("./data/LRT_pvalsNY.Rdata")
-NY_index <- which(populations$State == "NY")
-NY_index <- NY_index[(which(pvals < 0.05))]
+filename <- "figures/FL_signif_fittedcurves.pdf"
 
-signif_index <- c(NY_index, FL_index)
+pdf(filename, width = 6, height = 6)
 
-par(mfrow = c(2, 7))
-for (i in signif_index){
+par(mfrow = c(2, 3))
+
+for (i in FL_index){
   
-  mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations$population[i], 
-                   inds=1:length(new_cases_subset[i,]), df = 3)$mu
+  mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations_subset$population[i], 
+                   inds = 1:length(new_cases_subset[i,]), df = 3)$mu
   
-  plot(1:length(new_cases_subset[i,]), new_cases_subset[i,])
+  plot(1:length(new_cases_subset[i,]), new_cases_subset[i,], xlab = , ylab = )
   lines(mu, col=2)
 }
+
+dev.off()

### 2023-08-23 code/LRT_scan_figure.R
edit thetadiff surface

@@ -168,7 +168,7 @@ image(x, y, z,
       yaxt = "n",
       xlab = "",
       ylab = "Locations",
-      col = rev(mako(32)))
+      col = c("blue", "green", "purple", "pink", "red"))
 
 par(las = 2)
 axis(1, x, dates, tick = FALSE)

### 2023-08-28 noise_covid_v2.Rmd
edit thetadiff surface
edit text; move code for creating surface

--- 2023-08-11 noise_covid_v2.Rmd
@@
 title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
-author: "Rachael Aber"
+author: "Rachael Aber, Yanming Di, Ben Dalziel"
 header-includes: \usepackage{setspace}\doublespacing
@@
 For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy.
-Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.
+Similarly, the impacts of events/gatherings (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.
 
-Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable <-(exclude this sentence?**).
-We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.
+Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
+We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.
 
@@
 Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
-However, issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992).
-
 Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
@@
 However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data.
+
+We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic.
+
 In both the full data and the subset of data, rows that were all zeroes were omitted.
@@
 
+Issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992).
 A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
@@
 
-  $$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$ $$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
+![Fitted Curves With Three Spline Degrees of Freedom](figures/FL_signif_fittedcurves.pdf){width=30%, height=30%}
+
+$$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+$$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
   ,where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
@@
 In this case, we use a log link function:
+
 $$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
   $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+
   $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
@@
 
+```{r}
+load("data/dispersion_stats.Rdata")
+```
+
 To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion is present under this model and found that only a few dispersion statistics were less than or equal to one.
 Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.
+The proportion of dispersion statistics that are greater than one in US counties over the Thanksgiving period is `r length(which(disp_stats > 1))/length(disp_stats)`.
 
-```{r}
-load("data/dispersion_stats.Rdata")
-length(which(disp_stats > 1))/length(disp_stats)
-```
-
-Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical.
+Directly comparing between quasi-Poisson and Poisson models via deviance residuals isn't possible because the deviance residuals are identical.
 Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007).
@@
 
-We use a diagnostic plot approach in which we rely on the expected value of the residuals to infer "p" using the slope of a regression of log(Y) on log(X), where X is r^2 - Y.
-The expectation of X is phi * mu^p, so
- log(E(X)) ~ log(phi) + p log(E(Y))
+We use a diagnostic plot approach in which we rely on expected value of the residuals to infer "p" using the slope of a regression of log(Y) on log(X), where X is residual^2 - Y.
 
-![Diagnostic Plot](figures/diagnostic_plot.pdf)
+$$E[X] = \phi * \mu^p$$
+$$log(E(X)) \sim log(\phi) + p*log(E(Y))$$
+
+![Diagnostic Plot](figures/diagnostic_plot.pdf){width=30%, height=30%}
+
+We use a second diagnostic plot, squared residuals versus squared fitted values to look for a linear trend in the Florida counties with a significant LRT test result for the Thanksgiving hypothesis.
+The constant dispersion model (NB2) model is appropriate if we observe an approximately linear trend.
+
+![Diagnostic Plot 2](figures/diagnostic_plot.2.pdf){width=30%, height=30%}
 
@@
 
-![Empirical change in dispersion](./figures/empirical_increase.pdf)
+![Empirical change in dispersion](./figures/empirical_increase.pdf){width=30%, height=30%}
 
@@
 For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the two-sided alternative that $\theta$ before is different from theta after (different theta after Thanksgiving) was performed.
+A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.
 
@@
 So, we have the desired Gaussian function:
+
 $$ \mu_t = (n/10)/\sqrt{2\pi\sigma^2} e^{-(t-61)^2/(2\sigma^2)} $$
@@
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
-The resulting probabilities of rejection are visualized below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05).
+The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05).
 Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.
-We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (see Supplement).
+We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (Supplement).
 
@@
 
-![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf)
+Additionally, the relationship between the p-values resulting from the Wald test and the effect size is as expected.
+
+![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf){width=30%, height=30%}
 
@@
 
+The tabulated Type I error rate and power, as well as the relationship between p-values and effect size is presented below for the likelihood ratio test approach.
+
 ![Type I Error Rate and Power - LRT Test](figures/pvals_sim_LRT_table.pdf)
 
-![LRT  p-values v. Theta Difference](figures/pvals_sim_LRT.pdf)
+![LRT  p-values v. Theta Difference](figures/pvals_sim_LRT.pdf){width=30%, height=30%}
 
@@
 
-![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf)
+The spatial pattern of theta difference after Thanksgiving 2020 shows that there is spatial correlation in the degree of change.
+In particular, spatial clusters of increased dispersion occur on the East Coast and in the South.
 
-![LRT Scan Figure](./figures/LRT_scan_figure.pdf)
+![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf){width=30%, height=30%}
 
-![The LRT surface](./figures/lrt_surface_figure.pdf)
+Additionally, the likelihood ratio test statistic does not necessarily track with changes in mean incidence in a county, but rather captures dispersion changes.
+In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
 
-![The Theta Difference Surface](./figures/thetadiff_surface.pdf)
+![LRT Scan Figure](./figures/LRT_scan_figure.pdf){width=30%, height=30%}
 
-In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
+In visualizing a surface of the likelihood ratio test statistic and associated direction of the change in theta, it appears that
+
+![The LRT Surface](./figures/lrt_surface_figure.pdf){width=30%, height=30%}
+
+![The Theta Difference Surface](./figures/thetadiff_surface.pdf){width=30%, height=30%}
 

### 2023-08-28 code/create_surfaces.R

@@ -1,0 +1,70 @@
+# Plot likelihood ratios as a surface over counties and time
+
+rm(list = ls())
+
+graphics.off()
+
+library(viridis)
+
+# Load data
+
+load("data/lrt_lg_pops.Rdata")
+
+load("data/processed_long_dat.Rdata")   
+
+# Data setup
+
+dates <- dates[30:(length(dates) - 30 + 1)]
+z <- t(lrt_stats)
+x <- seq_len(nrow(z))
+y <- seq_len(ncol(z))
+
+filename <- "figures/lrt_surface_figure.pdf"
+
+pdf(filename, width = 8, height = 8)
+
+# Plot
+
+par(pin = c(7, 5))
+
+image(x, y, z,
+      xaxt = "n",
+      yaxt = "n",
+      xlab = "",
+      ylab = "Locations",
+      col = rev(mako(32)))
+
+par(las = 2)
+axis(1, x, dates, tick = FALSE)
+
+dev.off()
+
+# Dates of bands:
+# Just after 2021-03-01; just after 2021-05-15; 2021-04-05
+# 2021-01-10 seen across many of the counties
+# Color by direction of theta change by making same matrix w/
+# theta change
+
+load("data/thetadiff_lg_pops.Rdata")
+
+z <- t(thetadiffs)
+x <- seq_len(nrow(z))
+y <- seq_len(ncol(z))
+
+filename <- "figures/thetadiff_surface.pdf"
+
+pdf(filename, width = 6, height = 6)
+
+# Plot
+par(pin = c(7, 5))
+image(x, y, z,
+      xaxt = "n",
+      yaxt = "n",
+      xlab = "",
+      ylab = "Locations",
+      col = c("blue", "green", "purple", "pink", "red"))
+
+par(las = 2)
+axis(1, x, dates, tick = FALSE)
+
+dev.off()

### 2023-08-28 code/diagnostic_plot.2.R

@@ -26,7 +26,7 @@ colnames(state_fits) <- c("kappa", "beta", "mu", "l", "j", "theta", "SE.theta")
 
 filename <- "figures/diagnostic_plot.2.pdf"
 
-pdf(file = filename, width = 6, height = 6)
+pdf(file = filename, width = 3, height = 3)
 
 par(mfrow = c(2, 3))
 

### 2023-08-28 code/diagnostic_plot.R

@@ -41,7 +41,7 @@ diagnostic <- function(y = new_cases_subset[i, ],
 
 filename <- "figures/diagnostic_plot.pdf"
 
-pdf(filename, width = 6, height = 6)
+pdf(filename, width = 3, height = 3)
 
 par(mfrow = c(2,2))
 

### 2023-08-28 code/thetadifflrt_mat_LRTscan.R

@@ -34,9 +34,9 @@ lrt_stats <- matrix(NA, nrow = 51, ncol = 1124)
 
 thetadiffs <- matrix(NA, nrow = 51, ncol = 1124)
 
-      
-for (j in 1:nstate){
 
+for (j in 1:nstate){
+  
   series = new_cases_lg[j,]
   
   lrt_stat <- c()
@@ -44,7 +44,7 @@ for (j in 1:nstate){
   thetadiff <- c()
   
   for (i in 30:(length(series) - 30 + 1)){
-
+    
     Y = series[(i - 29):(i + 30)]
     
     test = tryCatch(lrt(y1 = Y[1:(length(Y)/2)], y2 = Y[((length(Y)/2) + 1):length(Y)],
@@ -58,18 +58,18 @@ for (j in 1:nstate){
     lrt_stat = c(lrt_stat, test)
     
     test.1 = tryCatch(W(y = Y, population_size = populations_lg$population[j], 
-                       breakpoint = 30, deg_free = 3, 
-                       fn = my_spl_fit, verbose = FALSE, return_theta_diff = TRUE),
+                        breakpoint = 30, deg_free = 3, 
+                        fn = my_spl_fit, verbose = FALSE, return_theta_diff = TRUE),
                       error = function(e) return(NA))
-
+    
     thetadiff = c(thetadiff, test.1)
     
- }
+  }
   
   lrt_stats[j,] <- lrt_stat
-
+  
   thetadiffs[j,] <- thetadiff
-
+  
 }
 
 # Save LRT statistic array; save thetadiff array
@@ -82,7 +82,13 @@ filename <- "data/thetadiff_lg_pops.Rdata"
 
 save(thetadiffs, file = filename)
 
-# Produce figure for one county - use first (of large populations) full time series
+# Load data
+
+load("data/lrt_lg_pops.Rdata")
+
+load("data/processed_long_dat.Rdata")  
+
+# Plot
 
 this_series <- new_cases_lg[1, ]
 
@@ -90,7 +96,9 @@ dates <- dates[30:(length(this_series) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"
 
-pdf(filename, width = 6, height = 6)
+pdf(filename, width = 3, height = 3)
+
+par(mfrow = c(2, 1))
 
[…]

### 2023-09-30 Supplement.Rmd
updates

---
title: Supplementary Information 
author: "Rachael Aber, Yanming Di, Ben Dalziel"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin = 2.54cm
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.align = "center", fig.width = 5,
                      fig.height = 5)
```

# Wald Test

In the Wald test framework, the estimates of $\theta$ in different segments were considered independent. This approach is possible because we used IRLS estimates of $\theta$ - the standard error of $\theta$ was found utilizing the observed information matrix.

$$Cov(\hat{\theta_1}, \hat{\theta_2}) = 0 $$
The form of the Wald test statistic is:

$$\frac{\hat{\theta_1} -\hat{\theta_2}}{SE(\hat{\theta_1} -\hat{\theta_2})}$$
$$ = \frac{\hat{\theta_1} -\hat{\theta_2}}{\sqrt{\widehat{Var(\hat{\theta_1})} + \widehat{Var(\hat{\theta_2})}}}$$

However, the dispersion parameter is biologically relevant on a log scale (i.e., the change in dispersion from a $\theta$ change of 0.3 to 3 is equivalent to the change from 3 to 30), a more appropriate hypothesis test statistic (using the Delta Method) is:

$$ = \frac{g(\hat{\theta_1},\hat{\theta_2}) - g(\theta_1, \theta_2)}{\sqrt{(\nabla(g(\theta_1,\theta_2))^T[\widehat{Var(\hat{\theta_1}, \hat{\theta_2}}]\nabla(g(\theta_1,\theta_2))}}$$

= $$\frac{(log(\hat{\theta_1)}-log(\hat{\theta_2})) - 0}{\sqrt{\widehat{Var(\hat\theta_1)}/\theta_1^2 + \widehat{Var(\hat\theta_2)}/\theta_2^2}}$$

![Type I Error Rate and Power - Wald Test](figures/pvals_sim_Wald_table.pdf)

![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf){width=30%, height=30%}

# Alternative Simulation Method

### 2023-09-30 noise_covid_v2.Rmd
updates

@@
     df_print: paged
-geometry: margin=2.54cm
+geometry: margin = 2.54cm
 fontsize: 12pt
@@
 For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy.
-Similarly, the impacts of events/gatherings (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.
+Similarly, the impacts of gatherings (such as those on national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.
 
+We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.
+This method may also be useful in identifying geographical areas that would benefit from targeted control strategies to reduce superspreading dynamics.
 Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
-We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.
 
@@
 
-Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009).
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
 Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
@@
 
-**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint.
+## Negative Binomial Regression
+
+The negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009).
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
+
+**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series (by fitting negative binomial models) and its relationship to a hypothesized changepoint.
 Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion changes significantly**.
@@
 
-We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic.
+We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic and visualizations of a dispersion metric.
 
@@
 
-## Proposed hypothesis test
+## Proposed Likelihood Ratio Test
 
 We tested the hypothesis that the incidence dispersion parameter changes at the Thanksgiving changepoint of interest: $\theta_1 \ne \theta_2$.
-Hypothesis tests were conducted to assess whether $\theta$ changes after Thanksgiving 2020.
-The Wald test was used, where the estimates of $\theta$ in different segments are considered independent.
-This approach is possible because we used IRLS estimates of $\theta$ - the standard error of $\theta$ was found utilizing the observed information matrix.
 
 ![Empirical change in dispersion](./figures/empirical_increase.pdf){width=30%, height=30%}
-
-$$Cov(\hat{\theta_1}, \hat{\theta_2}) = 0 $$
-The form of the Wald test statistic is:
-
-$$\frac{\hat{\theta_1} -\hat{\theta_2}}{SE(\hat{\theta_1} -\hat{\theta_2})}$$
-$$ = \frac{\hat{\theta_1} -\hat{\theta_2}}{\sqrt{\widehat{Var(\hat{\theta_1})} + \widehat{Var(\hat{\theta_2})}}}$$
-# Results
-
-## Thanksgiving 2020: Wald test
-
-For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
-A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.
-
-# Simulation
-
-## Validity and power of the Wald test
 
@@
 To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed.
+
 The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05).
@@
 We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (Supplement).
-
-![Type I Error Rate and Power - Wald Test](figures/pvals_sim_Wald_table.pdf)
-
-Additionally, the relationship between the p-values resulting from the Wald test and the effect size is as expected.
-
-![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf){width=30%, height=30%}
-
-## Validity and power of the LRT
 
@@
 
+# Results
+
 ## Biogeography of Superspreading Dynamics
+
+For every county, the negative binomial model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$.
+A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.
 
@@
 ![The Theta Difference Surface](./figures/thetadiff_surface.pdf){width=30%, height=30%}
+
+## Less dispersion over time
+
+In many US counties between 2020-09-27 and 2021-01-25, there is less burstiness (larger theta) in COVID-19 case counts over time.
+In other words, cases fluctuate less around a time-averaged trajectory later on, possibly indicating structure in the population with some groups experiencing different infection rates (overdispersed individual reproduction number), demographic/environmental stochasticity, or fast changes in propagation/reporting at the population level may have occurred earlier on.
+However, the trend of decreased burstiness over this period seems to be modulated by population size: among the largest population counties per state, those in the top half had an average theta change of approx.
+28.95604 (less burstiness), while the small populations had an average change in theta of approx.
+-3.741856 (more burstiness).
+This means that in the large populations, dynamics became "smoother", but this didn't happen in the smaller populations, where more burstiness might be a result of cases "stuttering out".
+This is a key observation, as smoother and more predictable case count time series may require less surge capacity in hospitals.
+
+We hypothesized that there would be an increase in burstiness (decrease in theta) around Thanksgiving 2020 due to people gathering and causing clusters of infection.
+However, we observed that in many counties, although there was more burstiness (a decrease in theta), it was a relatively small change, and not modulated by population size.
+The average difference between theta estimated in the window starting at 2020-11-26 and theta estimated in the window ending 2020-11-25 was -1.77944 in the top half and -1.104289 in the bottom half of population sizes.
+People may have avoided gathering due to personal preference, cultural norms, and other factors, contributing to less superspreading than anticipated
+
+![Compare](./figures/compare.pdf){width=30%, height=30%}
 

### 2023-09-30 code/W.R

@@ -13,19 +13,21 @@ my_spl_fit <- function(Y, population, inds, df) {
 
 
 W <- function(y, population_size, breakpoint, deg_free = 3, fn = my_spl_fit, verbose = FALSE, 
-              return_theta_diff = FALSE, return_theta = FALSE) {
+              return_theta_diff = FALSE, return_theta = FALSE, return_thetas = FALSE) {
 
-    pop <- population_size
-    df <- deg_free
-    inds1 <- 1:breakpoint
-    inds2 <- (breakpoint + 1):length(y)
+    pop = population_size
+    df = deg_free
+    inds1 = 1:breakpoint
+    inds2 = (breakpoint + 1):length(y)
+    indsall = 1:length(y)
 
-    spline_mod.1 <- fn(Y = y, population = pop, inds = inds1, df = df)
-    spline_mod.2 <- fn(Y = y, population = pop, inds = inds2, df = df)
+    spline_mod.1 = fn(Y = y, population = pop, inds = inds1, df = df)
+    spline_mod.2 = fn(Y = y, population = pop, inds = inds2, df = df)
+    spline_mod.all = fn(Y = y, population = pop, inds = indsall, df = df)
 
-    test_stat <- ((spline_mod.1$theta - spline_mod.2$theta) -
-        0) / sqrt((spline_mod.1$SE.theta^2 +
-        spline_mod.2$SE.theta^2))
+    test_stat = ((log10(spline_mod.1$theta) - log10(spline_mod.2$theta)) -
+        0)/sqrt(spline_mod.1$SE.theta^2/spline_mod.1$theta^2 +
+        spline_mod.2$SE.theta^2/spline_mod.2$theta^2)
 
     if (verbose == TRUE) {
         return(sprintf(
@@ -42,9 +44,15 @@ W <- function(y, population_size, breakpoint, deg_free = 3, fn = my_spl_fit, ver
     if (return_theta_diff) {
         return(theta1 - theta2)
     } 
+    
     if (return_theta) {
-      return(theta2)
+      return(spline_mod.all$theta)
     }
+    
+    if (return_thetas){
+      return(list(c(theta1, theta2)))
+    }
+    
     else {
         return(p)
     }

### 2023-09-30 code/compare.R

@@ -1,45 +1,62 @@
 
 load("data/processed_long_dat.Rdata")
+dates <- dates[30:(length(new_cases[1,]) - 30 + 1)]
 
-load("data/lrt_lg_pops.Rdata")
-load("data/thetadiff_lg_pops.Rdata")
+#load("data/lrt_lg_pops.Rdata")
+#load("data/thetadiff_lg_pops.Rdata")
 load("data/theta_lg_pops.Rdata")
 
 
-# Indices for largest county in each state
+# Indices for largest counties in each state
 
 unique_states <- unique(populations$State)
 
 nstate <- length(unique_states)
 
-keep <- rep(NA, times = nstate)
+keep <- c()
 
 for (i in 1:nstate){
   
   j <- which(populations$State == unique_states[i])
   
-  k <- which.max(populations$population[j])
+  k <- which(populations$population[j] > quantile(populations$population[j], 0.96))
   
-  keep[i] <- j[k]
+  keep <- c(keep, j[k])
 }
 
 new_cases_lg <- new_cases[keep,]
 
 populations_lg <- populations[keep,]
 
-filename <- "figures/compare.pdf"
+filename <- "figures/compare1.pdf"
 
 pdf(filename, height = 8, width = 8)
 
-for (i in 1:51){
+for (i in 1:length(keep)){
   
   series = new_cases_lg[i,]
-  
-  par(mfrow = c(2, 2))
+  par(mfrow = c(2, 1))
   plot(series[30:(length(series) - 30 + 1)], type = "l", main = "Incidence")
-  plot(lrt_stats[i, ], type = "l", col = 2, main = "LRT")
-  plot(thetadiffs[i, ], type = "l", col = 3, main = "Theta Difference")
-  plot(thetas[i, ], type = "l", col = 4, main = "Theta")
+  abline(v = 281, col = "red")
+  #plot(lrt_stats[i, ], type = "l", col = 2, main = "LRT")
+  #plot(thetadiffs[i, ], type = "l", col = 3, main = "Theta Difference")
+  plot(log(thetas[i, ]), type = "l", col = 4, main = "Theta")
+  abline(v = 281, col = "red")
 }
 
 dev.off()
+
+# Is the discrepancy in theta b/w two time periods different for the larger and smaller
+# population halves?
+
+thetasdiff = thetas[ ,380] - thetas[ ,250]
+
+thetasdiff_small <- thetasdiff[populations_lg$population <= 
[…]

### 2023-09-30 code/theta_sim_Ben.R

@@ -1,0 +1,44 @@
+# Ben’s quick code to think about what 
+# different values in theta mean:
+  
+# Simulate
+  
+mu <- 100
+  
+  
+  
+x30 <- rnbinom(n = 10000, mu = mu, size = 30)
+  
+x3 <- rnbinom(n = 10000, mu = mu, size = 3)
+  
+x.3 <- rnbinom(n = 10000, mu = mu, size = 0.3)
+  
+y <- rpois(n = 10000, lambda = mu)
+  
+  
+  
+# Plot CDFs
+  
+plot(ecdf(y), xlim = c(0, 500), main = "Empirical CDF", cex = 0.5)
+  
+plot(ecdf(x30), col = 2, pch = 21, cex = 0.5, add = T)
+  
+plot(ecdf(x3), col = 3, pch = 21, cex = 0.5, add = T)
+  
+plot(ecdf(x.3), col = 4, pch = 21, cex = 0.5, add = T)
+  
+  
+  
+legend('bottomright',
+         
+       pch = rep(21, 4),
+         
+       col = 1:4,
+         
+       legend = c("Poisson",
+                    
+                  "theta = 30",
+                    
+                  "theta = 3",
+                    
+                  "theta = 0.3"))

### 2023-09-30 code/thetadifflrt_mat_LRTscan.R

@@ -1,4 +1,4 @@
-# Compute LRT scan statistics and thetadiffs for large counties
+# Compute LRT scan statistics, thetadiffs, and thetas for large counties
 
 source("code/W.R")
 
@@ -6,21 +6,21 @@ source("code/lrt.R")
 
 load("data/processed_long_dat.Rdata")
 
-# Indices for largest county in each state
+# Indices for largest counties in each state
 
 unique_states <- unique(populations$State)
 
 nstate <- length(unique_states)
 
-keep <- rep(NA, times = nstate)
+keep <- c()
 
 for (i in 1:nstate){
   
   j <- which(populations$State == unique_states[i])
   
-  k <- which.max(populations$population[j])
+  k <- which(populations$population[j] > quantile(populations$population[j], 0.96))
   
-  keep[i] <- j[k]
+  keep <- c(keep, j[k])
 }
 
 new_cases_lg <- new_cases[keep,]
@@ -30,19 +30,21 @@ populations_lg <- populations[keep,]
 
 # Set up empty matrices 
 
-lrt_stats <- matrix(NA, nrow = 51, ncol = 1124)
+# lrt_stats <- matrix(NA, nrow = 51, ncol = 1124)
+# 
+# thetadiffs <- matrix(NA, nrow = 51, ncol = 1124)
 
-thetadiffs <- matrix(NA, nrow = 51, ncol = 1124)
+thetas <- matrix(NA, nrow = length(keep), ncol = 1124)
 
-thetas <- matrix(NA, nrow = 51, ncol = 1124)
-
-for (j in 1:nstate){
+for (j in 1:length(keep)){
+  
+  print(j)
   
   series = new_cases_lg[j,]
   
-  lrt_stat <- c()
-  
-  thetadiff <- c()
+  # lrt_stat <- c()
+  # 
+  # thetadiff <- c()
   
   theta <- c()
   
@@ -50,36 +52,35 @@ for (j in 1:nstate){
     
     Y = series[(i - 29):(i + 30)]
     
-    test = tryCatch(lrt(y1 = Y[1:(length(Y)/2)], y2 = Y[((length(Y)/2) + 1):length(Y)],
-                        s1 = populations_lg$population[j],
-                        s2 = populations_lg$population[j],
-                        i1 = 1:30,
-                        i2 = 31:60,
-                        df1 = 3,
-                        df2 = 3)$lambda, error = function(e) return(NA))
-    
-    lrt_stat = c(lrt_stat, test)
-    
[…]

### 2023-09-30 code/validity_and_power_Wald.R

@@ -55,6 +55,7 @@ grid.table(X)
 dev.off()
 
 # Visualize p-values
+
 dtheta <- curve_parms$theta2 - curve_parms$theta1
 
 filename <- "figures/pvals_sim_Wald.pdf"

### 2023-09-30 code/wald_counties.R

@@ -6,34 +6,36 @@ load(filename)
 rm(filename)
 
 # Loop through counties and calculate p values
-pvals <- rep(NA, nrow(populations))
-theta1 <- rep(NA, nrow(populations))
-theta2 <- rep(NA, nrow(populations))
-fips <- rep(NA, nrow(populations))
+pvals <- rep(NA, nrow(populations_subset))
+theta1 <- rep(NA, nrow(populations_subset))
+theta2 <- rep(NA, nrow(populations_subset))
+fips <- rep(NA, nrow(populations_subset))
 
-for (i in seq_len(nrow(populations))) {
-  out <- tryCatch(
-    W(new_cases_subset[i, ],
-      population_size = populations$population[i],
-      breakpoint = 30,
-      deg_free = 3,
-      fn = my_spl_fit,
-      verbose = FALSE,
-      return_thetas = TRUE
-    ),
-    error = function(e) {
-      return(NA)
-    }
-  )
+for (i in seq_len(nrow(populations_subset))) {
 
-  if (is.list(out)) {
-    pvals[i] <- out$p
-    theta1[i] <- out$theta1
-    theta2[i] <- out$theta2
-    fips[i] <- populations$countyFIPS[i]
-  }
+    pvals[i] <- tryCatch(W(new_cases_subset[i, ],
+                  population_size = populations_subset$population[i],
+                  breakpoint = 30,
+                  deg_free = 3,
+                  fn = my_spl_fit), 
+                  error = function(e) return(NA))
+    theta1[i] <- tryCatch(W(new_cases_subset[i, ],
+                   population_size = populations_subset$population[i],
+                   breakpoint = 30,
+                   deg_free = 3,
+                   fn = my_spl_fit,
+                   return_thetas = TRUE)[[1]][1], 
+                   error = function(e) return(NA))
+    theta2[i] <- tryCatch(W(new_cases_subset[i, ],
+                   population_size = populations_subset$population[i],
+                   breakpoint = 30,
+                   deg_free = 3,
+                   fn = my_spl_fit, return_thetas = TRUE)[[1]][2], 
+                    error = function(e) return(NA))
+   
+     fips[i] <- populations_subset$countyFIPS[i]
 
-  cat("County", i, "of", nrow(populations), "p-value:", pvals[i], "\n")
+  cat("County", i, "of", nrow(populations_subset), "p-value:", pvals[i], "\n")
 }
 
 # Save p-values

### 2023-11-11 noise_covid_v2.Rmd
changes

@@
 We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.
-This method may also be useful in identifying geographical areas that would benefit from targeted control strategies to reduce superspreading dynamics.
+This method may also be useful in identifying geographical areas that would benefit from targeted control strategies to reduce superspreading.
 Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable.
@@
 
-**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series (by fitting negative binomial models) and its relationship to a hypothesized changepoint.
+**Using county-level SARS-CoV-2 time series, we investigate dispersion in case count time series (by fitting negative binomial models) and its relationship to a hypothesized changepoint.
 Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion changes significantly**.
@@
 However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data.
-
-We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic and visualizations of a dispersion metric.
-
-In both the full data and the subset of data, rows that were all zeroes were omitted.
+We additionally use the full dataset to investigate variability breakpoints using a likelihood ratio test statistic and visualizations of other dispersion metrics.
+In both the full data and the subset of data, time series that were all zeroes were omitted.
 
@@
 $$g(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-$$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$
-  ,where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
+$$log(E[Y_i]/n_i) = \sum_m^M\beta_m h_m(t_i)$$, 
+where $n_i$ is the population size, $Y_i$ is incidence, and $t_i$ is the time point.
 
@@
 
-The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05).
+The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true.
 Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.
@@
 
-The spatial pattern of theta difference after Thanksgiving 2020 shows that there is spatial correlation in the degree of change.
-In particular, spatial clusters of increased dispersion occur on the East Coast and in the South.
+The spatial pattern of the difference in log(theta) after Thanksgiving 2020 shows that there is spatial correlation in the degree of change.
+For example, a spatial clusters of increased dispersion occurs in the South.
 
-![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf){width=30%, height=30%}
+![Mapping the Thanksgiving hypothesis across the US (Log Theta)](./figures/map_difftheta_US.pdf){width=30%, height=30%}
 
 Additionally, the likelihood ratio test statistic does not necessarily track with changes in mean incidence in a county, but rather captures dispersion changes.
-In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.
+In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of $\mu$ is a function of $\mu$ and dispersion, both of which are zero in a stretch of zeroes.
 
@@
 
-In visualizing a surface of the likelihood ratio test statistic and associated direction of the change in theta, it appears that
+Visualizations surfaces of the likelihood ratio test statistic and associated direction of the change in theta for large populations are presented below.
 
@@
 
-## Less dispersion over time
+## Patterns in dispersion over time
 
-In many US counties between 2020-09-27 and 2021-01-25, there is less burstiness (larger theta) in COVID-19 case counts over time.
-In other words, cases fluctuate less around a time-averaged trajectory later on, possibly indicating structure in the population with some groups experiencing different infection rates (overdispersed individual reproduction number), demographic/environmental stochasticity, or fast changes in propagation/reporting at the population level may have occurred earlier on.
-However, the trend of decreased burstiness over this period seems to be modulated by population size: among the largest population counties per state, those in the top half had an average theta change of approx.
-28.95604 (less burstiness), while the small populations had an average change in theta of approx.
--3.741856 (more burstiness).
-This means that in the large populations, dynamics became "smoother", but this didn't happen in the smaller populations, where more burstiness might be a result of cases "stuttering out".
-This is a key observation, as smoother and more predictable case count time series may require less surge capacity in hospitals.
+In many US counties after late fall of 2020, there was more burstiness (smaller log theta) in COVID-19 case counts over time.
+In other words, cases fluctuated more around a time-averaged trajectory later on, possibly indicating structure in the population with some groups experiencing different infection rates (overdispersed individual reproduction number), demographic/environmental stochasticity, or fast changes in propagation/reporting at the population level.
+I will investigate whether the trend of increased burstiness over this period is modulated by factors such population size.
+I predict that in very large populations, dynamics may be "smoother", but this may not be the case in other populations, where more burstiness might be a result of cases "stuttering out" or dispersion at the individual level affecting case count time series.
+This research avenue has practical implications: smoother and more predictable case count time series may require less surge capacity in hospitals.
 
 We hypothesized that there would be an increase in burstiness (decrease in theta) around Thanksgiving 2020 due to people gathering and causing clusters of infection.
-However, we observed that in many counties, although there was more burstiness (a decrease in theta), it was a relatively small change, and not modulated by population size.
-The average difference between theta estimated in the window starting at 2020-11-26 and theta estimated in the window ending 2020-11-25 was -1.77944 in the top half and -1.104289 in the bottom half of population sizes.
-People may have avoided gathering due to personal preference, cultural norms, and other factors, contributing to less superspreading than anticipated
+However, we observed that in many counties, this was not the case.
+People may have avoided gathering due to personal preference, cultural norms, and other factors, contributing to less superspreading than anticipated in some locations.
 
-![Compare](./figures/compare.pdf){width=30%, height=30%}
+![Compare](./figures/compare1.pdf){width=30%, height=30%}
 

### 2023-11-11 code/W.R

@@ -42,15 +42,15 @@ W <- function(y, population_size, breakpoint, deg_free = 3, fn = my_spl_fit, ver
     p = 2 * (1 - pnorm(abs(test_stat)))
 
     if (return_theta_diff) {
-        return(theta1 - theta2)
+        return(log(theta1) - log(theta2))
     } 
     
     if (return_theta) {
-      return(spline_mod.all$theta)
+      return(log(spline_mod.all$theta))
     }
     
     if (return_thetas){
-      return(list(c(theta1, theta2)))
+      return(list(c(log(theta1), log(theta2))))
     }
     
     else {

### 2023-11-11 code/compare.R

@@ -2,9 +2,9 @@
 load("data/processed_long_dat.Rdata")
 dates <- dates[30:(length(new_cases[1,]) - 30 + 1)]
 
-#load("data/lrt_lg_pops.Rdata")
-#load("data/thetadiff_lg_pops.Rdata")
-load("data/theta_lg_pops.Rdata")
+#load("data/lrt_lg_pops.Rdata") # doesn't need to be put on log scale
+#load("data/thetadiff_lg_pops.Rdata") # already on log scale
+load("data/theta_lg_pops.Rdata") # already on log scale
 
 
 # Indices for largest counties in each state
@@ -40,7 +40,7 @@ for (i in 1:length(keep)){
   abline(v = 281, col = "red")
   #plot(lrt_stats[i, ], type = "l", col = 2, main = "LRT")
   #plot(thetadiffs[i, ], type = "l", col = 3, main = "Theta Difference")
-  plot(log(thetas[i, ]), type = "l", col = 4, main = "Theta")
+  plot(log(thetas[i, ]), type = "l", col = 4, main = "Log Theta")
   abline(v = 281, col = "red")
 }
 

### 2023-11-11 code/create_surfaces.R

@@ -53,7 +53,7 @@ y <- seq_len(ncol(z))
 
 filename <- "figures/thetadiff_surface.pdf"
 
-pdf(filename, width = 6, height = 6)
+pdf(filename, width = 8, height = 8)
 
 # Plot
 par(pin = c(7, 5))
@@ -62,7 +62,7 @@ image(x, y, z,
       yaxt = "n",
       xlab = "",
       ylab = "Locations",
-      col = c("blue", "green", "purple", "pink", "red"))
+      col = rev(mako(32)))
 
 par(las = 2)
 axis(1, x, dates, tick = FALSE)

### 2023-11-11 code/map_thanksgiving_RA.R

@@ -18,7 +18,7 @@ dat$county <- gsub(" County", "", dat$County.Name)
 dat$county <- gsub(" Parish", "", dat$county)
 dat$county <- tolower(dat$county)
 dat$state <- dat$State
-dat$dtheta <- dat$theta1 - dat$theta2
+dat$dtheta <- dat$theta2 - dat$theta1
 
 ## a clamped version of dtheta
 clamp <- function(x, c) {
@@ -38,7 +38,7 @@ filename <- "./figures/map_difftheta_US.pdf"
 
 pdf(filename, width = 6, height = 6)
 
-plot(dat.clor["dtheta_clamped"], main = "dtheta_clamped")
+plot(dat.clor["dtheta_clamped"], main = "Change in dispersion parameter - Thanksgiving 2020")
 
 dev.off()
 

### 2023-11-11 code/thetadifflrt_mat_LRTscan.R

@@ -61,14 +61,14 @@ for (j in 1:length(keep)){
     #                     df2 = 3)$lambda, error = function(e) return(NA))
     # 
     # lrt_stat = c(lrt_stat, test)
+# 
+#     test.1 = tryCatch(W(y = Y, population_size = populations_lg$population[j],
+#                          breakpoint = 30, deg_free = 3,
+#                          fn = my_spl_fit, verbose = FALSE, return_theta_diff = TRUE),
+#                        error = function(e) return(NA))
+# 
+#     thetadiff = c(thetadiff, test.1)
 
-    # test.1 = tryCatch(W(y = Y, population_size = populations_lg$population[j],
-    #                      breakpoint = 30, deg_free = 3,
-    #                      fn = my_spl_fit, verbose = FALSE, return_theta_diff = TRUE),
-    #                    error = function(e) return(NA))
-    # 
-    # thetadiff = c(thetadiff, test.1)
-    
     test.2 = tryCatch(W(y = Y, population_size = populations_lg$population[j],
                         breakpoint = 30, deg_free = 3,
                         fn = my_spl_fit, verbose = FALSE, return_theta = TRUE),
@@ -82,7 +82,7 @@ for (j in 1:length(keep)){
   # 
   # thetadiffs[j,] <- thetadiff
   
-  thetas[j,] <- theta
+   thetas[j,] <- theta
   
 }
 
@@ -114,13 +114,13 @@ dates <- dates[30:(length(this_series) - 30 + 1)]
 
 filename <- "figures/LRT_scan_figure.pdf"
 
-pdf(filename, width = 3, height = 3)
+pdf(filename, width = 6, height = 6)
 
 par(mfrow = c(2, 1))
 
 plot(lrt_stats[1, ] ~ dates, type = "l", xlab = "Date", ylab = "LRT Statistic")
 
-plot(this_series[30:(length(this_series) - 30 + 1)] ~ dates, col = "red", type = "l")
+plot(this_series[30:(length(this_series) - 30 + 1)] ~ dates, col = "red", type = "l", xlab = "Date", ylab = "Incidence")
 
 abline(h = quantile(lrt_stats[1, ], probs = 0.95, na.rm =TRUE), col = "black")
 

### 2023-11-11 code/vis_empirical_increase.R

@@ -4,7 +4,7 @@ load("./data/simulated_curves.Rdata")
 
 filename <- "figures/empirical_increase.pdf"
 
-pdf(filename, width = 6, height = 6)
+pdf(filename, width=6, height=6)
 
 american_thanksgiving <- as.Date("2020-11-26")
 
@@ -16,19 +16,19 @@ days <- seq(start_date, end_date, by = "day")
 
 par(mfrow = c(2,2))
 
-plot(new_cases_subset[1862, ] ~ days,  ylab = "Daily Incidence in FIPS 36067 (Wald test p-value = 0.017)", 
+plot(new_cases_subset[1862, ] ~ days,  ylab = "Wald test p-value = 0.017", 
      xlab = "Date", ylim = c(0, 1400)) 
 abline(v = american_thanksgiving, col = 2)
 
-plot(new_cases_subset[1863, ] ~ days, ylab = "Daily Incidence in FIPS 36069 (Wald test p_value = 0.099)", 
+plot(new_cases_subset[1863, ] ~ days, ylab = "Wald test p_value = 0.099", 
      xlab ="Date", ylim = c(0, 1400)) 
 abline(v = american_thanksgiving, col = 2)
 
-plot(curves[220, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.004)", 
+plot(curves[220, ] ~ days, ylab = "Wald test p-value = 0.004", 
      xlab = "Date", ylim = c(0, 1400))
 abline(v = american_thanksgiving, col = 2)
 
-plot(curves[212, ] ~ days, ylab = "Simulated Daily Incidence (Wald test p-value = 0.260)", 
+plot(curves[212, ] ~ days, ylab = "Wald test p-value = 0.260", 
      xlab = "Date", ylim = c(0, 1400))
 abline(v = american_thanksgiving, col = 2)
 

### 2023-11-11 code/vis_signific_state.R

@@ -18,10 +18,10 @@ par(mfrow = c(2, 3))
 
 for (i in FL_index){
   
-  mu <- my_spl_fit(Y = new_cases_subset[i,], population = populations_subset$population[i], 
-                   inds = 1:length(new_cases_subset[i,]), df = 3)$mu
+  mu <- my_spl_fit(Y = new_cases_subset[i,1:30], population = populations_subset$population[i], 
+                   inds = 1:length(new_cases_subset[i,1:30]), df = 3)$mu
   
-  plot(1:length(new_cases_subset[i,]), new_cases_subset[i,], xlab = , ylab = )
+  plot(1:length(new_cases_subset[i,1:30]), new_cases_subset[i,1:30], xlab = "Day Index", ylab = "Cases")
   lines(mu, col=2)
 }
 

### 2023-11-11 code/wald_counties.R

@@ -54,7 +54,7 @@ sig <- pvals[keep] < p_max
 par(cex = 1.5)
 par(pin = c(4, 4))
 
-plot(log(theta1[keep]), log(theta2[keep]),
+plot(theta1[keep], theta2[keep],
   cex = ifelse(sig, 1, 0.5),
   col = ifelse(sig, 2, grey(0.5)),
   xlab = expression(paste("log ", theta[1])),

### 2024-01-19 noise_covid_v2.Rmd

@@
 
-Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
+Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.
+Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions.
 Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al.

### 2024-06-05 plos_latex_template.tex
changes

% Template for PLoS
% Version 3.6 Aug 2022
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Rachael Aber\textsuperscript{1,2\Yinyang},
Benjamin Dalziel\textsuperscript{2\Yinyang},
Yanming Di\textsuperscript{1\textcurrency},
Name4 Surname\textsuperscript{2},
Name5 Surname\textsuperscript{2\ddag},
Name6 Surname\textsuperscript{2\ddag},
Name7 Surname\textsuperscript{1,2,3*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Department of Integrative Biology, Oregon State University, Corvallis, Oregon, United States of America
\\
\textbf{2} Department of Statistics, Oregon State University, Corvallis, Oregon, United States of America
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* correspondingauthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Metrics of variability are often overlooked and useful ways to understand epidemic dynamics. For instance, superspreading of SARS-CoV-2 can be elucidated by utilizing such metrics. Our method identifies shifts in population-level incidence dispersion, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity in certain months. Although classical theory predicts that there will be less dispersion when incidence is higher, we consider a more general negative binomial regression framework to account for processes that may also affect the spread of cases. We investigate changes in dispersion and find that there are increases in dispersion around holiday periods in many US counties, concurrent with incidence increases. In addition, highly overdispersed patterns occur more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility, and reporting. Our method is robust to changes in incidence and to population size, allowing for quantification of dispersion-indicative of superspreading dynamics-without artifactual contributions from these features.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Time series of observed infectious disease incidence are, to varying degrees, "noisy", showing higher frequency oscillations around trends at broader temporal scales. Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, 'measurement error'), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R). Therefore, variability can contain information important to understanding epidemic dynamics and societal responses. Yet metrics of variability are often overlooked ways to understand these dynamics, and techniques based on variability in epidemic time series are still emerging. One area of interest is how variability is related to different phases of an epidemic. For instance, Graham et al. (2019) use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen. Similarly, it was recently found that 
, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity. Recently, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection from incidence data (Schneckenreither et al. 2023). Analyzing variability in terms of bursts of incidence is also important for planning surge capacity (Wallinga 2018). Sun et al. (2021) found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.

Dispersion around a rolling mean (moving window) of an incidence time series may contain information about the size and frequency of local outbreaks. From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. A 'mean crowding' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat (Lloyd 1967). This is a useful way to think about dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity. Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics (Lloyd-Smith 2005), so dispersion in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process. Individual-level variability in transmission is most often studied using contact-tracing data. However, contact tracing data requires intense investment of resources (Kretzschmar et al. 2020), so analysis of incidence data may often be more feasible.

One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. For instance, if we model a process as Poisson, we assume that if the mean of the process is large, the variance will be large as well. Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings. Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accomodated by a Poisson model. The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model. Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).

We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence. We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the pandemic. From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003).

\section*{Materials and methods}
\subsection*{Introduction to the method}

% For figure citations, please use "Fig" instead of "Figure".
Classical theory put forward by Grenfell et al. (2002) proposed that incidence can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion equal to previous incidence:

\begin{eqnarray}
\label{eq:schemeP}
	\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)
\end{eqnarray}

However, other processes besides the current number infected might affect dispersion, so we instead investigated changes in 
 to understand important processes that may leave a signal in dispersion. In other words, we estimated 
 over time. A persistent challenge in investigating changes in variability has been "spurious correlation" with population size. Since population size influences mean and variance in count data and thus could have an impact on estimates of dispersion, we robustly adjusted for population size using an offset in the model. In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size. The general framework is that incidence is drawn from a negative binomial distribution with time-varying mean and dispersion parameters (that vary more slowly than the mean). The model is formulated as follows:

 The linear predictor includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts. Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). b) There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual):

\begin{eqnarray}
\label{eq:schemeP}
    log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)

    log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)

    log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)
\end{eqnarray}

So, the form of the probability mass function is:
$$f_t(I) = {I + \theta - 1 \choose I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta$$
This form has expectation and variance as follows:
$$E(I) = \mu$$
$$Var(I) = \mu + \frac{\mu^2}{\theta}$$

\subsection*{Application to simulated data}

For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series--3,000 epidemic curves over 60 days each were produced, and the LRT test procedure was applied to each. Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach.

\subsection*{Application to empirical data}
To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007), along with evaluation of dispersion statistics.
We then estimated $\mu_t$ and $\theta$ using iterative reweighted least-squares (procedure implemented via the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al. (2011)) with a moving window approach. For each window, $\mu_t$ was estimated using a spline function in time, and a single value of $\theta$ was estimated for the window. By moving the window one time step at a time, a time series for $\theta_t$ was produced. We investigated large counties (top 4% in each state), due to the power constraints discussed above.

% Results and Discussion can be combined.
\section*{Results}
We found that the negative binomial method is robust to changes in population size (for population sizes of at least 10,000). For an analytical derivation, see Supplement 1. The criteria are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size. In row one and two of Figure 1, we illustrated that a drop in $\theta$ is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in $\theta$ corresponding to a decrease in variability around the trend in incidence.

Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. Interestingly, the most dispersed category reaches its highest proportion near the end of the timeframe (Figure 1.(a)) In addition, there are increases in dispersion around the holiday periods in the dataset (Figure 1.(b)), concurrent with increases in incidence (Figure 1.(c)). The evidence for a change in $\theta$ was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Figure 1.(d)).

% Place figure captions after the first paragraph in which they are cited.
\begin{figure}[!h]
\caption{{\bf Fig 1.}
Detecting dispersion changes in incidence time series in populations of different sizes. A: Simulated incidence when dispersion is constant. B: When dispersion changes during the epidemic. C: Daily COVID-19 cases in Jefferson County, AL between 2020-09-10 and 2020-11-09 (X-axis is days since beginning of inclusion period for the model fit). D: Constant dispersion used in generation of above. E: Changing dispersion used in generation of above. F: Dispersion estimates from model fit to the above series. G: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates similar p-value distribution across different population sizes (each pane is one population size). O and X mark the null and alternative hypotheses indicated in panels D and E. 
 }
\label{fig1}
\end{figure}

\begin{figure}[!h]
\caption{{\bf Fig 2.}
Incidence and dispersion between 2020-02-20 and 2023-03-19 in large counties in the US. A: Binned log of the dispersion parameter over time. B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis). C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis). D: LRT p-values over time as well as for each of the large counties (y-axis).
}
\label{fig1}
\end{figure}

The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variability than when incidence is lower. In other words, what makes a change in dispersion meaningful is how it affects variance and variance-mean relationships. For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. A change from $\theta = 1000$ to $\theta = 100$ is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
$$var = \mu + \mu^2/\theta$$
So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing (Dalziel et al. 2018), or increased locally experienced mean density (Lloyd 1967). To illustrate this using the empirical data, we observed an increase from $\theta$ = 2.707735e+16 to $\theta = 21.8711$, which, conditional on the observed case count of 3$$, would lead to an increase in variance from around 3 to around 3.4. In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.

\begin{figure}[!h]
\caption{{\bf Fig 3.}

}
\label{fig1}
\end{figure}

% Place tables after the first paragraph in which they are cited.
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
\subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
\subsubsection*{3rd level heading} 
Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat.

\begin{enumerate}
	\item{react}
	\item{diffuse free particles}
	\item{increment time by dt and go to 1}
\end{enumerate}

\subsection*{Sed ac quam id nisi malesuada congue}

Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{itemize}
	\item First bulleted item.
	\item Second bulleted item.
	\item Third bulleted item.
\end{itemize}

\section*{Discussion}
We presented an approach to compare variability in epidemic time series that does not detect artifacts based on population size and incidence. Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series. Burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window (Jo et al. 2020), allowing comparison of variability within one location over time . Similarly, spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments (Loo et al. 2021). Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public  goal can be achieved with limited resrouces. Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts (Lloyd-Smith et al. 2005), so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.

Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (peak incidence) and corresponds to significant increases in variance when incidence is high. Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties.

In addition, a big city with more hospitals could effectively be analogous to a collection of small towns, and therefore experience a benefit of reduced variance-mean relationships, especially during periods of peak incidence.
This implies that there may be merit to revising current public health strategies. Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series’ showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest (Dalziel et al. 2018). So, additional research is needed to understand the balance between burstiness of small communities and the potentially less severe effect of increased dispersion.

Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992), our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case. Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time (Earn et al. 1998).

\section*{Conclusion}

CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec, auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Appendix}.

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Lorem ipsum.} Analytical approach: robust to population size changes.

\paragraph*{S1 File.}
\label{S1_File}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Video.}
\label{S1_Video}
{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\begin{thebibliography}{10}

\bibitem{bib1}
Conant GC, Wolfe KH.
\newblock {{T}urning a hobby into a job: how duplicated genes find new
  functions}.
\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.

\bibitem{bib2}
Ohno S.
\newblock Evolution by gene duplication.
\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
  Springer-Verlag.; 1970.

\bibitem{bib3}
Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
  infection through a transposon insertion followed by a {D}uplication}.
\newblock PLoS Genet. 2011 Oct;7(10):e1002337.

\end{thebibliography}

\end{document}

### 2024-06-05 code/2-generate_time_series.R

@@ -14,9 +14,9 @@ rm(new_cases_lg_weekly)
 final_attack_rate <- 0.1
 peak_time <- 30
 peak_width <- 200
-population_min <- min(populations)
-population_mid <- round(mean(populations))
-population_max <- max(populations)
+population_min <- 10000
+population_mid <- signif(min(populations), 2)
+population_max <- signif(max(populations), 2)
 theta_min <- 3
 theta_max <- 30
 breakpoint_min <- 20

### 2024-06-05 code/5-figure1.R

@@ -23,27 +23,27 @@ dates <- dates[35: length(dates)]
 thetas <- thetas[1, 35:dim(thetas2)[2]]
 
 # Row 1
-plot(curves[2701,], xlab = "Day", type = "l", ylab = "Incidence")
-mtext("A", side = 3)
-plot(curves[2710,], xlab = "Day", type = "l", ylab = "Incidence")
-mtext("B", side = 3)
+plot(curves[1701,], xlab = "Day", type = "l", ylab = "Incidence")
+mtext("a", side = 3, line = 1, adj = 0)
+plot(curves[1791,], xlab = "Day", type = "l", ylab = "Incidence")
+mtext("b", side = 3, line = 1, adj = 0)
 plot(dates[170:230], series[170:230], xaxt = "n", xlab = "Day", type = "l", ylab = "Incidence")
 
 at <- seq.Date(from = min(dates[170:230]), max(dates[170:230]), by = '10 days')
 labels <- seq(0, 60, by = 10)
 
 axis.Date(1, at = at, labels = labels)
-mtext("C", side = 3)
+mtext("c", side = 3, line = 1, adj = 0 )
 
 
 # Row 2
-plot(rep(curve_parms$theta1[2701], times = length(curves[2701,])),
-     xlab = "Day", ylab = expression(theta), ylim = c(0, 30), col = 4, cex = .1)
+plot(rep(curve_parms$theta1[1701], times = length(curves[1701,])),
+     xlab = "Day", ylab = expression(theta), ylim = c(0, 30), col = 4, cex = .1, type = "l")
 mtext("d", side = 3, line = 1, adj = 0)
 mtext("O", side = 3, line = -2)
-plot(c(rep(curve_parms$theta1[2710], times = length(curves[2710,])/2), 
-       rep(curve_parms$theta2[2710], times = length(curves[2710,])/2)),
-     xlab = "Day", ylab = expression(theta), ylim = c(0, 30), col = 4, cex = .1)
+plot(c(rep(curve_parms$theta1[1791], times = length(curves[1791,])/2), 
+       rep(curve_parms$theta2[1791], times = length(curves[1791,])/2)),
+     xlab = "Day", ylab = expression(theta), ylim = c(0, 30), col = 4, cex = .1, type = "l")
 mtext("e", side = 3, line = 1, adj = 0)
 mtext("X", side = 3, line = -2)
[…]

### 2024-06-14 plos_latex_template.tex
add paragraph state variable

@@
 The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model.
+In distributions that form part of the exponential family, the dispersion parameter determines the mean-variance relationship.
+
+$$f(y_i; \theta_i, \phi) = exp\{\theta_i y_i - b(\theta_i)/a_i(\phi) - c(y_i; \phi)\}, i = 1,...,n$$
+$$Var(y_i) = b''(\theta_i) a_i(\phi) $$
+
+Where $\theta_i$ is the natural (canonical) parameter and $\phi$ is the dispersion parameter.
+
 Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
@@
 
-\begin{eqnarray}
-\label{eq:schemeP}
-	\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)
-\end{eqnarray}
+    $$\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)$$
 
@@
 
-\begin{eqnarray}
-\label{eq:schemeP}
-    log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)
-
-    log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)
-
-    log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)
-\end{eqnarray}
+    $$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+    $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
+    $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
 
 So, the form of the probability mass function is:
+
 $$f_t(I) = {I + \theta - 1 \choose I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta$$
+
 This form has expectation and variance as follows:
+
 $$E(I) = \mu$$
@@
 }
-\label{fig1}
+\label{fig2}
 \end{figure}
@@
 A change from $\theta = 1000$ to $\theta = 100$ is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
+
 $$var = \mu + \mu^2/\theta$$
+
 So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence.
@@
 2018), or increased locally experienced mean density (Lloyd 1967).
-To illustrate this using the empirical data, we observed an increase from $\theta$ = 2.707735e+16 to $\theta = 21.8711$, which, conditional on the observed case count of 3$$, would lead to an increase in variance from around 3 to around 3.4.
+To illustrate this using the empirical data, we observed an increase from $\theta = 2.707735e+16$ to $\theta = 21.8711$, which, conditional on the observed case count of 3, would lead to an increase in variance from around 3 to around 3.4.
 In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.
+
+Additionally, $\theta$ seems to behave like a state variable.
+That is, its rate of change is dependent on its value and the values of other state variables (like incidence), which determine the "state" of the system.
+State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics.
+This predictability and simplicity are crucial for both theoretical studies and practical applications, and creating a reliable description of this (epidemic) system is overdue.
 
@@
 \caption{{\bf Fig 3.}
-
+caption
 }
-\label{fig1}
+\label{fig3}
 \end{figure}

### 2024-06-24 manuscript/ms.tex
Update manuscript formatting

--- 2024-06-14 plos_latex_template.tex
@@
 % ligatures disabled
-\usepackage[nopatch=eqnum]{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
+% \usepackage[nopatch=eqnum]{microtype}
+% \DisableLigatures[f]{encoding = *, family = * }
 
@@
 \\
-Rachael Aber\textsuperscript{1,2\Yinyang},
-Benjamin Dalziel\textsuperscript{2\Yinyang},
-Yanming Di\textsuperscript{1\textcurrency},
-Name4 Surname\textsuperscript{2},
-Name5 Surname\textsuperscript{2\ddag},
-Name6 Surname\textsuperscript{2\ddag},
-Name7 Surname\textsuperscript{1,2,3*},
-with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
+Rachael Aber\textsuperscript{1,2},
+Yanming Di\textsuperscript{2},
+Benjamin Dalziel\textsuperscript{1, 3},
 \\
 \bigskip
-\textbf{1} Department of Integrative Biology, Oregon State University, Corvallis, Oregon, United States of America
+\textbf{1} Department of Integrative Biology, Oregon State University, Corvallis, Oregon, USA
 \\
-\textbf{2} Department of Statistics, Oregon State University, Corvallis, Oregon, United States of America
+\textbf{2} Department of Statistics, Corvallis, Oregon, Oregon State University, Corvallis, Oregon, USA
 \\
+\textbf{3} Department of Mathematics, Oregon State University, Corvallis, Oregon, USA
 \bigskip
@@
 % Primary Equal Contribution Note
-\Yinyang These authors contributed equally to this work.
+% \Yinyang These authors contributed equally to this work.
 
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors also contributed equally to this work.
+% \ddag These authors also contributed equally to this work.
 
 % Current address notes
-\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+%%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
 % \textcurrency b Insert second current address 
@@
 % Deceased author note
-\dag Deceased
+% \dag Deceased
 
 % Group/Consortium Author Note
-\textpilcrow Membership list can be found in the Acknowledgments section.
+%\textpilcrow Membership list can be found in the Acknowledgments section.
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* correspondingauthor@institute.edu
+* aberr@oregonstate.edu
 
@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. 
 Morbi consectetur est vel gravida pretium. 
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. 
 Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. 
@@
 One area of interest is how variability is related to different phases of an epidemic.
-For instance, Graham et al.
-(2019) use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
-Similarly, it was recently found that
-, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al.
-2022).
+For instance, \cite{graham2019measles} use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
+Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam2022time}.
 Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
@@
 Sun et al.
-(2021) found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
+\cite{sun2021transmission} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
 
@@
 
-$$f(y_i; \theta_i, \phi) = exp\{\theta_i y_i - b(\theta_i)/a_i(\phi) - c(y_i; \phi)\}, i = 1,...,n$$
-$$Var(y_i) = b''(\theta_i) a_i(\phi) $$
+\begin{align}
+  f(y_i| \theta_i, \phi) &= \exp\left( \dfrac{\theta_i y_i - b (\theta_i)}{a_i(\phi)} - c(y_i, \phi) \right) \\
+  \notag \\
+  Var(y_i | \phi) &= b''(\theta_i) a_i(\phi)
+\end{align}
 
@@
 
-    $$\mathrm{I_t+1} = NB(\mu = \lambda_t+1, \theta_t = I_t)$$
+\begin{equation}
+    \mathrm{I_t} = NB(\mu = \lambda_t, \theta_t = I_{t-1})
+\end{equation}
 
@@
  over time.
-A persistent challenge in investigating changes in variability has been "spurious correlation" with population size.
+A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size.
 Since population size influences mean and variance in count data and thus could have an impact on estimates of dispersion, we robustly adjusted for population size using an offset in the model.
@@
 
-    $$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-    $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
-    $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$
+\begin{align}
+  log(E[Y_i]/n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
+  \log(E[Y_i])-log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
+  log(E[Y_i]) &= \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i) 
+\end{align}
 
@@
 
-$$f_t(I) = {I + \theta - 1 \choose I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta$$
+\begin{equation}
+x  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
+\end{equation}
 
@@
 % 
-\begin{thebibliography}{10}
-
-\bibitem{bib1}
-Conant GC, Wolfe KH.
-\newblock {{T}urning a hobby into a job: how duplicated genes find new
-  functions}.
-\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-
-\bibitem{bib2}
-Ohno S.
-\newblock Evolution by gene duplication.
-\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-  Springer-Verlag.; 1970.
-
-\bibitem{bib3}
-Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-  infection through a transposon insertion followed by a {D}uplication}.
-\newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-
-\end{thebibliography}
+\bibliography{references}
 

### 2024-06-25 manuscript/ms.tex
add refs

@@
 One area of interest is how variability is related to different phases of an epidemic. 
-For instance, \cite{graham2019measles} use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
-Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam2022time}.
+For instance, \cite{graham_measles_2019} use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
+Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
 Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
-Recently, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection from incidence data (Schneckenreither et al.
-2023).
-Analyzing variability in terms of bursts of incidence is also important for planning surge capacity (Wallinga 2018).
+Recently, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection from incidence data \cite{schneckenreither_assessing_2023}. 
+Analyzing variability in terms of bursts of incidence is also important for planning surge capacity \cite{wallinga_metropolitan_2018}. 
 Sun et al.
-\cite{sun2021transmission} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
+\cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
 
@@
 From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. 
-A 'mean crowding' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat (Lloyd 1967).
+A 'mean crowding' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
 This is a useful way to think about dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
@@
 Individual-level variability in transmission is most often studied using contact-tracing data. 
-However, contact tracing data requires intense investment of resources (Kretzschmar et al.
-2020), so analysis of incidence data may often be more feasible.
+However, contact tracing data requires intense investment of resources \cite{kretzschmar_impact_2020}, so analysis of incidence data may often be more feasible.
 
@@
 Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accomodated by a Poisson model. 
-The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009).
+The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
 The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model.
@@
 
-\begin{align}
-  f(y_i| \theta_i, \phi) &= \exp\left( \dfrac{\theta_i y_i - b (\theta_i)}{a_i(\phi)} - c(y_i, \phi) \right) \\
-  \notag \\
-  Var(y_i | \phi) &= b''(\theta_i) a_i(\phi)
-\end{align}
-
-Where $\theta_i$ is the natural (canonical) parameter and $\phi$ is the dispersion parameter.
-
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992).
-
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion \cite{barron_analysis_1992}.
 We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence.
 We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the pandemic. 
-From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean (Benedetti-Cecchi 2003).
+From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
 
@@
 Classical theory put forward by Grenfell et al.
-(2002) proposed that incidence can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion equal to previous incidence:
+\cite{grenfell_dynamics_2002} proposed that incidence can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion equal to previous incidence:
 
@@
 
-However, other processes besides the current number infected might affect dispersion, so we instead investigated changes in 
- to understand important processes that may leave a signal in dispersion. In other words, we estimated 
- over time.
+However, other processes besides the current number infected might affect dispersion, so we instead investigated changes over time to understand important processes that may leave a signal in dispersion.
 A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
@@
 The general framework is that incidence is drawn from a negative binomial distribution with time-varying mean and dispersion parameters (that vary more slowly than the mean). 
-The model is formulated as follows:
-
- The linear predictor includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts.
-Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al.
-2019).
-A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009).
-b) There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual):
+The model is formulated with a linear predictor that includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts. 
+Natural splines are cubic splines which are linear outside of the boundary knots \cite{perperoglou_review_2019}. 
+A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence \cite{davis_negative_2009}. 
+There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual):
 
@@
 
-$$E(I) = \mu$$
-$$Var(I) = \mu + \frac{\mu^2}{\theta}$$
+\begin{align}
+E(I) &= \mu
+Var(I) &= \mu + \frac{\mu^2}{\theta}
+\end{align}
 
@@
 \subsection*{Application to empirical data}
-To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007), along with evaluation of dispersion statistics.
-We then estimated $\mu_t$ and $\theta$ using iterative reweighted least-squares (procedure implemented via the NBPSeq R package (https://CRAN.R-project.org/package=NBPSeq) and from Di et al.
-(2011)) with a moving window approach.
-For each window, $\mu_t$ was estimated using a spline function in time, and a single value of $\theta$ was estimated for the window.
-By moving the window one time step at a time, a time series for $\theta_t$ was produced.
-We investigated large counties (top 4% in each state), due to the power constraints discussed above.
+To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
+We then estimated \begin{math}\mu_t\end{math} and \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package (CITE) and from Di et al.
+\cite{yanming_nbp_2011} with a moving window approach.
+For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for the window.
+By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced.
+We investigated large counties (top 4\% in each state), due to the power constraints discussed above.
 
@@
 The criteria are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size. 
-In row one and two of Figure 1, we illustrated that a drop in $\theta$ is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in $\theta$ corresponding to a decrease in variability around the trend in incidence.
+In row one and two of Figure 1, we illustrated that a drop in \begin{math}\theta\end{math} is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
 
 Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
-Interestingly, the most dispersed category reaches its highest proportion near the end of the timeframe (Figure 1.(a)) In addition, there are increases in dispersion around the holiday periods in the dataset (Figure 1.(b)), concurrent with increases in incidence (Figure 1.(c)).
-The evidence for a change in $\theta$ was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Figure 1.(d)).
+Interestingly, the most dispersed category reaches its highest proportion near the end of the timeframe (Figure 1.(a)).
+In addition, there are increases in dispersion around the holiday periods in the dataset (Figure 1.(b)), concurrent with increases in incidence (Figure 1.(c)). 
+The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Figure 1.(d)).
 
@@
 For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. 
-A change from $\theta = 1000$ to $\theta = 100$ is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
+A change from \begin{math}\theta = 1000\end{math} to \begin{math}\theta = 100\end{math} is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
 
-$$var = \mu + \mu^2/\theta$$
+\begin{equation}
+var = \mu + \mu^2/\theta  
+\end{equation}
 
 So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. 
-Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e.
-localized surges) which may necessitate more surge capacity in hospitals and testing centers.
-Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing (Dalziel et al.
-2018), or increased locally experienced mean density (Lloyd 1967).
-To illustrate this using the empirical data, we observed an increase from $\theta = 2.707735e+16$ to $\theta = 21.8711$, which, conditional on the observed case count of 3, would lead to an increase in variance from around 3 to around 3.4.
+Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
+Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
+To illustrate this using the empirical data, we observed an increase from \begin{math}\theta = 2.707735e+16\end{math} to \begin{math}\theta = 21.8711\end{math}, which, conditional on the observed case count of 3, would lead to an increase in variance from around 3 to around 3.4.
 In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.
 
-Additionally, $\theta$ seems to behave like a state variable.
+Additionally, \begin{math}\theta\end{math} seems to behave like a state variable.
 That is, its rate of change is dependent on its value and the values of other state variables (like incidence), which determine the "state" of the system.
@@
 Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series.
-Burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window (Jo et al.
-2020), allowing comparison of variability within one location over time .
-Similarly, spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments (Loo et al.
-2021).
+Burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
+Similarly, spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}.
 Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public  goal can be achieved with limited resrouces.
-Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts (Lloyd-Smith et al.
-2005), so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
+Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
 
@@
 This implies that there may be merit to revising current public health strategies. 
-Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series’ showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest (Dalziel et al.
-2018).
+Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series’ showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
 So, additional research is needed to understand the balance between burstiness of small communities and the potentially less severe effect of increased dispersion.
 
-Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) (Barron 1992), our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.
-Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time (Earn et al.
-1998).
+Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.
+Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.
 

### 2024-06-26 manuscript/ms.tex
Update ms.tex

@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. 
-Morbi consectetur est vel gravida pretium. 
-Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. 
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. 
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id. 
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae. 
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Understanding how diseases spread is crucial for managing epidemics, but traditional metrics often overlook the variability in how diseases like COVID-19 spread. 
+We present a method to identify shifts in the spread patterns of SARS-CoV-2, the virus that causes COVID-19. 
+By examining how the virus spreads differently at various times and places, we can better predict and prepare for surges in cases.
+Our study uses negative binomial regression to account for various factors that might influence the spread of cases, allowing us to use a single parameter to infer the degree of clustering of cases. 
+We found that during holiday periods, the spread of COVID-19 becomes more erratic, with some individuals spreading the virus to many more people than others. 
+Additionally, as the pandemic progressed, the spread of the virus became erratic, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to it.
+Our method is able to accurately measure spreading heterogeneity, regardless of the number of cases or the size of the underlying population, providing a way to understand the dynamics of superspreading events across a range of locations.
+This can help public health officials better anticipate and manage outbreaks, especially during times when the virus spreads more unpredictably.
 
@@
 \section*{Introduction}
-Time series of observed infectious disease incidence are, to varying degrees, "noisy", showing higher frequency oscillations around trends at broader temporal scales.
-Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, 'measurement error'), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
+Time series of observed infectious disease incidence are, to varying degrees, ``noisy'', showing higher frequency oscillations around trends at broader temporal scales.
+Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, measurement error), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
 Therefore, variability can contain information important to understanding epidemic dynamics and societal responses. 
@@
 Additionally, \begin{math}\theta\end{math} seems to behave like a state variable.
-That is, its rate of change is dependent on its value and the values of other state variables (like incidence), which determine the "state" of the system.
+That is, its rate of change is dependent on its own value and the values of other state variables (like incidence), which determine the "state" of the system.
 State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics.

### 2024-07-03 manuscript/ms.tex
Minor LateX fixes in ms and refs

@@
 From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. 
-A 'mean crowding' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
+A ``mean crowding'' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
 This is a useful way to think about dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
@@
 So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. 
-Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
+Raising variance relative to mean implies spatiotemporal ``crowding'' of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
 Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
@@
 Additionally, \begin{math}\theta\end{math} seems to behave like a state variable.
-That is, its rate of change is dependent on its own value and the values of other state variables (like incidence), which determine the "state" of the system.
+That is, its rate of change is dependent on its own value and the values of other state variables (like incidence), which determine the ``state'' of the system.
 State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics.

### 2024-07-03 manuscript/ms.tex
add ms.pdf

--- 2024-06-25 manuscript/ms.tex
@@
 \usepackage{array}
+
+% graphicx package
+\usepackage{graphicx}
+
+%Path relative to the main .tex file 
+\graphicspath{ {./figures/} }
 
@@
 Metrics of variability are often overlooked and useful ways to understand epidemic dynamics. 
-For instance, superspreading of SARS-CoV-2 can be elucidated by utilizing such metrics.
-Our method identifies shifts in population-level incidence dispersion, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity in certain months.
-Although classical theory predicts that there will be less dispersion when incidence is higher, we consider a more general negative binomial regression framework to account for processes that may also affect the spread of cases.
-We investigate changes in dispersion and find that there are increases in dispersion around holiday periods in many US counties, concurrent with incidence increases. 
+For instance, superspreading of SARS-CoV-2 (the virus that causes COVID-19) can be elucidated by utilizing such metrics. 
+Our method identifies shifts in population-level dispersion in COVID-19 cases, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity in certain months.
+Although classical theory predicts that there will be less dispersion when incidence is higher, we considered a more general negative binomial regression framework to account for processes that may also affect the spread of cases.
+We investigated changes in dispersion and found that there are increases in dispersion around holiday periods in many US counties, concurrent with incidence increases. 
 In addition, highly overdispersed patterns occur more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility, and reporting. 
-Our method is robust to changes in incidence and to population size, allowing for quantification of dispersion-indicative of superspreading dynamics-without artifactual contributions from these features.
+Our method is robust to differences in population size and incidence, allowing for quantification of dispersion-indicative of superspreading dynamics-without artifactual contributions from other features.
 
@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. 
-Morbi consectetur est vel gravida pretium. 
-Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. 
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. 
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id. 
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae. 
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the variability in how diseases like COVID-19 spread. 
+We developed a method to identify shifts in the spread patterns of SARS-CoV-2 (the virus that causes COVID-19) within a single time series.
+By examining spatiotemporal differences in how the virus spreads, we can better predict and prepare for surges in cases.
+We used negative binomial regression to account for factors that might influence the spread of cases, allowing us to use a single parameter to infer the degree of clustering of cases. 
+We found that during holiday periods, the spread of SARS-CoV-2 becomes more erratic, with some individuals spreading the virus to many more people than others. 
+As the pandemic progressed, the spread of the virus became erratic, suggesting increased differences in factors such as how people transmit the virus or susceptibility.
+Our method accurately measures spreading heterogeneity regardless of population size and incidence, providing a way to understand superspreading across a range of locations. 
+This can help public health officials better anticipate and manage outbreaks, especially during times when the virus spreads unpredictably.
 
@@
 \section*{Introduction}
-Time series of observed infectious disease incidence are, to varying degrees, "noisy", showing higher frequency oscillations around trends at broader temporal scales.
-Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, 'measurement error'), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
+Time series of infectious disease incidence are, to varying degrees, ``noisy'', showing higher frequency oscillations around trends at broader temporal scales.
+Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, measurement error), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
 Therefore, variability can contain information important to understanding epidemic dynamics and societal responses. 
@@
 One area of interest is how variability is related to different phases of an epidemic. 
-For instance, \cite{graham_measles_2019} use the mean and interannual coefficient of variation of measles incidence to construct a metric indicative of where a location may be on the path to elimination of the pathogen.
-Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
+For instance, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen \cite{graham_measles_2019}.
+Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19 decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
 Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
-Recently, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection from incidence data \cite{schneckenreither_assessing_2023}. 
+To that end, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
 Analyzing variability in terms of bursts of incidence is also important for planning surge capacity \cite{wallinga_metropolitan_2018}. 
@@
 Dispersion around a rolling mean (moving window) of an incidence time series may contain information about the size and frequency of local outbreaks. 
+Dispersion may be a more useful metric than variance because the same value of the dispersion parameter will result in a larger variance/mean ratio when the mean is larger, thus capturing clustering in the true sense, and not just reflecting the statistical artifact that large mean results in large variance.
 From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. 
@@
 This is a useful way to think about dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
-Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics (Lloyd-Smith 2005), so dispersion in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
+Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so dispersion in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
 Individual-level variability in transmission is most often studied using contact-tracing data. 
@@
 For instance, if we model a process as Poisson, we assume that if the mean of the process is large, the variance will be large as well. 
-Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings. 
-Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accomodated by a Poisson model. 
+Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings by default. 
+Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accommodated by a Poisson model. 
 The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
@@
 Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion \cite{barron_analysis_1992}.
-We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence.
-We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the pandemic. 
+We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence.
+We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic. 
 From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
@@
 However, other processes besides the current number infected might affect dispersion, so we instead investigated changes over time to understand important processes that may leave a signal in dispersion.
-A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
+As mentioned above, a persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
 Since population size influences mean and variance in count data and thus could have an impact on estimates of dispersion, we robustly adjusted for population size using an offset in the model. 
@@
 \begin{equation}
-x  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
+  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
 \end{equation}
@@
 \begin{align}
-E(I) &= \mu
+  E(I) &= \mu\\
   Var(I) &= \mu + \frac{\mu^2}{\theta}
@@
 
-For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series--3,000 epidemic curves over 60 days each were produced, and the LRT test procedure was applied to each.
-Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach.
+For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series--3,000 epidemic curves over 60 days each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
+Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach \nameref{S1_Appendix}.
 
@@
 To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
-We then estimated \begin{math}\mu_t\end{math} and \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package (CITE) and from Di et al.
-\cite{yanming_nbp_2011} with a moving window approach.
+We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq} and from Di et al.
+\cite{yanming_nbp_2011}) with a moving window approach.
 For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for the window. 
-By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced.
-We investigated large counties (top 4\% in each state), due to the power constraints discussed above.
+By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced. We investigated large counties (top 4\% in each state), due to power constraints.
 
@@
 \section*{Results}
-We found that the negative binomial method is robust to changes in population size (for population sizes of at least 10,000). 
-For an analytical derivation, see Supplement 1.
-The criteria are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size. 
-In row one and two of Figure 1, we illustrated that a drop in \begin{math}\theta\end{math} is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
+We found that the negative binomial/LRT method is robust to differences in population size (for population sizes of at least 10,000) (Fig. 1).
+For an analytical derivation, see the Supplemental Materials \nameref{S1_Appendix}.
+The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size. 
+In row one and two of Fig.
+1, we illustrated that a drop in \begin{math}\theta\end{math} is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
 
-Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
-Interestingly, the most dispersed category reaches its highest proportion near the end of the timeframe (Figure 1.(a)).
-In addition, there are increases in dispersion around the holiday periods in the dataset (Figure 1.(b)), concurrent with increases in incidence (Figure 1.(c)). 
-The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Figure 1.(d)).
-
-% Place figure captions after the first paragraph in which they are cited.
 \begin{figure}[!h]
-\caption{{\bf Fig 1.}
+\includegraphics[width=0.6\textwidth]{fig1}
+\caption{
 Detecting dispersion changes in incidence time series in populations of different sizes.
@@
 
+Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
+Interestingly, the most dispersed category in Fig.2 (a) reaches its highest proportion near the end of the timeframe examined.
+In addition, there are increases in dispersion around the holiday periods in the dataset (Fig. 2.(b)), concurrent with increases in incidence (Fig. 2(c)). 
+The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Fig. 2 (d)).
+
 \begin{figure}[!h]
-\caption{{\bf Fig 2.}
+\includegraphics[width=0.6\textwidth]{fig2}
+\caption{
 Incidence and dispersion between 2020-02-20 and 2023-03-19 in large counties in the US.
@@
 
-The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variability than when incidence is lower. 
-In other words, what makes a change in dispersion meaningful is how it affects variance and variance-mean relationships. 
+The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variance/mean relationships than when incidence is lower. 
+In other words, what makes a change in dispersion meaningful is how it affects variance-mean relationships. 
 For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. 
@@
 
-So, a small increase in dispersion could have large impacts on the variance in large populations at times of peak incidence. 
 Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
 Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
-To illustrate this using the empirical data, we observed an increase from \begin{math}\theta = 2.707735e+16\end{math} to \begin{math}\theta = 21.8711\end{math}, which, conditional on the observed case count of 3, would lead to an increase in variance from around 3 to around 3.4.
-In contrast, if the observed case count were 300, variance would go from around 300 to around 4415.019.
-
 Additionally, \begin{math}\theta\end{math} seems to behave like a state variable. 
-That is, its rate of change is dependent on its value and the values of other state variables (like incidence), which determine the "state" of the system.
+That is, its rate of change is dependent on its own value and the values of other state variables (like incidence), which determine the "state" of the system (Fig. 3). 
 State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics.
@@
 \begin{figure}[!h]
-\caption{{\bf Fig 3.}
-caption
+\includegraphics[width=0.6\textwidth]{fig3}
+\caption{
+Log of the mean (over a single time series) dispersion parameter v. log of the mean incidence (over a single time series; normalized by population size).
 }
@@
 
-% Place tables after the first paragraph in which they are cited.
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-\begin{tabular}{|l+l|l|l|l|l|l|l|}
-\hline
-\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
-$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-\end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{table}
-
-%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
-\subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
-\subsubsection*{3rd level heading} 
-Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-
-\begin{enumerate}
-	\item{react}
-	\item{diffuse free particles}
-	\item{increment time by dt and go to 1}
-\end{enumerate}
-
-\subsection*{Sed ac quam id nisi malesuada congue}
-
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
-
-\begin{itemize}
-	\item First bulleted item.
-	\item Second bulleted item.
-	\item Third bulleted item.
-\end{itemize}
-
 \section*{Discussion}
-We presented an approach to compare variability in epidemic time series that does not detect artifacts based on population size and incidence. 
+We presented an approach to compare dispersion in epidemic time series that does not detect artifacts based on population size and incidence. 
 Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series.
@@
 Similarly, spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}. 
-Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public  goal can be achieved with limited resrouces.
+Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
 Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
-
 Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (peak incidence) and corresponds to significant increases in variance when incidence is high.
 Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties. 
-
-In addition, a big city with more hospitals could effectively be analogous to a collection of small towns, and therefore experience a benefit of reduced variance-mean relationships, especially during periods of peak incidence.
+In addition, a big city with more (smaller) hospitals could effectively be analogous to a collection of small towns, and therefore experience a benefit of reduced variance-mean relationships, especially during periods of peak incidence.
 This implies that there may be merit to revising current public health strategies. 
@@
 
-Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.
+%Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.
 Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.

### 2024-07-04 manuscript/ms.tex
Update ms.tex

@@
 \section*{Introduction}
-Time series of infectious disease incidence are, to varying degrees, ``noisy'', showing higher frequency oscillations around trends at broader temporal scales.
+Time series of observed infectious disease incidence are, to varying degrees, ``noisy'', showing higher frequency oscillations around trends at broader temporal scales.
 Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, measurement error), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
@@
 For instance, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen \cite{graham_measles_2019}.
-Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19 decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
+Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
 Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
@@
 \begin{equation}
-  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
+x  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
 \end{equation}
@@
 \section*{Discussion}
-We presented an approach to compare dispersion in epidemic time series that does not detect artifacts based on population size and incidence. 
+We presented an approach to compare variability in epidemic time series that does not detect artifacts based on population size and incidence. 
 Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series.
@@
 So, additional research is needed to understand the balance between burstiness of small communities and the potentially less severe effect of increased dispersion.
-
 %Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.

### 2024-07-10 code/5-figure1.R
fig1 edits

@@ -57,16 +57,19 @@ axis.Date(1, at = at, labels = labels)
 mtext("f", side = 3, line = 1, adj = 0)
 
 
-
 # Row 3
 pops <- c(1.0e+03, 1.0e+04, 1.0e+07)
 dtheta <- curve_parms$theta2 - curve_parms$theta1
 
 for (i in 1:length(pops)) {
-  current_dtheta <- dtheta[which(curve_parms$population == pops[i])]
-  current_pvals <- pvals[which(curve_parms$population == pops[i])]
   
-  plot(current_pvals ~ jitter(abs(current_dtheta), 1.5),
+  current_dtheta = dtheta[which(curve_parms$population == pops[i])]
+  current_pvals = pvals[which(curve_parms$population == pops[i])]
+  
+  current_dtheta_sm = current_dtheta[abs(current_dtheta) <= 27]
+  current_pvals_sm = current_pvals[abs(current_dtheta) <= 27]
+  
+  plot(current_pvals_sm ~ jitter(abs(current_dtheta_sm), 1.5),
        pch = 21, col = "darkgrey", cex = 0.2,
        bg = rgb(0.4, 0.4, 0.4, 0.3),
        #main = paste("Population size", pops[i]),
@@ -75,30 +78,48 @@ for (i in 1:length(pops)) {
        ylab = "p-value",
        xaxt = "n", yaxt = "n",
        ylim = c(0, 1),
-       xlim = c(-0.5, 27.5)
   )
   
   mtext(letters[7:9][i], side = 3, line = 1, adj = 0)
-
+  
   axis(1, seq(0, 27, 3))
   axis(2, seq(0, 1, 0.25))
   
-  ag <- aggregate(current_pvals ~ abs(current_dtheta), FUN = median)
+  ag <- aggregate(current_pvals_sm ~ abs(current_dtheta_sm), FUN = median)
   points(ag[, 1], ag[, 2], pch = 19, col = 2, cex = 1)
   
-  ag <- aggregate(current_pvals ~ abs(current_dtheta), FUN = quantile, probs = c(0.25, 0.75))
+  ag <- aggregate(current_pvals_sm ~ abs(current_dtheta_sm), FUN = quantile, probs = c(0.25, 0.75))
+  lo <- ag[, 2][, 1]
+  hi <- ag[, 2][, 2]
[…]

### 2024-07-10 manuscript/ms.tex

@@
 F: Dispersion estimates from model fit to the above series.
-G: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates similar p-value distribution across different population sizes (each pane is one population size).
+G: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
 O and X mark the null and alternative hypotheses indicated in panels D and E.

### 2024-07-16 manuscript/ms.tex
Update ms.tex

@@
 The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model.
-In distributions that form part of the exponential family, the dispersion parameter determines the mean-variance relationship.
 
@@
 B: When dispersion changes during the epidemic.
-C: Daily COVID-19 cases in Jefferson County, AL between 2020-09-10 and 2020-11-09 (X-axis is days since beginning of inclusion period for the model fit).
-D: Constant dispersion used in generation of above.
+C:  D: Constant dispersion used in generation of above.
 E: Changing dispersion used in generation of above.

### 2024-07-17 code/1c - assemble_weekly_data.R
Create 1c - assemble_weekly_data.R

@@ -1,0 +1,166 @@
+rm(list = ls())
+graphics.off()
+
+
+
+# Load raw epi data for each year and combine
+# Note cases is cumulative
+
+# csvs are not on github and can be downloaded from
+# https://github.com/nytimes/covid-19-data/us-counties-2020.csv
+# https://github.com/nytimes/covid-19-data/us-counties-2021.csv
+# https://github.com/nytimes/covid-19-data/us-counties-2022.csv
+# https://github.com/nytimes/covid-19-data/us-counties-2023.csv
+
+epi <- read.csv("data/raw/us-counties-2020.csv")
+
+temp <- read.csv("data/raw/us-counties-2021.csv")
+epi <- rbind(epi, temp)
+
+temp <- read.csv("data/raw/us-counties-2022.csv")
+epi <- rbind(epi, temp)
+
+temp <- read.csv("data/raw/us-counties-2023.csv")
+epi <- rbind(epi, temp)
+
+rm(temp)
+
+
+# Remove regions we will not analyze
+# Alaska is excluded due to fips mismatches between census and epi data
+excluded_regions <- c("Alaska",
+                      "American Samoa",
+                      "District of Columbia",
+                      "Guam",
+                      "Puerto Rico",
+                      "Northern Mariana Islands",
+                      "Virgin Islands")
+epi <- epi[!epi$state %in% excluded_regions, ]
+epi <- epi[!is.na(epi$fips), ]
+
+
+# Convert dates
+epi$date <- as.Date(epi$date, format = "%Y-%m-%d")
+
+
+
+# Load population data
+# csv is not on github and can be downloaded from
+# https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/totals/  #nolint
+temp <- read.csv("data/raw/co-est2021-alldata.csv")
+pop <- data.frame(fips = temp$STATE * 1000 + temp$COUNTY,
+                  population = temp$POPESTIMATE2020,
+                  state = temp$STNAME,
+                  county = temp$CTYNAME)
+pop <- pop[pop$state != pop$county, ] # remove state rows
+rm(temp)
+
+
+
+# Merge epi and pop size data
+dat <- merge(epi, pop, by = "fips", all.x = TRUE)
+
+
+
+# Investiage unmatched fips
+unmatched <- subset(dat, !complete.cases(dat))
+
+
+# Keep needed columns
[…]

### 2024-07-30 manuscript/ms.tex
Update ms.tex

@@
 Although classical theory predicts that there will be less dispersion when incidence is higher, we considered a more general negative binomial regression framework to account for processes that may also affect the spread of cases.
-We investigated changes in dispersion and found that there are increases in dispersion around holiday periods in many US counties, concurrent with incidence increases. 
+We investigated changes in dispersion and found that there are increases in dispersion around peaks in incidence in many US counties.
 In addition, highly overdispersed patterns occur more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility, and reporting. 
@@
 We used negative binomial regression to account for factors that might influence the spread of cases, allowing us to use a single parameter to infer the degree of clustering of cases. 
-We found that during holiday periods, the spread of SARS-CoV-2 becomes more erratic, with some individuals spreading the virus to many more people than others. 
+We found that around incidence peaks, the spread of SARS-CoV-2 becomes more erratic, with some individuals spreading the virus to many more people than others. 
 As the pandemic progressed, the spread of the virus became erratic, suggesting increased differences in factors such as how people transmit the virus or susceptibility.
@@
 Dispersion around a rolling mean (moving window) of an incidence time series may contain information about the size and frequency of local outbreaks. 
-Dispersion may be a more useful metric than variance because the same value of the dispersion parameter will result in a larger variance/mean ratio when the mean is larger, thus capturing clustering in the true sense, and not just reflecting the statistical artifact that large mean results in large variance.
+Dispersion may be a useful metric because it forms a part of a way to more flexibly model variance as a function of the mean, as opposed to the equality of mean and variance implied by the Poisson distribution.
 From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. 
@@
 One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. 
-For instance, if we model a process as Poisson, we assume that if the mean of the process is large, the variance will be large as well. 
 Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings by default. 
-Furthermore, the rate of the case-count-generating process is often changing in an epidemic, which cannot be accommodated by a Poisson model. 
-The negative binomial distribution may accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
-The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance of a Poisson model.
+Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model. 
+The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
+The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance.
 
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion \cite{barron_analysis_1992}.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
 We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence.
-We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic. 
+We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic. 
 From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
@@
 
-For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series--3,000 epidemic curves over 60 days each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
+For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series-- epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
 Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach \nameref{S1_Appendix}.
@@
 For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for the window. 
-By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced. We investigated large counties (top 4\% in each state), due to power constraints.
+By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced.
+We investigated large counties (largest three counties in each state), due to power constraints.
 
@@
 \section*{Results}
-We found that the negative binomial/LRT method is robust to differences in population size (for population sizes of at least 10,000) (Fig. 1).
+We found that the negative binomial/LRT method is robust to differences in population size (for population sizes included in the empirical data) (Fig. 1).
 For an analytical derivation, see the Supplemental Materials \nameref{S1_Appendix}.
@@
 In row one and two of Fig.
-1, we illustrated that a drop in \begin{math}\theta\end{math} is associated with increased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
+1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
+This pattern is independent of whether incidence increases or decreases, as seen in Fig.
+1.5.
 
@@
 Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
-Interestingly, the most dispersed category in Fig.2 (a) reaches its highest proportion near the end of the timeframe examined.
-In addition, there are increases in dispersion around the holiday periods in the dataset (Fig. 2.(b)), concurrent with increases in incidence (Fig. 2(c)). 
-The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values concurrent with peak incidence) (Fig. 2 (d)).
+The most dispersed category in Fig.2 (a) reaches its highest proportion near the end of the timeframe examined.
+In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 2.(b,c)).
+The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) (Fig. 2 (d)).
 
@@
 \caption{
-Incidence and dispersion between 2020-02-20 and 2023-03-19 in large counties in the US.
+Incidence and dispersion between 2020-01-04 and 2023-03-18 in large counties in the US.
 A: Binned log of the dispersion parameter over time.
@@
 
-The occurrence of high dispersion at times of peak incidence is of interest because it has more impact on variance/mean relationships than when incidence is lower. 
-In other words, what makes a change in dispersion meaningful is how it affects variance-mean relationships. 
-For instance, if dispersion is high in a high incidence setting, the variance-mean ratio would be larger than for the same dispersion in a smaller incidence setting. 
-A change from \begin{math}\theta = 1000\end{math} to \begin{math}\theta = 100\end{math} is operationally significant for large populations during times of peak incidence due to this variance-mean scaling:
+What makes a change in dispersion meaningful is how it affects variance-mean relationships.
 
@@
 Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
-Additionally, \begin{math}\theta\end{math} seems to behave like a state variable. 
-That is, its rate of change is dependent on its own value and the values of other state variables (like incidence), which determine the "state" of the system (Fig. 3). 
-State variables depend only on the current state and not on the history or the process, so they allow for the formulation of predictable relationships, such as those found in the laws of thermodynamics.
-This predictability and simplicity are crucial for both theoretical studies and practical applications, and creating a reliable description of this (epidemic) system is overdue.
-
-\begin{figure}[!h]
-\includegraphics[width=0.6\textwidth]{fig3}
-\caption{
-Log of the mean (over a single time series) dispersion parameter v. log of the mean incidence (over a single time series; normalized by population size).
-}
-\label{fig3}
-\end{figure}
 
 \section*{Discussion}
-We presented an approach to compare variability in epidemic time series that does not detect artifacts based on population size and incidence. 
+We presented an approach to quantify variability in epidemic time series that does not detect artifacts based on population size and incidence. 
 Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series.
@@
 Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
-Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (peak incidence) and corresponds to significant increases in variance when incidence is high.
+Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (near peak incidence).
 Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties. 
-In addition, a big city with more (smaller) hospitals could effectively be analogous to a collection of small towns, and therefore experience a benefit of reduced variance-mean relationships, especially during periods of peak incidence.
-This implies that there may be merit to revising current public health strategies. 
 Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series’ showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
-So, additional research is needed to understand the balance between burstiness of small communities and the potentially less severe effect of increased dispersion.
-%Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case.
+So, additional research is needed to understand the balance between burstiness of small communities and the potential impact of temporally changing dispersion in these areas.
+Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case, which is caused by these factors operating within a set period.
 Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.

### 2024-07-31 jsm/final_slides.tex
JSM materials

\documentclass{beamer}
%Information to be included in the title page:
\title{\textbf{Detecting changes in dispersion in COVID-19 case counts using a negative binomial model}}
\subtitle{JSM 2024}
\author{Rachael Aber}
\date{2024}
\logo{\includegraphics[height=1.5cm]{osu_logo}}

% import citation package
\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

\AtBeginSection[]
{
	\begin{frame}
		\frametitle{Table of Contents}
		\tableofcontents[currentsection]
	\end{frame}
}

\begin{document}

\frame{\titlepage}

\begin{frame}
	\frametitle{Table of Contents}
	\tableofcontents
\end{frame}

\section{Background}

\begin{frame}
\frametitle{Why study variability?}
\begin{itemize}[<+-| alert@+>] % other than alert it's also okay to insert \pause manually.
			\item Highly variable case count time series suggest transmission heterogeneity(bc), demographic/environmental heterogeneity, or changes in R
			\item Metrics of variability are often an overlooked way to understand systems (How is variability related to different phases of an epidemic? \cite{graham_measles_2019}
			\item Adam et al. \cite{adam_time-varying_2022} found that COVID-19 transmission heterogeneity decreased over time and was associated with interventions to slow spread
			\item Information about what phase/dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales (in)
	\end{itemize}
\end{frame}

\begin{frame}{\textbf{Why study variability?}}
	\begin{itemize}[<+-| alert@+>] 
		\item Dispersion around a rolling mean (moving window) of a case count time series may be a useful metric because it forms part of a framework that models variance more flexibly than a Poisson model
		\item A 'mean crowding' parameter \cite{lloyd_mean_1967} was proposed, which is the mean number per individual of other individuals in the same quadrat 
		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases (rel to crowdin parm) (from the perspective)
	\end{itemize}
\end{frame}

\section{Methods}
\begin{frame}{\textbf{Introduction to the method}}
	\begin{itemize}[<+-| alert@+>] 
		\item Investigated changes in dispersion over time (many processes may affect dispersion)
		\item Adjusted for population size using an offset in the model (directly model)
		\item Dispersion allowed to vary more slowly than the process mean
		\item Linear predictor includes a natural spline in time to account for autocorrelation in case counts (ns are)
		\begin{align}
			log(E[Y_i]/n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
			log(E[Y_i])-log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
			log(E[Y_i]) &= \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i) 
		\end{align}
	\end{itemize}
\end{frame}

\begin{frame}{\textbf{Negative binomial model}}
	\begin{itemize}[<+-| alert@+>] 
		\item \begin{equation}
			f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
		\end{equation}
		\item \begin{align}
		E(I) &= \mu\\
		Var(I) &= \mu + \frac{\mu^2}{\theta}
	\end{align}
	\end{itemize}
\end{frame}

\begin{frame}{\textbf{Application to simulated data}}
\begin{itemize}[<+-| alert@+>]
	\item For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 
	\item Epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each
	\item Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach
\end{itemize}
\end{frame}

\begin{frame}{\textbf{Application to empirical data}}
	\begin{itemize}[<+-| alert@+>]
		\item Mention diagnostic plots/dispersion statistics (take out?)
		\item We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta_t\end{math} using an iterative reweighted least-squares (procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011} with a moving window approach (for each window)
		\item We investigated large counties (largest three counties in each state), due to power constraints
	\end{itemize}
\end{frame}

\section{Results}
\begin{frame}{\textbf{Correspondence simulated and empirical}}
	\begin{itemize}[<+-| alert@+>]
		\item We found that the negative binomial/LRT method is robust to differences in population size (for population sizes examined) (the criteria)
		\item In row one and two of Fig. 1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series  (same relationship is observable in the empirical time series), with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence
		\begin{figure}[!h]
			\includegraphics[width=0.6\textwidth]{fig1}
			\caption{
				Detecting dispersion changes in incidence time series in populations of different sizes. A: Simulated incidence when dispersion is constant. B: When dispersion changes during the epidemic. C: Constant dispersion used in generation of above. D: Changing dispersion used in generation of above. E:  Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size). O and X mark the null and alternative hypotheses indicated in panels C and D. 
			}
			\label{fig1}
		\end{figure}
	\end{itemize}
\end{frame}

\begin{frame}{\textbf{Applied to counties}}
	\begin{itemize}[<+-| alert@+>]
		\item Highly overdispersed incidence patterns were observed more frequently later in time series (consistent) 
		\item The most dispersed category in Fig.3 (a) reaches its highest proportion near the end of the timeframe examined, and there are increases in dispersion around the peaks in incidence  (Fig. 3.(b,c))
		\item The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) (Fig. 3 (d))
		\begin{figure}[!h]
			\includegraphics[width=0.6\textwidth]{fig2}
			\caption{
				Incidence and dispersion between 2020-01-04 and 2023-03-18 in large counties in the US. A: Binned log of the dispersion parameter over time. B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis). C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis). D: LRT p-values over time as well as for each of the large counties (y-axis).
			}
			\label{fig2}
		\end{figure}
		\item What makes a change in dispersion meaningful is how it affects variance-mean relationships
		\begin{equation}
			var = \mu + \mu^2/\theta  
		\end{equation}
		\item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
		\item Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing\cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}. 
	\end{itemize}
\end{frame}

\section{Conclusion}
\begin{frame}{\textbf{Concluding remarks}}
	\begin{itemize}[<+-| alert@+>]
		\item approach to quantify variability in epidemic time series that does not detect artifacts based on population size and incidence (forms)
		\item Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources
		\item Population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005} (catalyze the development of more efficient control strategies)
		\item Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (near peak incidence)
	\end{itemize}
\end{frame}

\section{References}

\begin{frame}
	\begin{center}
		{\Huge Thanks!}
	\end{center}
\end{frame}

\end{document}

### 2024-08-01 jsm/final_slides.tex
jsm changes

@@
 \documentclass{beamer}
+
 %Information to be included in the title page:
@@
 \subtitle{JSM 2024}
-\author{Rachael Aber}
+\author{Rachael Aber, Yanming Di, Ben Dalziel}
 \date{2024}
@@
 
-\section{Background}
+\section{Why study variability?}
 
@@
 \begin{itemize}[<+-| alert@+>] % other than alert it's also okay to insert \pause manually.
-			\item Highly variable case count time series suggest transmission heterogeneity(bc), demographic/environmental heterogeneity, or changes in R
-			\item Metrics of variability are often an overlooked way to understand systems (How is variability related to different phases of an epidemic? \cite{graham_measles_2019}
-			\item Adam et al. \cite{adam_time-varying_2022} found that COVID-19 transmission heterogeneity decreased over time and was associated with interventions to slow spread
-			\item Information about what phase/dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales (in)
+			\item\includegraphics[height=2cm]{var}
+			\item Highly variable case count time series suggest transmission heterogeneity, demographic/environmental heterogeneity, or changes in R
+			\item Metrics of variability are overlooked (way to): "How is variability related to different phases of an epidemic?"\cite{graham_measles_2019}
+			\item Adam et al.\cite{adam_time-varying_2022} found that COVID-19 transmission heterogeneity decreased over time; associated with interventions to slow spread
+			\item Information about what phase/dynamic regime an epidemic is in, as well as indicating the level of heterogeneity at finer spatial and temporal scales (in)
 	\end{itemize}
@@
 	\begin{itemize}[<+-| alert@+>] 
-		\item Dispersion around a rolling mean (moving window) of a case count time series may be a useful metric because it forms part of a framework that models variance more flexibly than a Poisson model
+		\item Dispersion (around) of a case count time series may be a useful metric--part of a framework that models variance (more) flexibly
 		\item A 'mean crowding' parameter \cite{lloyd_mean_1967} was proposed, which is the mean number per individual of other individuals in the same quadrat 
-		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases (rel to crowdin parm) (from the perspective)
+		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases (from the perspective)
+		\item \includegraphics[height=2cm]{sup}
 	\end{itemize}
@@
 
-\section{Methods}
+\section{Negative binomial model}
 \begin{frame}{\textbf{Introduction to the method}}
 	\begin{itemize}[<+-| alert@+>] 
-		\item Investigated changes in dispersion over time (many processes may affect dispersion)
+		\item Let \begin{math}\lambda_t\end{math} 
+		\item \begin{equation}
+			\mathrm{I_t = NB(\mu = \lambda_t, \theta_t = I_{t-1})}
+		\end{equation}
 		\item Adjusted for population size using an offset in the model (directly model)
@@
 			log(E[Y_i])-log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
-			log(E[Y_i]) &= \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i) 
+			log(E[Y_i]) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i) 
 		\end{align}
@@
 	\end{align}
+	\item LRT framework with null that dispersion stays the same - applied on a rolling basis
 	\end{itemize}
@@
 	\item For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 
-	\item Epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each
+	\item Epidemic curves over 60 timesteps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each
 	\item Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach
@@
 	\begin{itemize}[<+-| alert@+>]
-		\item Mention diagnostic plots/dispersion statistics (take out?)
 		\item We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta_t\end{math} using an iterative reweighted least-squares (procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011} with a moving window approach (for each window)
-		\item We investigated large counties (largest three counties in each state), due to power constraints
+		\item We investigated large counties (largest three counties in each state)
 	\end{itemize}
@@
 	\begin{itemize}[<+-| alert@+>]
-		\item We found that the negative binomial/LRT method is robust to differences in population size (for population sizes examined) (the criteria)
+		\item We found that the negative binomial/LRT method is robust to differences in population size (for population sizes examined) (for, row)
 		\item In row one and two of Fig.
 1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series  (same relationship is observable in the empirical time series), with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{\textbf{Correspondence simulated and empirical}}
 		\begin{figure}[!h]
-			\includegraphics[width=0.6\textwidth]{fig1}
+			\includegraphics[width=0.8\textwidth]{fig1}
 			\caption{
@@
 		\end{figure}
+\end{frame}
+
+\begin{frame}{\textbf{}}
+		\begin{figure}[!h]
+			\includegraphics[width=0.4\textwidth]{compare}
+			\caption{
+				Method applied to case counts between 2020-01-04 and 2023-03-18 for Jefferson County, AL. A: Case counts . B: LRT statistic C: Log dispersion parameter.
+			}
+			\label{fig2}
+		\end{figure}
+\end{frame}
+
+\begin{frame}{\textbf{Applied to counties}}
+	\begin{itemize}[<+-| alert@+>]
+		\item Highly overdispersed incidence patterns observed more frequently later in time series (consistent) (Most dispersed category in Fig reaches) 
+		\item Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) 
+		\item High dispersion may indicate less diffuse epidemics that are potentially more subject to climate forcing\cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}
 	\end{itemize}
@@
 \begin{frame}{\textbf{Applied to counties}}
-	\begin{itemize}[<+-| alert@+>]
-		\item Highly overdispersed incidence patterns were observed more frequently later in time series (consistent) 
-		\item The most dispersed category in Fig.3 (a) reaches its highest proportion near the end of the timeframe examined, and there are increases in dispersion around the peaks in incidence  (Fig.
-3.(b,c))
-		\item The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) (Fig. 3 (d))
 	\begin{figure}[!h]
-			\includegraphics[width=0.6\textwidth]{fig2}
+	\includegraphics[width=0.4\textwidth]{fig2}
 	\caption{
@@
 \end{figure}
-		\item What makes a change in dispersion meaningful is how it affects variance-mean relationships
-		\begin{equation}
-			var = \mu + \mu^2/\theta  
-		\end{equation}
-		\item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
-		\item Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing\cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
-	\end{itemize}
 \end{frame}
 
-\section{Conclusion}
+\section{Concluding remarks}
 \begin{frame}{\textbf{Concluding remarks}}
 	\begin{itemize}[<+-| alert@+>]
-		\item approach to quantify variability in epidemic time series that does not detect artifacts based on population size and incidence (forms)
-		\item Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources
+		\item Quantify variability in epidemic time series that does not detect artifacts based on population size and incidence (forms)
+		\item Methods that use time series are crucial  (due to) 
+		\item Timing/allocation of public health resources can be achieved (with) 
 		\item Population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005} (catalyze the development of more efficient control strategies)
-		\item Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (near peak incidence)
+		\item Our results imply (revise): dispersion is high at unexpected times (near peak incidence)
+		\item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers
 	\end{itemize}

### 2024-08-01 code/4.5-Fig1.5_compare.R

@@ -21,6 +21,6 @@ dates <- dates[8:(length(cases[1,]) - 8 + 1)]
 par(mfrow = c(3, 1))
 plot(dates, series, type = "h", ylab = "", main = "Case counts")
 plot(dates, lrt_stats[1,], type = "h", col = 2, ylab = "", main = "LRT")
-plot(dates, log(thetas[1,]), type = "l", col = 4, ylab = "", main = "Log Theta")
+plot(dates, log(thetas[1,]), type = "l", col = 4, ylab = "", main = expression(log(theta)))
 
 dev.off()

### 2024-08-02 jsm/final_slides.tex
jsm materials

@@
 \author{Rachael Aber, Yanming Di, Ben Dalziel}
-\date{2024}
-\logo{\includegraphics[height=1.5cm]{osu_logo}}
+\logo{\includegraphics[height=1.1cm]{osu_logo}}
 
@@
 \addbibresource{references.bib}
+
+% font size package
+\usepackage[font=footnotesize,labelfont=bf]{caption}
 
@@
 \frametitle{Why study variability?}
+\includegraphics[height=4cm]{var}
 \begin{itemize}[<+-| alert@+>] % other than alert it's also okay to insert \pause manually.
-			\item\includegraphics[height=2cm]{var}
-			\item Highly variable case count time series suggest transmission heterogeneity, demographic/environmental heterogeneity, or changes in R
-			\item Metrics of variability are overlooked (way to): "How is variability related to different phases of an epidemic?"\cite{graham_measles_2019}
-			\item Adam et al.\cite{adam_time-varying_2022} found that COVID-19 transmission heterogeneity decreased over time; associated with interventions to slow spread
-			\item Information about what phase/dynamic regime an epidemic is in, as well as indicating the level of heterogeneity at finer spatial and temporal scales (in)
+			\item Highly variable case count time series suggest transmission heterogeneity or changes in R
+			\item Metrics of variability are overlooked: "How is variability related to different phases of an epidemic?"\cite{graham_measles_2019}
+			\item Adam et al.\cite{adam_time-varying_2022} found that COVID-19 transmission heterogeneity decreased over time
 	\end{itemize}
@@
 	\begin{itemize}[<+-| alert@+>] 
-		\item Dispersion (around) of a case count time series may be a useful metric--part of a framework that models variance (more) flexibly
-		\item A 'mean crowding' parameter \cite{lloyd_mean_1967} was proposed, which is the mean number per individual of other individuals in the same quadrat 
+		\item Dispersion of a case count time series may be a useful metric--part of a framework that models variance flexibly
+		\item A 'mean crowding' parameter \cite{lloyd_mean_1967} was proposed: mean number per individual of other individuals in the same quadrat 
 		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases (from the perspective)
@@
 	\begin{itemize}[<+-| alert@+>] 
-		\item Let \begin{math}\lambda_t\end{math} 
+		\item Let \begin{math}\lambda_t\end{math} be epidemic intensity, I\begin{math}_t\end{math} be incidence at time t, and \begin{math}\theta_t\end{math} be dispersion at time t
 		\item \begin{equation}
-			\mathrm{I_t = NB(\mu = \lambda_t, \theta_t = I_{t-1})}
+			\mathrm{I_t = NB(\mu = \lambda_t, \theta_t = I_{t-1})} \cite{grenfell_dynamics_2002}
 		\end{equation}
-		\item Adjusted for population size using an offset in the model (directly model)
-		\item Dispersion allowed to vary more slowly than the process mean
+		\item Adjusted for population size using an offset in the model
 		\item Linear predictor includes a natural spline in time to account for autocorrelation in case counts (ns are)
@@
 		\end{align}
+            \item Model fit on a rolling basis to each time series (one estimate)
 	\end{itemize}
@@
 	\begin{itemize}[<+-| alert@+>] 
-		\item \begin{equation}
-			f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
+		\item 
+            \begin{equation}
+                f_t(I) = \binom{I + \theta - 1}{I} \left( \frac{\mu}{\mu+\theta} \right)^I \left( \frac{\theta}{\mu +\theta} \right)^\theta
             \end{equation}
@@
 	\end{align}
-	\item LRT framework with null that dispersion stays the same - applied on a rolling basis
+	\item LRT framework: at each time step along a time series, fit null and full(\begin{math}\theta\end{math} change) model, conduct LRT to produce p-value at each time point
 	\end{itemize}
@@
 
-\begin{frame}{\textbf{Application to simulated data}}
+\begin{frame}{\textbf{Simulated time series data set}}
+\includegraphics[height=3cm]{sim}
 \begin{itemize}[<+-| alert@+>]
-	\item For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 
-	\item Epidemic curves over 60 timesteps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each
-	\item Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach
+	\item Validity/power simulations: Gaussian and uniform epidemic curves with an attack rate of 0.1 
+ 	\item Varying magnitude of \begin{math}\theta\end{math} change, location of the change, underlying population size, and epidemic curve shape (allowed)
+	\item Epidemic curves over 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each
 \end{itemize}
@@
 	\begin{itemize}[<+-| alert@+>]
-		\item We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta_t\end{math} using an iterative reweighted least-squares (procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011} with a moving window approach (for each window)
+            \item Weekly case counts from US counties between 2020-01-04 and 2023-03-18
+		\item Estimated \begin{math}\theta_t\end{math} for 154 time steps for 144 US counties (IRLS procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011}) 
 		\item We investigated large counties (largest three counties in each state)
@@
 \section{Results}
-\begin{frame}{\textbf{Correspondence simulated and empirical}}
+\begin{frame}{\textbf{Results: simulated data}}
 	\begin{itemize}[<+-| alert@+>]
-		\item We found that the negative binomial/LRT method is robust to differences in population size (for population sizes examined) (for, row)
-		\item In row one and two of Fig.
-1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series  (same relationship is observable in the empirical time series), with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence
+		\item Negative binomial/LRT method is robust to differences in population size (for population sizes examined)
+		\item Illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series  (same relationship is observable in the empirical time series), with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence
 	\end{itemize}
@@
 
-\begin{frame}{\textbf{Correspondence simulated and empirical}}
+\begin{frame}{\textbf{Results: simulated data}}
 		\begin{figure}[!h]
-			\includegraphics[width=0.8\textwidth]{fig1}
+			\includegraphics[width=0.6\textwidth]{fig1}
 			\caption{
 				Detecting dispersion changes in incidence time series in populations of different sizes.
-A: Simulated incidence when dispersion is constant.
-B: When dispersion changes during the epidemic.
-C: Constant dispersion used in generation of above.
-D: Changing dispersion used in generation of above.
-E:  Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
+A/B: Simulated incidence when dispersion is constant/changes.
+C/D: Constant/changing dispersion used in generation of above.
+E: Performance of LRT with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
 O and X mark the null and alternative hypotheses indicated in panels C and D.
@@
 		\begin{figure}[!h]
-			\includegraphics[width=0.4\textwidth]{compare}
+			\includegraphics[width=0.6\textwidth]{compare}
 			\caption{
@@
 
-\begin{frame}{\textbf{Applied to counties}}
+\begin{frame}{\textbf{Results: empirical data}}
 	\begin{itemize}[<+-| alert@+>]
-		\item Highly overdispersed incidence patterns observed more frequently later in time series (consistent) (Most dispersed category in Fig reaches) 
+		\item Highly overdispersed incidence observed more frequently later in time series (consistent) (Most dispersed category in Fig reaches) 
 		\item Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) 
 		\item High dispersion may indicate less diffuse epidemics that are potentially more subject to climate forcing\cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}
+            \item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers
+
 	\end{itemize}
@@
 
-\begin{frame}{\textbf{Applied to counties}}
+\begin{frame}{\textbf{Results: empirical data}}
 	\begin{figure}[!h]
-	\includegraphics[width=0.4\textwidth]{fig2}
+	\includegraphics[width=0.6\textwidth]{fig2}
 	\caption{
-		Incidence and dispersion between 2020-01-04 and 2023-03-18 in large counties in the US.
-A: Binned log of the dispersion parameter over time.
-B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis).
-C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis).
-D: LRT p-values over time as well as for each of the large counties (y-axis).
+		Incidence and dispersion in large counties in the US.
+A: Binned log of the dispersion parameter.
+B: Log of the dispersion parameter for each of the large counties (y-axis).
+C: Log incidence for each of the large counties (y-axis).
+D: LRT p-values for each of the large counties (y-axis).
 	}
@@
 	\begin{itemize}[<+-| alert@+>]
-		\item Quantify variability in epidemic time series that does not detect artifacts based on population size and incidence (forms)
-		\item Methods that use time series are crucial  (due to) 
-		\item Timing/allocation of public health resources can be achieved (with) 
-		\item Population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005} (catalyze the development of more efficient control strategies)
-		\item Our results imply (revise): dispersion is high at unexpected times (near peak incidence)
-		\item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers
+            \item LRT procedure performs well for relevant range of population sizes
+            \item Dispersion parameter and LRT statistic don't simply reflect changes in process mean
+		\item Dispersion is high at unexpected times (near peak incidence)(changes)
+            \item Methods that use time series are crucial (due to); timing/allocation of public health resources can be achieved (with, pop less effective) 
 	\end{itemize}
@@
 \section{References}
+
+\begin{frame}
+    \printbibliography
+\end{frame}
 

### 2024-08-04 jsm/final_slides.tex
final jsm changes

@@
 \author{Rachael Aber, Yanming Di, Ben Dalziel}
+\date{August 5, 2024}
 \logo{\includegraphics[height=1.1cm]{osu_logo}}
@@
 % import citation package
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber, doi=false, url=false]{biblatex}
 \addbibresource{references.bib}
@@
 }
+
+\setbeamertemplate{caption}[numbered]
 
@@
 \includegraphics[height=4cm]{var}
-\begin{itemize}[<+-| alert@+>] % other than alert it's also okay to insert \pause manually.
+\begin{itemize}[<+->] % using only the pause overlay specification
 			\item Highly variable case count time series suggest transmission heterogeneity or changes in R
@@
 \begin{frame}{\textbf{Why study variability?}}
-	\begin{itemize}[<+-| alert@+>] 
+    \begin{itemize}[<+->] % using only the pause overlay specification
 		\item Dispersion of a case count time series may be a useful metric--part of a framework that models variance flexibly
 		\item A 'mean crowding' parameter \cite{lloyd_mean_1967} was proposed: mean number per individual of other individuals in the same quadrat 
-		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases (from the perspective)
-		\item \includegraphics[height=2cm]{sup}
+		\item Useful way to think about dispersion in case count time series, degree of dispersion is degree of clustering/crowding of cases
     \end{itemize}
+		\includegraphics[height=2cm]{sup} \cite{nielsen_counterintuitive_2023}
 \end{frame}
@@
 \begin{frame}{\textbf{Introduction to the method}}
-	\begin{itemize}[<+-| alert@+>] 
-		\item Let \begin{math}\lambda_t\end{math} be epidemic intensity, I\begin{math}_t\end{math} be incidence at time t, and \begin{math}\theta_t\end{math} be dispersion at time t
-		\item \begin{equation}
-			\mathrm{I_t = NB(\mu = \lambda_t, \theta_t = I_{t-1})} \cite{grenfell_dynamics_2002}
-		\end{equation}
+    \begin{itemize}[<+->] % using only the pause overlay specification
+		\item Let \begin{math}\lambda_t\end{math} be epidemic intensity at time t, I\begin{math}_t\end{math} be incidence at time t, and \begin{math}\theta_t\end{math} be dispersion at time t
+		\item \begin{equation}\mathrm{I_t = NB(\mu = \lambda_t, \theta_t = I_{t-1})} 
+		\end{equation} \cite{grenfell_dynamics_2002}
 		\item Adjusted for population size using an offset in the model
-		\item Linear predictor includes a natural spline in time to account for autocorrelation in case counts (ns are)
+		\item Linear predictor includes a natural spline in time to account for autocorrelation in case counts
 		\begin{align}
@@
 		\end{align}
-            \item Model fit on a rolling basis to each time series (one estimate)
+            \item Model fit on a rolling basis to each time series
 	\end{itemize}
@@
 \begin{frame}{\textbf{Negative binomial model}}
-	\begin{itemize}[<+-| alert@+>] 
+\begin{itemize}[<+->] % using only the pause overlay specification
 		\item 
@@
 	\end{align}
-	\item LRT framework: at each time step along a time series, fit null and full(\begin{math}\theta\end{math} change) model, conduct LRT to produce p-value at each time point
+	\item LRT framework: at each time step along a time series, fit null model and \begin{math}\theta\end{math} change model, conduct LRT to produce p-value
 	\end{itemize}
@@
 \includegraphics[height=3cm]{sim}
-\begin{itemize}[<+-| alert@+>]
+    \begin{itemize}[<+->] % using only the pause overlay specification
 	\item Validity/power simulations: Gaussian and uniform epidemic curves with an attack rate of 0.1 
- 	\item Varying magnitude of \begin{math}\theta\end{math} change, location of the change, underlying population size, and epidemic curve shape (allowed)
-	\item Epidemic curves over 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each
+ 	\item Varied magnitude of \begin{math}\theta\end{math} change, location of the change, underlying population size, and epidemic curve shape
+	\item Epidemic curves over 60 time steps each were produced, and the LRT procedure was applied to each
 \end{itemize}
@@
 
-\begin{frame}{\textbf{Application to empirical data}}
-	\begin{itemize}[<+-| alert@+>]
-            \item Weekly case counts from US counties between 2020-01-04 and 2023-03-18
-		\item Estimated \begin{math}\theta_t\end{math} for 154 time steps for 144 US counties (IRLS procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011}) 
-		\item We investigated large counties (largest three counties in each state)
-	\end{itemize}
-\end{frame}
-
-\section{Results}
-\begin{frame}{\textbf{Results: simulated data}}
-	\begin{itemize}[<+-| alert@+>]
-		\item Negative binomial/LRT method is robust to differences in population size (for population sizes examined)
-		\item Illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series  (same relationship is observable in the empirical time series), with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{\textbf{Results: simulated data}}
+\begin{frame}{\textbf{Performance: simulated data}}
 		\begin{figure}[!h]
-			\includegraphics[width=0.6\textwidth]{fig1}
+			\includegraphics[width=0.5\textwidth]{fig1}
 			\caption{
-				Detecting dispersion changes in incidence time series in populations of different sizes.
+				Detecting dispersion changes in incidence time series.
 A/B: Simulated incidence when dispersion is constant/changes.
-C/D: Constant/changing dispersion used in generation of above.
-E: Performance of LRT with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
+C/D: Constant/changing dispersion used above.
+E: Performance of LRT with simulated data that has different absolute differences in theta (x-axis of each pane) illustrates p-value distribution across different population sizes.
 O and X mark the null and alternative hypotheses indicated in panels C and D.
@@
 
+\begin{frame}{\textbf{Application to empirical data}}
+    \begin{itemize}[<+->] % using only the pause overlay specification
+            \item Weekly case counts from US counties between 2020-01-04 and 2023-03-18
+		\item Estimated \begin{math}\theta_t\end{math} for 154 time steps for 144 US counties (IRLS procedure implemented via the NBPSeq package\cite{NBPSeq} and from Di et al.\cite{yanming_nbp_2011}) 
+		\item Investigated large counties (largest three counties in each state)
+	\end{itemize}
+\end{frame}
+
+\section{Results}
 \begin{frame}{\textbf{}}
 		\begin{figure}[!h]
-			\includegraphics[width=0.6\textwidth]{compare}
+			\includegraphics[width=0.5\textwidth]{compare}
 			\caption{
@@
 \begin{frame}{\textbf{Results: empirical data}}
-	\begin{itemize}[<+-| alert@+>]
-		\item Highly overdispersed incidence observed more frequently later in time series (consistent) (Most dispersed category in Fig reaches) 
-		\item Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) 
-		\item High dispersion may indicate less diffuse epidemics that are potentially more subject to climate forcing\cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}
-            \item Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers
-
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{\textbf{Results: empirical data}}
 	\begin{figure}[!h]
-	\includegraphics[width=0.6\textwidth]{fig2}
+	\includegraphics[width=0.5\textwidth]{fig2}
 	\caption{
@@
 \begin{frame}{\textbf{Concluding remarks}}
-	\begin{itemize}[<+-| alert@+>]
-            \item LRT procedure performs well for relevant range of population sizes
-            \item Dispersion parameter and LRT statistic don't simply reflect changes in process mean
-		\item Dispersion is high at unexpected times (near peak incidence)(changes)
-            \item Methods that use time series are crucial (due to); timing/allocation of public health resources can be achieved (with, pop less effective) 
+    \begin{itemize}[<+->] % using only the pause overlay specification
+            \item LRT procedure performed well for relevant range of population sizes
+		\item Dispersion is high at unexpected times (near peak incidence)
+            \item Concentration of changes in dispersion parameter near peak incidence
+            \item Methods that use population-level data are important: timing/allocation of public health resources can be achieved with limited resources
 	\end{itemize}
@@
 \begin{frame}
-    \printbibliography
+{\tiny\printbibliography}
 \end{frame}

### 2024-08-04 code/5-figure1.R

@@ -82,6 +82,7 @@ for (i in 1:length(pops)) {
   segments(ag[, 1], lo, ag[, 1], hi, col = 2, lwd = 3)
   
   text(0, 0.9, "O", cex = 1.2)
+  text(9, 0.9, "X", cex = 1.2)
 
 }
 

### 2024-08-09 manuscript/ms.tex
manuscript

@@
 For instance, superspreading of SARS-CoV-2 (the virus that causes COVID-19) can be elucidated by utilizing such metrics. 
-Our method identifies shifts in population-level dispersion in COVID-19 cases, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity in certain months.
-Although classical theory predicts that there will be less dispersion when incidence is higher, we considered a more general negative binomial regression framework to account for processes that may also affect the spread of cases.
+Our method identifies shifts in population-level dispersion (a measure of case clustering) in COVID-19 cases, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity at certain points in an epidemic.
+Although classical theory predicts that there will be less dispersion (less clustering) when incidence is higher, we considered a more general negative binomial regression framework to investigate processes that may also affect the spread of cases.
 We investigated changes in dispersion and found that there are increases in dispersion around peaks in incidence in many US counties.
 In addition, highly overdispersed patterns occur more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility, and reporting. 
-Our method is robust to differences in population size and incidence, allowing for quantification of dispersion-indicative of superspreading dynamics-without artifactual contributions from other features.
+Our method is robust to differences in population size and incidence, allowing for quantification of dispersion-potentially indicative of superspreading dynamics-without artifactual contributions from other features.
 
@@
 \section*{Author summary}
-Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the variability in how diseases like COVID-19 spread. 
-We developed a method to identify shifts in the spread patterns of SARS-CoV-2 (the virus that causes COVID-19) within a single time series.
+Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread. 
+We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series.
 By examining spatiotemporal differences in how the virus spreads, we can better predict and prepare for surges in cases.
-We used negative binomial regression to account for factors that might influence the spread of cases, allowing us to use a single parameter to infer the degree of clustering of cases. 
-We found that around incidence peaks, the spread of SARS-CoV-2 becomes more erratic, with some individuals spreading the virus to many more people than others. 
-As the pandemic progressed, the spread of the virus became erratic, suggesting increased differences in factors such as how people transmit the virus or susceptibility.
-Our method accurately measures spreading heterogeneity regardless of population size and incidence, providing a way to understand superspreading across a range of locations. 
+We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion (clustering) of cases.
+We found that around incidence peaks, the spread of SARS-CoV-2 became more erratic, with some individuals potentially spreading the virus to many more people than others. 
+Additionally, as the pandemic progressed, the spread of the virus also became more erratic, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
+Our method accurately measures disperison regardless of population size and incidence, providing a way to understand case count clustering across a range of locations. 
 This can help public health officials better anticipate and manage outbreaks, especially during times when the virus spreads unpredictably.
@@
 For instance, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen \cite{graham_measles_2019}.
-Similarly, it was recently found that the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
-Variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
-To that end, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
+It may be possible to use other variability metrics to determine what dynamic regime of an epidemic is taking place. 
+Additionally, it was recently found that the time-varying transmission heterogeneity for COVID-19 decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
+Individual-level variability in transmission is most often studied using contact-tracing data. 
+However, contact tracing data requires intense investment of resources \cite{kretzschmar_impact_2020}, so analysis of incidence data may often be more feasible. 
+Therefore, it is desirable to detect signatures of a change in transmission heterogeneity using case count time series.
+Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
+In sum, variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
+\noindent To that end, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
 Analyzing variability in terms of bursts of incidence is also important for planning surge capacity \cite{wallinga_metropolitan_2018}. 
@@
 \cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-
-Dispersion around a rolling mean (moving window) of an incidence time series may contain information about the size and frequency of local outbreaks. 
-Dispersion may be a useful metric because it forms a part of a way to more flexibly model variance as a function of the mean, as opposed to the equality of mean and variance implied by the Poisson distribution.
-From the perspective of an infectious individual, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high. 
-A ``mean crowding'' parameter was proposed, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
-This is a useful way to think about dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
-Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so dispersion in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
-Individual-level variability in transmission is most often studied using contact-tracing data. 
-However, contact tracing data requires intense investment of resources \cite{kretzschmar_impact_2020}, so analysis of incidence data may often be more feasible.
-
-One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. 
+\noindent One potentially useful variability metric in the context of case count time series is dispersion. 
+The dispersion of a case count time series forms part of a negative binomial model used to model case counts at a given time step. 
+The negative binomial model provides a way to more flexibly model the variance as a function of the mean, as opposed to the equality of mean and variance implied by the Poisson distribution. 
+Importantly, dispersion is related to a previously proposed ``mean crowding'' parameter, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
+If we instead think of mean crowding from the perspective of an infectious individual in an epidemic system, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high.
+It is useful to consider dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
+\noindent One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. 
 Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings by default. 
@@
 The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
-The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful variability in settings with differing base population/incidence, not the variance.
-
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
-We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in variability that are not due to changes in population size or overall burden of incidence.
+The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.
+\noindent Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
+We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
 We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic. 
-From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
+%From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
 
@@
 Classical theory put forward by Grenfell et al.
-\cite{grenfell_dynamics_2002} proposed that incidence can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion equal to previous incidence:
+\cite{grenfell_dynamics_2002} proposed that incidence at a time step can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion parameter equal to previous incidence:
 
@@
 
-However, other processes besides the current number infected might affect dispersion, so we instead investigated changes over time to understand important processes that may leave a signal in dispersion.
+\noindent However, other processes besides the current number infected might affect dispersion, so we instead investigated changes over time to understand important processes that may leave a signal in dispersion.
 As mentioned above, a persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
@@
 In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size. 
-The general framework is that incidence is drawn from a negative binomial distribution with time-varying mean and dispersion parameters (that vary more slowly than the mean). 
+The general framework is that incidence at a time step is drawn from a negative binomial distribution with time-varying mean and dispersion parameters (that vary more slowly than the mean). 
 The model is formulated with a linear predictor that includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts. 
@@
 
-So, the form of the probability mass function is:
+So, the form of the probability mass function for incidence at a time step is:
 
@@
 \end{equation}
+
+Here, \begin{math}\mu\end{math} for the time step is estimated via the linear predictor outlined above.
 
@@
 
+\noindent Note that high variability/dispersion corresponds to low values of the dispersion parameter, \begin{math}\theta\end{math}.
+\noindent In addition to fitting the model at each time step, we developed a straightforward likelihood-ratio test (LRT) that could be applied at each time step.
+This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math} change) model.
+
 \subsection*{Application to simulated data}
 
-For validity/power simulations, we used both Gaussian and uniform epidemic curves with an attack rate of 0.1 in the simulated set of time series-- epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
-Varying the effect size, location of the breakpoint, population size, and curve shape allowed us to test the validity and power of our approach \nameref{S1_Appendix}.
+To test the validity and power of the LRT, we simulated both Gaussian and uniform epidemic curves with an attack rate of 0.1 -- epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
+Varying the magnitude of the \begin{math}\theta\end{math} change, location of the change in the curve, population size underlying the curve, and curve shape (as mentioned above) allowed us to test the validity and power of our approach across a range of situations.
+
+\nameref{S1_Appendix}.
 
 \subsection*{Application to empirical data}
-To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
-We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq} and from Di et al.
-\cite{yanming_nbp_2011}) with a moving window approach.
-For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for the window. 
+%To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
+We estimated \begin{math}\mu_t\end{math} and one \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq} and from Di et al.
+\cite{yanming_nbp_2011}) using a window around each time step.
+For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and the single value of \begin{math}\theta\end{math} was estimated for the window. 
 By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced.
@@
 \section*{Results}
-We found that the negative binomial/LRT method is robust to differences in population size (for population sizes included in the empirical data) (Fig. 1).
-For an analytical derivation, see the Supplemental Materials \nameref{S1_Appendix}.
+We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig. 1 e, f).
 The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size. 
-In row one and two of Fig.
-1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series, and that the same relationship is observable in the empirical time series, with an increase in \begin{math}\theta\end{math} corresponding to a decrease in variability around the trend in incidence.
-This pattern is independent of whether incidence increases or decreases, as seen in Fig.
-1.5.
-
 \begin{figure}[!h]
@@
 B: When dispersion changes during the epidemic.
-C:  D: Constant dispersion used in generation of above.
-E: Changing dispersion used in generation of above.
-F: Dispersion estimates from model fit to the above series.
-G: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
-O and X mark the null and alternative hypotheses indicated in panels D and E.
+C: Constant dispersion used in generation of above.
+D: Changing dispersion used in generation of above.
+E: Dispersion estimates from model fit to the above series.
+E: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
+O and X mark the null and alternative hypotheses indicated in panels C and D.
  }
@@
 \end{figure}
+The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined (see Supplemental Materials \nameref{S1_Appendix}).
+In row one and two of Fig.
+1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series.
+This pattern is observed in the empirical data, and is independent of whether incidence increases or decreases, as seen in Fig.
+2(a, c).
+\begin{figure}[!h]
+\includegraphics[width=0.6\textwidth]{compare.pdf}
+\caption{Method applied to case counts between 2020-01-04 and 2023-03-18
+for Jefferson County, AL. A: Case counts . B: LRT statistic C: Log dispersion
+parameter.
+}
+\label{fig2}
+\end{figure}
 
-Highly overdispersed incidence patterns were observed more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
-The most dispersed category in Fig.2 (a) reaches its highest proportion near the end of the timeframe examined.
-In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 2.(b,c)).
-The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by concentration of low p-values around peak incidence) (Fig. 2 (d)).
+\noindent Highly overdispersed incidence patterns were observed for many counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting.
+The most dispersed category in Fig.3 (a) reaches its highest proportion near the end of the timeframe examined.
+In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 3 b, c).
+The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig. 3 d).
 
@@
 
-What makes a change in dispersion meaningful is how it affects variance-mean relationships.
-
-\begin{equation}
-var = \mu + \mu^2/\theta  
-\end{equation}
-
-Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
+\noindent Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e.
+localized surges) which may necessitate more surge capacity in hospitals and testing centers.
+Therefore, it may be the case that there are more surges on the way up to or on the way down from peak incidence.
 Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
@@
 \section*{Discussion}
-We presented an approach to quantify variability in epidemic time series that does not detect artifacts based on population size and incidence. 
-Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series.
-Burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
-Similarly, spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}.
-Methods that use incidence time series are a crucial part of this research area due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
+We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
+Our method forms part of a larger push to investigate variability in incidence time series as an important attribute of epidemic time series using novel metrics.
+For instance, burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
+Spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our variability metric and indicators of high risk of superspreading.
+Methods that use incidence time series are crucial part due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
 Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
-Our results imply that we can revise our understanding of case count dispersion: dispersion is high at unexpected times (near peak incidence).
+Our results imply that we can revise our understanding of case count dispersion: dispersion/crowding is high near peak incidence, suggesting that low dispersion is not simply governed by high incidence.
 Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties. 
-Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series’ showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
-So, additional research is needed to understand the balance between burstiness of small communities and the potential impact of temporally changing dispersion in these areas.
-Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case, which is caused by these factors operating within a set period.
-Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.
+Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
+So, additional research is needed to understand the correspondence between burstiness of small communities and the potential impact of temporally changing dispersion in these areas.
+%Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case, which is caused by these factors operating within a set period.
+%Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.
 
 \section*{Conclusion}
-
-CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
-Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-Nam condimentum sem eget mollis euismod.
-Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
-Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
-Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
-For more information, see \nameref{S1_Appendix}.
 
@@
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
 

### 2024-08-20 manuscript/ms.tex
change jitter

@@
 D: Changing dispersion used in generation of above.
-E: Dispersion estimates from model fit to the above series.
 E: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
@@
 \end{figure}
-The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined (see Supplemental Materials \nameref{S1_Appendix}).
+The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined (see \nameref{S1_Fig}).
 In row one and two of Fig.
@@
 \label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+{\bf Estimates of the dispersion parameter from simulated data.} Points colored red for underlying population sizes less than 50,000.
+\begin{figure}[!h]
+\includegraphics[width=0.6\textwidth]{thetaest_v_theta.pdf}
+\label{S1}
+\end{figure}
 

### 2024-08-20 code/5-figure1.R

@@ -39,12 +39,12 @@ mtext("b", side = 3, line = 1, adj = 0)
 # Row 2
 plot(rep(curve_parms$theta1[11188], times = length(curves[11188,])),
      xlab = "Day", ylab = expression(theta), col = 4, cex = .1, type = "l")
-mtext("d", side = 3, line = 1, adj = 0)
+mtext("c", side = 3, line = 1, adj = 0)
 mtext("O", side = 3, line = -2)
 plot(c(rep(curve_parms$theta1[11314], times = length(curves[11314,])/2), 
        rep(curve_parms$theta2[11314], times = length(curves[11314,])/2)),
      xlab = "Day", ylab = expression(theta), col = 4, cex = .1, type = "l")
-mtext("e", side = 3, line = 1, adj = 0)
+mtext("d", side = 3, line = 1, adj = 0)
 mtext("X", side = 3, line = -2)
 
 
@@ -57,7 +57,7 @@ for (i in 1:length(pops)) {
   current_dtheta = dtheta[which(curve_parms$population == pops[i])]
   current_pvals = pvals[which(curve_parms$population == pops[i])]
   
-  plot(current_pvals ~ jitter(abs(current_dtheta), 1.5),
+  plot(current_pvals ~ jitter(abs(current_dtheta), 0.2),
        pch = 21, col = "darkgrey", cex = 0.2,
        bg = rgb(0.4, 0.4, 0.4, 0.3),
        main = "",
@@ -68,14 +68,14 @@ for (i in 1:length(pops)) {
        xaxt = "n", yaxt = "n",
   )
   
-  mtext(letters[7:8][i], side = 3, line = 1, adj = 0)
+  mtext(letters[5:6][i], side = 3, line = 1, adj = 0)
   
   axis(1, seq(0, 9, 3))
   axis(2, seq(0, 1, 0.25))
   
   ag <- aggregate(current_pvals ~ abs(current_dtheta), FUN = median)
   points(ag[, 1], ag[, 2], pch = 19, col = 2, cex = 1)
-  
+
   ag <- aggregate(current_pvals ~ abs(current_dtheta), FUN = quantile, probs = c(0.25, 0.75))
   lo <- ag[, 2][, 1]
   hi <- ag[, 2][, 2]

### 2024-08-20 manuscript/ms.tex
add conclusion

@@
 \section*{Conclusion}
+We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
+While our estimation framework facilitates comparision of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
+Application of these methods to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time window and near peaks in COVID-19 incidence.
+These findings will assist in the allocation of public health resources, especially in the planning of surge capacity.
+Since there are regimes of the COVID-19 epidemic that are subject to increased dispersion across large counties, these may be candidate time periods for rolling out extra capacity.
+However, further research is needed to understand this phenomemon in small counties, as well as for other pathogens.
 
@@
 
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem ipsum.} Analytical approach: robust to population size changes.
+%\paragraph*{S2 Fig.}
+%\label{S2_Fig}
+%{\bf Lorem ipsum.} Analytical approach: robust to population size changes.
 
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+%\paragraph*{S1 File.}
+%\label{S1_File}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod %ligula.
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+%\paragraph*{S1 Video.}
+%\label{S1_Video}
+%{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod %ligula.
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+%\paragraph*{S1 Appendix.}
+%\label{S1_Appendix}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod %ligula.
 Curabitur fringilla pulvinar lectus consectetur pellentesque.
 
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+%\paragraph*{S1 Table.}
+%\label{S1_Table}
+%{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod %ligula.
 Curabitur fringilla pulvinar lectus consectetur pellentesque.

### 2024-09-18 manuscript/ms.tex
Upload Ben edits

@@
 % \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}
 \hline
@@
 {\Large
-\textbf\newline{Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Time-series modeling of epidemics in complex populations: detecting changes in incidence dispersion over time} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \\
-\textbf{2} Department of Statistics, Corvallis, Oregon, Oregon State University, Corvallis, Oregon, USA
+\textbf{2} Department of Statistics, Oregon, Oregon State University, Corvallis, Oregon, USA
 \\
@@
 \section*{Abstract}
-Metrics of variability are often overlooked and useful ways to understand epidemic dynamics. 
-For instance, superspreading of SARS-CoV-2 (the virus that causes COVID-19) can be elucidated by utilizing such metrics. 
-Our method identifies shifts in population-level dispersion (a measure of case clustering) in COVID-19 cases, allowing a more complete and predictive understanding at both the individual and population level, and allowing practitioners to prepare surge capacity at certain points in an epidemic.
-Although classical theory predicts that there will be less dispersion (less clustering) when incidence is higher, we considered a more general negative binomial regression framework to investigate processes that may also affect the spread of cases.
-We investigated changes in dispersion and found that there are increases in dispersion around peaks in incidence in many US counties.
-In addition, highly overdispersed patterns occur more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility, and reporting. 
-Our method is robust to differences in population size and incidence, allowing for quantification of dispersion-potentially indicative of superspreading dynamics-without artifactual contributions from other features.
+Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
+Higher-frequency variation around incidence trends can shed light on the processes driving transmission in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can directly impact volatility in observed incidence.
+However, measures of incidence variability, and how it is changing over time, are often overlooked in population-level analyses of incidence data.
+Here we present a statistical framework to quantify temporal changes in case dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
+We apply the method to COVID-19 incidence data in X US counties from 20xx to 20yy.
+While standard theory predicts that dispersion should be inversely proportional to incidence, our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone but which are replicated across counties.
+Dispersion increased around the major surge in 20xx, and highly overdispersed patterns became more frequent later in the time series.
+These findings suggest that heterogeneity in transmission, susceptibility, and reporting may play causal roles in driving large surges and extending epidemic duration. 
+The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmision, with applications as leading indicators for public health response.
 
@@
 \section*{Author summary}
-Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread. 
+Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread. 
 We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series.
@@
 \section*{Introduction}
-Time series of observed infectious disease incidence are, to varying degrees, ``noisy'', showing higher frequency oscillations around trends at broader temporal scales.
-Highly variable incidence that characterizes noisy data arise from imperfect and variable reporting (i.e, measurement error), but also suggests transmission heterogeneity (superspreading), demographic/environmental heterogeneity, or changes in population effective reproduction number (R).
-Therefore, variability can contain information important to understanding epidemic dynamics and societal responses. 
-Yet metrics of variability are often overlooked ways to understand these dynamics, and techniques based on variability in epidemic time series are still emerging. 
+Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
+Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, and/or changes in susceptiblility~\cite{lloyd-smith_superspreading_2005, kirkegaard2021superspreading, sun_transmission_2021,guo2023statistical,ko2023time}.
+Metrics of variability in incidence time series could therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
+
+By ``metrics of variability'' we mean estimates of the extent to which incidence is dispersed relative to its expected value.
+The expectation could be provided by a moving average of past incidence, a process-based model, or another type of population-level model. 
+While the expectation itself will vary over time, there may also me important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
+In particular, contact tracing data has revealed temporal variation in transmission heterogeneity. 
+Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter varied during different phases of the COVID-19 epidemic in Hong Kong, suggesting that variations in the dispersion parameter are important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
+Similarly, \cite{ko2023time} found the dispersion parameter varied over time as new variants emerged.
+
+But tradiational contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate the impacts of heteoregeneity using incidence data, which is more widely available.
+Sun et al.
+\cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability.
+Statistical techniques to dispersion in epidemic time series at the population level are emerging.
 One area of interest is how variability is related to different phases of an epidemic.
-For instance, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen \cite{graham_measles_2019}.
-It may be possible to use other variability metrics to determine what dynamic regime of an epidemic is taking place. 
-Additionally, it was recently found that the time-varying transmission heterogeneity for COVID-19 decreased over time and was significantly associated with interventions to slow spread in Hong Kong \cite{adam_time-varying_2022}.
-Individual-level variability in transmission is most often studied using contact-tracing data. 
-However, contact tracing data requires intense investment of resources \cite{kretzschmar_impact_2020}, so analysis of incidence data may often be more feasible. 
-Therefore, it is desirable to detect signatures of a change in transmission heterogeneity using case count time series.
-Specifically, individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process.
-In sum, variability in population-level incidence time series may therefore provide information about what phase or dynamic regime an epidemic is in, as well as potentially indicating the level of heterogeneity at finer spatial and temporal scales, in transmission, susceptibility, reporting and/or that resulting from environmental/demographic stochasticity.
-\noindent To that end, an index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
-Analyzing variability in terms of bursts of incidence is also important for planning surge capacity \cite{wallinga_metropolitan_2018}. 
-Sun et al.
-\cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, further highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-\noindent One potentially useful variability metric in the context of case count time series is dispersion. 
-The dispersion of a case count time series forms part of a negative binomial model used to model case counts at a given time step. 
-The negative binomial model provides a way to more flexibly model the variance as a function of the mean, as opposed to the equality of mean and variance implied by the Poisson distribution. 
-Importantly, dispersion is related to a previously proposed ``mean crowding'' parameter, which is the mean number per individual of other individuals in the same quadrat \cite{lloyd_mean_1967}. 
-If we instead think of mean crowding from the perspective of an infectious individual in an epidemic system, incidence experienced in their local spatial-temporal neighborhood will be higher than the global mean when dispersion is high.
-It is useful to consider dispersion in case count time series, as (absent other processes that influence dispersion such as variation in reporting) one can think of degree of dispersion as degree of clustering/crowding of cases, which is related to concepts like transmission heterogeneity.
-\noindent One of the reasons why studying variability in incidence time series is not more widely done is because it is difficult to disentangle the effects of population size/incidence on variance. 
-Since mean case counts is directly related to population size/incidence, we would infer that variance in case counts is large in large population/incidence settings by default. 
+For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}.
+Analyzing variability in terms of bursts of incidence is also important for planning surge capacity in public health systems\cite{wallinga_metropolitan_2018}. 
+An index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}. 
+Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
+\cite{kirkegaard2021superspreading} estimated the dispersion parameter from population data.
+
+We now discuss in more detail incidence dispersion dynamics, how we measure them, and what they mean.
+Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious.
+Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954)
+
+$$
+I_t \sim NB \left( \mu_t, \theta_t \right)
+$$
+
+where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter, which is related to the variance by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$.
+Importantly smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
+Conversely, the distribution of $I_t$ tends to a Poisson distribution (where variance equals mean) as $\theta_t$ becomes large.
+This model is common in population biology (e.g.
+May's host-parisitoid work, Grenfell's papers, funk superspreading).
+
+An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967} (see appendix).
+For example $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2.
+In a simple example relavent to surge capacity in the health care system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.
+
+That covers the interpretation and it is clear that a lot of modeling epidemic dynamics comes down to modeling $\mu_t$ and $\theta_t$.
+So how do we model $\mu_t$ and $\theta_t$?
+A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time t-1 produces new infections at time $t$ and $\theta_t = I_{t-1}$.
+Why is $\theta_t = I_{t-1}$?
+This comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter.
+Under these conditions $\theta_t = I_{t-1}$ is right for sufficiently large populations.
+(Kendall 1954, Bjorstand 2002).
+But this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another.
+That is, it assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
+
+Since we know that is probably not true, we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
+Instead of assuming $\theta_t = I_{t-1}$, let's learn about the relationship and what drives $\theta_t$, first by quantifying how it changes over time.
+(e.g., does it vary with $I_t$?
+that's the null hypothesis.) To do so we will also adopt a more general / phenominological model for $\mu_t$.
+Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we will use a spline which can approximate the solution to a wider range of autoregressive and other models.
+
 Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model.
+
 The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
 The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.
+
 \noindent Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
+
 We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
@@
 \begin{align}
-  log(E[Y_i]/n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
-  \log(E[Y_i])-log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
-  log(E[Y_i]) &= \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i) 
+  \log(E[Y_i]/n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
+  \log(E[Y_i])-\log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
+  \log(E[Y_i]) &= \beta_1 h_1(t_i) + \beta_2h_2(t_i) + \beta_3 h_3(t_i) + log(n_i) 
 \end{align}
@@
 \begin{equation}
-  f_t(I) = \binom{I + \theta - 1}{I} \frac{\mu}{\mu+\theta}^I \frac{\theta}{\mu +\theta}^\theta
+  f_t(I) = \binom{I + \theta - 1}{I} \dfrac{\mu}{\mu+\theta}^I \dfrac{\theta}{\mu +\theta}^\theta
 \end{equation}
@@
   E(I) &= \mu\\
-  Var(I) &= \mu + \frac{\mu^2}{\theta}
+  Var(I) &= \mu + \dfrac{\mu^2}{\theta}
 \end{align}
@@
 \begin{figure}[!h]
-\includegraphics[width=0.6\textwidth]{fig1}
+\includegraphics[width=1\textwidth]{fig1}
 \caption{
@@
 \begin{figure}[!h]
-\includegraphics[width=0.6\textwidth]{compare.pdf}
+\includegraphics[width=1\textwidth]{compare.pdf}
 \caption{Method applied to case counts between 2020-01-04 and 2023-03-18
@@
 \begin{figure}[!h]
-\includegraphics[width=0.6\textwidth]{fig2}
+\includegraphics[width=1\textwidth]{fig2}
 \caption{

### 2024-09-25 manuscript/ms.tex
Update ms.tex

@@
 Here we present a statistical framework to quantify temporal changes in case dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
-We apply the method to COVID-19 incidence data in X US counties from 20xx to 20yy.
+We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
 While standard theory predicts that dispersion should be inversely proportional to incidence, our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone but which are replicated across counties.
-Dispersion increased around the major surge in 20xx, and highly overdispersed patterns became more frequent later in the time series.
+Dispersion increased around the major surge in 2022, and highly overdispersed patterns became more frequent later in the time series.
 These findings suggest that heterogeneity in transmission, susceptibility, and reporting may play causal roles in driving large surges and extending epidemic duration. 
@@
 \section*{Author summary}
-Understanding disease spread is crucial for managing epidemics, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread. 
+Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread.
+Quantifying this variability (higher-frequency variation around incidence trends) can elucidate important underlying processes such as transmission heterogeneity.
 We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series.
-By examining spatiotemporal differences in how the virus spreads, we can better predict and prepare for surges in cases.
 We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion (clustering) of cases.
-We found that around incidence peaks, the spread of SARS-CoV-2 became more erratic, with some individuals potentially spreading the virus to many more people than others. 
-Additionally, as the pandemic progressed, the spread of the virus also became more erratic, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
-Our method accurately measures disperison regardless of population size and incidence, providing a way to understand case count clustering across a range of locations. 
-This can help public health officials better anticipate and manage outbreaks, especially during times when the virus spreads unpredictably.
+We apply the method to incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023, and find that conspicuous shifts in dispersion occur across counties concurrently, and that these shifts are not explained by incidence alone.
+Around incidence peaks, dispersion increased spread, with some individuals potentially spreading the virus to many more people than others. 
+Additionally, as the pandemic progressed, dispersion also increased, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
+Shifts in dispersion can indicate shifts in epidemic phase and information about underlying drivers of transmission, so our method to accurately measure disperison regardless of population size and incidence, provides a way to help public health officials anticipate and manage outbreaks.
 
@@
 \begin{figure}[!h]
-\includegraphics[width=1\textwidth]{fig2}
+\includegraphics[width=1\textwidth]{fig3}
 \caption{
@@
 }
-\label{fig2}
+\label{fig3}
 \end{figure}

### 2024-09-25 manuscript/ms.tex
Update ms.tex

@@
 Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread.
-Quantifying this variability (higher-frequency variation around incidence trends) can elucidate important underlying processes such as transmission heterogeneity.
+Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes such as transmission heterogeneity.
 We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series.
 We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion (clustering) of cases.
-We apply the method to incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023, and find that conspicuous shifts in dispersion occur across counties concurrently, and that these shifts are not explained by incidence alone.
-Around incidence peaks, dispersion increased spread, with some individuals potentially spreading the virus to many more people than others. 
+We apply the method to incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023, and find that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts are not explained by incidence alone.
+Around incidence peaks, dispersion increased, potentially indicating that some individuals spread the virus to many more people than others. 
 Additionally, as the pandemic progressed, dispersion also increased, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
-Shifts in dispersion can indicate shifts in epidemic phase and information about underlying drivers of transmission, so our method to accurately measure disperison regardless of population size and incidence, provides a way to help public health officials anticipate and manage outbreaks.
+Shifts in dispersion can indicate shifts in epidemic phase as well as provide information about underlying drivers of transmission, so our method to accurately measure disperison regardless of population size and incidence provides a way to help public health officials anticipate and manage outbreaks.
 
@@
 \cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-Statistical techniques to dispersion in epidemic time series at the population level are emerging.
+Statistical techniques to estimate dispersion in epidemic time series at the population level are emerging.
 One area of interest is how variability is related to different phases of an epidemic.

### 2024-09-25 manuscript/ms.tex
Update ms.tex

--- 2024-09-25 manuscript/ms.tex
@@
 \section*{Author summary}
-Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses, but traditional metrics often overlook the inherent variability in how viruses like SARS-CoV-2 (the virus that causes COVID-19) spread.
-Quantifying this variability (higher-frequency variation around incidence trends) can elucidate important underlying processes such as transmission heterogeneity.
-We developed a method to identify shifts in the degree of dispersion (case clustering) patterns of SARS-CoV-2  within a single time series.
-We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion (clustering) of cases.
-We apply the method to incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023, and find that conspicuous shifts in dispersion occur across counties concurrently, and that these shifts are not explained by incidence alone.
-Around incidence peaks, dispersion increased spread, with some individuals potentially spreading the virus to many more people than others. 
-Additionally, as the pandemic progressed, dispersion also increased, suggesting increased differences in factors such as how people transmit the virus or how susceptible they are to the virus.
-Shifts in dispersion can indicate shifts in epidemic phase and information about underlying drivers of transmission, so our method to accurately measure disperison regardless of population size and incidence, provides a way to help public health officials anticipate and manage outbreaks.
+Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses.
+However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic in incidence time series.
+Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes, including transmission heterogeneity or changes in the susceptibility landscape.
+We developed a likelihood-ratio testing framework to identify shifts in the degree of dispersion (case clustering) of COVID-19  within a single time series.
+We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion of cases.
+We applied the method to COVID-19 incidence data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
+Dispersion increased around peaks in incidence such as the major surge in 2022, and dispersion also increased as the pandemic progressed.
+These increases potentially indicate that some individuals spread SARS-CoV-2 (the virus that causes COVID-19) to many more people than others, that the susceptibility landscape changed, or that there were changes in reporting.
+Shifts in dispersion can indicate shifts in epidemic phase as well as provide information about important underlying biological processes.
+Therefore, our method to estimate dispersion regardless of population size and incidence provides a way to help public health officials anticipate and manage changes in epidemic regime and the drivers of transmission.
 
@@
 While the expectation itself will vary over time, there may also me important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
-In particular, contact tracing data has revealed temporal variation in transmission heterogeneity. 
-Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter varied during different phases of the COVID-19 epidemic in Hong Kong, suggesting that variations in the dispersion parameter are important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
-Similarly, \cite{ko2023time} found the dispersion parameter varied over time as new variants emerged.
+Individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process (transmission heterogeneity).
+Contact tracing data has revealed temporal variation in transmission heterogeneity, which is quantified using the dispersion parameter of the offspring distribution.
+Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter of the offspring distribution varied during different phases of the COVID-19 epidemic in Hong Kong, which is important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
+Similarly, \cite{ko2023time} found the dispersion parameter of the offspring distribution varied over time as new variants emerged. 
+However, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate heteoregeneity using incidence data, which is more widely available.
 
-But tradiational contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate the impacts of heteoregeneity using incidence data, which is more widely available.
 Sun et al.
 \cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-Statistical techniques to dispersion in epidemic time series at the population level are emerging.
-One area of interest is how variability is related to different phases of an epidemic.
+Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
+\cite{kirkegaard2021superspreading} estimated the offspring distribution dispersion parameter from population data.
+Statistical techniques that make inference based on dispersion of epidemic time series at the population level as opposed to dispersion in the offspring distribution are emerging.
+One area of interest is how incidence variability is related to different phases of an epidemic.
 For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}.
@@
 An index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}.
-Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
-\cite{kirkegaard2021superspreading} estimated the dispersion parameter from population data.
 
-We now discuss in more detail incidence dispersion dynamics, how we measure them, and what they mean.
+We now discuss in more detail incidence dispersion dynamics (distinct from but related to the dispersion parameter of the offspring distribution), how we measure them, and what they indicate.
 Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious.
-Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954)
+Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954).
 
@@
 where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter, which is related to the variance by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$.
-Importantly smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
+Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
 Conversely, the distribution of $I_t$ tends to a Poisson distribution (where variance equals mean) as $\theta_t$ becomes large.
-This model is common in population biology (e.g.
-May's host-parisitoid work, Grenfell's papers, funk superspreading).
+This model is common in population biology (e.g., in the study of measles \cite{grenfell_dynamics_2002}).
 
-An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967} (see appendix).
-For example $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2.
-In a simple example relavent to surge capacity in the health care system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.
+An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967}.
+For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2.
+In a simple example relevent to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.
 
-That covers the interpretation and it is clear that a lot of modeling epidemic dynamics comes down to modeling $\mu_t$ and $\theta_t$.
-So how do we model $\mu_t$ and $\theta_t$?
-A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time t-1 produces new infections at time $t$ and $\theta_t = I_{t-1}$.
-Why is $\theta_t = I_{t-1}$?
-This comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter.
-Under these conditions $\theta_t = I_{t-1}$ is right for sufficiently large populations.
-(Kendall 1954, Bjorstand 2002).
-But this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another.
-That is, it assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
+Therefore, modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics.
+A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$.
+The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter.
+Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (Kendall 1954, Bjorstand 2002).
+However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another.
+That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
 
-Since we know that is probably not true, we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
-Instead of assuming $\theta_t = I_{t-1}$, let's learn about the relationship and what drives $\theta_t$, first by quantifying how it changes over time.
-(e.g., does it vary with $I_t$?
-that's the null hypothesis.) To do so we will also adopt a more general / phenominological model for $\mu_t$.
-Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we will use a spline which can approximate the solution to a wider range of autoregressive and other models.
+In practice, the above assumptions may not hold, so we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
+Instead of assuming $\theta_t = I_{t-1}$, drivers of $\theta_t$ can be investigated, first by quantifying how it changes over time and to what degree it is governed by $I_t$.
+We adopt a more general/phenominological model for $\mu_t$.
+Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we use a spline which is able to approximate the solution to a wider range of autoregressive and other models.
 
 Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model. 
-
 The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
-The result is that we should use the dispersion parameter of a negative binomial distribution to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.
-
-\noindent Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
+This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
 
@@
 We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
-%From a modern theoretical ecology perspective, investigating beyond the first moment of a process has also been identified as important: ecological experiments are typically geared towards assessing the impacts of the mean strength of causal processes, however the variance about mean effects have been mostly ignored as a driver in biological assemblages, but may be as important as the mean \cite{benedetti-cecchi_importance_2003}.
 

### 2024-09-25 manuscript/ms.tex
Update ms.tex

@@
 Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious.
-Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954).
+Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution \cite{kendall_stochastic_1949}.
 
@@
 The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter.
-Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (Kendall 1954, Bjorstand 2002).
+Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (\cite{kendall_stochastic_1949}, \cite{bjornstad_dynamics_nodate}).
 However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another.

### 2024-11-13 manuscript/ms.tex

--- 2024-09-25 manuscript/ms.tex
@@
 Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
-Higher-frequency variation around incidence trends can shed light on the processes driving transmission in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can directly impact volatility in observed incidence.
-However, measures of incidence variability, and how it is changing over time, are often overlooked in population-level analyses of incidence data.
-Here we present a statistical framework to quantify temporal changes in case dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
-We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
-While standard theory predicts that dispersion should be inversely proportional to incidence, our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone but which are replicated across counties.
-Dispersion increased around the major surge in 2022, and highly overdispersed patterns became more frequent later in the time series.
+Higher-frequency variation around incidence trends can shed light on the processes driving transmission in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can directly impact the volatility of observed case counts.
+However, measures of incidence variability--and how they change over time--are often overlooked in population-level analyses of incidence data.
+Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
+We apply the method to COVID-19 case count data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
+While standard theory predicts that dispersion should be inversely proportional to incidence, our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties.
+Dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
 These findings suggest that heterogeneity in transmission, susceptibility, and reporting may play causal roles in driving large surges and extending epidemic duration. 
-The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmision, with applications as leading indicators for public health response.
+The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with applications as leading indicators for public health response.
 
@@
 Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses. 
-However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic in incidence time series.
+However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
 Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes, including transmission heterogeneity or changes in the susceptibility landscape.
-We developed a likelihood-ratio testing framework to identify shifts in the degree of dispersion (case clustering) of COVID-19  within a single time series.
-We used negative binomial regression to account for factors that might confound the estimation of case count dispersion, allowing us to use a single parameter to infer the degree of dispersion of cases.
-We applied the method to COVID-19 incidence data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
-Dispersion increased around peaks in incidence such as the major surge in 2022, and dispersion also increased as the pandemic progressed.
+We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series. 
+We used negative binomial regression to account for factors that might confound the estimation of dispersion, allowing us to use a single parameter to infer the degree of case count dispersion. 
+We applied the method to COVID-19 case count data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
+Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
 These increases potentially indicate that some individuals spread SARS-CoV-2 (the virus that causes COVID-19) to many more people than others, that the susceptibility landscape changed, or that there were changes in reporting.
 Shifts in dispersion can indicate shifts in epidemic phase as well as provide information about important underlying biological processes. 
-Therefore, our method to estimate dispersion regardless of population size and incidence provides a way to help public health officials anticipate and manage changes in epidemic regime and the drivers of transmission.
+Therefore, our method to estimate dispersion provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.
 
@@
 Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
-Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, and/or changes in susceptiblility~\cite{lloyd-smith_superspreading_2005, kirkegaard2021superspreading, sun_transmission_2021,guo2023statistical,ko2023time}.
+Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, and/or changes in susceptiblility~\cite{lloyd-smith_superspreading_2005, kirkegaard_superspreading_2021, sun_transmission_2021,guo2023statistical,ko2023time}.
 Metrics of variability in incidence time series could therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
 
-By ``metrics of variability'' we mean estimates of the extent to which incidence is dispersed relative to its expected value.
+``Metrics of variability'' are estimates of the extent to which incidence is dispersed relative to its expected value.
 The expectation could be provided by a moving average of past incidence, a process-based model, or another type of population-level model. 
-While the expectation itself will vary over time, there may also me important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
+While the expectation itself will vary over time, there may also be important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
 Individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process (transmission heterogeneity).
+
 Contact tracing data has revealed temporal variation in transmission heterogeneity, which is quantified using the dispersion parameter of the offspring distribution.
-Working from contacting tracing data, \cite{guo2023statistical} found that the dispersion parameter of the offspring distribution varied during different phases of the COVID-19 epidemic in Hong Kong, which is important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
-Similarly, \cite{ko2023time} found the dispersion parameter of the offspring distribution varied over time as new variants emerged. 
-However, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020} there is a need for complementary population-level analyses that can estimate heteoregeneity using incidence data, which is more widely available.
+Working from contact tracing data, researchers found that the dispersion parameter of the offspring distribution varied during different phases of the COVID-19 epidemic in Hong Kong \cite{guo2023statistical}, which is important for understanding and managing the population-level dynamics of disease transmission and outbreak control.
+Similarly, the dispersion parameter of the offspring distribution varied over time as new variants emerged \cite{ko2023time}. 
+However, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability \cite{kretzschmar_impact_2020}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
+The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control \cite{sun_transmission_2021}.
 
-Sun et al.
-\cite{sun_transmission_2021} found a combination of individual-based and population-based strategies was required for SARS-CoV-2 control, highlighting the importance of considering population-level variability and its relationship to individual-level variability.
-Working from population-level incidence data, \cite{kirkegaard2021superspreading}.
-\cite{kirkegaard2021superspreading} estimated the offspring distribution dispersion parameter from population data.
-Statistical techniques that make inference based on dispersion of epidemic time series at the population level as opposed to dispersion in the offspring distribution are emerging.
-One area of interest is how incidence variability is related to different phases of an epidemic.
+Statistical techniques that make inference based on the variability of epidemic time series at the population level are emerging.
+Working from population-level incidence data, the offspring distribution dispersion parameter was estimated \cite{kirkegaard_superspreading_2021}.
+Another area of interest is how incidence variability is related to different phases of an epidemic.
 For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}.
 Analyzing variability in terms of bursts of incidence is also important for planning surge capacity in public health systems\cite{wallinga_metropolitan_2018}. 
+Dispersion is a way to specifically capture clustering of cases in epidemic time series.
 An index of effective aggregate dispersion (EffDI) was proposed to elucidate clusters of infection directly from incidence data \cite{schneckenreither_assessing_2023}.
+In the current work, our interest is in the dispersion of epidemic time series.
 
-We now discuss in more detail incidence dispersion dynamics (distinct from but related to the dispersion parameter of the offspring distribution), how we measure them, and what they indicate.
+Incidence dispersion dynamics are distinct from, but related to, the dispersion parameter of the offspring distribution. 
 Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious. 
-Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954).
-
-$$
-I_t \sim NB \left( \mu_t, \theta_t \right)
-$$
-
-where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter, which is related to the variance by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$.
+Under a wide range of configurations for a branching process model, including the vast majority of parsimonious ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution \cite{kendall_stochastic_1949}: $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter.
+The dispersion parameter is related to the variance of incidence by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. 
 Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
-Conversely, the distribution of $I_t$ tends to a Poisson distribution (where variance equals mean) as $\theta_t$ becomes large.
+Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
 This model is common in population biology (e.g., in the study of measles \cite{grenfell_dynamics_2002}).
 
-An intepretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as "crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967}.
-For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same timestep as a randomly selected case will exceed the Poisson expectation by a factor of 2.
-In a simple example relevent to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals with the same pathogen than expected for a Poisson process with the same incidence rate.
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded" in time relative to a Poisson process with the same mean \cite{lloyd_mean_1967}.
+For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two.
+In a simple example relevant to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
 
-Therefore, modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics.
+Modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics.
 A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$.
-The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter.
-Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (Kendall 1954, Bjorstand 2002).
-However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another.
+\begin{equation}
+    \mathrm{I_t} = NB(\mu = \lambda I_{t-1}, \theta_t = I_{t-1})
+\end{equation}
+The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ``independent lineages'' with identically distributed ``local'' rate parameter.
+Under these conditions, $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (\cite{kendall_stochastic_1949}, \cite{bjornstad_dynamics_nodate}).
+However, this model assumes that subpopulations of infectious individuals emanating from one infected individuals are independent and statistically identical to those from another.
 That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
 
-In practice, the above assumptions may not hold, so we can use estimates of $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
+In practice, the above assumptions may not hold, so we instead estimate $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
 Instead of assuming $\theta_t = I_{t-1}$, drivers of $\theta_t$ can be investigated, first by quantifying how it changes over time and to what degree it is governed by $I_t$. 
-We adopt a more general/phenominological model for $\mu_t$.
+We adopt a more general/phenomenological model for $\mu_t$. 
 Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we use a spline which is able to approximate the solution to a wider range of autoregressive and other models. 
+Furthermore, the rate of the case-count-generating process is often changing within a time step, which cannot be accommodated by a Poisson model. 
+The negative binomial distribution may accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
+This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion \cite{barron_analysis_1992}.
 
-Furthermore, the rate of the case-count-generating process is often changing within a timestep, which cannot be accommodated by a Poisson model. 
-The negative binomial distribution may accurately model a time series if there is a changing process mean within a timestep: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
-This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence, not the variance.
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a timestep that all lead to overdispersion \cite{barron_analysis_1992}.
-
-We developed a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
-We applied the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
+We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
+We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
 
@@
 % For figure citations, please use "Fig" instead of "Figure".
-Classical theory put forward by Grenfell et al.
-\cite{grenfell_dynamics_2002} proposed that incidence at a time step can be modeled by a negative binomial variable with expectation equal to the epidemic intensity and dispersion parameter equal to previous incidence:
+Classical approaches \cite{grenfell_dynamics_2002} model incidence at a time step using a negative binomial variable with expectation equal to the epidemic intensity and dispersion parameter equal to previous incidence.
+However, other processes besides the current number infected might affect dispersion, so we instead investigate changes in dispersion over time to understand processes that may leave a signal in dispersion.
 
-\begin{equation}
-    \mathrm{I_t} = NB(\mu = \lambda_t, \theta_t = I_{t-1})
-\end{equation}
-
-\noindent However, other processes besides the current number infected might affect dispersion, so we instead investigated changes over time to understand important processes that may leave a signal in dispersion.
-As mentioned above, a persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
-Since population size influences mean and variance in count data and thus could have an impact on estimates of dispersion, we robustly adjusted for population size using an offset in the model. 
-In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size. 
-The general framework is that incidence at a time step is drawn from a negative binomial distribution with time-varying mean and dispersion parameters (that vary more slowly than the mean). 
+Our general framework is that incidence at a time step is drawn from a negative binomial distribution with time-varying mean and dispersion parameter (the dispersion parameter varies more slowly than the mean parameter).
 The model is formulated with a linear predictor that includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts. 
@@
 A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence \cite{davis_negative_2009}.
-There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual):
+
+A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size. 
+Since population size influences the mean and variance of case count data and thus could have an impact on estimates of dispersion, we adjust for population size using an offset in the model. 
+There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual).
 
@@
 
-So, the form of the probability mass function for incidence at a time step is:
+Thus, the form of the probability mass function for incidence at a time step is:
 
 \begin{equation}
-  f_t(I) = \binom{I + \theta - 1}{I} \dfrac{\mu}{\mu+\theta}^I \dfrac{\theta}{\mu +\theta}^\theta
+  f_t(I) = \binom{I + \theta - 1}{I} \left(\dfrac{\mu}{\mu+\theta}\right)^I \left(\dfrac{\theta}{\mu +\theta}\right)^\theta
 \end{equation}
 
-Here, \begin{math}\mu\end{math} for the time step is estimated via the linear predictor outlined above.
+where \begin{math}\mu\end{math} for the time step is estimated via the linear predictor outlined above.
 
-This form has expectation and variance as follows:
+The form of the expectation and variance from this model is below.
 
@@
 
-\noindent Note that high variability/dispersion corresponds to low values of the dispersion parameter, \begin{math}\theta\end{math}.
-\noindent In addition to fitting the model at each time step, we developed a straightforward likelihood-ratio test (LRT) that could be applied at each time step.
-This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math} change) model.
+Note that high variability/dispersion corresponds to low values of the dispersion parameter, \begin{math}\theta\end{math}.
+In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size. 
+In addition to fitting the model at each time step, we developed a straightforward likelihood-ratio test (LRT) that could be applied at each time step.
+This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
 
@@
 
-To test the validity and power of the LRT, we simulated both Gaussian and uniform epidemic curves with an attack rate of 0.1 -- epidemic curves over 60 timesteps each were produced, and a likelihood-ratio test (LRT) procedure was applied to each.
+To test the validity and power of the LRT, we simulated both Gaussian and uniform epidemic curves with an attack rate of 0.1--epidemic curves over 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each.
 Varying the magnitude of the \begin{math}\theta\end{math} change, location of the change in the curve, population size underlying the curve, and curve shape (as mentioned above) allowed us to test the validity and power of our approach across a range of situations.
-
-\nameref{S1_Appendix}.
 
@@
 %To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
-We estimated \begin{math}\mu_t\end{math} and one \begin{math}\theta\end{math} using iterative reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq} and from Di et al.
-\cite{yanming_nbp_2011}) using a window around each time step.
-For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and the single value of \begin{math}\theta\end{math} was estimated for the window. 
-By moving the window one time step at a time, a time series for \begin{math}\theta_t\end{math} was produced.
-We investigated large counties (largest three counties in each state), due to power constraints.
+We estimated \begin{math}\mu_t\end{math} and one \begin{math}\theta\end{math} using iteratively reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq}) using a window around each time step.
+For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and the single value of \begin{math}\theta\end{math} was estimated for that window. 
+By moving the window one time step at a time, a time series of \begin{math}\theta_t\end{math} was produced.
+We investigated large counties (the largest three counties in each state), due to power constraints.
 
@@
 \section*{Results}
-We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig. 1 e, f).
+We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig.\ 1 e, f).
 The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size.
+
 \begin{figure}[!h]
@@
 Detecting dispersion changes in incidence time series in populations of different sizes.
-A: Simulated incidence when dispersion is constant.
-B: When dispersion changes during the epidemic.
-C: Constant dispersion used in generation of above.
-D: Changing dispersion used in generation of above.
-E: Performance of the method with simulated data that has different absolute differences in theta (horizontal axis of each pane) illustrates p-value distribution across different population sizes (each pane is one population size).
-O and X mark the null and alternative hypotheses indicated in panels C and D.
+a: Simulated incidence when dispersion is constant.
+b: Simulated incidence when dispersion changes.
+c: Constant dispersion parameter used in generation of the above.
+d: Changing dispersion parameter used in generation of the above.
+e, f: Performance of the LRT with simulated data that have different absolute differences in \begin{math}\theta\end{math} (horizontal axis of each panel) illustrates p-value distribution in population size of 50,000 (e) and population size of 10,000,000 (f), which represent the range in the empirical data.
+O and X mark the null and alternative hypotheses indicated in panels c and d.
+Red vertical lines represent the interquartile range.
  }
@@
 \end{figure}
-The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined (see \nameref{S1_Fig}).
-In row one and two of Fig.
-1, we illustrated that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series.
-This pattern is observed in the empirical data, and is independent of whether incidence increases or decreases, as seen in Fig.
-2(a, c).
+
+The negative binomial framework for estimating \begin{math}\theta\end{math} is robust across the population sizes examined, and a large portion of inaccurate estimates correspond to small counties outside of the range considered (\nameref{S1_Fig}).
+In row one and two of Fig.\ 1, we illustrate that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series.
+This pattern is observed in the empirical data, and is independent of whether incidence increases or decreases, as seen in Fig.\ 2.
+
 \begin{figure}[!h]
-\includegraphics[width=1\textwidth]{compare.pdf}
-\caption{Method applied to case counts between 2020-01-04 and 2023-03-18
-for Jefferson County, AL. A: Case counts . B: LRT statistic C: Log dispersion
-parameter.
+\includegraphics[width=1\textwidth]{fig2.pdf}
+\caption{Negative binomial regression/LRT method applied to case counts between 2020-01-04 and 2023-03-18 from Jefferson County, AL.
+a: Case counts.
+b: Time series of $log_{10}\theta$ colored by whether the associated LRT statistic is greater than the Bonferroni-corrected level 0.05 quantile of a chi-square random variable with one degree of freedom (red indicates areas of putative change in $\theta$).
+The black line indicates the dispersion parameter under the model from Equation 1.
 }
@@
 
-\noindent Highly overdispersed incidence patterns were observed for many counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting.
-The most dispersed category in Fig.3 (a) reaches its highest proportion near the end of the timeframe examined.
-In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 3 b, c).
-The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig. 3 d).
+Highly overdispersed incidence patterns were observed for many counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
+The most dispersed category in Fig.\ 3 (e) reaches its highest proportion near the end of the time frame examined.
+In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 3 c, d).
+The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig. 3 f).
 
 \begin{figure}[!h]
-\includegraphics[width=1\textwidth]{fig3}
+\centering
+\includegraphics[width=0.85\textwidth]{fig3.pdf}
 \caption{
 Incidence and dispersion between 2020-01-04 and 2023-03-18 in large counties in the US.
-A: Binned log of the dispersion parameter over time.
-B: Log of the dispersion parameter over time as well as for each of the large counties (y-axis).
-C: Log incidence (new cases per individual) over time as well as for each of the large counties (y-axis).
-D: LRT p-values over time as well as for each of the large counties (y-axis).
+a: Mean COVID-19 cases of the 144 US counties over time.
+b: Mean $\theta$ of the 144 US counties over time.
+c: Case counts over time for each of the large counties (y-axis).
+d: $log_{10}\theta$ over time for each of the large counties (y-axis).
+e: Binned $log_{10}\theta$ over time.
+f: LRT p-values over time for each of the large counties (y-axis).
 }
@@
 
-\noindent Raising variance relative to mean implies spatiotemporal "crowding" of cases (i.e.
+Raising variance relative to mean (higher dispersion) implies spatiotemporal "crowding" of cases (i.e.
 localized surges) which may necessitate more surge capacity in hospitals and testing centers.
-Therefore, it may be the case that there are more surges on the way up to or on the way down from peak incidence.
+Therefore, it may be the case that there were more surges on the way up to or on the way down from peak incidence.
 Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.

### 2024-11-14 manuscript/ms (1).tex
offset versions

--- 2024-11-13 manuscript/ms.tex
@@
 b: Time series of $log_{10}\theta$ colored by whether the associated LRT statistic is greater than the Bonferroni-corrected level 0.05 quantile of a chi-square random variable with one degree of freedom (red indicates areas of putative change in $\theta$).
-The black line indicates the dispersion parameter under the model from Equation 1.
+The black line indicates the dispersion parameter under the model from Equation 1 assuming a reporting rate of 0.10.
 }
@@
 b: Mean $\theta$ of the 144 US counties over time.
-c: Case counts over time for each of the large counties (y-axis).
+c: log10(case counts) - log10(reporting rate) over time for each of the large counties (y-axis).
 d: $log_{10}\theta$ over time for each of the large counties (y-axis).
@@
 We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
-Our method forms part of a larger push to investigate variability in incidence time series as an important attribute of epidemic time series using novel metrics.
-For instance, burst-tree decomposition of time series’ has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
-Spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our variability metric and indicators of high risk of superspreading.
-Methods that use incidence time series are crucial part due to the ease of obtaining incidence data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
-Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so models that incorporate transmission heterogeneity may catalyze the development of more efficient control strategies.
-Our results imply that we can revise our understanding of case count dispersion: dispersion/crowding is high near peak incidence, suggesting that low dispersion is not simply governed by high incidence.
+Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series using novel metrics.
+For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
+Spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our dispersion metric and indicators of high risk of superspreading.
+Methods that use incidence time series are crucial due to the ease of obtaining this type of data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
+Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so identifying candidate time periods when transmission heterogeneity is high may catalyze the development of more efficient control strategies.
+Importantly, our results imply that we can revise our understanding of case count dispersion: dispersion is high near peak incidence, not simply when incidence is low. 
 Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties. 
@@
 We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
-While our estimation framework facilitates comparision of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
+While our estimation framework facilitates comparison of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
 Application of these methods to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time window and near peaks in COVID-19 incidence.
@@
 Since there are regimes of the COVID-19 epidemic that are subject to increased dispersion across large counties, these may be candidate time periods for rolling out extra capacity.
-However, further research is needed to understand this phenomemon in small counties, as well as for other pathogens.
+However, further research is needed to understand this phenomenon in small counties, as well as for other pathogens.
 

### 2024-11-15 manuscript/ms (2).tex
Add files via upload

--- 2024-11-14 manuscript/ms (1).tex
@@
 Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
-We apply the method to COVID-19 case count data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
-While standard theory predicts that dispersion should be inversely proportional to incidence, our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties.
+We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
+While standard theory predicts that dispersion should be inversely proportional to incidence (i.e., that the dispersion parameter should be equal to the incidence at the previous time step), our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties.
 Dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
@@
 Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes, including transmission heterogeneity or changes in the susceptibility landscape.
-We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series. 
-We used negative binomial regression to account for factors that might confound the estimation of dispersion, allowing us to use a single parameter to infer the degree of case count dispersion. 
-We applied the method to COVID-19 case count data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
+We developed a statistical framework to quantify temporal changes in incidence dispersion within a single time series. 
+We used negative binomial regression to account for factors that might confound the estimation of dispersion, allowing us to use a single parameter to infer the degree of incidence dispersion. 
+We applied the method to COVID-19 incidence data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
 Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
@@
 
-``Metrics of variability'' are estimates of the extent to which incidence is dispersed relative to its expected value.
+``Metrics of variability'' refers to estimates of the extent to which incidence is dispersed relative to its expected value.
 The expectation could be provided by a moving average of past incidence, a process-based model, or another type of population-level model. 
-While the expectation itself will vary over time, there may also be important variation in the level of dispersion around the deterministic skeleton of the epidemic. 
+While the expectation itself will vary over time, there may also be important changes in the level of dispersion around the deterministic skeleton of the epidemic. 
 Individual-level heterogeneity in transmission scales up to affect population-level dynamics \cite{lloyd-smith_superspreading_2005}, so variability in epidemic trajectories at the population level may provide information about individual-level variability in the transmission process (transmission heterogeneity).
@@
 Statistical techniques that make inference based on the variability of epidemic time series at the population level are emerging.
-Working from population-level incidence data, the offspring distribution dispersion parameter was estimated \cite{kirkegaard_superspreading_2021}.
+Working from population-level incidence data, the offspring distribution dispersion parameter was recovered \cite{kirkegaard_superspreading_2021}.
 Another area of interest is how incidence variability is related to different phases of an epidemic.
-For example, the mean and interannual coefficient of variation of measles incidence was used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}.
+For example, the mean and interannual coefficient of variation of measles incidence were used to construct a metric indicative of where a location may be on the path to elimination of a pathogen~\cite{graham_measles_2019}.
 Analyzing variability in terms of bursts of incidence is also important for planning surge capacity in public health systems\cite{wallinga_metropolitan_2018}. 
@@
 The dispersion parameter is related to the variance of incidence by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. 
-Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, increasing amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
+Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
 Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
@@
 
-Modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics.
-A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$.
+Modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemic trajectories.
+A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which infections at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$.
 \begin{equation}
@@
 
-In practice, the above assumptions may not hold, so we instead estimate $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
+In practice, the above assumptions may not hold, and thus we instead estimate $\theta_t$ to understand how heterogeneity in transmission, susceptibility and reporting (and other factors) are driving epidemic dynamics.
 Instead of assuming $\theta_t = I_{t-1}$, drivers of $\theta_t$ can be investigated, first by quantifying how it changes over time and to what degree it is governed by $I_t$. 
 We adopt a more general/phenomenological model for $\mu_t$. 
-Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we use a spline which is able to approximate the solution to a wider range of autoregressive and other models. 
+Instead of $\mu_t = \lambda I_{t-1}$, which is a special kind of moving window, we use a natural spline in time to model the mean, which is able to approximate a wider range of autoregressive and other models.
+
 Furthermore, the rate of the case-count-generating process is often changing within a time step, which cannot be accommodated by a Poisson model. 
 The negative binomial distribution may accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial \cite{cook_notes_nodate}.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion \cite{barron_analysis_1992}.
 This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence.
-Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion \cite{barron_analysis_1992}.
 
 We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
-We apply the method to COVID-19 incidence data in US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
+We apply the method to COVID-19 \emph{weekly} new cases (to avoid irregularities in daily data reporting, such as when all weekend cases are reported on Monday) in large US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
 
@@
 Note that high variability/dispersion corresponds to low values of the dispersion parameter, \begin{math}\theta\end{math}.
+In addition to fitting the model at each time step, we developed a straightforward likelihood-ratio test (LRT) that can be applied at each time step.
+This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
 In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size.
-In addition to fitting the model at each time step, we developed a straightforward likelihood-ratio test (LRT) that could be applied at each time step.
-This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
 
@@
 
-To test the validity and power of the LRT, we simulated both Gaussian and uniform epidemic curves with an attack rate of 0.1--epidemic curves over 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each.
+To test the validity and power of the LRT, we simulated both Gaussian and uniform epidemic curves with an attack rate (proportion of the population who become infected) of 0.1.
+Epidemic curves spanning 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each at its known change point in \begin{math}\theta\end{math}.
 Varying the magnitude of the \begin{math}\theta\end{math} change, location of the change in the curve, population size underlying the curve, and curve shape (as mentioned above) allowed us to test the validity and power of our approach across a range of situations.
@@
 \subsection*{Application to empirical data}
-%To evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible \cite{ver_hoef_quasi-poisson_2007}, along with evaluation of dispersion statistics.
-We estimated \begin{math}\mu_t\end{math} and one \begin{math}\theta\end{math} using iteratively reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq}) using a window around each time step.
-For each window, \begin{math}\mu_t\end{math} was estimated using a spline function in time, and the single value of \begin{math}\theta\end{math} was estimated for that window. 
+
+We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta_t\end{math} using iteratively reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq}) using a window around each time step.
+For each window, a series of \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for that window. 
 By moving the window one time step at a time, a time series of \begin{math}\theta_t\end{math} was produced.
-We investigated large counties (the largest three counties in each state), due to power constraints.
+We investigated large counties (the three counties with the largest populations in each state), due to power constraints and convergence issues.
 
@@
 We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig.\ 1 e, f).
-The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and low average p-values are observed with increasing effect size.
+The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and lower average p-values are observed with increasing effect sizes.
 
@@
 Detecting dispersion changes in incidence time series in populations of different sizes.
-a: Simulated incidence when dispersion is constant.
-b: Simulated incidence when dispersion changes.
+a: Simulated case counts when dispersion is constant.
+b: Simulated case counts when dispersion changes.
 c: Constant dispersion parameter used in generation of the above.
@@
 
-The negative binomial framework for estimating \begin{math}\theta\end{math} is robust across the population sizes examined, and a large portion of inaccurate estimates correspond to small counties outside of the range considered (\nameref{S1_Fig}).
-In row one and two of Fig.\ 1, we illustrate that an increase in \begin{math}\theta\end{math} is associated with decreased variability in simulated incidence time series.
-This pattern is observed in the empirical data, and is independent of whether incidence increases or decreases, as seen in Fig.\ 2.
+The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined, and a large portion of inaccurate estimates correspond to counties with small population sizes outside of the range considered (\nameref{S1_Fig}).
+In row one and two of Fig.\ 1, we illustrate that an increase in \begin{math}\theta\end{math} is associated with decreased dispersion in simulated case count time series.
+This pattern is observed in the empirical data, and is independent of whether incidence is high or low, as seen in Fig.\ 2, where an example time series of $log_{10}\theta$ estimated from empirical data in Jefferson County, AL shows decreasing values around periods in which cases increase in volatility.
+For example, consider the period of time around the beginning of 2022, in which high case count variability coincides with low values of $log_{10}\theta$.
+Additionally, the dispersion parameter changed as new variability regimes began ($log_{10}\theta$ values colored red in Fig.\ 2 b).
+There were departures from the dispersion parameter under the model in Equation 1, especially around the beginning of 2022 and the end of the time frame examined (Fig.\ 2 b).
+The dispersion parameter under the model in Equation 1 can be plotted by taking a series of previous incidence and adding an offset to account for the reporting rate (log10($\theta_t$) in the model = log10(I_{t-1}) = log10(cases_{t-1}/reporting rate) = log10(cases_{t-1}) - log10(reporting rate)).
 
@@
 b: Time series of $log_{10}\theta$ colored by whether the associated LRT statistic is greater than the Bonferroni-corrected level 0.05 quantile of a chi-square random variable with one degree of freedom (red indicates areas of putative change in $\theta$).
-The black line indicates the dispersion parameter under the model from Equation 1 assuming a reporting rate of 0.10.
+The black line represents the dispersion parameter under the model from Equation 1, assuming a reporting rate of 0.10.
 }
@@
 
-Highly overdispersed incidence patterns were observed for many counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
+Highly overdispersed incidence patterns were observed for many of the examined counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting.
 The most dispersed category in Fig.\ 3 (e) reaches its highest proportion near the end of the time frame examined.
 In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 3 c, d).
-The evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig. 3 f).
+Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig.\ 3 f).
+Note that these p-values should be corrected for multiple testing when used for inference instead of visualization.
 
@@
 c: log10(case counts) - log10(reporting rate) over time for each of the large counties (y-axis).
+This represents $log_{10}\theta$ under the model in Equation 1.
 d: $log_{10}\theta$ over time for each of the large counties (y-axis).
-e: Binned $log_{10}\theta$ over time.
+e: Binned $log_{10}\theta$ estimates for all counties over time.
 f: LRT p-values over time for each of the large counties (y-axis).
@@
 Therefore, it may be the case that there were more surges on the way up to or on the way down from peak incidence.
-Additionally, it may indicate less diffuse epidemics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, or increased locally experienced mean density \cite{lloyd_mean_1967}.
+It appears that there are more surges on the way up to, and on the way down from, peak incidence.
+This indicates less diffuse epidemic dynamics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, and increased locally experienced mean density \cite{lloyd_mean_1967}.
 
 \section*{Discussion}
-We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
-Our method forms part of a larger push to investigate variability as an important attribute of epidemic time series using novel metrics.
+
+We developed an approach to quantify clustering of cases in epidemic time series that is less susceptible to detecting artifacts based on population size and incidence. 
+Our method forms part of a larger initiative to investigate variability in incidence as an important attribute of epidemic time series using novel metrics.
 For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
-Spatial variation in superspreading potential has been investigated through e.g., risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our dispersion metric and indicators of high risk of superspreading.
-Methods that use incidence time series are crucial due to the ease of obtaining this type of data, so the timing/geographical allocation of public health resources can be achieved with limited resources.
+Spatial variation in superspreading potential has been investigated through risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our dispersion metric and indicators of a high risk of superspreading.
+
+Methods that use incidence time series are crucial due to the ease of obtaining this type of data, so the timing and geographical allocation of public health resources can be achieved with limited resources.
 Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd-smith_superspreading_2005}, so identifying candidate time periods when transmission heterogeneity is high may catalyze the development of more efficient control strategies.
-Importantly, our results imply that we can revise our understanding of case count dispersion: dispersion is high near peak incidence, not simply when incidence is low. 
-Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large counties. 
-Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series showed that epidemics in smaller communities are concentrated on narrower windows of the influenza season - the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
-So, additional research is needed to understand the correspondence between burstiness of small communities and the potential impact of temporally changing dispersion in these areas.
-%Though some kinds of time dependence in the rate can cause autocorrelation, as can contagion (if it occurs outside of set periods), and heterogeneity (if an omitted variable is correlated in time) \cite{barron_analysis_1992}, our focus on dispersion makes sense because we are concerned with the clustering of cases from the point of view of an individual case, which is caused by these factors operating within a set period.
-%Also, demographic structure (e.g., age structure) has the potential to affect temporal autocorrelation in transmission rate - the effects of age structure can be captured by a model that includes an infection rate that varies over time \cite{earn_dynamic_nodate}.
+
+Application of the methods we developed to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time frame considered and near peaks in COVID-19 incidence.
+These results imply that we can revise our understanding of case count dispersion: dispersion is high \emph{near} peak incidence, suggesting that dispersion is not simply governed by incidence.
+These departures from the level of dispersion expected under the model represented in Equation 1 represent time periods in which the assumptions may be violated and cases may be highly clustered in time.
+
+Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large populations.
+
+Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series showed that epidemics in smaller communities were concentrated on narrower windows of the influenza season; the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
+So, additional research is needed to understand the relationship between burstiness in small communities and temporally changing dispersion in these areas.
+
+Since there are regimes of the COVID-19 epidemic that are subject to increased dispersion across large counties, these may be candidate time periods for rolling out extra surge capacity in large counties.
+But, as mentioned above, further research is needed to understand whether this phenomenon occurs in small and mid-sized counties.
 
@@
 While our estimation framework facilitates comparison of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
-Application of these methods to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time window and near peaks in COVID-19 incidence.
+Application of these methods to empirical incidence time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time window and near peaks in COVID-19 incidence.
 These findings will assist in the allocation of public health resources, especially in the planning of surge capacity.
@@
 \begin{figure}[!h]
+\centering
 \includegraphics[width=0.6\textwidth]{thetaest_v_theta.pdf}

### 2024-12-06 manuscript/ms (2).tex
Update ms (2).tex

@@
 
+We model broad-scale change in incidence as $log(E[I_t / N])  = \sum_j \beta_j h_j(t)$.
 A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size.
 Since population size influences the mean and variance of case count data and thus could have an impact on estimates of dispersion, we adjust for population size using an offset in the model. 
-There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual).
+There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual). Let $I_i$ be the number of infections at time step i and $h_i$
 
 \begin{align}
-  \log(E[Y_i]/n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
-  \log(E[Y_i])-\log(n_i) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
-  \log(E[Y_i]) &= \beta_1 h_1(t_i) + \beta_2h_2(t_i) + \beta_3 h_3(t_i) + log(n_i) 
+  \log(E[I]/N) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
+  \log(E[I])-\log(N) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
+  \log(E[I_i]) &= \beta_1 h_1(t_i) + \beta_2h_2(t_i) + \beta_3 h_3(t_i) + log(n_i) 
 \end{align}
 
-Thus, the form of the probability mass function for incidence at a time step is:
+The form of the probability mass function for incidence at a time step is:
 

### 2024-12-06 manuscript/ms (2).tex
Update incidence equations

@@
 
-We model broad-scale change in incidence as $log(E[I_t / N])  = \sum_j \beta_j h_j(t)$.
-A persistent challenge in investigating changes in variability has been ``spurious correlation'' with population size.
-Since population size influences the mean and variance of case count data and thus could have an impact on estimates of dispersion, we adjust for population size using an offset in the model. 
-There is an offset term in order to directly model counts (here, COVID-19 cases) per unit of observation (here, per individual). Let $I_i$ be the number of infections at time step i and $h_i$
-
-\begin{align}
-  \log(E[I]/N) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\
-  \log(E[I])-\log(N) &= \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) \\ 
-  \log(E[I_i]) &= \beta_1 h_1(t_i) + \beta_2h_2(t_i) + \beta_3 h_3(t_i) + log(n_i) 
-\end{align}
+We model broad-scale change in incidence as 
+\begin{equation}
+  \log(E[I_t / N])  = \sum_j \beta_j h_j(t)
+\end{equation}
+A persistent challenge in investigating changes in incidence dispersion the statistical impact of population size on dispersion. 
+Since population size influences the mean and variance of case count data it impacts estimates of dispersion even in populations that are otherwise identical.
+Accordingly, population size appears as an offset in our model of broad-scale incidence changes.
+Letting $I_t$ be the number of infections at time $t$ in a population of size $N$, equation (x) implies
+\begin{equation}
+  \log(E[I_t]) &= \sum_j \beta_j h_j(t) + \log(N) 
+\end{equation}
 

### 2024-12-08 manuscript/ms (2).tex
Update ms (2).tex

@@
 Under a wide range of configurations for a branching process model, including the vast majority of parsimonious ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution \cite{kendall_stochastic_1949}: $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expected value for $I_t$ and $\theta_t$ is the dispersion parameter.
-The dispersion parameter is related to the variance of incidence by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. 
+The dispersion parameter is related to the variance of infections by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$. 
 Importantly, smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
@@
 Modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemic trajectories.
-A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which infections at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$.
+A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which infections at time $t-1$ produce new infections at time $t$ and $\theta_t = I_{t-1}$.
 \begin{equation}
@@
 Under these conditions, $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (\cite{kendall_stochastic_1949}, \cite{bjornstad_dynamics_nodate}). 
-However, this model assumes that subpopulations of infectious individuals emanating from one infected individuals are independent and statistically identical to those from another.
+However, this model assumes that subpopulations of infectious individuals emanating from one infected individual are independent and statistically identical to those from another. 
 That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
@@
 Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion \cite{barron_analysis_1992}.
-This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base population/incidence.
+This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base populations/incidence.
 
-We develop a method that quantifies the evolution of dispersion along incidence time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
+We develop a method that quantifies the evolution of dispersion along case count time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.
 We apply the method to COVID-19 \emph{weekly} new cases (to avoid irregularities in daily data reporting, such as when all weekend cases are reported on Monday) in large US counties to investigate the relationships between incidence, dispersion and epidemic dynamic regimes over a portion of the COVID-19 pandemic.
@@
 % For figure citations, please use "Fig" instead of "Figure".
-Classical approaches \cite{grenfell_dynamics_2002} model incidence at a time step using a negative binomial variable with expectation equal to the epidemic intensity and dispersion parameter equal to previous incidence.
+Classical approaches \cite{grenfell_dynamics_2002} model infections at a time step using a negative binomial variable with expectation equal to the epidemic intensity and dispersion parameter equal to previous infections.
 However, other processes besides the current number infected might affect dispersion, so we instead investigate changes in dispersion over time to understand processes that may leave a signal in dispersion.
 
-Our general framework is that incidence at a time step is drawn from a negative binomial distribution with time-varying mean and dispersion parameter (the dispersion parameter varies more slowly than the mean parameter).
-The model is formulated with a linear predictor that includes a natural spline in time with three degrees of freedom to account for autocorrelation in case counts. 
+Our general framework is that the number of infections at a time step is drawn from a negative binomial distribution with time-varying mean and dispersion parameter (the dispersion parameter varies more slowly than the mean parameter).
+The model is formulated with a linear predictor that includes a natural spline in time to account for autocorrelation in case counts. 
 Natural splines are cubic splines which are linear outside of the boundary knots \cite{perperoglou_review_2019}. 
@@
 
-We model broad-scale change in incidence as 
+Letting $I_t$ be the number of infections at time $t$ in a population of size $N$, we model broad-scale change in incidence as 
 \begin{equation}
@@
 \end{equation}
-A persistent challenge in investigating changes in incidence dispersion the statistical impact of population size on dispersion. 
-Since population size influences the mean and variance of case count data it impacts estimates of dispersion even in populations that are otherwise identical.
-Accordingly, population size appears as an offset in our model of broad-scale incidence changes.
-Letting $I_t$ be the number of infections at time $t$ in a population of size $N$, equation (x) implies
+where $h_j(t)$ are basis functions for the natural spline and $\beta_j$ are their associated parameters.
+A persistent challenge in investigating changes in the dispersion of case counts is the statistical impact of population size. 
+Since population size influences the mean and variance of case count data, it impacts estimates of dispersion even in populations that are otherwise identical.
+Accordingly, population size appears as an offset in our model of broad-scale incidence changes. That is, Eq 2 implies
 \begin{equation}
@@
 
-The form of the probability mass function for incidence at a time step is:
+The form of the probability mass function for infections at a time step is:
 
@@
 This involves fitting both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
-In sum, our method identifies shifts in population-level dispersion in incidence while accounting for population size.
+In sum, our method identifies shifts in population-level dispersion in infections while accounting for population size.
 
@@
 Epidemic curves spanning 60 time steps each were produced, and the likelihood-ratio test (LRT) procedure was applied to each at its known change point in \begin{math}\theta\end{math}.
-Varying the magnitude of the \begin{math}\theta\end{math} change, location of the change in the curve, population size underlying the curve, and curve shape (as mentioned above) allowed us to test the validity and power of our approach across a range of situations.
+Varying the magnitude of the \begin{math}\theta\end{math} change, location of the change in the curve, population size underlying the curve, and curve shape (as mentioned above) allowed us to test the validity and power of our approach across a range of scenarios.
+
+\subsection*{Data availability and processing}
+
+COVID-19 cumulative case data are from [The New York Times](https://github.com/nytimes/covid-19-data/) and county population sizes are from [The United States Census Bureau](https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/totals/.
+Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
 
@@
 \section*{Results}
-We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig.\ 1 e, f).
+We found that the LRT method is robust across population sizes (for population sizes included in the empirical data) (Fig\ 1 e, f).
 The criteria for adequate test performance are that the average p-value is 0.5 when the effect size is zero, and lower average p-values are observed with increasing effect sizes.
@@
 The negative binomial framework for estimating \begin{math}\theta\end{math} is also robust across the population sizes examined, and a large portion of inaccurate estimates correspond to counties with small population sizes outside of the range considered (\nameref{S1_Fig}).
-In row one and two of Fig.\ 1, we illustrate that an increase in \begin{math}\theta\end{math} is associated with decreased dispersion in simulated case count time series.
-This pattern is observed in the empirical data, and is independent of whether incidence is high or low, as seen in Fig.\ 2, where an example time series of $log_{10}\theta$ estimated from empirical data in Jefferson County, AL shows decreasing values around periods in which cases increase in volatility.
+In row one and two of Fig\ 1, we illustrate that an increase in \begin{math}\theta\end{math} is associated with decreased dispersion in simulated case count time series.
+This pattern is observed in the empirical data, and is independent of whether incidence is high or low, as seen in Fig\ 2, where an example time series of $log_{10}\theta$ estimated from empirical data in Jefferson County, AL shows decreasing values around periods in which cases increase in volatility.
 For example, consider the period of time around the beginning of 2022, in which high case count variability coincides with low values of $log_{10}\theta$.
-Additionally, the dispersion parameter changed as new variability regimes began ($log_{10}\theta$ values colored red in Fig.\ 2 b).
-There were departures from the dispersion parameter under the model in Equation 1, especially around the beginning of 2022 and the end of the time frame examined (Fig.\ 2 b).
-The dispersion parameter under the model in Equation 1 can be plotted by taking a series of previous incidence and adding an offset to account for the reporting rate (log10($\theta_t$) in the model = log10(I_{t-1}) = log10(cases_{t-1}/reporting rate) = log10(cases_{t-1}) - log10(reporting rate)).
+Additionally, the dispersion parameter changed as new variability regimes began ($log_{10}\theta$ values colored red in Fig\ 2 b). 
+There were departures from the dispersion parameter expected under the model in Eq 1, especially around the beginning of 2022 and the end of the time series (Fig\ 2 b).
+The dispersion parameter under the model in Eq 1 can be plotted by taking a series of previous cases and adding an offset to account for the reporting rate (log10($\theta_t$) in the model in Eq 1 = log10(I_{t-1}) = log10(reported_cases_{t-1}/reporting rate) = log10(reported_cases_{t-1}) - log10(reporting rate)).
 
@@
 b: Time series of $log_{10}\theta$ colored by whether the associated LRT statistic is greater than the Bonferroni-corrected level 0.05 quantile of a chi-square random variable with one degree of freedom (red indicates areas of putative change in $\theta$).
-The black line represents the dispersion parameter under the model from Equation 1, assuming a reporting rate of 0.10.
+The black line represents the dispersion parameter under the model from Eq 1, assuming a reporting rate of 0.10.
 }
@@
 
-Highly overdispersed incidence patterns were observed for many of the examined counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting.
-The most dispersed category in Fig.\ 3 (e) reaches its highest proportion near the end of the time frame examined.
-In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig. 3 c, d).
-Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig.\ 3 f).
+Highly overdispersed patterns were observed for many of the examined counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
+The most dispersed category in Fig\ 3 (e) reaches its highest proportion near the end of the time frame examined.
+In addition, there are increases in dispersion around the peaks in incidence in the dataset (Fig 3 c, d).
+Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig\ 3 f).
 Note that these p-values should be corrected for multiple testing when used for inference instead of visualization.
@@
 
-Raising variance relative to mean (higher dispersion) implies spatiotemporal "crowding" of cases (i.e.
-localized surges) which may necessitate more surge capacity in hospitals and testing centers.
-Therefore, it may be the case that there were more surges on the way up to or on the way down from peak incidence.
+Raising variance relative to mean (higher dispersion) implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
 It appears that there are more surges on the way up to, and on the way down from, peak incidence. 
@@
 We developed an approach to quantify clustering of cases in epidemic time series that is less susceptible to detecting artifacts based on population size and incidence. 
-Our method forms part of a larger initiative to investigate variability in incidence as an important attribute of epidemic time series using novel metrics.
+Our method forms part of a larger initiative to investigate variability in infections as an important attribute of epidemic time series using novel metrics.
 For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
@@
 
-Application of the methods we developed to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time frame considered and near peaks in COVID-19 incidence.
+Application of the methods we developed to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time series considered and near peaks in COVID-19 incidence.
 These results imply that we can revise our understanding of case count dispersion: dispersion is high \emph{near} peak incidence, suggesting that dispersion is not simply governed by incidence. 
-These departures from the level of dispersion expected under the model represented in Equation 1 represent time periods in which the assumptions may be violated and cases may be highly clustered in time.
-
-Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large populations.
+These departures from the level of dispersion expected under the model represented in Eq 1 represent time periods in which the assumptions may be violated and cases may be highly clustered in time.
 
 Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series showed that epidemics in smaller communities were concentrated on narrower windows of the influenza season; the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel_urbanization_2018}.
-So, additional research is needed to understand the relationship between burstiness in small communities and temporally changing dispersion in these areas.
+Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large populations.
+Additional research is needed to understand the relationship between burstiness in small communities and temporally changing dispersion in these areas.
 
 Since there are regimes of the COVID-19 epidemic that are subject to increased dispersion across large counties, these may be candidate time periods for rolling out extra surge capacity in large counties.
-But, as mentioned above, further research is needed to understand whether this phenomenon occurs in small and mid-sized counties.
+But, as mentioned above, further research is needed to understand whether this phenomenon occurs in small counties.
 
 \section*{Conclusion}
-We presented an approach to quantify clustering of cases in epidemic time series that does not detect artifacts based on population size and incidence. 
+We presented an approach to quantify clustering of cases in epidemic time series that is less susceptible to detecting artifacts based on population size and incidence. 
 While our estimation framework facilitates comparison of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
-Application of these methods to empirical incidence time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time window and near peaks in COVID-19 incidence.
+Application of these methods to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time series and near peaks in COVID-19 incidence.
 These findings will assist in the allocation of public health resources, especially in the planning of surge capacity. 

### 2024-12-13 manuscript/ms (2).tex
Shorten Author Summary

@@
 However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
-Quantifying higher-frequency variation around incidence trends can elucidate important underlying processes, including transmission heterogeneity or changes in the susceptibility landscape.
-We developed a statistical framework to quantify temporal changes in incidence dispersion within a single time series. 
-We used negative binomial regression to account for factors that might confound the estimation of dispersion, allowing us to use a single parameter to infer the degree of incidence dispersion. 
-We applied the method to COVID-19 incidence data from 144 US counties from January 1st, 2020 to March 23rd, 2023, and found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
+Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
+We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data.
+We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone.
 Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
-These increases potentially indicate that some individuals spread SARS-CoV-2 (the virus that causes COVID-19) to many more people than others, that the susceptibility landscape changed, or that there were changes in reporting.
-Shifts in dispersion can indicate shifts in epidemic phase as well as provide information about important underlying biological processes. 
-Therefore, our method to estimate dispersion provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.
+These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
+Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.
 

### 2024-12-13 manuscript/ms (2).tex
Update ms (2).tex

@@
 This is an additional reason that the dispersion parameter of a negative binomial distribution should be used to measure changes in meaningful case count clustering in settings with differing base populations/incidence.
-
+[The Introduction should conclude with a brief statement of the overall aim of the experiments and a comment about whether that aim was achieved.]
 We develop a method that quantifies the evolution of dispersion along case count time series, allowing for the detection of changes in clustering that are not due to changes in population size or overall burden of incidence.

### 2024-12-16 ms-3/ms.tex
Start clean version

\documentclass[11pt,letterpaper]{article}

% Packages
\usepackage{authblk} % For author affiliations
\usepackage{graphicx} % For including graphics
\usepackage{amsmath, amssymb} % For mathematical symbols
\usepackage{hyperref} % For hyperlinks
\usepackage{natbib} % For bibliography
\usepackage[margin=1in]{geometry} % Standard margins
\usepackage{lineno} % For line numbers

% Metadata
\title{Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time}
\author[1]{Rachael Aber}
\author[2]{Yanming Di}
\author[1,3]{Benjamin D. Dalziel}
\affil[1]{Department of Integrative Biology, Oregon State University, Corvallis, Oregon, USA}
\affil[2]{Department of Statistics, Oregon, Oregon State University, Corvallis, Oregon, USA}
\affil[3]{Department of Mathematics, Oregon State University, Corvallis, Oregon, USA}
\date{} % No date for submissions

% Begin Document
\begin{document}

\linenumbers\maketitle

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how it changes over time, is often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses. 
However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data. We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.

\section*{Introduction}
Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, and/or changes in susceptiblility~\cite{lloyd2005superspreading, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.

Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\cite{guo2023statistical,ko2023time}.
However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\cite{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\cite{sun2021transmission}. 
An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.

\section*{Acknowledgments}
RA work on this project was supported by X and Y. BDD work on this project was supported by the National Science Foundation (NSF) grants (X, Y) and by the David and Lucile Packard Foundation.

\section*{References}
\bibliographystyle{plainnat} % Use plainnat or another appropriate style
\bibliography{references} % Create a references.bib file

% End Document
\end{document}

### 2024-12-16 ms-3/ms.tex
Update ms.tex

@@
 
-Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\cite{guo2023statistical,ko2023time}.
+Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers (i.e., the expected number of secondary infections that will result if a particular individual becomes infected), quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\cite{guo2023statistical,ko2023time}.
 However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.

### 2024-12-19 ms-3/ms.tex
Complete intro

@@
 Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
-However, measures of temporal volatility in incidence, and how it changes over time, is often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
+However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
 Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
@@
 
+Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
+For example, a recently-developed method estimates the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual, directly from population-level incidence data~\cite{kirkegaard2021superspreading}.
+Clustering of cases has also been estimated directly from incidence data~\cite{schneckenreither2023assessing}.
+Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\cite{graham2019measles}.
+Analysis of the shape of epidemic curves for influenza in cities also aids public health planning by identifying contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for surge capacity in the health system~\cite{dalziel2018urbanization, wallinga2018metropolitan}.
+
+What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
+Under a wide range of configurations for a branching process model the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\cite{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
+The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
+Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
+The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\cite{barron1992analysis}.
+
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\cite{lloyd1967mean} (see Supplemental Information).
+For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
+In a simple example relevant to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
+
+In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, incidence is often modeled as
+\begin{equation}
+    \mathrm{I_{t+1}} = NB(\mu = \lambda R_t I_t, \theta_t = I_t)
+\end{equation}
+where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\cite{kendall1949stochastic, bjornstad2002dynamics}.
+Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
+However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.
+
+In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach, which makes data-driven estimates of $\theta_t$, including identifying probably timepoints when $\theta$ is changing rapidly, which may help to reveal how heterogeneity in transmission, susceptibility and reporting, are impacting incidence volatlity.
+
+\section*{Methods}
+
 \section*{Acknowledgments}

### 2024-12-19 ms-3/ms.tex
Update intro

@@
 Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\cite{graham2019measles}.
-Analysis of the shape of epidemic curves for influenza in cities also aids public health planning by identifying contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for surge capacity in the health system~\cite{dalziel2018urbanization, wallinga2018metropolitan}.
+Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for the sensitivity of cities to forcing from exogenous drivers of transmission (e.g., climate) and for surge capacity in the health system~\cite{dalziel2018urbanization, wallinga2018metropolitan}.
 
@@
 For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
-In a simple example relevant to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
+In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
 
@@
 \begin{equation}
-    \mathrm{I_{t+1}} = NB(\mu = \lambda R_t I_t, \theta_t = I_t)
+    \mathrm{I_{t+1}} = NB(\mu = R_t I_t, \theta_t = I_t)
 \end{equation}

### 2024-12-19 manuscript/S1_equations.tex
Create S1_equations.tex

Let $m^*$ be mean crowding and $m$ be the mean. By definition, we have $m^* = m +(\sigma^2/m -1)$. 
Therefore, we have that $m^*- m = \sigma^2/m -1$.

We also have that $\sigma^2 = m + m^2/\theta$, by definition of negative binomial variance, where $\theta$ is the dispersion parameter. This is equivalent to $\sigma^2/m = 1 + m/\theta$.
Therefore, we have that $m/\theta = \sigma^2/m - 1$.

By equating these, we have $m^*- m = m/\theta$.

Dividing both sides by $m$, we have $m^*/m - 1 = 1/\theta$

Finally, we have that $m^*= m/\theta + m$

### 2024-12-19 ms-3/ms.tex
Add methods

--- 2024-12-16 ms-3/ms.tex
@@
 % Packages
-\usepackage{authblk} % For author affiliations
-\usepackage{graphicx} % For including graphics
-\usepackage{amsmath, amssymb} % For mathematical symbols
+\usepackage{authblk} % author affiliations
+\usepackage{graphicx} % graphics
+\usepackage{amsmath, amssymb} % mathematical symbols
+\usepackage{natbib} % bibliography
+\usepackage[margin=1in]{geometry} % margins
+\usepackage{lineno} % line numbers
+\usepackage{setspace} % linespacing
 \usepackage{hyperref} % For hyperlinks
-\usepackage{natbib} % For bibliography
-\usepackage[margin=1in]{geometry} % Standard margins
-\usepackage{lineno} % For line numbers
+
+% Set 1.5 spacing
+\setstretch{1.5}
+
+% Set citation style
+\setcitestyle{numbers}
 
@@
 Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
-However, measures of temporal volatility in incidence, and how it changes over time, is often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
-Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
+However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
+Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
 We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
@@
 
+\clearpage
 \section*{Author summary}
@@
 
+\clearpage
 \section*{Introduction}
 Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
-Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, and/or changes in susceptiblility~\cite{lloyd2005superspreading, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
+Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, or changes in susceptiblility~\citep{lloyd2005superspreading, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
 Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
 
-Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers (i.e., the expected number of secondary infections that will result if a particular individual becomes infected), quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\cite{guo2023statistical,ko2023time}.
+Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers (i.e., the expected number of secondary infections that will result if a particular individual becomes infected), quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}.
 However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
-In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\cite{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
-The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\cite{sun2021transmission}.
+In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\citep{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
+The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\citep{sun2021transmission}.
 An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.
+
+Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
+For example, a recently-developed method estimates the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual, directly from population-level incidence data~\citep{kirkegaard2021superspreading}.
+Clustering of cases has also been estimated directly from incidence data~\citep{schneckenreither2023assessing}.
+Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}.
+Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for the sensitivity of cities to forcing from exogenous drivers of transmission (e.g., climate) and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
+
+What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
+Under a wide range of configurations for a branching process model the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
+The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
+Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
+The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial.
+Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\citep{barron1992analysis}.
+
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information).
+For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
+In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
+
+In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, incidence is often modeled as
+\begin{equation}
+    I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
+\end{equation}
+where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}.
+Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
+However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.
+
+In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach, which makes data-driven estimates of $\theta_t$, including identifying timepoints when $\theta$ is changing rapidly, which may help to reveal the impacts of heterogeneity in transmission, susceptibility, and reporting.
+
+\section*{Methods}
+By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
+Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review} fitted within a moving window
+\begin{equation}
+  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
+\end{equation}
+where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$.
+The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
+Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
+
+Modeling the underlying epidemic dynamics based on log-transformed incidence allows us to address the statistical effects of population size on the relationship between the mean and variance in count data, which would otherwise confound our analysis.
+Specifically, since population size influences the mean and variance of case count data, it impacts estimates of dispersion in populations of different sizes that are otherwise identical.
+Accordingly, population size appears as an offset in our model of broad-scale incidence changes. That is,
+\begin{equation}
+  \log(\mu_t) = \sum_{j=1}^J \beta_j^{(t)} h_j(t) + \log(N) 
+\end{equation}
+
+The form of the probability mass function for infections at a time step is:
+\begin{equation}
+  f_t(I) = \binom{I + \theta - 1}{I} {\left( \frac{\mu}{\mu + \theta} \right)}^I {\left( \frac{\theta}{\mu + \theta} \right)}^\theta
+\end{equation}
+where $\mu$ is estimated via the linear predictor outlined above.
+
+We estimate $\theta_t$ given observed incidence using an iteratively reweighted least-squares procedure implemented via the NBPSeq R package~\citep{di2015nbpseq}. 
+For each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ is estimated using the spline, and a single value of $\theta_t$ is fitted. [more detail needed]
+
+In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that can be applied at each time step to test the hypothesis that $\theta$ has changed.
+This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
+[more detail needed]
+
+\subsection*{Application to simulated data}
+
+To test the validity and power of the methods, we epidemic curves with known values of $\mu$ and $\theta$.
+We varyied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
+Epidemic curves were created from a deterministic skeleton (template) that either assumed a flat distribution of incidence over time or a Gaussian-shaped distribution, with parameters controlling the total epidemic size, peak time, peak width, and duration.
+Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $theta_2$ were used in each respective piece, before and after the breakpoint.
+This allowed us to simulate epidemic curves with variation in dispersion, as well as the special case when $\theta_1 = \theta_2$.
+
+\subsection*{Data availability and processing}
+We applied our method to cumulative COVID-19 case countes in US counties~\citep{nytimes_covid19}.
+Population sizes were from The United States Census Bureau~\citep{us_census_county_population_2021}.
+Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
+Missing values for new cases were all at the beginning of the pandemic and were imputed as zero.
+Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero.
 
@@
 \section*{References}
-\bibliographystyle{plainnat} % Use plainnat or another appropriate style
+\bibliographystyle{vancouver} % Use plainnat or another appropriate style
 \bibliography{references} % Create a references.bib file

### 2024-12-19 ms-3/ms.tex
Update

@@
 By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
-Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review} fitted within a moving window
+Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
+To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
 \begin{equation}
@@
 We estimate $\theta_t$ given observed incidence using an iteratively reweighted least-squares procedure implemented via the NBPSeq R package~\citep{di2015nbpseq}. 
-For each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ is estimated using the spline, and a single value of $\theta_t$ is fitted. [more detail needed]
+For each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ is estimated using the spline, and a single value of $\theta_t$ is fitted.
 

### 2024-12-20 ms-3/ms.tex
Add methods text

@@
 
-We estimate $\theta_t$ given observed incidence using an iteratively reweighted least-squares procedure implemented via the NBPSeq R package~\citep{di2015nbpseq}. 
-For each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ is estimated using the spline, and a single value of $\theta_t$ is fitted.
+Since the 'glm.nb' function in the MASS R package \cite{MASS} doesn't accomodate an offset term, we estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$.
+That is, for each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ was estimated using the spline and offset term via IRLS in the NBPSeq R package~\citep{di2015nbpseq}.
+Then, a single value of $\theta_t$ for the time window was computed via an optimization procedure.
 
-In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that can be applied at each time step to test the hypothesis that $\theta$ has changed.
+In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that could be applied at each time step to test the hypothesis that $\theta$ has changed.
 This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
-[more detail needed]
+For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the \begin{math}\theta\end{math}-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.
 

### 2024-12-20 ms-3/ms.tex
Update manuscript and format some code

@@
 \section*{Author summary}
-Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health responses. 
-However, traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
+Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series.
 Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
@@
 
-Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers (i.e., the expected number of secondary infections that will result if a particular individual becomes infected), quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}.
+Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}, indicating temporal changes in the level of variability around the expected number of secondary infections that will result if a particular individual becomes infected.
 However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
@@
 What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
-Under a wide range of configurations for a branching process model the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
+Under a wide range of configurations for a branching process model of contagion spread the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
 The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
@@
 
-In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, incidence is often modeled as
+In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, the number of infections each generation is often modeled as
 \begin{equation}
@@
 Modeling the underlying epidemic dynamics based on log-transformed incidence allows us to address the statistical effects of population size on the relationship between the mean and variance in count data, which would otherwise confound our analysis.
-Specifically, since population size influences the mean and variance of case count data, it impacts estimates of dispersion in populations of different sizes that are otherwise identical.
+Specifically, since population size influences the mean and variance of case count data, it impacts dispersion in different-sized populations that are otherwise identical.
 Accordingly, population size appears as an offset in our model of broad-scale incidence changes. That is,
@@
 
-Since the 'glm.nb' function in the MASS R package \cite{MASS} doesn't accomodate an offset term, we estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$.
+We estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$.
 That is, for each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ was estimated using the spline and offset term via IRLS in the NBPSeq R package~\citep{di2015nbpseq}.
@@
 In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that could be applied at each time step to test the hypothesis that $\theta$ has changed.
-This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part (\begin{math}\theta\end{math}-change) model.
-For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the \begin{math}\theta\end{math}-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.
+This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part ($\theta$-change) model.
+For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the $\theta$-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.
 
@@
 To test the validity and power of the methods, we epidemic curves with known values of $\mu$ and $\theta$.
-We varyied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
+We varied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
 Epidemic curves were created from a deterministic skeleton (template) that either assumed a flat distribution of incidence over time or a Gaussian-shaped distribution, with parameters controlling the total epidemic size, peak time, peak width, and duration.
-Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $theta_2$ were used in each respective piece, before and after the breakpoint.
+Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $\theta_2$ were used in each respective piece, before and after the breakpoint.
 This allowed us to simulate epidemic curves with variation in dispersion, as well as the special case when $\theta_1 = \theta_2$.
@@
 Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
+We used a width paramter of $\delta = 8$ and set the degrees of freedom parameter for the natural splines to $J = 3$.
 

### 2024-12-20 ms-3/ms.tex
Update manuscript and fig1 code

@@
 Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
-We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
+We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
 Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties.
@@
 \subsection*{Data availability and processing}
-We applied our method to cumulative COVID-19 case countes in US counties~\citep{nytimes_covid19}.
-Population sizes were from The United States Census Bureau~\citep{us_census_county_population_2021}.
+We applied our method to cumulative COVID-19 case counts in United States (US) counties~\citep{nytimes_covid19}.
+Population sizes were from taken the US Census Bureau~\citep{us_census_county_population_2021}.
 Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
@@
 Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
-We used a width paramter of $\delta = 8$ and set the degrees of freedom parameter for the natural splines to $J = 3$.
+We used a width paramter of $\delta = 8$ week and set the degrees of freedom parameter for the natural splines to $J = 3$.  
+[This seems like a wide window and a low-df spline.
+Why did we choose these and what happens if we use different values?
+Some sort of sensitivity analysis is probably needed, even if it is just to run the analysis using a smaller window and larger df]
+
+\section*{Results and Discussion}
+Our estimates of $\theta$ were robust across the population sizes included in the empirical analysis of U.S. counties. 
+However, inaccuracies were more common in counties with small populations outside the examined range.%(\nameref{S1_Fig}). 
+We found that the LRT method of testing for sharp changes in $\theta$ is also robust across population sizes represented in the empirical data (Fig\ 1 e, f). 
+Adequate test performance is characterized by an average \( p \)-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
+
+\begin{figure}[!ht]
+\includegraphics[width=1\textwidth]{fig1}
+\caption{
+Detecting changes in dispersion within incidence time series for populations of different sizes.
+(a, b) Simulated case counts with constant (a) and changing (b) dispersion parameters.
+(c, d) Dispersion parameters used to generate the data in (a) and (b), respectively.
+(e, f) LRT performance under varying absolute differences in \(\theta\) (horizontal axis), showing \( p \)-value distributions for population sizes of 50,000 (e) and 10,000,000 (f), representing the range in the empirical data.
+The null and alternative hypotheses (O and X) correspond to the parameters in (c) and (d).
+Red vertical lines indicate interquartile ranges.
+}\label{fig1}
+\end{figure}
 

### 2024-12-20 code/2-generate_time_series.R

@@ -16,9 +16,9 @@ population_max <- signif(max(pops), 1)
 breakpoint_min <- 20
 breakpoint_max <- 40
 
-nbreakpoint <- 5  # Number of unique breakpoint values
+nbreakpoint <- 5 # Number of unique breakpoint values
 
-nstep <- 60       # Number of time steps in simulated trajectories
+nstep <- 60 # Number of time steps in simulated trajectories
 
 # Theta, breakpoint, population size, and curve type sequences
 # Max theta could be based on mu^2, where the alg performs well, alg boundaries, or emp. data
@@ -29,11 +29,13 @@ population <- c(10000, seq(population_min, population_max, length.out = 5))
 curve_type <- c(1, 2)
 
 # Expand grid and add final size
-curve_parms <- expand.grid(theta1 = theta1,
-                           theta2 = theta2,
-                           breakpoint = breakpoint,
-                           population = population,
-                           curve_type = curve_type)
+curve_parms <- expand.grid(
+  theta1 = theta1,
+  theta2 = theta2,
+  breakpoint = breakpoint,
+  population = population,
+  curve_type = curve_type
+)
 
 curve_parms$final_size <- final_attack_rate * curve_parms$population
 
@@ -43,27 +45,27 @@ rm(theta1, theta2, breakpoint, population, curve_type)
 ncurve <- nrow(curve_parms)
 curves <- matrix(NA, nrow = ncurve, ncol = nstep)
 for (i in 1:ncurve) {
+  breakpoint <- curve_parms$breakpoint[i]
+  theta1 <- curve_parms$theta1[i]
+  theta2 <- curve_parms$theta2[i]
+  final_size <- curve_parms$final_size[i]
+  curve_type <- curve_parms$curve_type[i]
 
-    breakpoint <- curve_parms$breakpoint[i]
-    theta1 <- curve_parms$theta1[i]
-    theta2 <- curve_parms$theta2[i]
-    final_size <- curve_parms$final_size[i]
-    curve_type <- curve_parms$curve_type[i]
+  means <- get_template(
+    param = c(final_size, peak_time, peak_width, nstep),
+    type = curve_type
+  )
 
-    means <- get_template(param = c(final_size, peak_time, peak_width, nstep),
-                          type = curve_type)
+  means1 <- means[1:breakpoint]
[…]

### 2024-12-20 code/5-figure1.R

@@ -4,7 +4,7 @@ source("code/lrt.R")
 
 
 load("data/processed/simulated_curves.Rdata")
-load("data/processed/nyt_weekly.Rdata") 
+load("data/processed/nyt_weekly.Rdata")
 load("data/processed/theta_lg_pops.Rdata")
 load("data/processed/pvals_sim_LRT.Rdata")
 
@@ -15,35 +15,40 @@ pdf(filename)
 par(mfrow = c(3, 2))
 
 # Trim empirical incidence to same length as thetas2
-series <- cases[10, 8:(length(cases[10,]) - 8 + 1)]
-dates <- dates[8:(length(cases[10,]) - 8 + 1)]
+series <- cases[10, 8:(length(cases[10, ]) - 8 + 1)]
+dates <- dates[8:(length(cases[10, ]) - 8 + 1)]
 
-# Remove first observations from all
-# series <- series[35:length(series)]
-# dates <- dates[35: length(dates)]
-# thetas <- thetas[1, 35:dim(thetas2)[2]]
 
 # Row 1
 
-plot(curves[which(curve_parms$theta1 == curve_parms$theta2 & curve_parms$breakpoint == 30 
-                  & curve_parms$final_size == 1000000 & curve_parms$curve_type == 2 & curve_parms$theta1 == 1.1),], 
-     xlab = "Day", type = "l", ylab = "Incidence", cex.lab = 1.7, cex.axis = 1.7)
+plot(
+  curves[which(curve_parms$theta1 == curve_parms$theta2 & curve_parms$breakpoint == 30 &               #nolint 
+    curve_parms$final_size == 1000000 & curve_parms$curve_type == 2 & curve_parms$theta1 == 1.1), ],   #nolint 
+  xlab = "Day", type = "l", ylab = "Incidence", cex.lab = 1.7, cex.axis = 1.7
+)
 mtext("a", side = 3, line = 1, adj = 0, cex = 1.2)
-plot(curves[which(curve_parms$theta1 == 1.1 & curve_parms$theta2 == 10.1 & curve_parms$breakpoint == 30 
-                  & curve_parms$final_size == 1000000 & curve_parms$curve_type == 2),], 
-     xlab = "Day", type = "l", ylab = "Incidence", cex.lab = 1.7, cex.axis = 1.7)
+plot(
[…]

### 2024-12-20 ms-3/ms.tex

@@
 We found that the LRT method of testing for sharp changes in $\theta$ is also robust across population sizes represented in the empirical data (Fig\ 1 e, f). 
-Adequate test performance is characterized by an average \( p \)-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
+Adequate test performance for the LRT is characterized by an average \( p \)-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
 
-\begin{figure}[!ht]
-\includegraphics[width=1\textwidth]{fig1}
-\caption{
-Detecting changes in dispersion within incidence time series for populations of different sizes.
+%\begin{figure}[!ht]
+%\includegraphics[width=1\textwidth]{fig1}
+%\caption{
+%Detecting changes in dispersion within incidence time series for populations of different sizes.
 (a, b) Simulated case counts with constant (a) and changing (b) dispersion parameters.
@@
 Red vertical lines indicate interquartile ranges.
-}\label{fig1}
-\end{figure}
+%}\label{fig1}
+%\end{figure}
 

### 2024-12-20 ms-3/ms.tex
Reveal a strange phenomenon in Fig 1e

@@
 We found that the LRT method of testing for sharp changes in $\theta$ is also robust across population sizes represented in the empirical data (Fig\ 1 e, f). 
-Adequate test performance for the LRT is characterized by an average \( p \)-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
+Adequate test performance for the LRT is characterized by an average $p$-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
 

### 2024-12-20 code/5-figure1.R

@@ -20,41 +20,83 @@ dates <- dates[8:(length(cases[10, ]) - 8 + 1)]
 
 
 # Row 1
+# Define conditions for filtering curves
+condition_a <- curve_parms$theta1 == curve_parms$theta2 &
+  curve_parms$breakpoint == 30 &
+  curve_parms$final_size == 1e6 &
+  curve_parms$curve_type == 2 &
+  curve_parms$theta1 == 1.1
 
+condition_b <- curve_parms$theta1 == 10.1 &
+  curve_parms$theta2 == 1.1 &
+  curve_parms$breakpoint == 30 &
+  curve_parms$final_size == 1e6 &
+  curve_parms$curve_type == 2
+
+# Plot for condition_a
 plot(
-  curves[which(curve_parms$theta1 == curve_parms$theta2 & curve_parms$breakpoint == 30 &               #nolint 
-    curve_parms$final_size == 1000000 & curve_parms$curve_type == 2 & curve_parms$theta1 == 1.1), ],   #nolint 
-  xlab = "Day", type = "l", ylab = "Incidence", cex.lab = 1.7, cex.axis = 1.7
+  curves[which(condition_a), ],
+  xlab = "Day", ylab = "Incidence", type = "l",
+  cex.lab = 1.7, cex.axis = 1.7
 )
 mtext("a", side = 3, line = 1, adj = 0, cex = 1.2)
+
+# Plot for condition_b
 plot(
-  curves[which(curve_parms$theta1 == 1.1 & curve_parms$theta2 == 10.1 & curve_parms$breakpoint == 30 & #nolint 
-    curve_parms$final_size == 1000000 & curve_parms$curve_type == 2), ],                               #nolint 
-  xlab = "Day", type = "l", ylab = "Incidence", cex.lab = 1.7, cex.axis = 1.7
+  curves[which(condition_b), ],
+  xlab = "Day", ylab = "Incidence", type = "l",
+  cex.lab = 1.7, cex.axis = 1.7
 )
 mtext("b", side = 3, line = 1, adj = 0, cex = 1.2)
 
 
 
-# Row 2
-plot(rep(curve_parms$theta1[which(curve_parms$theta1 == curve_parms$theta2 & curve_parms$breakpoint == 30 & curve_parms$final_size == 1000000 & curve_parms$curve_type == 2 & curve_parms$theta1 == 1.1)], 
[…]

### 2025-01-08 ms-3/ms.tex
add text

@@
 
+Since very large $\theta$ correspond to processes that are close to a Poisson process, the test will not produce a p-value if any of the three values are greater than 1/(1e-3). 
+Very large $\theta$ estimates are often produced in small populations when the process can't be distinguished from a Poisson process.
+Therefore, we have very small $\phi = 1/\theta$  values be considered zero, at which point we fail to reject the Poisson hypothesis.
+
+In addition, we imposed restrictions on $\theta$ estimates such that $\theta = 0$ (point mass) results in no p-value being produced in the test result.
+So, large $\phi$ are considered values where we fail to reject the point mass hypothesis.
+
+Additionally, we included an option to retrieve the likelihood of a range of theta values in addition to these restrictions.
+
 \subsection*{Application to simulated data}
+
+Our simulation samples start points and theta values, and incidence is modeled according to overall incidence trends.
 

### 2025-01-14 manuscript/ms (2).tex
Improve data availability statement
Update figure captions

@@
 Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
+Note that missing values for new cases were all at the beginning of the pandemic and were imputed as zero.
+Apprxoimately 0.24% of new cases were negative due to corrections in the cumulative data.
+These were also imputed as zero.
 
@@
 \caption{
-Detecting dispersion changes in incidence time series in populations of different sizes.
-a: Simulated case counts when dispersion is constant.
-b: Simulated case counts when dispersion changes.
-c: Constant dispersion parameter used in generation of the above.
-d: Changing dispersion parameter used in generation of the above.
-e, f: Performance of the LRT with simulated data that have different absolute differences in \begin{math}\theta\end{math} (horizontal axis of each panel) illustrates p-value distribution in population size of 50,000 (e) and population size of 10,000,000 (f), which represent the range in the empirical data.
-O and X mark the null and alternative hypotheses indicated in panels c and d.
-Red vertical lines represent the interquartile range.
+Detecting dispersion changes in case count time series.
+a: Sampled time period (red line) from overall US COVID-19 incidence per 1,000 people between 2020-01-04 and 2023-03-18.
+b: Cases per 1,000 people in Douglas County, NE over the sample time period with estimated mean values (red) and estimated dispersion values on either side of the midpoint.
+c: Estimated $\theta$ v.
+true $\theta$  with values outside of tolerance plotted in purple (close to Poisson) and blue (close to collapsing to zero).
+d: p-value versus absolute difference in $log_{10}\theta$.
+Red curve and red shading depict ***, respectively.
  }
@@
 a: Mean COVID-19 cases of the 144 US counties over time.
-b: Mean $\theta$ of the 144 US counties over time.
+b: Mean $log_{10}\theta$ of the 144 US counties over time.
+Dashed line represents expected value of $log_{10}\theta_0$ under the null model in Equation 1.
 c: log10(case counts) - log10(reporting rate) over time for each of the large counties (y-axis).
-This represents $log_{10}\theta$ under the model in Equation 1.
+This again represents $log_{10}\theta_0$ under the model in Equation 1.
 d: $log_{10}\theta$ over time for each of the large counties (y-axis).
-e: Binned $log_{10}\theta$ estimates for all counties over time.
+e: Binned $log_{10}\theta_0$ for all counties over time.
 f: LRT p-values over time for each of the large counties (y-axis).

### 2025-01-14 code/4-figure2.R

@@ -20,7 +20,7 @@ plot(dates, series, type = "h", xlab = "Dates", ylab = "", main = "Case counts",
 mtext("a", side = 3, line = 1, adj = 0, cex = 1.3)
 
 
-is_sig <- lrt_stats[1, ] > qchisq(0.9996753, df = 1) # w/o correcting for multiple testing
+is_sig <- lrt_stats[1, ] > qchisq(0.9996732, df = 1) 
 plot(dates, log10(thetas[1,]), type = "n", 
      xlab = "Dates", ylab = "", 
      main = expression(bold(log10(theta))), 

### 2025-01-14 code/6-figure3.R

@@ -11,11 +11,6 @@ load("data/processed/lrtps_lg_pops.Rdata")
 load('data/processed/ftr_poiss_lg_pops.Rdata')
 load('data/processed/ctzs_lg_pops.Rdata')
 
-mean(ftr_poiss)
-mean(ctzs)
-mean(thetas)
-mean(thetas1)
-mean(thetas2)
 
 # Parameters
 reporting_rate <- 0.10

### 2025-01-16 ms-3/ms.tex
Continue finalizing manuscript

@@
 Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
-Short-term fluctuations in incidence time series are often caused in part by variable reporting, but also reflect the population-level impacts of transmission heterogeneity, or changes in susceptiblility~\citep{lloyd2005superspreading, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
+Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity, and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
 Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
 
-Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}, indicating temporal changes in the level of variability around the expected number of secondary infections that will result if a particular individual becomes infected.
+Contact tracing data has revealed temporal changes in the variability of individual reproductive numbers, quantified by shifts in the dispersion parameter of the offspring distribution in branching process models~\citep{guo2023statistical,ko2023time}.
+Similar evidence has been recovered through statistical reconstruction of transmission networks, indicating temporal trends in the level of dispersion at different phases of an epidemic~\citep{lau2017spatial}.
 However, the scaling from individual-level transmission heterogeneity to population-level epidemic dynamics is not fully understood.
-In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and availability~\citep{kretzschmar2020impact}, there is a need for complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
-The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies was required for SARS-CoV-2 control~\citep{sun2021transmission}.
+In addition, traditional contact tracing is very resource intensive, and although new approaches using digital technologies may improve its speed and scalability~\citep{kretzschmar2020impact}, it would be helpful to have complementary population-level analyses that can estimate heterogeneity using incidence data, which is more widely available.
+The importance of considering population-level variability and its relationship to individual-level variability is further highlighted by the finding that a combination of individual-based and population-based strategies were required for SARS-CoV-2 control during the early phases of the pandemic in China~\citep{sun2021transmission}.
 An important challenge therefore is to develop methods that can detect changes in population-level variability in incidence time series, and to interpret these changes in terms of underlying transmission processes.
@@
 Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
-For example, a recently-developed method estimates the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual, directly from population-level incidence data~\citep{kirkegaard2021superspreading}.
+For example, a recently-developed method uses population-level incidence data to the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual~\citep{kirkegaard2021superspreading}.
+It is also possible to estimate the dispersion parameter from the distribution of the final size of a series of localized outbreaks~\citep{blumberg2013inference}.
 Clustering of cases has also been estimated directly from incidence data~\citep{schneckenreither2023assessing}.
 Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}.
-Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for the sensitivity of cities to forcing from exogenous drivers of transmission (e.g., climate) and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
+Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with implications on the sensitivity of cities to climate forcing and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
 
 What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
-Under a wide range of configurations for a branching process model of contagion spread the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
+Under a wide range of configurations for a branching process model of contagion spread, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
 The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$.
@@
 
-Since very large $\theta$ correspond to processes that are close to a Poisson process, the test will not produce a p-value if any of the three values are greater than 1/(1e-3). 
+Since very large $\theta$ correspond to processes that are operationally nearly identical to a Poisson process, the test will not produce a p-value if any of the three values are greater than a user-specified threshold.
+In the applicatio below we set this threshold at $10^3$, meaning that theta estimates with temporal crowding that was within $0.1\%$ of that expected for a Poisson process was considered effectively Poisson.
+
 Very large $\theta$ estimates are often produced in small populations when the process can't be distinguished from a Poisson process.

### 2025-01-24 ms-3/ms.tex
Complete up to application to simulated data

@@
 
-The form of the probability mass function for infections at a time step is:
+The form of the probability mass function (PMF) for infections at a time step is:
 \begin{equation}
@@
 
-We estimated $\theta_t$ given observed incidence using an iteratively reweighted least-squares (IRLS) procedure for mean estimation in conjunction with an optimization procedure to compute $\theta_t$.
-That is, for each time window, a series of $\mu_s$ from $s = t-\Delta$ to $s = t + \Delta$ was estimated using the spline and offset term via IRLS in the NBPSeq R package~\citep{di2015nbpseq}.
-Then, a single value of $\theta_t$ for the time window was computed via an optimization procedure.
+We estimate $\theta_t$ from observed incidence data using an iteratively reweighted least-squares (IRLS) procedure for mean estimation, combined with a maximum likelihood optimization to compute $\theta_t$.
+Specifically, within each time window, the spline model with an offset term was used to estimate a series of $\mu_s$ values for $s = t-\Delta$ to $s = t + \Delta$ via IRLS, as implemented in the \texttt{NBPSeq} R package~\citep{di2015nbpseq}.
+A single value of $\theta_t$ was then calculated for the entire time window by maximizing the likelihood function, which is based on the negative binomial probability mass function defined above.
 
-In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) that could be applied at each time step to test the hypothesis that $\theta$ has changed. 
-This involves fitting and comparing both a null model (no \begin{math}\theta\end{math} change) and a two-part ($\theta$-change) model. 
-For the null model, an optimization for a single $\theta$ value for the time window was performed, and for the $\theta$-change model, an optimization for a $\theta$ value during the first half of the window was performed, as well as an optimization for the second half.
+In addition to fitting the model at each time step, we developed a likelihood-ratio test (LRT) to test the hypothesis that $\theta$ has changed at each time step. 
+This test involves fitting and comparing two models: a null model (no $\theta$ change) and a two-part model (with a $\theta$ change). 
+For the null model, a single $\theta$ value was fitted for the entire time window. 
+For the $\theta$-change model, separate $\theta$ values were fitted for the left (from $t-\Delta$ to $t$) and right (from $t$ to $t + \Delta$) halves of the time window.
 
-Since very large $\theta$ correspond to processes that are operationally nearly identical to a Poisson process, the test will not produce a p-value if any of the three values are greater than a user-specified threshold.
-In the applicatio below we set this threshold at $10^3$, meaning that theta estimates with temporal crowding that was within $0.1\%$ of that expected for a Poisson process was considered effectively Poisson.
+Very large $\theta$ values correspond to processes that are operationally identical to a Poisson process. 
+Accordingly, the test does not produce a $p$-value if any of the three $\theta$ estimates exceed a user-specified threshold. 
+In the application below, we set this threshold at $10^3$, meaning that $\theta$ estimates with temporal crowding within $0.1\%$ of that expected for a Poisson process were considered effectively Poisson.
 
-Very large $\theta$ estimates are often produced in small populations when the process can't be distinguished from a Poisson process.
-Therefore, we have very small $\phi = 1/\theta$  values be considered zero, at which point we fail to reject the Poisson hypothesis.
+Similarly, values of $\theta$ very close to 0 focus all of the mass of the PMF on 0, representing a scenario where the probability of observing any infections approaches zero.
+As with the Poisson-like tolerance described in the previous paragraph, our algorithm does not produce a $p$-value if any of the three $\theta$ estimates are below a user-specified threshold.
+This threshold will depend on the presence of contiguous sections of the time series being analyzed during which no cases are observed.
+In the application below, we set this threshold to $10^{-3}$, because $\theta$ values below this level correspond to 0 frequencies that greatly exceed those in the data.
 
-In addition, we imposed restrictions on $\theta$ estimates such that $\theta = 0$ (point mass) results in no p-value being produced in the test result.
-So, large $\phi$ are considered values where we fail to reject the point mass hypothesis.
-
-Additionally, we included an option to retrieve the likelihood of a range of theta values in addition to these restrictions.
+With both upper and lower $\theta$ thresholds--—corresponding to Poisson-like and zero tolerances, respectively—--maximum likelihood estimates (MLEs) of $\theta$ beyond these thresholds exhibited unbounded behavior.
+When $\theta$ exceeded the upper threshold, corresponding to processes operationally identical to a Poisson process, the MLE tended to grow arbitrarily large, with the likelihood surface reaching its maximum at the upper boundary of the calculated domain.
+Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood surface peaked at the lower boundary of the domain.
+This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds
+(Fig. 1a)
 
@@
 Our simulation samples start points and theta values, and incidence is modeled according to overall incidence trends.
-
-To test the validity and power of the methods, we epidemic curves with known values of $\mu$ and $\theta$.
-We varied the magnitude of the $\theta$ change, location of the change in the curve, population size underlying the curve, and curve shape to test the validity and power of our approach across a range of scenarios.
-Epidemic curves were created from a deterministic skeleton (template) that either assumed a flat distribution of incidence over time or a Gaussian-shaped distribution, with parameters controlling the total epidemic size, peak time, peak width, and duration.
-Stochastic noise was then added to these templates by sampling from a negative binomial distribution with mean given by the template values.Each curve was split into two equal pieces and separate dispersion parameters $\theta_1$ and $\theta_2$ were used in each respective piece, before and after the breakpoint.
-This allowed us to simulate epidemic curves with variation in dispersion, as well as the special case when $\theta_1 = \theta_2$.
-
-\subsection*{Data availability and processing}
-We applied our method to cumulative COVID-19 case counts in United States (US) counties~\citep{nytimes_covid19}.
-Population sizes were from taken the US Census Bureau~\citep{us_census_county_population_2021}.
-Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
-Missing values for new cases were all at the beginning of the pandemic and were imputed as zero.
-Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
-We used a width paramter of $\delta = 8$ week and set the degrees of freedom parameter for the natural splines to $J = 3$.  
-[This seems like a wide window and a low-df spline.
-Why did we choose these and what happens if we use different values?
-Some sort of sensitivity analysis is probably needed, even if it is just to run the analysis using a smaller window and larger df]
-
-\section*{Results and Discussion}
-Our estimates of $\theta$ were robust across the population sizes included in the empirical analysis of U.S. counties. 
-However, inaccuracies were more common in counties with small populations outside the examined range.%(\nameref{S1_Fig}). 
-We found that the LRT method of testing for sharp changes in $\theta$ is also robust across population sizes represented in the empirical data (Fig\ 1 e, f). 
-Adequate test performance for the LRT is characterized by an average $p$-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.
-
-%\begin{figure}[!ht]
-%\includegraphics[width=1\textwidth]{fig1}
-%\caption{
-%Detecting changes in dispersion within incidence time series for populations of different sizes.
-(a, b) Simulated case counts with constant (a) and changing (b) dispersion parameters.
-(c, d) Dispersion parameters used to generate the data in (a) and (b), respectively.
-(e, f) LRT performance under varying absolute differences in \(\theta\) (horizontal axis), showing \( p \)-value distributions for population sizes of 50,000 (e) and 10,000,000 (f), representing the range in the empirical data.
-The null and alternative hypotheses (O and X) correspond to the parameters in (c) and (d).
-Red vertical lines indicate interquartile ranges.
-%}\label{fig1}
-%\end{figure}
 

### 2025-01-24 ms-3/ms.tex
Update ms.tex
optimization info

@@
 
-We estimate $\theta_t$ from observed incidence data using an iteratively reweighted least-squares (IRLS) procedure for mean estimation, combined with a maximum likelihood optimization to compute $\theta_t$.
+We estimate $\theta_t$ from observed incidence data using an iteratively reweighted least-squares (IRLS) procedure for mean estimation, combined with the optimize function in R, which uses a combination of golden section search and successive parabolic interpolation, to compute $\theta_t$.
 Specifically, within each time window, the spline model with an offset term was used to estimate a series of $\mu_s$ values for $s = t-\Delta$ to $s = t + \Delta$ via IRLS, as implemented in the \texttt{NBPSeq} R package~\citep{di2015nbpseq}.

### 2025-01-24 ms-3/ms.tex
Update ms.tex
Finish new app to sim data

@@
 
-Our simulation samples start points and theta values, and incidence is modeled according to overall incidence trends.
+To test the fidelity of our $\theta$ estimates and the validity and power of our LRT, we simulated epidemic curves spanning 16 time steps each.
+Our simulation framework samples the start point of each curve and a $\theta$ value on each side of the midpoint.
+Incidence modeled according to overall incidence trends ((Fig\ 1 a) is scaled by the population size for that curve (population sizes used ranged from $10^3$ to $10^7$).
+Equal $\theta$ values on both sides of the midpoint were set for 10% of the epidemic curves.
+The estimation framework and LRT was applied to each epidemic curve in the simulated data in order to produce estimates of $\theta$ on either side of the midpoint, as well as a p-value from the LRT.
+This approach allowed us to test the robustness of our framework over a range of $\theta$ difference magnitudes, underlying population sizes, and point of the epidemic curve at which the procedure is applied.
 

### 2025-01-24 ms-3/ms.tex
Update ms.tex
Transfer relevant results and discussion section through end of new Fig 1.

@@
 The estimation framework and LRT was applied to each epidemic curve in the simulated data in order to produce estimates of $\theta$ on either side of the midpoint, as well as a p-value from the LRT.
-This approach allowed us to test the robustness of our framework over a range of $\theta$ difference magnitudes, underlying population sizes, and point of the epidemic curve at which the procedure is applied.
+This approach allowed us to test the robustness of our framework over a range of $\theta$ difference magnitudes, underlying population sizes, and point of the epidemic curve at which the procedure was applied.
+
+\section*{Results and Discussion}
+Our LRT framework is robust across population sizes.
+That is, when the effect size is zero, the p-values are on average 0.5 and as the effect size increases, the p-values shrink toward zero (Fig\ 1 d).
+
+Additionally, the $\theta$ estimates produced are close to the true values used in simulations (Fig\ 1 c), regardless of the underlying population size.
+The estimates that are not close correspond to instances when the boundary is reached.
+
+Need to discuss Figure 1 b somewhere.
+
+\begin{figure}[!h]
+\includegraphics[width=1\textwidth]{fig1}
+\caption{
+Detecting dispersion changes in case count time series.
+a: Sampled time period (red line) from overall US COVID-19 incidence per 1,000 people between 2020-01-04 and 2023-03-18.
+b: Cases per 1,000 people in Douglas County, NE over the sample time period with estimated mean values (red) and estimated dispersion values on either side of the midpoint.
+c: Estimated $\theta$ v.
+true $\theta$  with values outside of tolerance plotted in purple (close to Poisson) and blue (close to collapsing to zero).
+d: p-value versus absolute difference in $log_{10}\theta$.
+Red curve and red shading depict ***, respectively.
+ }
+\label{fig1}
+\end{figure}
 

### 2025-01-24 manuscript/ms (2).tex
Update ms (2).tex
Update methods

@@
 
-We estimated \begin{math}\mu_t\end{math} and \begin{math}\theta_t\end{math} using iteratively reweighted least-squares (procedure implemented via the NBPSeq R package \cite{NBPSeq}) using a window around each time step.
-For each window, a series of \begin{math}\mu_t\end{math} was estimated using a spline function in time, and a single value of \begin{math}\theta\end{math} was estimated for that window. 
-By moving the window one time step at a time, a time series of \begin{math}\theta_t\end{math} was produced.
-We investigated large counties (the three counties with the largest populations in each state), due to power constraints and convergence issues.
+For each time step in each empirical epidemic curve, a series of \begin{math}\mu_t\end{math} and a single value of \begin{math}\theta\end{math} was estimated for the corresponding window. 
+By moving the window one time step at a time, time series of \begin{math}\theta\end{math} estimates and p-values were produced.
+We investigated large counties (the three counties with the largest populations in each state).
 

### 2025-01-24 ms-3/ms.tex
Update ms.tex
Last of methods transfer

@@
 
+\subsection*{Data availability and processing}
+
+COVID-19 cumulative case data are from [The New York Times](https://github.com/nytimes/covid-19-data/) and county population sizes are from [The United States Census Bureau](https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/totals/.
+Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases.
+Note that missing values for new cases were all at the beginning of the pandemic and were imputed as zero.
+Apprxoimately 0.24% of new cases were negative due to corrections in the cumulative data.
+These were also imputed as zero.
+
+\subsection*{Application to empirical data}
+
+For each time step in each empirical epidemic curve, a series of \begin{math}\mu_t\end{math} and a single value of \begin{math}\theta\end{math} was estimated for the corresponding window. 
+By moving the window one time step at a time, time series of \begin{math}\theta\end{math} estimates and p-values were produced.
+We investigated large counties (the three counties with the largest populations in each state).
+
 \section*{Results and Discussion}

### 2025-01-26 ms-3/ms.tex
Complete up to Figure 2

--- 2025-01-24 ms-3/ms.tex
@@
 Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
+In applications this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
 However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.
@@
 \subsection*{Application to simulated data}
+We evaluated the robustness of our framework to a range of population sizes, magnitudes of dispersion changes, and shapes of underlying incidence trends by generating 2,000 simulated epidemic curves with known parameters.
+Epidemic trends were modeled as smoothed incidence series derived from 16-week sections randomly selected from US COVID-19 data (described below), scaled to reflect different population sizes ranging from $10^3$ to $10^7$.
+For each simulated trajectory, dispersion parameters ($\theta_1$ and $\theta_2$) were assigned to the two halves of the selected 16-week window, and case counts were simulated using a negative binomial distribution, where the mean ($\mu$) was based on the smoothed incidence trend scaled by the population size.
+The values of $\theta_1$ and $\theta_2$ were drawn from a uniform distribution spanning $10^{-2}$ to $10^{2}$, with 10\% of simulations set to have no change in dispersion ($\theta_1 = \theta_2$). 
+Extremely large differences in dispersion (absolute log-ratio $>3$) were capped by setting $\theta_2 = \theta_1$.
 
-To test the fidelity of our $\theta$ estimates and the validity and power of our LRT, we simulated epidemic curves spanning 16 time steps each.
-Our simulation framework samples the start point of each curve and a $\theta$ value on each side of the midpoint.
-Incidence modeled according to overall incidence trends ((Fig\ 1 a) is scaled by the population size for that curve (population sizes used ranged from $10^3$ to $10^7$).
-Equal $\theta$ values on both sides of the midpoint were set for 10% of the epidemic curves.
-The estimation framework and LRT was applied to each epidemic curve in the simulated data in order to produce estimates of $\theta$ on either side of the midpoint, as well as a p-value from the LRT.
-This approach allowed us to test the robustness of our framework over a range of $\theta$ difference magnitudes, underlying population sizes, and point of the epidemic curve at which the procedure is applied.
+\subsection*{Application to empirical data}
+We applied our framework to COVID-19 case data for the United States at the administrative level of counties, compiled by The New York Times, based on reports from state and local health agencies between Jan 4, 2020, and March 18, 2023~\citep{nytimes_covid19_2021}, and using county population sizes estimated for 2021 from the United States Census Bureau~\citep{us_census_county_estimates_2021}.
+Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differencing to compute new cases.
+Occasionally, reported cumulative case counts were not monotonically increasing due to corrections posted by local agencies as they resolved incoming data.
+As a result, approximately 0.24\% of estimated new cases across all counties in the dataset were negative and these were set to zero. 
+For each county, we analyzed overlapping 16-week windows, shifting one week (i.e., one timestep) at a time. 
+Within each window, the framework estimated the dispersion parameter ($\theta$) using a natural spline with three degrees of freedom to model the broad-scale trend in incidence. 
+Outputs included estimated dispersion parameters ($\theta_1$, $\theta_2$, for the left and right halves of each window, and $\theta$ for the entire window), likelihood ratio test statistics, $p$-values for changes in dispersion at the midpoint of the window, and flags for boundary conditions such as failure to reject Poisson-like dispersion or collapse to extreme overdispersion.
+
+\section*{Results and Discussion}
+The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and shrinking toward zero as the effect size increases (Fig.~\ref{fig1}).
+The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$.
+
+\clearpage
+\begin{figure}[!ht]
+\includegraphics[width=1\textwidth]{fig1}
+\caption{
+Detecting dispersion changes in case count time series. 
+a: Weekly incidence of COVID-19 in the United States, with time measured in weeks since January 4, 2020, showing an example of a randomly-selected 16 week period used as an incidence trend when in simulation-based validation of the LRT test (red).
+b: Cases in in one county (Douglas County, Nebraska) over the sample time period with estimated incidence trend (red) and estimated dispersion values on either side of the midpoint. 
+c: Estimated $\theta$ versus true $\theta$ in simulation studies combining a randomly-selected section of the national incidence curve with a random population size and set of dispersion values.
+Estimated values outside of tolerance plotted in purple (close to Poisson) and blue (close to collapsing to zero).
+d: Statistical power of the LRT test.}\label{fig1}
+\end{figure}
+\clearpage
+
+Paragraph on results shown in figure 2.
+
+\clearpage
+\begin{figure}[!ht]
+\includegraphics[width=1\textwidth]{fig2}
+\caption{Dispersion analysis of weekly COVID-19 case data for one county. 
+a: Weekly incidence in Jefferson County, Alabama. 
+b: Estimated dispersion parameter over time. 
+c: Comparison of estimated dispersion (grey) with predicted values under the standard model of $\theta_{t+1} = C_{t} / \rho_t$ where $C_t$ represents reported cases and $\rho_t$ the reporting rate, for constant values of $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue).
+This wide range of constant values is intended to enclose the $theta$ values that would be predicted by the standard model under under variable $\rho$.
+d: Likelihood ratio test (LRT) statistic over time, highlighting statistically large changes in dispersion (red), defined as p-values less than the Bonferroni-corrected 5\% quantile of a chi-square random variable with one degree of freedom.}\label{fig2}
+\end{figure}
+\clearpage
 
@@
 
+\clearpage
 \section*{References}
-\bibliographystyle{vancouver} % Use plainnat or another appropriate style
-\bibliography{references} % Create a references.bib file
+\bibliographystyle{vancouver} 
+\bibliography{references}
 

### 2025-01-26 code/3-figure1.R

@@ -9,6 +9,8 @@ ww <- 8
 ptol <- 1e-4
 ztol <- 1e4
 nsim <- 2 * 10^3
+poprange <- c(4, 7)   # range of population sizes, log10 scale
+thetamag <- 2         # largest value of abs(log10(theta))
 
 
 # Load data and functions
@@ -67,9 +69,9 @@ sim <- function(theta1, theta2, s, a = NULL, b = NULL) {
 
 # Apply
 simdata <- data.frame(
-  s = 10^runif(nsim, 3, 7),
-  theta1 = 10^runif(nsim, -2.5, 2.5),
-  theta2 = 10^runif(nsim, -2.5, 2.5)
+  s = 10^runif(nsim, poprange[1], poprange[2]),
+  theta1 = 10^runif(nsim, -thetamag, thetamag),
+  theta2 = 10^runif(nsim, -thetamag, thetamag)
 )
 
 i <- sample.int(nsim, ceiling(0.1 * nsim)) # set some theta2 = theta1
@@ -118,7 +120,8 @@ plot(incidence_smooth * 1000,
   type = "l",
   lwd = 3,
   xlab = "Week",
-  ylab = "Cases per 1000"
+  ylab = "Cases per 1000",
+  main = "a"
 )
 
 u <- 25:(144 - 2 * ww)
@@ -153,7 +156,8 @@ ylim[1] <- 0
 plot(a:b, cases_snippet * 1e-3,
   xlab = "Week",
   ylab = "Cases per 1000",
-  ylim = ylim
+  ylim = ylim,
+  main = "b"
 )
 
 lines(a:b, y * 1e-3, col = 2, lwd = 3)
@@ -232,10 +236,10 @@ plot(
   y = log10(simdata$theta1_est),
   xlab = expression(theta[1]),
   ylab = expression(hat(theta)[1]),
-  cex = 1,
   col = col,
   xaxt = "n",
   yaxt = "n",
+  main = "c"
 )
 axis(1, at = c(-2, 0, 2), labels = expression(10^-2, 10^0, 10^2))
 axis(2,
@@ -258,7 +262,8 @@ plot(
   xlab = expression(abs(log[10] ~ (theta[2] / theta[1]))),
   ylab = "p",
   xaxt = "n",
-  yaxt = "n"
+  yaxt = "n",
+  main = "d"
 )
 axis(1, at = c(0, 1, 2, 3, 4, 5))
 axis(2,

### 2025-01-26 code/4-figure2.R

@@ -8,10 +8,8 @@ ww <- 8 # window half-width
 
 # Load data
 load("data/processed/nyt_weekly.Rdata")
-load("data/processed/lrt_lg_pops.Rdata")
 load("data/processed/theta_lg_pops.Rdata")
-load("data/processed/ftr_poiss_lg_pops.Rdata")
-load("data/processed/ctzs_lg_pops.Rdata")
+
 
 filename <- "figures/fig2.pdf"
 pdf(filename, width = 6, height = 6)
@@ -81,8 +79,10 @@ lines(dates, log10(theta0b), col = "blue")
 lines(dates, log10(thetas), col = "grey")
 mtext("c", side = 3, line = 1, adj = 0, cex = 1.3)
 
-legend("topleft", legend=c("Reporting rate 0.1", "Reporting rate 0.9", "Estimated"),
-       col=c("black", "blue", "grey"), cex=0.7, lty = 1, bt = "n")
+legend("topleft",
+  legend = c("Reporting rate 0.1", "Reporting rate 0.9", "Estimated"),
+  col = c("black", "blue", "grey"), cex = 0.7, lty = 1, bt = "n"
+)
 
 # d
 is_sig <- lrt_stats[1, ] > qchisq(0.9996732, df = 1)
@@ -97,7 +97,12 @@ plot(dates, lrt_stats[target_county, ],
 )
 mtext("d", side = 3, line = 1, adj = 0, cex = 1.3)
 
-legend("topleft", legend=c(expression("X ">chi[1-alpha/153]^2), expression("X "<=chi[1-alpha/153]^2)),
-       col=c("red", "black"), cex=0.7, pch = c(19, 1), bt = "n")
+legend("topleft",
+  legend = c(
+    expression("X " > chi[1 - alpha / 153]^2),
+    expression("X " <= chi[1 - alpha / 153]^2)
+  ),
+  col = c("red", "black"), cex = 0.7, pch = c(19, 1), bt = "n"
+)
 
 dev.off()

### 2025-01-27 ms-3/ms.tex
Update ms.tex

@@
 
-\begin{abstract}
+\begin{Abstract}
 Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
@@
 
-An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information).
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text).
 For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
@@
 The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
-Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
+Explanation of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
 

### 2025-01-28 ms-3/ms.tex
RA all good up through start of Methods

@@
 \section*{Author summary}
-Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series.
+Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health policy.
+Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
 Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
@@
 Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
-Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity, and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
+Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
 Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
@@
 Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
-For example, a recently-developed method uses population-level incidence data to the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual~\citep{kirkegaard2021superspreading}.
-It is also possible to estimate the dispersion parameter from the distribution of the final size of a series of localized outbreaks~\citep{blumberg2013inference}.
+For example, a recently-developed method uses population-level incidence data to estimate the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual~\citep{kirkegaard2021superspreading}.
+It is also possible to estimate the dispersion parameter of the offspring distribution from the distribution of the final size of a series of localized outbreaks~\citep{blumberg2013inference}.
 Clustering of cases has also been estimated directly from incidence data~\citep{schneckenreither2023assessing}.
 Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}.
-Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with implications on the sensitivity of cities to climate forcing and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
+Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with implications for the sensitivity of cities to climate forcing and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
 
@@
 For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
-In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
+In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) as expected for a Poisson process with the same incidence rate.
 
@@
 Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
-In applications this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
+In applications, this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
 However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.
 
-In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach, which makes data-driven estimates of $\theta_t$, including identifying timepoints when $\theta$ is changing rapidly, which may help to reveal the impacts of heterogeneity in transmission, susceptibility, and reporting.
+In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach which produces data-driven estimates of $\theta_t$, including identifying timepoints when $\theta$ is changing rapidly, which may help to reveal the impacts of heterogeneity in transmission, susceptibility, and reporting.
 

### 2025-01-29 ms-3/ms.tex
Update ms.tex

@@
 By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
-Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
-To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
+Consequently, in order to estimate incidence volatility we first modeled incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
+To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling was conducted within a moving window
 \begin{equation}

### 2025-01-29 ms-3/ms.tex
Draft of para on Figure 2

--- 2025-01-27 ms-3/ms.tex
@@
 
-\begin{Abstract}
+\begin{abstract}
 Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
@@
 
-An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text).
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information).
 For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
@@
 The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
-Explanation of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
+Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
 
@@
 \section*{Results and Discussion}
-The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and shrinking toward zero as the effect size increases (Fig.~\ref{fig1}).
-The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$.
+The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and decreasibg toward zero as the effect size increases (Fig.~\ref{fig1}).
+The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$, which encompasses all operationally relevant values of $\theta$ for COVID-19 incidence data and many other infectious diseases: values lower than this range would focus the PMF for cases almost entirely on 0 while values higher than this range are effectively Poisson.
 
@@
 
-Paragraph on results shown in figure 2.
+Applying the method to COVID-19 cases in US counties enabled investigation of changes in the $\theta$ in relation to both observed trends and model expectations (Figure 2). 
+Periods of increased case count variability corresponded with decreases in the dispersion parameter, indicating that $\log_{10}\theta$ was dynamic and displayed both expected and unexpected patterns of variation.
+In some instances, $\theta$ appeared to vary inversely with incidence, aligning with standard epidemic theory, while in other periods, deviations from that expectation occurred, potentially highlighting shifts in the underlying transmission dynamics.
+Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, coinciding with heightened volatility in reported cases, and at the end of the time series, during the approach to endemicity as the landscape of susceptibility was changing and potentially becoming more complex~\citep{lavine2021immunological}.
+In other words, as a larger portion of incidence involved repeat infections (albiet potentially with different variants), prior immunity became a larger factor shaping epidemic dynamics, and potentially impacting the demography of burden, in particular the burden as it depends on age, and potentially particularly in children~\citep{lavine2021immunological}.
+These findings underscore the complex behavior of the dispersion parameter, which both varied with changes in case count regimes and highlighted departures from the model expectations described by Eq.
+1.
 
@@
 \caption{Dispersion analysis of weekly COVID-19 case data for one county. 
+Analysis for all counties together is shown in Fig 3.
 a: Weekly incidence in Jefferson County, Alabama. 

### 2025-01-30 ms-3/ms.tex
Update results and discussion

@@
 \section*{Results and Discussion}
-The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and decreasibg toward zero as the effect size increases (Fig.~\ref{fig1}).
-The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$, which encompasses all operationally relevant values of $\theta$ for COVID-19 incidence data and many other infectious diseases: values lower than this range would focus the PMF for cases almost entirely on 0 while values higher than this range are effectively Poisson.
+Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}).
+The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million.
+Across this range, the method produced accurate estimates for $\theta$ within $10^{-2} \leq \theta \leq 10^2$, encompassing all operationally relevant values for COVID-19 incidence data and many other infectious diseases.
+Lower values would concentrate the probability mass function (PMF) for cases almost entirely on 0, while higher values effectively correspond to a Poisson distribution.
 
@@
 
-Applying the method to COVID-19 cases in US counties enabled investigation of changes in the $\theta$ in relation to both observed trends and model expectations (Figure 2). 
-Periods of increased case count variability corresponded with decreases in the dispersion parameter, indicating that $\log_{10}\theta$ was dynamic and displayed both expected and unexpected patterns of variation.
-In some instances, $\theta$ appeared to vary inversely with incidence, aligning with standard epidemic theory, while in other periods, deviations from that expectation occurred, potentially highlighting shifts in the underlying transmission dynamics.
-Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, coinciding with heightened volatility in reported cases, and at the end of the time series, during the approach to endemicity as the landscape of susceptibility was changing and potentially becoming more complex~\citep{lavine2021immunological}.
-In other words, as a larger portion of incidence involved repeat infections (albiet potentially with different variants), prior immunity became a larger factor shaping epidemic dynamics, and potentially impacting the demography of burden, in particular the burden as it depends on age, and potentially particularly in children~\citep{lavine2021immunological}.
-These findings underscore the complex behavior of the dispersion parameter, which both varied with changes in case count regimes and highlighted departures from the model expectations described by Eq.
-1.
+Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations (Figure~\ref{fig2}).  
+Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
+Changes in dispersion exhibited both expected and unexpected patterns of variation relative to standard theory.  
+In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.
+Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.
+The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
+These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.
+
+The next paragraph should look across all counties and reference Fig 3.
 
@@
 \includegraphics[width=1\textwidth]{fig2}
-\caption{Dispersion analysis of weekly COVID-19 case data for one county. 
-Analysis for all counties together is shown in Fig 3.
-a: Weekly incidence in Jefferson County, Alabama. 
-b: Estimated dispersion parameter over time. 
-c: Comparison of estimated dispersion (grey) with predicted values under the standard model of $\theta_{t+1} = C_{t} / \rho_t$ where $C_t$ represents reported cases and $\rho_t$ the reporting rate, for constant values of $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue).
-This wide range of constant values is intended to enclose the $theta$ values that would be predicted by the standard model under under variable $\rho$.
-d: Likelihood ratio test (LRT) statistic over time, highlighting statistically large changes in dispersion (red), defined as p-values less than the Bonferroni-corrected 5\% quantile of a chi-square random variable with one degree of freedom.}\label{fig2}
+\caption{Dispersion analysis of weekly COVID-19 case data for Jefferson County, Alabama. 
+Results for all counties are shown in Figure 3  %.~\ref{fig3}.  
+a: Weekly reported COVID-19 incidence.  
+b: Estimated dispersion parameter ($\theta$) over time.  
+c: Comparison of estimated dispersion (gray) with predicted values from the standard model $\theta_{t+1} = C_t / \rho_t$, where $C_t$ is reported cases and $\rho_t$ is the reporting rate.
+Predictions are shown for fixed $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue), chosen to encompass the range of $\theta$ expected under variable $\rho$.
+d: Likelihood ratio test (LRT) statistic over time.
+Statistically significant changes in dispersion (red) correspond to p-values below the Bonferroni-corrected 5\% threshold of a chi-square distribution with one degree of freedom.
+}\label{fig2}
 \end{figure}

### 2025-01-30 ms-3/ms.tex
Update ms.tex

--- 2025-01-29 ms-3/ms.tex
@@
 Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
-To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
+To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling is conducted in a moving window such that for each half of the window
 \begin{equation}
@@
 \end{equation}
-where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$.
-The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
-Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
+where $N$ represents population size, $h_j(t)$ are basis functions, the degrees of freedom is equal to the number of knots $k$ for the natural spline, $J = k + d +1$, where $d$ is the degree of the polynomial, and $\beta_j^{(t)}$ are fitted parameters.
+The window has half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$.
+The degrees of freedom (number of knots) to be used for the natural splines, and the width of the moving window will depend on the application.
+Explaination of the specific choices we used for our application to COVID-19 cases in US counties is described below.
 
@@
 With both upper and lower $\theta$ thresholds--—corresponding to Poisson-like and zero tolerances, respectively—--maximum likelihood estimates (MLEs) of $\theta$ beyond these thresholds exhibited unbounded behavior.
-When $\theta$ exceeded the upper threshold, corresponding to processes operationally identical to a Poisson process, the MLE tended to grow arbitrarily large, with the likelihood surface reaching its maximum at the upper boundary of the calculated domain.
-Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood surface peaked at the lower boundary of the domain.
-This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds
-(Fig. 1a)
+When $\theta$ exceeded the upper threshold, corresponding to processes operationally identical to a Poisson process, the MLE tended to grow arbitrarily large, with the likelihood function reaching its maximum at the upper boundary of the calculated domain.
+Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood function peaked at the lower boundary of the domain.
+This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig.
+1c).
 
@@
 For each county, we analyzed overlapping 16-week windows, shifting one week (i.e., one timestep) at a time. 
-Within each window, the framework estimated the dispersion parameter ($\theta$) using a natural spline with three degrees of freedom to model the broad-scale trend in incidence. 
+Within each window, the framework estimated the dispersion parameter ($\theta$) using a natural spline with three degrees of freedom for each half of the window to model the broad-scale trend in incidence.
 Outputs included estimated dispersion parameters ($\theta_1$, $\theta_2$, for the left and right halves of each window, and $\theta$ for the entire window), likelihood ratio test statistics, $p$-values for changes in dispersion at the midpoint of the window, and flags for boundary conditions such as failure to reject Poisson-like dispersion or collapse to extreme overdispersion.
@@
 \section*{Results and Discussion}
-The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and decreasibg toward zero as the effect size increases (Fig.~\ref{fig1}).
+The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and decreasing toward zero as the effect size increases (Fig.
+1d ~\ref{fig1}).
 The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$, which encompasses all operationally relevant values of $\theta$ for COVID-19 incidence data and many other infectious diseases: values lower than this range would focus the PMF for cases almost entirely on 0 while values higher than this range are effectively Poisson.
@@
 Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, coinciding with heightened volatility in reported cases, and at the end of the time series, during the approach to endemicity as the landscape of susceptibility was changing and potentially becoming more complex~\citep{lavine2021immunological}.
-In other words, as a larger portion of incidence involved repeat infections (albiet potentially with different variants), prior immunity became a larger factor shaping epidemic dynamics, and potentially impacting the demography of burden, in particular the burden as it depends on age, and potentially particularly in children~\citep{lavine2021immunological}.
+In other words, as a larger portion of incidence involved repeat infections (albeit potentially with different variants), prior immunity became a larger factor shaping epidemic dynamics, and potentially impacting the demography of burden, in particular the burden as it depends on age, and potentially particularly in children~\citep{lavine2021immunological}.
 These findings underscore the complex behavior of the dispersion parameter, which both varied with changes in case count regimes and highlighted departures from the model expectations described by Eq.

### 2025-01-31 ms-3/ms.tex

--- 2025-01-30 ms-3/ms.tex
@@
 \section*{Results and Discussion}
+
 Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}).

### 2025-01-31 ms-3/ms.tex
Update ms.tex

@@
 
-The next paragraph should look across all counties and reference Fig 3.
+Highly overdispersed patterns were observed for many of the examined counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
+In addition, there are increases in dispersion around the peaks in incidence (Figure~\ref{fig3}).
+Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig\ 3 f).
+Note that these p-values should be corrected for multiple testing when used for inference instead of visualization.
+Raising variance relative to mean (higher dispersion) implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
+It appears that there are more surges on the way up to, and on the way down from, peak incidence, some of which departs from the dispersion expected under standard epidemic theory. 
+This indicates less diffuse epidemic dynamics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, and increased locally experienced mean density \cite{lloyd_mean_1967} at unexpected points in the COVID-19 pandemic.
 

### 2025-01-31 ms-3/ms.tex
Update ms.tex
Standardize in-text figure references.

@@
 Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood function peaked at the lower boundary of the domain.
-This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig.
-1c).
+This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig.~\ref{fig1}).
 
@@
 In addition, there are increases in dispersion around the peaks in incidence (Figure~\ref{fig3}).
-Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Fig\ 3 f).
+Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Figure~\ref{fig3}).
 Note that these p-values should be corrected for multiple testing when used for inference instead of visualization.

### 2025-02-05 ms-3/ms.tex

@@
 
-An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information).
+An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text).
 For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
-In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.
+In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) as expected for a Poisson process with the same incidence rate.
 
@@
 By definition incidence volatility is fast relative to broadscale epidemic dynamics.
-Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
-To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling is conducted in a moving window such that for each half of the window
+
+Consequently, in order to estimate incidence volatility we first modeled incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}.
+To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling was conducted within a moving window such that for each half of the window
 \begin{equation}
@@
 The window has half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$.
-The degrees of freedom (number of knots) to be used for the natural splines, and the width of the moving window will depend on the application.
-Explaination of the specific choices we used for our application to COVID-19 cases in US counties is described below.
+The degrees of freedom (number of knots) to be used for the splines, and the width of the moving window will depend on the application.
+Explanation of the specific choices we used for our application to COVID-19 cases in US counties is described below.
 

### 2025-02-07 ms-3/ms.tex
Update manuscript

@@
 
-Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations (Figure~\ref{fig2}).  
+Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Figure~\ref{fig2}).
 Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
@@
 In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.
-Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.
+Notably, significant changes in the dispersion parameter were observed during major epidemic transitions.
+For example, during the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.
 The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
 These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.
-
-Highly overdispersed patterns were observed for many of the examined counties more frequently later in time series, consistent with more heterogeneity in transmission, susceptibility and reporting. 
-In addition, there are increases in dispersion around the peaks in incidence (Figure~\ref{fig3}).
-Evidence for a change in \begin{math}\theta\end{math} was observed across many counties (evidenced by a concentration of low p-values around peak incidence) (Figure~\ref{fig3}).
-Note that these p-values should be corrected for multiple testing when used for inference instead of visualization.
-Raising variance relative to mean (higher dispersion) implies spatiotemporal "crowding" of cases (i.e. localized surges) which may necessitate more surge capacity in hospitals and testing centers. 
-It appears that there are more surges on the way up to, and on the way down from, peak incidence, some of which departs from the dispersion expected under standard epidemic theory. 
-This indicates less diffuse epidemic dynamics that are potentially more subject to climate forcing \cite{dalziel_urbanization_2018}, and increased locally experienced mean density \cite{lloyd_mean_1967} at unexpected points in the COVID-19 pandemic.
 
@@
 
+Dispersion increased markedly around the peaks in incidence during the major 2022 wave, from late December 2021 to early February 2022 (Figure~\ref{fig3}).
+This is in strong contrast to standard epidemic theory which predicts that dispersion should decrease as incidence rises.
+A high concentration of low $p$-values around peak incidence (Figure~\ref{fig3}) corroborates widespread changes in $\theta$ across counties, reinforcing the statistical significance of this pattern.
+While these $p$-values should be corrected for multiple testing if used for inference rather than visualization, the overall trend suggests a systematic departure from theoretical expectations.
+Highly overdispersed patterns were also observed more frequently later in the time series, pointing to increasing heterogeneity in transmission, susceptibility, and reporting during the later phases of the pandemic.
+In both the 2022 wave and later in the pandemic, localized surges indicated by higher dispersion may have played a larger role in pandemic dynamics than expected, including potentially placing increased demand for surge capacity in hospitals and testing centers.
+
+\clearpage
+\begin{figure}[!ht]
+\includegraphics[width=1\textwidth]{fig3}
+\caption{caption.}\label{fig3}
+\end{figure}
+\clearpage
+
 \section*{Acknowledgments}

### 2025-02-07 ms-3/ms.tex
Update ms.tex
Added fig.3 caption with updated final discussion points.

@@
 \includegraphics[width=1\textwidth]{fig3}
-\caption{caption.}\label{fig3}
+\caption{Incidence and dispersion between 2020-01-04 and 2023-03-18 in large counties in the US. 
+a: Mean COVID-19 cases of the 144 US counties over time. 
+b: Mean $log_{10}\theta$ of the 144 US counties over time.  
+c: $log_{10}(cases)$ over time for each of the 144 counties (y-axis).
+d: $log_{10}\theta$ over time for each of the 144 counties (y-axis).
+e: Expected value of $log_{10}\theta_0$ under the null model in Equation 1 ($log_{10}(cases)$ - log_{10}(reporting rate)$) over time for each of the 144 counties (y-axis).
+f: LRT p-values over time for each of the 144 counties (y-axis).}
+\label{fig3}
 \end{figure}
 \clearpage
+
+We presented an approach to quantify clustering of cases in epidemic time series that is less susceptible to detecting artifacts based on population size and incidence. 
+While our estimation framework facilitates comparison of dispersion between counties and also over time in a given county, our LRT framework additionally allows for detection of changing dispersion. 
+Our method forms part of a larger interest in investigating variability in infections as an important attribute of epidemic time series using novel metrics.
+For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window \cite{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
+Spatial variation in superspreading potential has been investigated through risk maps of superspreading environments \cite{loo_identification_2021}, so future work could investigate the correspondence between our dispersion metric and indicators of a high risk of superspreading.
+Methods that use incidence time series are crucial due to the ease of obtaining this type of data, so the timing and geographical allocation of public health resources can be achieved with limited resources.
+Additionally, population-wide disease control approaches are often less effective than those which are targeted to individuals in high-transmission contexts \cite{lloyd2005superspreading}, so identifying candidate time periods when transmission heterogeneity is high may catalyze the development of more efficient control strategies.
+
+Application of the methods we developed to empirical case count time series spanning from the beginning of 2020 to early 2023 revealed distinct increases in dispersion both near the end of the time series considered and near peaks in COVID-19 incidence.
+These results imply that dispersion is high \emph{near} peak incidence, suggesting that high dispersion is not simply associated with low incidence. 
+Observed departures from the level of dispersion expected under the model represented in Eq 1 represent time periods in which the assumptions of the null model may be violated and cases may be highly clustered in time.
+
+Previous research to evaluate bursty dynamics based on Influenza-like Illness (ILI) times series showed that epidemics in smaller communities were concentrated on narrower windows of the influenza season; the proportion of disease incidence that occurred in a given week was a metric of interest \cite{dalziel2018urbanization}.
+Though large cities may be subject to more "smooth" epidemic dynamics, our contribution highlights the circumstances under which dynamics are less smooth in large populations.
+
+Since there are regimes of the COVID-19 epidemic that are subject to increased dispersion across large counties, these may be candidate time periods for rolling out extra surge capacity in large counties.
+However, further research is needed to understand whether this phenomenon occurs in smaller counties and for other pathogens.
 

### 2025-02-13 ms-3/ms.tex
Complete figure 3 and writing

--- 2025-02-07 ms-3/ms.tex
@@
 \includegraphics[width=1\textwidth]{fig3}
-\caption{caption.}\label{fig3}
+\caption{Incidence and dispersion between Jan 4, 2020 and March 18, 2023, in large counties in the US.
+a: Mean COVID-19 cases of the 144 US counties over time. 
+b: Mean $\log_{10}\theta$ of the 144 US counties over time.  
+c: $\log_{10}(cases)$ over time for each of the 144 counties.
+d: $\log_{10}\theta$ over time for each of the 144 counties.
+e: Expected value of $\log_{10}\theta_0$ under the null model, assuming a reporting rate of 0.5.
+f: LRT p-values over time for each location.}\label{fig3}
 \end{figure}
 \clearpage
+
+Our method forms part of a larger interest in investigating variability in infections as an important attribute of epidemic time series using novel metrics.
+For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window~\citep{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
+Spatial variation in superspreading potential has been investigated through risk maps of superspreading environments~\citep{loo_identification_2021}, and future work could investigate the correspondence between dispersion in case count time series, as quantified here, and indicators of a high risk of superspreading, with the potential to further illucidate drivers of transmission risk across scales, and more finely resolve landscapes of susceptibility.
+Additionally, as population-wide disease control may be less effective than those which are focussed to individuals in high-transmission contexts~\citep{lloyd2005superspreading}, identifying candidate time periods when transmission heterogeneity is high may catalyze the development of more effective control strategies, particularly those that connect vulnerable populations with resources at critical times.
+The finding that dispersion increased rather than decreased during the 2022 surge challenges theoretical expectations and suggests that fundamental assumptions about the scaling of transmission dynamics may require reevaluation.
+One hypothesis is that transmission heterogeneity could play a role in driving large surges, amplifying incidence beyond what homogeneous models predict. 
+Future work could investigate whether bursts of highly clustered transmission events generate feedback that accelerates epidemic spread, which, if true, could refine predictive models of contagion dynamics in complex populations.
 

### 2025-02-13 code/6-figure3.R

@@ -20,52 +20,73 @@ cases <- cases[, keep]
 incidence <- incidence[, keep]
 
 
+# Define common x-axis ticks and labels
+start_date_label <- as.Date("2020-07-01")
+end_date_label <- as.Date("2022-07-01")
+at <- seq.Date(
+  from = start_date_label,
+  to = end_date_label,
+  by = "year"
+)
+labels <- format(at, format = "%b %y") 
+
+
+
 # Plot
 filename <- "figures/fig3.pdf"
 pdf(filename)
 par(mfrow = c(3, 2))
 
-# a. Mean cases
+# Cases time series
 par(mar = c(4, 4, 3, 6))
 
-plot(dates, colMeans(cases),
-  type = "l", lwd = 2, col = "black", xlab = "Date",
-  ylab = "", main = "Mean cases", cex.main = 1.7, cex.lab = 1.4
+plot(dates, colSums(cases) / sum(pops) * 1000,
+  type = "l", lwd = 2, col = "black", xlab = "",
+  xaxt = "n",
+  ylab = "",
+  ylim = c(0, 20),
+  main = "Cases per 1000 pop",
+  cex.main = 1.7, cex.lab = 1.4
 )
 
 par(xpd = TRUE)
 mtext("a", side = 3, line = 1, adj = 0, cex = 1.5)
 par(xpd = FALSE)
 
+axis(1, at = at, labels = labels, cex.axis = 1.3)
 
-# b. Mean log10(theta)
+
+# Mean theta time series
 par(mar = c(4, 4, 3, 6))
 
-plot(dates, colMeans(log10(thetas), na.rm = TRUE),
-  type = "l", lwd = 2, col = "black", xlab = "Date",
+plot(dates, log10(colMeans(thetas, na.rm = TRUE)),
+  type = "l", lwd = 2, col = "black", xlab = "",
   ylab = "",
   main = expression(bold("Mean " * log[10](theta))),
   cex.main = 1.7,
   cex.lab = 1.4,
   yaxt = "n",
-  ylim = c(0.8, 2.2)
+  xaxt = "n",
+  ylim = c(0.8, 3)
 )
-axis(2, c(1, 1.5, 2))
+axis(2, c(1, 2, 3))
 
 par(xpd = TRUE)
 mtext("b", side = 3, line = 1, adj = 0, cex = 1.5)
 
+axis(1, at = at, labels = labels, cex.axis = 1.3)
 
[…]

### 2025-02-14 ms-3/ms.tex
Update ms.tex

@@
 
+% Redefine the \figurename to make it bold
+\renewcommand{\figurename}{\textbf{Figure}}
+
 % Set 1.5 spacing
@@
 \title{Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time}
-\author[1]{Rachael Aber}
+\author[1,2,3]{Rachael Aber}
 \author[2]{Yanming Di}
-\author[1,3]{Benjamin D. Dalziel}
+\author[1,4]{Benjamin D. Dalziel}
 \affil[1]{Department of Integrative Biology, Oregon State University, Corvallis, Oregon, USA}
 \affil[2]{Department of Statistics, Oregon, Oregon State University, Corvallis, Oregon, USA}
-\affil[3]{Department of Mathematics, Oregon State University, Corvallis, Oregon, USA}
+\affil[3]{Exponent, Inc., Bellevue, Washington, USA}
+\affil[4]{Department of Mathematics, Oregon State University, Corvallis, Oregon, USA}
 \date{} % No date for submissions
@@
 Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood function peaked at the lower boundary of the domain.
-This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig.~\ref{fig1}).
+This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig~\ref{fig1}).
 
@@
 
-Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}).
+Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig~\ref{fig1}).
 The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million.
@@
 
-Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Figure~\ref{fig2}).
+Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Fig~\ref{fig2}).  
 Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
@@
 
-Dispersion increased markedly around the peaks in incidence during the major 2022 wave, from late December 2021 to early February 2022 (Figure~\ref{fig3}).
+Dispersion increased markedly around the peaks in incidence during the major 2022 wave, from late December 2021 to early February 2022 (Fig~\ref{fig3}).
 This is in strong contrast to standard epidemic theory which predicts that dispersion should decrease as incidence rises.
-A high concentration of low $p$-values around peak incidence (Figure~\ref{fig3}) corroborates widespread changes in $\theta$ across counties, reinforcing the statistical significance of this pattern.
+A high concentration of low $p$-values around peak incidence (Fig~\ref{fig3}) corroborates widespread changes in $\theta$ across counties, reinforcing the statistical significance of this pattern.
 While these $p$-values should be corrected for multiple testing if used for inference rather than visualization, the overall trend suggests a systematic departure from theoretical expectations.
@@
 
+\section*{Supporting Information}
+
+\paragraph*{S1 Text.}
+\label{S1_Text}
+{\bf Derivation of the relationship between the disperion parameter and the mean crowding parameter.}
+
 \section*{Acknowledgments}

### 2025-02-14 ms-3/ms.tex
Update ms.tex

@@
 
-% Redefine the \figurename to make it bold
-\renewcommand{\figurename}{\textbf{Figure}}
-
 % Set 1.5 spacing
@@
 
+% Remove parentheses around equation numbers
+\renewcommand{\eqref}[1]{Eq.~\ref{#1}}
+
+% Change figure references to "Fig. X" instead of "Figure X"
+\renewcommand{\figurename}{Fig.}
+
 % Set citation style
-\setcitestyle{numbers}
+\setcitestyle{square, numbers}
 
@@
 \includegraphics[width=1\textwidth]{fig1}
-\caption{
-Detecting dispersion changes in case count time series. 
+\caption{{\bf Fig 1. Detecting dispersion changes in case count time series.}
 a: Weekly incidence of COVID-19 in the United States, with time measured in weeks since January 4, 2020, showing an example of a randomly-selected 16 week period used as an incidence trend when in simulation-based validation of the LRT test (red).
@@
 The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
-These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.
+These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq ~1, which are consistent with changes in the underlying drivers of transmission.
 
@@
 \includegraphics[width=1\textwidth]{fig2}
-\caption{Dispersion analysis of weekly COVID-19 case data for Jefferson County, Alabama. 
+\caption{{\bf Fig 2. Dispersion analysis of weekly COVID-19 case data for Jefferson County, Alabama.} 
 Results for all counties are shown in Figure 3  %.~\ref{fig3}.  
@@
 \includegraphics[width=1\textwidth]{fig3}
-\caption{Incidence and dispersion between Jan 4, 2020 and March 18, 2023, in large counties in the US.
+\caption{{\bf Fig 3. Incidence and dispersion between Jan 4, 2020 and March 18, 2023, in large counties in the US.}
 a: Mean COVID-19 cases of the 144 US counties over time. 

### 2025-02-14 ms-3/ms.tex
Update ms.tex

@@
 
-% Remove parentheses around equation numbers
-\renewcommand{\eqref}[1]{Eq.~\ref{#1}}
-
-% Change figure references to "Fig. X" instead of "Figure X"
-\renewcommand{\figurename}{Fig.}
-
 % Set citation style
-\setcitestyle{square, numbers}
+\setcitestyle{numbers}
 
@@
 The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
-These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq ~1, which are consistent with changes in the underlying drivers of transmission.
+These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.
 

### 2025-02-14 ms-3/ms_final.tex
Add formatted tex file

--- 2025-02-14 ms-3/ms.tex
@@
 \usepackage{amsmath, amssymb} % mathematical symbols
-\usepackage{natbib} % bibliography
+\usepackage[numbers,square]{natbib}
 \usepackage[margin=1in]{geometry} % margins
@@
 % Set citation style
-\setcitestyle{numbers}
+\setcitestyle{numbers,square}
 
@@
     I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
+    \label{eq1}
 \end{equation}
@@
 The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
-These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.
+These findings underscore the complex behavior of the dispersion parameter, 
+which not only varied with changes in case count regimes but also revealed 
+departures from the model expectations described by Eq (\ref{eq1}), which are 
+consistent with changes in the underlying drivers of transmission.
 

### 2025-02-15 ms-3/S1_text.tex
Update S1_text.tex

--- 2024-12-19 manuscript/S1_equations.tex
@@
-Let $m^*$ be mean crowding and $m$ be the mean. By definition, we have $m^* = m +(\sigma^2/m -1)$. 
-Therefore, we have that $m^*- m = \sigma^2/m -1$.
+\documentclass{article}
+\usepackage{amsmath}
+\begin{document}
 
-We also have that $\sigma^2 = m + m^2/\theta$, by definition of negative binomial variance, where $\theta$ is the dispersion parameter. This is equivalent to $\sigma^2/m = 1 + m/\theta$.
-Therefore, we have that $m/\theta = \sigma^2/m - 1$.
+{\noindent \bf S1 Text. Derivation of the relationship between the disperion parameter and the mean crowding parameter.}\\
 
-By equating these, we have $m^*- m = m/\theta$.
+\noindent 1. Let $m^*$ be mean crowding and $m$ be the mean.\\
+$m^* = m +(\frac{\sigma^2}{m} -1)$ (by definition of mean crowding)\\
+$ \implies m^*- m = \frac{\sigma^2}{m} -1$\\
 
-Dividing both sides by $m$, we have $m^*/m - 1 = 1/\theta$
+\noindent 2. Let $\theta$ be the dispersion parameter and $\sigma^2$ be the variance of the negative binomial distribution.\\ 
+$\sigma^2 = m + \frac{m^2}{\theta}$ (by definition of negative binomial variance)\\ $\implies \frac{\sigma^2}{m} = 1 + \frac{m}{\theta}$\\
+$\implies \frac{m}{\theta} = \frac{\sigma^2}{m} - 1$\\
 
-Finally, we have that $m^*= m/\theta + m$
+\noindent 3. $\frac{m^*}{m} - 1 = \frac{1}{\theta}$ (by equating the left-hand sides of the results in 1 and 2 and dividing by $m$) \\
+$\implies m^*= m + \frac{m}{\theta}$
+
+\end{document}

### 2025-02-16 ms-3/ms.tex
Update ms.tex
Add files via upload

@@
 
+\usepackage{caption} % Ensure the caption package is included
+\captionsetup{labelformat=empty} % Suppress automatic "Figure X" label
+
 % Set 1.5 spacing
@@
 \title{Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time}
-\author[1,2,3]{Rachael Aber}
+\author[1,2,3*]{Rachael Aber}
 \author[2]{Yanming Di}
@@
 \section*{Introduction}
-Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
-Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
+Time series of infectious disease incidence appear, to varying degrees, 
+``noisy'', showing higher frequency fluctuations (e.g., day-to-day or 
+week-to-week fluctuations) around trends at the broader temporal ranges typical 
+for epidemic curves (e.g., months or years).
+Short-term fluctuations in incidence time series are caused in part by variable 
+reporting, but may also reflect the population-level impacts of transmission 
+heterogeneity and changes in the landscape of 
+susceptibility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, 
+dalziel2018urbanization, kirkegaard2021superspreading, 
+sun2021transmission,guo2023statistical,ko2023time}.
 Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.
@@
 Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\citep{graham2019measles}.
-Analysis of the shape of epidemic curves for influenza in cities may identify contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with implications for the sensitivity of cities to climate forcing and for surge capacity in the health system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
+Analysis of the shape of epidemic curves for influenza in cities may identify 
+contexts where incidence is focused more intensely (proportionally more 
+infections in a smaller span of time) with implications for the sensitivity of 
+cities to climate forcing and for surge capacity in the health 
+system~\citep{dalziel2018urbanization, wallinga2018metropolitan}.
 
-What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
+What drives incidence dispersion and how does it relate to the underlying 
+branching process of transmission, and to observations of cases? 
 Under a wide range of configurations for a branching process model of contagion spread, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\citep{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter.
@@
 \end{equation}
-where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}.
+where time-varying reproductive number $R_t$ gives the expected number of 
+secondary infections acquired from an infected individual at time $t$, and the 
+generation time is set to 1 without loss of 
+generality~\citep{kendall1949stochastic, bjornstad2002dynamics}. 
 Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
-In applications, this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
+In applications, this model for $theta$ becomes $\theta_t = C_t / \rho_t$ where 
+$C_t$ represents reported cases and $\rho$ the reporting rate, which relates 
+reported cases to the true number of infections as $C_t = \rho_t I_t$.
 However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.
@@
 a: Weekly incidence of COVID-19 in the United States, with time measured in weeks since January 4, 2020, showing an example of a randomly-selected 16 week period used as an incidence trend when in simulation-based validation of the LRT test (red).
-b: Cases in in one county (Douglas County, Nebraska) over the sample time period with estimated incidence trend (red) and estimated dispersion values on either side of the midpoint. 
+b: Cases in one county (Douglas County, Nebraska) over the sample time period 
+with estimated incidence trend (red) and estimated dispersion values on either 
+side of the midpoint. 
 c: Estimated $\theta$ versus true $\theta$ in simulation studies combining a randomly-selected section of the national incidence curve with a random population size and set of dispersion values.
@@
 For instance, burst-tree decomposition of time series has also facilitated computation of a burst-size distribution for a series given a specified time window~\citep{jo_burst-tree_2020}, allowing comparison of variability within one location over time.
-Spatial variation in superspreading potential has been investigated through risk maps of superspreading environments~\citep{loo_identification_2021}, and future work could investigate the correspondence between dispersion in case count time series, as quantified here, and indicators of a high risk of superspreading, with the potential to further illucidate drivers of transmission risk across scales, and more finely resolve landscapes of susceptibility.
-Additionally, as population-wide disease control may be less effective than those which are focussed to individuals in high-transmission contexts~\citep{lloyd2005superspreading}, identifying candidate time periods when transmission heterogeneity is high may catalyze the development of more effective control strategies, particularly those that connect vulnerable populations with resources at critical times.
+Spatial variation in superspreading potential has been investigated through 
+risk maps of superspreading environments~\citep{loo_identification_2021}, and 
+future work could investigate the correspondence between dispersion in case 
+count time series, as quantified here, and indicators of a high risk of 
+superspreading, with the potential to further elucidate drivers of transmission 
+risk across scales, and more finely resolve landscapes of susceptibility.
+Additionally, as population-wide disease control may be less effective than 
+those which are focused to individuals in high-transmission 
+contexts~\citep{lloyd2005superspreading}, identifying candidate time periods 
+when transmission heterogeneity is high may catalyze the development of more 
+effective control strategies, particularly those that connect vulnerable 
+populations with resources at critical times.
 The finding that dispersion increased rather than decreased during the 2022 surge challenges theoretical expectations and suggests that fundamental assumptions about the scaling of transmission dynamics may require reevaluation.
@@
 \label{S1_Text}
-{\bf Derivation of the relationship between the disperion parameter and the mean crowding parameter.}
-
-\section*{Acknowledgments}
-RA work on this project was supported by X and Y. BDD work on this project was supported by the National Science Foundation (NSF) grants (X, Y) and by the David and Lucile Packard Foundation.
+{\bf Derivation of the relationship between the dispersion parameter and the 
+mean crowding parameter.}
 
 \clearpage
-\section*{References}
 \bibliographystyle{vancouver} 
